
(c.f. @dobson4e §2.4.4)

---

::: notes
We can create ordinal variables in R using the `ordered()` function^[or equivalently, `factor(ordered = TRUE)`].
:::

:::{#exm-ordinal-variable}

```{r}
#| eval: false
url <- paste0(
  "https://regression.ucsf.edu/sites/g/files/tkssra6706/",
  "f/wysiwyg/home/data/hersdata.dta"
)
library(haven)
hers <- read_dta(url)
```

```{r}
#| include: false
hers <- rmb::hers |> haven::zap_labels()
```

```{r}
#| tbl-cap: "HERS dataset"
#| label: tbl-HERS
hers |> head()
```

:::

---

::: notes
For ordinal variables, 
we might want to use contrasts that respect the ordering of the categories.
The default treatment contrasts don't account for ordering.
:::

::: {.callout-note}
## Working with ordinal variables

When working with ordinal covariates in linear models:

1. Use `ordered()` or `factor(ordered = TRUE)` to create the variable
2. Consider using polynomial contrasts (`contr.poly`) which respect ordering
3. Alternatively, treat the ordinal variable as numeric if equal spacing is reasonable
4. Check `?codingMatrices::contr.diff` for additional contrast options

See @dobson4e §2.4.4 for more details on contrasts for ordinal variables.
:::
