:::{#exm-bw-inference}

#### MLEs for birthweight data

In model [-@eq-BW-lm-interact] above, $\heinf(\beta)$ is:

:::{#tbl-vcov}

```{r}
#| code-fold: show
bw_lm2 |> vcov()
```

Covariance matrix of $\hvb$ for `birthweight` model [-@eq-BW-lm-interact] (with interaction term)

:::

#### Interpreting the layout of the covariance matrix

The covariance matrix $\widehat{\Covt}(\hvb)$ is a $p \times p$ symmetric matrix,
where $p$ is the number of regression coefficients
(including the intercept, if present).
Its rows and columns correspond to those $p$ coefficient estimates.
When an intercept is included,
the coefficients are typically written
$\hat\b_0, \hat\b_1, \ldots$,
with $\hat\b_0$ denoting the intercept.
The matrix entries themselves are still indexed by position,
so matrix row/column index $1$ corresponds to the intercept term when one is included.

The general layout is:

$$
\widehat{\Covt}(\hvb) =
\begin{array}{c|cccc}
  & \hat\b_0 & \hat\b_1 & \cdots & \hat\b_{p-1} \\
\hline
\hat\b_0     & \Var{\hat\b_0}                    & \Covs{\hat\b_0}{\hat\b_1}     & \cdots & \Covs{\hat\b_0}{\hat\b_{p-1}}     \\
\hat\b_1     & \Covs{\hat\b_1}{\hat\b_0}         & \Var{\hat\b_1}                & \cdots & \Covs{\hat\b_1}{\hat\b_{p-1}}     \\
\vdots       & \vdots                             & \vdots                        & \ddots & \vdots \\
\hat\b_{p-1} & \Covs{\hat\b_{p-1}}{\hat\b_0}     & \Covs{\hat\b_{p-1}}{\hat\b_1} & \cdots & \Var{\hat\b_{p-1}}
\end{array}
$$

That is:

- The **diagonal** entries are the **variances** of the individual coefficient estimates:
  the entry in matrix position $(j+1, j+1)$ is $\Var{\hat\b_j}$,
  for $j = 0, 1, \ldots, p - 1$.
- The **off-diagonal** entries are the **covariances** between pairs of coefficient estimates:
  the entry in matrix position $(i+1, j+1)$ is $\Covs{\hat\b_i}{\hat\b_j}$,
  for $i \neq j$.
- The matrix is **symmetric**: $\Covs{\hat\b_i}{\hat\b_j} = \Covs{\hat\b_j}{\hat\b_i}$.

:::{#exm-vcov-bw}

#### Schematic for the birthweight model

For model [-@eq-BW-lm-interact],
which has four coefficients
$(\hat\b_0, \hat\b_M, \hat\b_A, \hat\b_{AM})$,
the covariance matrix has the following layout:

$$
\widehat{\Covt}(\hvb) =
\begin{array}{c|cccc}
  & \hat\b_0 & \hat\b_M & \hat\b_A & \hat\b_{AM} \\
\hline
\hat\b_0   & \Var{\hat\b_0}                & \Covs{\hat\b_0}{\hat\b_M}    & \Covs{\hat\b_0}{\hat\b_A}    & \Covs{\hat\b_0}{\hat\b_{AM}}    \\
\hat\b_M   & \Covs{\hat\b_M}{\hat\b_0}    & \Var{\hat\b_M}               & \Covs{\hat\b_M}{\hat\b_A}    & \Covs{\hat\b_M}{\hat\b_{AM}}    \\
\hat\b_A   & \Covs{\hat\b_A}{\hat\b_0}    & \Covs{\hat\b_A}{\hat\b_M}    & \Var{\hat\b_A}               & \Covs{\hat\b_A}{\hat\b_{AM}}    \\
\hat\b_{AM}& \Covs{\hat\b_{AM}}{\hat\b_0} & \Covs{\hat\b_{AM}}{\hat\b_M} & \Covs{\hat\b_{AM}}{\hat\b_A} & \Var{\hat\b_{AM}}
\end{array}
$$

::: notes
The square roots of the diagonal entries give the standard errors of the coefficient estimates:
:::

```{r}
#| code-fold: show
bw_lm2 |>
  vcov() |>
  diag() |>
  sqrt()
```

```{r}
#| label: tbl-mod-intx
#| tbl-cap: "Estimated model for `birthweight` data with interaction term"
bw_lm2 |>
  parameters() |>
  print_md()
```

::: notes
So we can do confidence intervals, hypothesis tests, and p-values exactly as in the one-variable case we looked at previously.
:::

:::

:::
