
To look for problems with our model,
we can check whether the residuals $\resid_i$
and standardized residuals $\stdresid_i$
look like they have the distributions
that they are supposed to have,
according to the model.

---

#### Standardized residuals in R

```{r}
h_ii <- hatvalues(bw_lm2)
manual_std_resid <- resid(bw_lm2) / (sigma(bw_lm2) * sqrt(1 - h_ii))
manual_std_resid
rstandard(bw_lm2)
```

::: notes
These are nearly the same.
Any differences are from numerical rounding.

:::

---

```{r}
rstandard_compare_plot <-
  tibble(
    x = resid(bw_lm2) / (sigma(bw_lm2) * sqrt(1 - h_ii)),
    y = rstandard(bw_lm2)
  ) |>
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  theme_bw() +
  coord_equal() +
  xlab("resid(bw_lm2)/(sigma(bw_lm2)*sqrt(1-h_ii))") +
  ylab("rstandard(bw_lm2)") +
  geom_abline(
    aes(
      intercept = 0,
      slope = 1,
      col = "x=y"
    )
  ) +
  labs(colour = "") +
  scale_colour_manual(values = "red")

print(rstandard_compare_plot)
```

---

Let's add these residuals to the `tibble` of our dataset:

```{r}
bw <-
  bw |>
  mutate(
    fitted_lm2 = fitted(bw_lm2),
    resid_lm2 = resid(bw_lm2),
    resid_lm2_alt = weight - fitted_lm2,
    std_resid_lm2 = rstandard(bw_lm2),
    std_resid_lm2_alt =
    resid_lm2 / (sigma(bw_lm2) * sqrt(1 - h_ii))
  )

bw |>
  select(
    sex,
    age,
    weight,
    fitted_lm2,
    resid_lm2,
    std_resid_lm2
  )
```

---

::: notes

Now let's build histograms:

:::

```{r}
#| fig-cap: "Marginal distribution of (nonstandardized) residuals"
#| label: fig-marg-dist-resid
resid_marginal_hist <-
  bw |>
  ggplot(aes(x = resid_lm2)) +
  geom_histogram()

print(resid_marginal_hist)
```

::: notes
Hard to tell with this small amount of data, but I'm a bit concerned that the histogram doesn't show a bell-curve shape.

:::

---


```{r}
#| fig-cap: "Marginal distribution of standardized residuals"
#| label: fig-marg-stresd
std_resid_marginal_hist <-
  bw |>
  ggplot(aes(x = std_resid_lm2)) +
  geom_histogram()

print(std_resid_marginal_hist)
```

::: notes
This looks similar, although the scale of the x-axis got narrower, because we divided by $\hat\sigma$ (roughly speaking).

Still hard to tell if the distribution is Gaussian.

:::

---

