We can also construct **prediction intervals** for
the value of a new observation $Y^*$,
given a covariate pattern $\vxs$.

A new observation $Y^*$ at $\vxs$ follows the same linear model as the training data:

$$Y^* = \tp{(\vxs)}\vb + \eps^*$$

where $\eps^* \sim \ndist{0, \sigma^2}$ is independent of the training data
(and therefore independent of $\hb$).

The predicted value of $Y^*$ is:

$$\hat{Y}^* \eqdef \hat\mu(\vxs) = \tp{(\vxs)}\hb$$

We base the prediction interval on the **prediction error** $Y^* - \hat{Y}^*$:

$$
\ba
Y^* - \hat{Y}^*
&= \tp{(\vxs)}\vb + \eps^* - \tp{(\vxs)}\hb
\\
&= \eps^* - \tp{(\vxs)}(\hb - \vb)
\ea
$$

Assuming the model is correctly specified, the predictions will be unbiased:

$$
\ba
\E{Y^* - \hat{Y}^*}
&= \E{\eps^*} - \tp{(\vxs)}\E{\hb - \vb}
\\
&= 0 - \tp{(\vxs)} \cdot 0
\\
&= 0
\ea
$$

The variance of the prediction error is:

$$
\ba
\Var{Y^* - \hat{Y}^*}
&= \Var{\eps^* - \tp{(\vxs)}(\hb - \vb)}
\\
&= \Var{\eps^*} + \Var{\tp{(\vxs)}\hb}
\quad \text{(since } \eps^* \ind \hb\text{)}
\\
&= \sigma^2 + \tp{(\vxs)}\Var{\hb}\vxs
\\
&= \sigma^2 + \tp{(\vxs)}\paren{\sigma^2(\mX'\mX)^{-1}}\vxs
\\
&= \sigma^2 + \sigma^2\tp{(\vxs)}(\mX'\mX)^{-1}\vxs
\\
&= \sigma^2\paren{1 + \tp{(\vxs)}(\mX'\mX)^{-1}\vxs}
\ea
$${#eq-prediction-var}

::: notes
See @hoggtanis2015 §7.6 (p. 340).
:::

The variance (@eq-prediction-var) has two components:

- $\Var{\eps^*} = \sigma^2$:
the variance of the future observation's own noise,
which is irreducible.

- $\Var{\dprodf{\vxs}{\hb}} = \tp{(\vxs)}\Var{\hb}\vxs$:
the variance due to estimating the mean $\dprodf{\vxs}{\vb}$ from the data.

The first component is not present in the confidence interval for $\hat\mu(\vxs)$,
which only accounts for estimation uncertainty.
This is why prediction intervals are always wider than the corresponding confidence intervals:
a prediction interval must additionally account for
the irreducible noise $\eps^*$ in a new observation.

The **standard error** of the prediction error is the square root of its variance:

$$
\SE{Y^* - \hat{Y}^*}
= \sqrt{\Var{Y^* - \hat{Y}^*}}
= \sigma\sqrt{1 + \tp{(\vxs)}(\mX'\mX)^{-1}\vxs}
$$

Replacing $\sigma$ with the residual standard error $\hs$ gives the estimated standard error:

$$
\HSE{Y^* - \hat{Y}^*}
= \hs\sqrt{1 + \tp{(\vxs)}(\mX'\mX)^{-1}\vxs}
$${#eq-prediction-se}

Since $\eps^* \sim \ndist{0, \sigma^2}$ and $\hb \sim \ndist{\vb, \sigma^2(\mX'\mX)^{-1}}$
are both normally distributed,
and since $\eps^* \ind \hb$ (as noted above in the variance derivation),
their difference $Y^* - \hat{Y}^* = \eps^* - \tp{(\vxs)}(\hb - \vb)$
is also normally distributed,
with mean 0 (as shown above) and variance given by @eq-prediction-var.
Standardizing by dividing by $\sigma\sqrt{1 + \tp{(\vxs)}(\mX'\mX)^{-1}\vxs}$
yields a standard normal random variable.
Replacing $\sigma$ with $\hs$,
which is estimated with $n-p$ degrees of freedom and is independent of $Y^* - \hat{Y}^*$,
the standardized prediction error follows a $t$ distribution:

$$
\frac{Y^* - \hat{Y}^*}{\HSE{Y^* - \hat{Y}^*}} \sim t_{n-p}
$$

Therefore, a $(1-\alpha)$ **prediction interval** for $Y^*$ is:

$$
Y^* \in
\paren{
\hat{Y}^*
\pm
t_{n-p}\paren{1 - \frac{\alpha}{2}}
\cdot
\HSE{Y^* - \hat{Y}^*}
}
$${#eq-prediction-interval}

```{r}
#| code-fold: show
x <- tibble(age = 40, sex = "male")
bw_lm2 |>
  predict(newdata = x, interval = "predict")
```

---

If you don't specify `newdata`, you get a warning:

```{r}
#| code-fold: show
#| warning: true
bw_lm2 |>
  predict(interval = "predict") |>
  head()
```

The warning from the last command is: "predictions on current data refer to *future* responses" (since you already know what happened to the current data, and thus don't need to predict it).

See `?predict.lm` for more.

---

:::{#fig-pred-int-bw}
```{r}
plot_PIs_and_CIs(bw, bw_lm2)
```

Confidence and prediction intervals for `birthweight` model [-@eq-BW-lm-interact]

:::

---
