
## General form of a Gaussian linear regression model {.smaller}

:::{#def-lm-general-form}
#### General form of a Gaussian linear regression model

A Gaussian linear regression model has two components:

1. The **random part**

    (the **outcome distribution model**):

    $$
    Y_i|\vx_i \simind \Normal(\mu_i,\ss)
    $$

    Here $\vx_i$ is the observed covariate vector for unit $i$.
    Here $\simind$ means independently distributed.
    Conditional on $\vx_1,\ldots,\vx_n$,
    the responses are independent across units,
    and each $Y_i|\vx_i$ follows the same conditional distributional form
    with common variance $\ss$
    while allowing $\mu_i$ to vary by observation.

2. The **systematic part**

    (the **linear predictor component**):

    $$
    \ba
    \mu_i
    &= \dprod{\vb}{\vx_i}
    \\
    &= \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip}
    \ea
    $$

:::

::: notes
For the systematic part of the model,
here $\vb \eqdef (\beta_0,\beta_1,\ldots,\beta_p)$
is the coefficient vector.

Here $p$ is the number of covariates
excluding the intercept.
The vector $\vx_i$ includes an intercept entry equal to 1,
so $\vx_i$ has length $p+1$.

Terminology varies across texts.
[@dobson4e, p. 36] uses this two-part formulation explicitly
(probability distribution
+
mean/link equation/linear component).
[@vittinghoff2e, pp. 72-73]
uses the labels
**systematic part**
and
**random part**.
[@dunn2018generalized, p. 11]
uses the labels
**random component**
and
**systematic component**.
:::

---

## Motivating example: birthweights and gestational age {.smaller}

Suppose we want to learn about the distributions of birthweights (*outcome* $Y$) 
for (human) babies born at different gestational ages (*covariate* $A$) 
and with different chromosomal sexes (*covariate* $S$) (@dobson4e Example 2.2.2).

## Dobson `birthweight` data {.smaller}

::::: {.panel-tabset}

### Data as table

```{r}
#| label: tbl-birthweight-data1
#| tbl-cap: "`birthweight` data (@dobson4e Example 2.2.2)"
library(dobson)
data("birthweight", package = "dobson")
birthweight
```

### Reshape data for graphing

```{r}
#| label: tbl-birthweight-data2
#| tbl-cap: "`birthweight` data reshaped"
library(tidyverse)
bw <-
  birthweight |>
  pivot_longer(
    cols = everything(),
    names_to = c("sex", ".value"),
    names_sep = "s "
  ) |>
  rename(age = `gestational age`) |>
  mutate(
    id = row_number(),
    sex = sex |>
      case_match(
        "boy" ~ "male",
        "girl" ~ "female"
      ) |>
      factor(levels = c("female", "male")),
    male = sex == "male",
    female = sex == "female"
  )

bw
```

### Data as graph

```{r}
#| label: fig-plot-birthweight1
#| fig-cap: "`birthweight` data (@dobson4e Example 2.2.2)"
#| fig-height: 5
#| fig-width: 7
plot1 <- bw |>
  ggplot(aes(
    x = age,
    y = weight,
    shape = sex,
    col = sex
  )) +
  theme_bw() +
  xlab("Gestational age (weeks)") +
  ylab("Birthweight (grams)") +
  theme(legend.position = "bottom") +
  # expand_limits(y = 0, x = 0) +
  geom_point(alpha = .7)
print(plot1 + facet_wrap(~sex))
```

:::::

---

### Data notation {.smaller}

::: notes
Let's define some notation to represent this data:
:::

-   $Y$: birthweight (measured in grams)
-   $S$: chromosomal sex: "male" (XY) or "female" (XX)
-   $M$: indicator variable for $S$ = "male"^[$M$ is implicitly a deterministic function of $S$]
-   $M = 0$ if $S$ = "female"
-   $M = 1$ if $S$ = "male"
-   $F$: indicator variable for $S$ = "female"^[$F$ is implicitly a deterministic function of $S$]
-   $F = 1$ if $S$ = "female"
-   $F = 0$ if $S$ = "male"
-   $A$: estimated gestational age at birth (measured in weeks).

::: notes
Female is the **reference level** for the categorical variable $S$ 
(chromosomal sex) and corresponding indicator variable $M$ . 
The choice of a reference level is arbitrary and does not limit what 
we can do with the resulting model; 
it only makes it more computationally convenient to make inferences 
about comparisons involving that reference group.

$M$ and $F$ are called **dummy variables**; 
together, they are a numeric representation of the categorical variable $S$.
Dummy variables with values 0 and 1 are also called **indicator variables**.
There are other ways to construct dummy variables, 
such as using the values -1 and 1
(see @dobson4e §2.4 for details).
:::

### Motivating example: `hers` data {.smaller}

{{< include _subfiles/Linear-models-overview/_sec_hers_data.qmd >}}

## Parallel lines regression {.smaller}

::: notes
c.f. @dunn2018generalized 
[§2.10.3](https://link.springer.com/chapter/10.1007/978-1-4419-0118-7_2#Sec31).
:::

::: notes

We don't have enough data to model the distribution of birth weight 
separately for each combination of gestational age and sex, 
so let's instead consider a (relatively) simple model for how that 
distribution varies with gestational age and sex:
:::


$$
\ba
Y|M,A &\sciid N(\mu(M,A), \sigma^2)\\
\mu(m,a) &= \beta_0 + \beta_M m + \beta_A a
\ea
$$ {#eq-lm-parallel}

:::{.notes}

@tbl-lm-parallel shows the parameter estimates from R.
@fig-parallel-fit1 shows the estimated model, superimposed on the data.

:::

::: {.column width=40%}

::: {#tbl-lm-parallel}

```{r}
bw_lm1 <- lm(
  formula = weight ~ sex + age,
  data = bw
)

library(parameters)
bw_lm1 |>
  parameters::parameters() |>
  parameters::print_md(
    include_reference = include_reference_lines,
    select = "{estimate}"
  )
```

Regression parameter estimates for [Model @eq-lm-parallel] of `birthweight` data

:::

:::

:::{.column width=10%}
:::

:::{.column width=50%}

:::{#fig-parallel-fit1}

```{r}
bw <-
  bw |>
  mutate(`E[Y|X=x]` = fitted(bw_lm1)) |>
  arrange(sex, age)

plot2 <-
  plot1 %+% bw +
  geom_line(aes(y = `E[Y|X=x]`))

print(plot2)
```

Graph of [Model @eq-lm-parallel] for `birthweight` data

:::

:::

---

### Model assumptions and predictions

::: notes
To learn what this model is assuming, let's plug in a few values.
:::

---

::: {#exr-pred-fem-parallel}

What's the mean birthweight for a female born at 36 weeks?

```{r}
#| tbl-cap: "Estimated coefficients for [model @eq-lm-parallel]"
#| code-fold: false
#| echo: false
bw_lm1 |>
  parameters::parameters() |>
  parameters::print_md(
    select = "{estimate}"
  )
```

:::

---

:::{.solution .smaller}
\ 

```{r}
#| tbl-cap: "Estimated coefficients for [model @eq-lm-parallel]"
#| code-fold: false
#| echo: false
bw_lm1 |>
  parameters::parameters() |>
  parameters::print_md(
    select = "{estimate}"
  )
```

```{r}
pred_female <- coef(bw_lm1)["(Intercept)"] + coef(bw_lm1)["age"] * 36
## or using built-in prediction:
pred_female_alt <- predict(bw_lm1, newdata = tibble(sex = "female", age = 36))
```

$$
\ba
E[Y|M = 0, A = 36] 
&= \beta_0 + \paren{\beta_M \cdot 0}+ \paren{\beta_A \cdot 36} \\
&= `r coef(bw_lm1)["(Intercept)"]` + \paren{`r coef(bw_lm1)["sexmale"]` \cdot 0} 
+ \paren{`r coef(bw_lm1)["age"]` \cdot 36} \\
&=  `r pred_female`
\ea
$$
:::

---

:::{#exr-pred-male-parallel}

What's the mean birthweight for a male born at 36 weeks?

```{r}
#| tbl-cap: "Estimated coefficients for [model @eq-lm-parallel]"
#| code-fold: false
#| echo: false
bw_lm1 |>
  parameters::parameters() |>
  parameters::print_md(
    select = "{estimate}"
  )
```

:::

---

:::{.solution}
\ 

```{r}
#| tbl-cap: "Estimated coefficients for [model @eq-lm-parallel]"
#| code-fold: false
#| echo: false
bw_lm1 |>
  parameters::parameters() |>
  parameters::print_md(
    select = "{estimate}"
  )
```

```{r}
pred_male <-
  coef(bw_lm1)["(Intercept)"] +
  coef(bw_lm1)["sexmale"] +
  coef(bw_lm1)["age"] * 36
```

$$
\ba
E[Y|M = 1, A = 36] 
&= \beta_0 + \beta_M \cdot 1+ \beta_A \cdot 36 \\
&=  `r pred_male`
\ea
$$

:::

---

:::{#exr-diff-sex-parallel-1}
What's the difference in mean birthweights between males born at 36 weeks and females born at 36 weeks?
:::

```{r}
coef(bw_lm1)
```

---

:::{.solution}

$$
\begin{aligned}
& E[Y|M = 1, A = 36] - E[Y|M = 0, A = 36]\\
&= 
`r pred_male` - `r pred_female`\\
&=
`r pred_male - pred_female`
\end{aligned}
$$

Shortcut:

$$
\begin{aligned}
& E[Y|M = 1, A = 36] - E[Y|M = 0, A = 36]\\
&= (\beta_0 + \beta_M \cdot 1+ \beta_A \cdot 36) - 
(\beta_0 + \beta_M \cdot 0+ \beta_A \cdot 36) \\
&= \beta_M \\ 
&=  `r coef(bw_lm1)["sexmale"]`
\end{aligned}
$$

:::

:::{.notes}

Age cancels out in this difference.
In other words, 
according to this model, 
the difference between females and males with the same gestational age 
is the same for every age.

This characteristic is an assumption of the model
specified by @eq-lm-parallel.
It's hardwired into the parametric model structure, 
even before we estimated values for those parameters.

:::

---

### Coefficient Interpretation

{{< include _subfiles/Linear-models-overview/_sec_linear_coef_interp_no_intxn.qmd >}}

### Parallel planes regression for `hers` data {.smaller}

{{< include _subfiles/Linear-models-overview/_sec_hers_lm_parallel.qmd >}}

## Interactions {.smaller}

::: notes
What if we don't like that parallel lines assumption?

Then we need to allow an "interaction" between  age $A$ and sex $S$:
:::

$$E[Y|S=s,A=a] = \beta_0 + \beta_A a+ \beta_M m + \beta_{AM} (a \cdot m)$$ {#eq-BW-lm-interact}

::: notes
Now, the slope of mean birthweight $E[Y|A,S]$ 
with respect to gestational age $A$ 
depends on the value of sex $S$.
:::

::: {.column width=40% .smaller}

```{r}
#| label: tbl-bw-model-coefs-interact
#| tbl-cap: "Birthweight model with interaction term"

bw_lm2 <- lm(weight ~ sex + age + sex:age, data = bw)
bw_lm2 |>
  parameters() |>
  parameters::print_md(
    include_reference = include_reference_lines,
    select = "{estimate}"
  )
```

:::

:::{.column width=5%}
:::

:::{.column width=55%}

```{r}
#| label: fig-bw-interaction
#| fig-cap: "Birthweight model with interaction term"
bw <-
  bw |>
  mutate(
    predlm2 = predict(bw_lm2)
  ) |>
  arrange(sex, age)

plot1_interact <-
  plot1 %+% bw +
  geom_line(aes(y = predlm2))

print(plot1_interact)
```

:::

::: {.notes}
Now we can see that the lines aren't parallel.
:::

---

Here's another way we could rewrite this model (by collecting terms involving $S$):

$$
E[Y|M,A] = \beta_0 + \beta_M M+ (\beta_A + \beta_{AM} M) A
$$

::: notes
If you want to understand a coefficient in a model with interactions, 
collect terms for the corresponding variable, 
and you will see which other covariates interact with the variable 
whose coefficient you are interested in.
In this case, 
the association between
$A$ (age) varies between males and females (that is, by sex $S$).
^[some call this kind of variation "interaction" or "effect modification",
but "act", "effect", "modify", and "by" all suggest causality, 
which we are not prepared to assess here;
let's try to avoid using causal terms, 
unless we are constructing a causal model.]
So the slope of $Y$ with respect to $A$ depends on the value of $M$.
According to this model, there is no such thing as 
"*the* slope of birthweight with respect to age". 
There are two slopes, one for each sex.
We can only talk about 
"the slope of birthweight with respect to age among males" 
and "the slope of birthweight with respect to age among females".
Then: 
each non-interaction slope coefficient is the difference in means 
per unit difference in its corresponding variable, 
when all interacting variables are set to 0.
:::

---

::: notes
To learn what this model is assuming, let's plug in a few values.
:::

:::{#exr-pred-fem-interact}
According to this model, what's the mean birthweight for a female born at 36 weeks?
:::

```{r}
#| code-fold: false
#| echo: false
bw_lm2 |>
  parameters() |>
  print_md(
    include_reference = include_reference_lines,
    select = "{estimate}"
  )
```


---

::: {.solution}
\ 

```{r}
#| code-fold: false
#| echo: false
bw_lm2 |>
  parameters() |>
  print_md(
    include_reference = include_reference_lines,
    select = "{estimate}"
  )
```

```{r}
pred_female <- coef(bw_lm2)["(Intercept)"] + coef(bw_lm2)["age"] * 36
```

$$
E[Y|M = 0, A = 36] = 
\beta_0 + \beta_M \cdot 0+ \beta_A \cdot 36 + \beta_{AM} \cdot (0 * 36) 
=  `r pred_female`
$$ 

:::

---

:::{#exr-pred-interact-male_36}
What's the mean birthweight for a male born at 36 weeks?

```{r}
#| code-fold: false
#| echo: false
bw_lm2 |>
  parameters() |>
  print_md(
    include_reference = include_reference_lines,
    select = "{estimate}"
  )
```

:::

---

::: solution
\ 

```{r}
#| code-fold: false
#| echo: false
bw_lm2 |>
  parameters() |>
  print_md(
    include_reference = include_reference_lines,
    select = "{estimate}"
  )
```

```{r}
pred_male <-
  coef(bw_lm2)["(Intercept)"] +
  coef(bw_lm2)["sexmale"] +
  coef(bw_lm2)["age"] * 36 +
  coef(bw_lm2)["sexmale:age"] * 36
```

$$
\ba
E[Y|M = 1, A = 36]
&= \beta_0 + \beta_M \cdot 1+ \beta_A \cdot 36 + \beta_{AM} \cdot 1 \cdot 36\\
&= `r pred_male`
\ea
$$

:::

---

:::{#exr-diff-gender-interact}
What's the difference in mean birthweights between males born at 36 weeks and females born at 36 weeks?
:::

---

:::{.solution}

$$
\begin{aligned}
& E[Y|M = 1, A = 36] - E[Y|M = 0, A = 36]\\ 
&= (\beta_0 + \beta_M \cdot 1+ \beta_A \cdot 36 + \beta_{AM} \cdot 1 \cdot 36)\\ 
&\ \ \ \ \  -(\beta_0 + \beta_M \cdot 0+ \beta_A \cdot 36 + \beta_{AM} \cdot 0 \cdot 36) \\
&= \beta_M + \beta_{AM}\cdot 36\\
&=  `r pred_male - pred_female`
\end{aligned}
$$
:::

:::{.notes}
Note that age now does show up in the difference: in other words, according to this model, the difference in mean birthweights between females and males with the same gestational age can vary by gestational age.

That's how the lines in the graph ended up non-parallel.

:::

---

### Coefficient Interpretation

{{< include _subfiles/Linear-models-overview/_sec_linear_coef_interp_intxn.qmd >}}

--- 

### Compare coefficient interpretations

:::{#tbl-compare-coef-interps}
  $\mu(m,a)$    | $\beta_0 + \beta_M m + \beta_A a$  | $\beta_0 + \beta_M m + \beta_A a + \red{\beta_{AM} ma}$
----------------|------------------------------------|---------------------------------------------------
  $\beta_0$     | $\mu(0,0)$                         | $\mu(0,0)$ 
  $\beta_A$     | $\deriv{a}\mu(\blue{m},a)$         | $\deriv{a}\mu(\red{0},a)$
  $\beta_M$     | $\mu(1, \blue{a}) - \mu(0, \blue{a})$            | $\mu(1, \red{0}) - \mu(0,\red{0})$
  $\beta_{AM}$  |                                    | $\deriv{a}\mu(1,a) - \deriv{a}\mu(0,a)$

Coefficient interpretations, by model structure

:::

::: notes
In the model with an interaction term multiplying $A\times M$, 
the interpretation of $\b_A$ involves the reference level of $M$, 
and interpretation of $\beta_M$ involves the reference level of $A$ (@tbl-compare-coef-interps).
:::

{{< slidebreak >}}

### General procedure for interpreting coefficients {#sec-coef-interp-procedure}

{{< include _subfiles/Linear-models-overview/_sec_linreg_coef_interp_procedure.qmd >}}

### Interactions in `hers` data {.smaller}

{{< include _subfiles/Linear-models-overview/_sec_hers_lm_interact.qmd >}}

## Stratified regression {.smaller}

:::{.notes}
We could re-write the interaction model as a stratified model, with a slope and intercept for each sex:
:::

$$
\E{Y|A=a, S=s} = 
\beta_M m + \beta_{AM} (a \cdot m) + 
\beta_F f + \beta_{AF} (a \cdot f)
$$ {#eq-model-strat}

Compare this stratified model (@eq-model-strat) with our interaction model, @eq-BW-lm-interact:

$$
\E{Y|A=a, S=s} = 
\beta_0 + \beta_A a + \beta_M m + \beta_{AM} (a \cdot m)
$$

::: notes

In the stratified model, the intercept term $\beta_0$ has been relabeled as $\beta_F$.

:::

::: {.column width=45%}
```{r}
#| label: tbl-bw-model-coefs-interact2
#| tbl-cap: "Birthweight model with interaction term"

bw_lm2 <- lm(weight ~ sex + age + sex:age, data = bw)
bw_lm2 |>
  parameters() |>
  print_md(
    include_reference = include_reference_lines,
    select = "{estimate}"
  )
```

:::

:::{.column width=10%}
:::

:::{.column width=45%}

```{r}
#| label: tbl-bw-model-coefs-strat
#| tbl-cap: "Birthweight model - stratified betas"

bw_lm_strat <-
  bw |>
  lm(
    formula = weight ~ sex + sex:age - 1,
    data = _
  )

bw_lm_strat |>
  parameters() |>
  print_md(
    select = "{estimate}"
  )
```

:::

### Stratified regression for `hers` data {.smaller}

{{< include _subfiles/Linear-models-overview/_sec_hers_lm_stratified.qmd >}}

## Curved-line regression

::: notes
If we transform some of our covariates ($X$s) 
and plot the resulting model on the original covariate scale, 
we end up with curved regression lines:
:::

```{r}
#| label: fig-birthweight-log-x
#| fig-cap: "`birthweight` model with `age` entering on log scale"
bw_lm3 <- lm(weight ~ sex:log(age) - 1, data = bw)

ggbw <-
  bw |>
  ggplot(
    aes(x = age, y = weight)
  ) +
  geom_point() +
  xlab("Gestational Age (weeks)") +
  ylab("Birth Weight (g)")

ggbw2 <- ggbw +
  stat_smooth(
    method = "lm",
    formula = y ~ log(x),
    geom = "smooth"
  ) +
  xlab("Gestational Age (weeks)") +
  ylab("Birth Weight (g)")


ggbw2 |> print()
```

---

::: notes
Below is an example with a slightly more obvious curve.
:::

```{r}
#| label: fig-penguins-log-x
#| fig-cap: "`palmerpenguins` model with `bill_length` entering on log scale"
library(palmerpenguins)

ggpenguins <-
  palmerpenguins::penguins |>
  dplyr::filter(species == "Adelie") |>
  ggplot(
    aes(x = bill_length_mm, y = body_mass_g)
  ) +
  geom_point() +
  xlab("Bill length (mm)") +
  ylab("Body mass (g)")

ggpenguins2 <- ggpenguins +
  stat_smooth(
    method = "lm",
    formula = y ~ log(x),
    geom = "smooth"
  ) +
  xlab("Bill length (mm)") +
  ylab("Body mass (g)")


ggpenguins2 |> print()
```

{{< include _subfiles/Linear-models-overview/_sec_linreg_rescaling.qmd >}}


## Categorical covariates with more than two levels

{{< include _subfiles/Linear-models-overview/_sec_linreg_categorical.qmd >}}

## Ordinal covariates

{{< include _subfiles/Linear-models-overview/_sec_linreg_ordinal.qmd >}}
