
:::{.callout-tip}
This section is adapted from @dobson4e [§6.2-6.3] and 
@dunn2018generalized [Chapter 3](https://link.springer.com/chapter/10.1007/978-1-4419-0118-7_3).
:::

### Assumptions in linear regression models {.smaller .scrollable}

{{< include _subfiles/Linear-models-overview/_sec-linreg-assumptions.qmd >}}

### Direct visualization

::: notes
The most direct way to examine the fit of a model is 
to compare it to the raw observed data.
:::

```{r}
#| label: fig-bw-interaction2
#| fig-cap: "Birthweight model with interaction term"
bw <-
  bw |>
  mutate(
    predlm2 = predict(bw_lm2)
  ) |>
  arrange(sex, age)

plot1_interact <-
  plot1 %+% bw +
  geom_line(aes(y = predlm2))

print(plot1_interact)
```

::: notes
It's not easy to assess these assumptions from this model.
If there are multiple continuous covariates, 
it becomes even harder to visualize the raw data.
:::

---

#### Fitted model for `hers` data {.smaller}

::: notes
Consider the `hers` data from @vittinghoff2e.

{{< include _subfiles/shared/_sec_hers_intro.qmd >}}

:::

::: {.landscape}

:::{#tbl-hers-data}

```{r}
hers <- fs::path_package("rme", "extdata/hersdata.dta") |> 
  haven::read_dta()
hers
```

`hers` data
:::

:::

---

::: notes
Suppose we consider models with and without intercept terms (i.e.,
possibly forcing the intercept to go through 0):
:::

```{r}
hers_lm_with_int <- lm(
  na.action = na.exclude,
  LDL ~ smoking * age, data = hers
)


library(equatiomatic)
equatiomatic::extract_eq(hers_lm_with_int)
```

```{r}
hers_lm_no_int <- lm(
  na.action = na.exclude,
  LDL ~ age + smoking:age - 1, data = hers
)

library(equatiomatic)
equatiomatic::extract_eq(hers_lm_no_int)
```


::: {.landscape}

:::{#tbl-hers-maybe-int layout-ncol=2}

:::{#tbl-hers-with-int}


```{r}
library(gtsummary)
hers_lm_with_int |> 
  tbl_regression(intercept = TRUE)
```
With intercept
:::

:::{#tbl-hers-no-int}

```{r}
hers_lm_no_int |> 
  tbl_regression(intercept = TRUE)
```

No intercept
:::

`hers` data models with and without intercepts

:::

:::

---

:::{#fig-hers-maybe-int layout-ncol=2}


:::{#fig-hers-no-int}

```{r}
library(sjPlot)

hers_plot1 <- hers_lm_no_int |>
  sjPlot::plot_model(
    type = "pred",
    terms = c("age", "smoking"),
    show.data = TRUE
  ) +
  facet_wrap(~group_col, ncol = 1) +
  expand_limits(y = 0) +
  theme(legend.position = "bottom")

hers_plot1
```

No intercept

:::

:::{#fig-hers-with-intercept}

```{r}
library(sjPlot)

hers_plot2 <- hers_lm_with_int |>
  sjPlot::plot_model(
    type = "pred",
    terms = c("age", "smoking"),
    show.data = TRUE
  ) +
  facet_wrap(~group_col, ncol = 1) +
  expand_limits(y = 0) +
  theme(legend.position = "bottom")

hers_plot2
```

With intercept
:::

`hers` data models with and without intercepts

:::

### Residuals {#sec-lm-residuals}

::: notes
Maybe we can transform the data and model in some way 
to make it easier to inspect.
:::

{{< include _subfiles/Linear-models-overview/_def-residual-deviation.qmd >}}

{{< include _subfiles/Linear-models-overview/_def-signal.qmd >}}

---

:::{#thm-gaussian-resid-deviation}
#### Deviations in Gaussian models

If $Y$ has a Gaussian distribution, 
then $\devn(Y)$ also has a Gaussian distribution, and vice versa.
:::

:::{.proof}
Left to the reader.
:::

---

:::{#def-resid-fitted}
#### Residuals

{{< include _subfiles/shared/_def-residual-observed-minus-fitted.qmd >}}
:::

---

:::{#exm-resid-bw}

#### residuals in `birthweight` data

:::{#fig-bw-resids1}

```{r}
plot1_interact +
  facet_wrap(~sex) +
  geom_segment(
    aes(
      x = age,
      y = predlm2,
      xend = age,
      yend = weight,
      col = sex,
      group = id
    ),
    linetype = "dotted"
  )
```

Fitted values and residuals for interaction model for `birthweight` data

:::

:::

---

#### Residuals of fitted values vs subpopulation-mean deviations

The residual is a plug-in estimate
of the deviation of an observation
from its subpopulation mean:

$$
\ba
\devn(y_i) &= y_i - \E{Y_i \mid X_i}
\\
\widehat{\devn(y_i)} &= y_i - \hat y_i
\\
&= \resid_i
\ea
$$

Different sources are not fully consistent
about these terms.
For terminology in this course,
we use:

- **deviation**
  for deviation of an observation from its population mean (typically conditional on covariates).
  This quantity is often called **error** in other sources,
  even though it doesn't directly involve an estimate;
  cf., <https://en.wikipedia.org/wiki/Errors_and_residuals>;
- **estimation error** for deviation of an estimate from its estimand;
- **residual** for deviation of an observed value from a fitted value.

---

#### General characteristics of residuals

:::{#def-hat-matrix}
#### Hat matrix

For an ordinary least squares linear model
with design matrix $\mX$,
the **hat matrix** is

$$H = \mX(\tp{\mX}\mX)^{-1}\tp{\mX}$$

so the fitted values satisfy

$$\hat{\vY} = H\vY$$
:::

:::{#thm-resid-unbiased}
For an ordinary least squares linear model
with fitted values $\hat y = \dprodf{\vx}{\vb}$,
if $\hExpf{Y}$ is an [unbiased](estimation.qmd#sec-unbiased-estimators)
estimator of the mean $\Expf{Y}$, so that:

$$ \Expf{\hExpf{Y}} = \Expf{Y}$$

and if the errors are homoskedastic so that
$\Var{\vY \mid \mX} = \ss \idmat_n$ where $\idmat_n$ is the $n \times n$ identity matrix,
then,
treating $\mX$ as fixed in this derivation,
the residual moments are:

$$\E{\resid \mid \mX} = 0$$ {#eq-mean-resid-unbiased}
$$\Var{\resid_i \mid \mX} = \ss(1-h_{ii})$$ {#eq-var-resid-leverage}

The conditional mean result (@eq-mean-resid-unbiased)
uses unbiasedness of fitted values.
The conditional variance result (@eq-var-resid-leverage)
uses the homoskedasticity condition
with the OLS hat-matrix representation.

When leverage is roughly uniform,
$h_{ii} \approx k/n$,
where $k = \mathrm{rank}(\mX)$
(so $k = p + 1$ for a full-rank model
with an intercept and $p$ predictors),
so $\Var{\resid_i} \approx \ss$
as $n$ grows.

:::

---

:::{.proof}
\ 

@eq-mean-resid-unbiased:

$$
\ba
\E{\resid \mid \mX} &= \E{Y - \hat Y \mid \mX}
\\ &= \E{Y \mid \mX} - \E{\hat Y \mid \mX}
\\ &= \E{Y \mid \mX} - \E{Y \mid \mX}
\\ &= 0
\ea
$$

@eq-var-resid-leverage:

$$
\ba
\resid &= (\idmat_n-H)\vY
\\ \Var{\resid \mid \mX} &= \Var{(\idmat_n-H)\vY \mid \mX}
\\ &= (\idmat_n-H)\Var{\vY \mid \mX}\tp{(\idmat_n-H)}
\\ &= (\idmat_n-H)\ss \idmat_n\tp{(\idmat_n-H)}
\\ &= \ss(\idmat_n-H)\tp{(\idmat_n-H)}
\\ &= \ss(\idmat_n-H)^2
\\ &= \ss(\idmat_n - 2H + H^2)
\\ &= \ss(\idmat_n-H) \qquad (\text{since } H^2 = H)
\\ \Var{\resid_i \mid \mX} &= \ss(1-h_{ii})
\ea
$$
:::

---

#### Characteristics of residuals in Gaussian models

In a correct Gaussian linear model,
the deviations from the model-implied means are i.i.d. Gaussian:

$$
\devn(Y_i) \siid N(0, \sigma^2)
$$

Residuals $\resid_i$ are observed approximations
to these deviations.
Their spread is often summarized using $\hat\sigma^2$.

---

#### Computing residuals in R

R provides a function for residuals:

```{r}
resid(bw_lm2)
```


---

:::{#exr-calc-resids}
Check R's output by computing the residuals directly.
:::

:::{.solution}
\ 
```{r}
bw$weight - fitted(bw_lm2)
```

This matches R's output!
:::

---

#### Graphing the residuals

:::{#fig-resid-bw layout-ncol=2}

:::{#fig-bw-resids1-c2}

```{r}
plot1_interact +
  facet_wrap(~sex) +
  geom_segment(
    aes(
      x = age,
      y = predlm2,
      xend = age,
      yend = weight,
      col = sex,
      group = id
    ),
    linetype = "dotted"
  )
```

fitted values

:::

:::{#fig-resids-intxn-bw}
```{r}

bw <- bw |>
  mutate(
    resids_intxn =
      weight - fitted(bw_lm2)
  )

plot_bw_resid <-
  bw |>
  ggplot(aes(
    x = age,
    y = resids_intxn,
    linetype = sex,
    shape = sex,
    col = sex
  )) +
  theme_bw() +
  xlab("Gestational age (weeks)") +
  ylab("residuals (grams)") +
  theme(legend.position = "bottom") +
  geom_hline(aes(
    yintercept = 0,
    col = sex
  )) +
  geom_segment(
    aes(yend = 0),
    linetype = "dotted"
  ) +
  geom_point()
# expand_limits(y = 0, x = 0) +
geom_point(alpha = .7)
print(plot_bw_resid + facet_wrap(~sex))
```

Residuals

:::

Fitted values and residuals for interaction model for `birthweight` data

:::

---

#### Residuals versus predictors

```{r}
hers <- hers |>
  mutate(
    resids_no_intcpt =
      LDL - fitted(hers_lm_no_int),
    resids_with_intcpt =
      LDL - fitted(hers_lm_with_int)
  )
```

:::{#fig-resids-hers-1 layout-ncol=2}

:::{#fig-resid-hers-pred-no-intcpt} 

```{r}
hers |>
  arrange(age) |>
  ggplot() +
  aes(x = age, y = resids_no_intcpt, col = factor(smoking)) +
  geom_point() +
  geom_hline(aes(yintercept = 0, col = factor(smoking))) +
  facet_wrap(~smoking, labeller = "label_both") +
  theme(legend.position = "bottom") +
  geom_smooth(col = "blue")
```

no intercept
:::

:::{#fig-resid-hers-pred-no-intcpt}

```{r}
hers |>
  arrange(age) |>
  ggplot() +
  aes(x = age, y = resids_with_intcpt, col = factor(smoking)) +
  geom_point() +
  geom_hline(aes(yintercept = 0, col = factor(smoking))) +
  facet_wrap(~smoking, labeller = "label_both") +
  theme(legend.position = "bottom") +
  geom_smooth(col = "blue")
```


with intercept
:::

Residuals of `hers` data vs predictors

:::

---

#### Residuals versus fitted values

::: notes
If the model contains multiple continuous covariates,
how do we check for errors in the mean structure assumption?
:::


```{r}
#| label: fig-resids-intxn-hers
#| fig-cap: "Residuals of interaction model for `hers` data"

library(ggfortify)
hers_lm_no_int |>
  update(na.action = na.omit) |>
  autoplot(
    which = 1,
    ncol = 1,
    smooth.colour = NA
  ) +
  geom_hline(yintercept = 0, col = "red")
```

---

We can add a LOESS smooth to visualize where the residual mean is nonzero:

```{r}
#| label: fig-resids-intxn-hers-smooth-no-int
#| fig-cap: "Residuals of interaction model for `hers` data, no intercept term"

library(ggfortify)
hers_lm_no_int |>
  update(na.action = na.omit) |>
  autoplot(
    which = 1,
    ncol = 1
  ) +
  geom_hline(yintercept = 0, col = "red")
```

---

:::{#fig-resid-hers-compare layout-ncol=2}

:::{#fig-resids-intxn-hers-smooth-no-int-2}

```{r}
library(ggfortify)
hers_lm_no_int |>
  update(na.action = na.omit) |>
  autoplot(
    which = 1,
    ncol = 1
  ) +
  geom_hline(yintercept = 0, col = "red")
```

no intercept term

:::

:::{#fig-resids-intxn-hers-smooth-with-int}

```{r}
hers_lm_with_int |>
  update(na.action = na.omit) |>
  autoplot(
    which = 1,
    ncol = 1
  ) +
  geom_hline(yintercept = 0, col = "red")
```

with intercept term

:::

Residuals of interaction model for `hers` data, with and without intercept term

:::

---

:::{#def-stred}

#### Standardized residuals

For an ordinary linear model,
the estimated standard deviation of the $i$th residual is

$$
\hSD{\resid_i} = \hat{\sigma}\sqrt{1-h_{ii}}
$$

so the standardized residual is

$$
\stdresid_i = \frac{\resid_i}{\hSD{\resid_i}}
= \frac{\resid_i}{\hat{\sigma}\sqrt{1-h_{ii}}}
$$
:::

Hence, with enough data and a correct model,
each standardized residual has an approximately standard Gaussian **marginal** distribution; that is,

$$
\stdresid_i \approx N(0,1)
$$

but they are not independent.
If $H$ is the hat matrix with entries $h_{ij}$,
then

$$
\Cov{\resid_i,\resid_j \mid \mX}
=
\begin{cases}
\ss(1-h_{ii}), & i=j,\\
-\ss h_{ij}, & i\neq j,
\end{cases}
$$

and therefore

$$
\Cov{\stdresid_i,\stdresid_j \mid \mX}
\approx
\begin{cases}
1, & i=j,\\
\dfrac{-h_{ij}}{\sqrt{(1-h_{ii})(1-h_{jj})}}, & i\neq j,
\end{cases}
$$

In particular,
for $i \neq j$,
these covariances are generally nonzero.

### Marginal distributions of residuals

To look for problems with our model,
we can check whether the residuals $\resid_i$
and standardized residuals $\stdresid_i$
look like they have the distributions
that they are supposed to have,
according to the model.

---

#### Standardized residuals in R

```{r}
h_ii <- hatvalues(bw_lm2)
manual_std_resid <- resid(bw_lm2) / (sigma(bw_lm2) * sqrt(1 - h_ii))
manual_std_resid
rstandard(bw_lm2)
```

::: notes
These are nearly the same.
Any differences are from numerical rounding.

:::

---

```{r}
rstandard_compare_plot <-
  tibble(
    x = resid(bw_lm2) / (sigma(bw_lm2) * sqrt(1 - h_ii)),
    y = rstandard(bw_lm2)
  ) |>
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  theme_bw() +
  coord_equal() +
  xlab("resid(bw_lm2)/(sigma(bw_lm2)*sqrt(1-h_ii))") +
  ylab("rstandard(bw_lm2)") +
  geom_abline(
    aes(
      intercept = 0,
      slope = 1,
      col = "x=y"
    )
  ) +
  labs(colour = "") +
  scale_colour_manual(values = "red")

print(rstandard_compare_plot)
```

---

Let's add these residuals to the `tibble` of our dataset:

```{r}
bw <-
  bw |>
  mutate(
    fitted_lm2 = fitted(bw_lm2),
    resid_lm2 = resid(bw_lm2),
    resid_lm2_alt = weight - fitted_lm2,
    std_resid_lm2 = rstandard(bw_lm2),
    std_resid_lm2_alt =
    resid_lm2 / (sigma(bw_lm2) * sqrt(1 - h_ii))
  )

bw |>
  select(
    sex,
    age,
    weight,
    fitted_lm2,
    resid_lm2,
    std_resid_lm2
  )
```

---

::: notes

Now let's build histograms:

:::

```{r}
#| fig-cap: "Marginal distribution of (nonstandardized) residuals"
#| label: fig-marg-dist-resid
resid_marginal_hist <-
  bw |>
  ggplot(aes(x = resid_lm2)) +
  geom_histogram()

print(resid_marginal_hist)
```

::: notes
Hard to tell with this small amount of data, but I'm a bit concerned that the histogram doesn't show a bell-curve shape.

:::

---


```{r}
#| fig-cap: "Marginal distribution of standardized residuals"
#| label: fig-marg-stresd
std_resid_marginal_hist <-
  bw |>
  ggplot(aes(x = std_resid_lm2)) +
  geom_histogram()

print(std_resid_marginal_hist)
```

::: notes
This looks similar, although the scale of the x-axis got narrower, because we divided by $\hat\sigma$ (roughly speaking).

Still hard to tell if the distribution is Gaussian.

:::

---

### QQ plot of standardized residuals

::: notes
Another way to assess whether the standardized residuals follow a standard normal distribution
is a **quantile-quantile (QQ) plot**.

A QQ plot compares the empirical distribution of a dataset to a theoretical reference distribution.
Each point corresponds to one observation:

- The **x-axis** (Theoretical Quantiles) shows the $N(0,1)$ quantile
  corresponding to the plotting position $p_i$
  used for the $i$th ordered residual.
  In R, `qqnorm()` obtains these plotting positions
  from `ppoints(n)`.
- The **y-axis** (Standardized Residuals) shows the observed standardized residuals,
  plotted in increasing order as $r_{(1)}\le\cdots\le r_{(n)}$.
  In QQ-plot terminology,
  these ordered values are [sample quantiles](nonparametric-models.qmd#def-sample-quantile).
  The QQ plot uses points
  $$\left(\Phi^{-1}(p_i),\, r_{(i)}\right), \quad i = 1, \ldots, n,$$
  where $p_i$ denotes the plotting positions returned by `ppoints(n)`.
  For the common midpoint choice $p_i=(i-0.5)/n$,
  the sample quantile at $p_i$ equals $r_{(i)}$.
  More generally,
  `qqnorm()` pairs the $i$th ordered residual
  with the corresponding theoretical quantile
  computed from its plotting position.
  So both labels are correct:
  `qqnorm()` emphasizes the quantile role ("Sample Quantiles"),
  while `autoplot.lm()` emphasizes the underlying ordered values
  ("Standardized residuals").

If the standardized residuals approximately follow $N(0,1)$,
the points should fall approximately on a straight line.
:::

```{r}
library(ggfortify)
# needed to make ggplot2::autoplot() work for `lm` objects

qqplot_lm2_auto <-
  bw_lm2 |>
  autoplot(
    which = 2, # options are 1:6; can do multiple at once
    ncol = 1
  ) +
  theme_classic()

print(qqplot_lm2_auto)
```

::: notes
If the Gaussian model were correct, the points should fall approximately on the reference line.

The **reference line** passes through the pairs of theoretical and empirical quantiles
at probabilities 0.25 and 0.75:
it connects $(\Phi^{-1}(0.25),\, \hat{Q}_{0.25})$ and $(\Phi^{-1}(0.75),\, \hat{Q}_{0.75})$,
where $\Phi^{-1}$ is the standard normal quantile function
and $\hat{Q}_p$ is the $p$-th sample quantile of the standardized residuals.
If the standardized residuals are approximately $N(0,1)$,
this line should be close to the identity line $y = x$,
and the points should fall roughly along a straight line.

Systematic deviations from the line suggest departures from normality:

- **S-shaped curves**: heavier or lighter tails than the normal distribution
- **Concave or convex curves**: skewness
- **Individual points far from the line**: potential outliers

Fig 2.4 panel (c) in @dobson4e is a little different; they didn't specify how they produced it, but other statistical analysis systems do things differently from R.

See also @dunn2018generalized [§3.5.4](https://link.springer.com/chapter/10.1007/978-1-4419-0118-7_3#Sec14:~:text=3.5.4%20Q%E2%80%93Q%20Plots%20and%20Normality).

:::

---

#### QQ plot: typical patterns

:::{#fig-qqplot-patterns}

```{r}
#| fig-height: 5
#| fig-width: 10
set.seed(36) # seed chosen for Fig. 3.6 reproduction
n <- 150     # sample size as in @dunn2018generalized Fig. 3.6

sims <- list(
  "Right skewed"    = rchisq(n, df = 3),
  "Left skewed"     = -rchisq(n, df = 3),
  "Tails too heavy" = rt(n, df = 3),
  "Tails too light" = runif(n, -sqrt(3), sqrt(3)),
  "True normal"     = rnorm(n),
  "Bimodal"         = c(rnorm(n / 2, -2.5), rnorm(n / 2, 2.5))
)

old_par <- par(
  mfcol = c(2, 6),
  mar = c(3, 3, 2, 0.5),
  oma = c(0, 0, 0, 0)
)

for (nm in names(sims)) {
  qqnorm(
    sims[[nm]],
    main = nm,
    xlab = "Theoretical Quantiles",
    ylab = "Sample Quantiles",
    pch = 1,
    cex = 0.6
  )
  qqline(sims[[nm]])
  hist(
    sims[[nm]],
    main = "",
    xlab = "Residuals",
    ylab = "Frequency",
    col = "grey70",
    border = "white",
    breaks = 15
  )
}

par(old_par)
```

Typical QQ-plot patterns for six types of residual distribution.
Adapted from @dunn2018generalized [Fig. 3.6], with thanks to the authors;
reproduced here using simulated data ($n = 150$).
Each column shows one distribution scenario:
a QQ plot (top) paired with a histogram of the simulated data (bottom).

:::

---

#### QQ plot - how it's built

::: notes
To construct a QQ plot by hand:

1. Assign each standardized residual a **plotting position** $p_i$
   based on its rank among the $n$ observations.
   R's default formula for $n > 10$ is:
   $$p_i = \frac{\text{rank}_i - 0.5}{n}$$
2. Compute the **theoretical quantile**: $q_i = \Phi^{-1}(p_i)$,
   the value from $N(0,1)$ that corresponds to probability $p_i$.
3. Plot $(q_i, \stdresid_i)$ — theoretical quantile $q_i$ on the x-axis,
   observed standardized residual $\stdresid_i$ (see @def-stred) on the y-axis.
4. Add the **reference line** through the pairs of theoretical and empirical quantiles
   at probabilities 0.25 and 0.75.
:::

```{r}
bw <- bw |>
  mutate(
    # plotting position: (rank - 0.5) / n
    p = (rank(std_resid_lm2) - 1 / 2) / n(),
    expected_quantiles_lm2 = qnorm(p)
  )

qqplot_lm2 <-
  bw |>
  ggplot(
    aes(
      x = expected_quantiles_lm2,
      y = std_resid_lm2,
      col = sex,
      shape = sex
    )
  ) +
  geom_point() +
  theme_classic() +
  theme(legend.position = "none") + # removing the plot legend
  ggtitle("Normal Q-Q") +
  xlab("Theoretical Quantiles") +
  ylab("Standardized residuals")

# find the reference line through the (theoretical, empirical) quantile pairs
# at probabilities 0.25 and 0.75:

ps <- c(.25, .75) # reference probabilities
a <- quantile(rstandard(bw_lm2), ps) # empirical quantiles
b <- qnorm(ps) # theoretical quantiles

qq_slope <- diff(a) / diff(b)
qq_intcpt <- a[1] - b[1] * qq_slope

qqplot_lm2 <-
  qqplot_lm2 +
  geom_abline(slope = qq_slope, intercept = qq_intcpt)
```

:::{#fig-qqplot-comparison layout-ncol=3}

:::{#fig-qqplot-manual}

```{r}
print(qqplot_lm2)
```

`ggplot2` (manual construction)

:::

:::{#fig-qqplot-baser}

```{r}
std_resid_lm2 <- rstandard(bw_lm2)

qqnorm(
  std_resid_lm2,
  main = "Normal Q-Q",
  xlab = "Theoretical Quantiles",
  ylab = "Standardized residuals"
)
qqline(std_resid_lm2)
```

base R (`qqnorm` + `qqline`)

:::

:::{#fig-qqplot-autoplot}

```{r}
library(ggfortify)
autoplot(
  bw_lm2,
  which = 2,
  ncol = 1
) +
  theme_classic()
```

`ggfortify::autoplot()`

:::

Three equivalent ways to produce a QQ plot of the standardized residuals
for the `birthweight` model (@eq-BW-lm-interact).
All three plots show the same data and reference line.

:::

---

### Conditional distributions of residuals

If our Gaussian linear regression model is correct, the residuals $\resid_i$ and standardized residuals $\stdresid_i$ should have:

-   an approximately Gaussian distribution, with:
-   a mean of 0
-   a constant variance

This should be true **for every** value of $x$.

---

If we didn't correctly guess the functional form of the linear component of the mean,
for covariate vector $x = (x_1,\ldots,x_p)$,
$$\E{Y \mid X=x} = \b_0 + \b_1 x_1 + ... + \b_p x_p$$

Then the residuals might have nonzero mean.

Regardless of whether we guessed the mean function correctly, the variance of the residuals might differ between values of $x$.

---

#### Residuals versus fitted values

::: notes
To look for these issues, we can plot the residuals $\resid_i$ against the fitted values $\hat y_i$ (@fig-bw_lm2-resid-vs-fitted).
:::

:::{#fig-bw_lm2-resid-vs-fitted}

```{r}
autoplot(bw_lm2, which = 1, ncol = 1) |> print()
```

`birthweight` model  (@eq-BW-lm-interact): residuals versus fitted values
:::

::: notes
If the model is correct, the blue line should stay flat and close to 0, and the cloud of dots should have the same vertical spread regardless of the fitted value.

If not,
we probably need to change the functional form
of the linear component of the mean,
$$\E{Y \mid X=x} = \b_0 + \b_1 x_1 + ... + \b_p x_p$$

:::

---

{{< include exm-plot-fits.qmd >}}
---

#### Scale-location plot

::: notes
We can also plot the square roots of the absolute values of the standardized residuals against the fitted values (@fig-bw-scale-loc).
:::

```{r}
#| label: fig-bw-scale-loc
#| fig-cap: "Scale-location plot of `birthweight` data"
autoplot(bw_lm2, which = 3, ncol = 1) |> print()
```

::: notes
Here, the blue line doesn't need to be near 0, 
but it should be flat. 
If not, the residual variance $\sigma^2$ might not be constant, 
and we might need to transform our outcome $Y$ 
(or use a model that allows non-constant variance).
:::

---

#### Residuals versus leverage

::: notes

We can also plot our standardized residuals against "leverage", which roughly speaking is a measure of how unusual each $x_i$ value is. Very unusual $x_i$ values can have extreme effects on the model fit, so we might want to remove those observations as outliers, particularly if they have large residuals.

:::

::: {#fig-bw_lm2_resid-vs-leverage}

```{r}
autoplot(bw_lm2, which = 5, ncol = 1) |> print()
```

`birthweight` model with interactions (@eq-BW-lm-interact): residuals versus leverage
:::

::: notes
The blue line should be relatively flat and close to 0 here.
:::

---

### Diagnostics constructed by hand

```{r}
bw <-
  bw |>
  mutate(
    predlm2 = predict(bw_lm2),
    residlm2 = weight - predlm2,
    std_resid = residlm2 / sigma(bw_lm2),
    # std_resid_builtin = rstandard(bw_lm2), # uses leverage
    sqrt_abs_std_resid = std_resid |> abs() |> sqrt()
  )
```

##### Residuals vs fitted

```{r}
resid_vs_fit <- bw |>
  ggplot(
    aes(x = predlm2, y = residlm2, col = sex, shape = sex)
  ) +
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0)

print(resid_vs_fit)
```


##### Standardized residuals vs fitted

```{r}
bw |>
  ggplot(
    aes(x = predlm2, y = std_resid, col = sex, shape = sex)
  ) +
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0)
```

##### Standardized residuals vs gestational age

```{r}
bw |>
  ggplot(
    aes(x = age, y = std_resid, col = sex, shape = sex)
  ) +
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0)
```

##### `sqrt(abs(rstandard()))` vs fitted

Compare with `autoplot(bw_lm2, 3)`

```{r}
bw |>
  ggplot(
    aes(x = predlm2, y = sqrt_abs_std_resid, col = sex, shape = sex)
  ) +
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0)
```
