::: notes
Just as the birthweight parallel-lines model assumes that sex shifts the intercept
but not the age slope (parallel lines in 2D),
the parallel-planes model for the HERS data assumes a flat regression surface
over the age-BMI space: the slope of mean LDL with respect to BMI
is the same at every age, and the slope with respect to age is the same at every BMI.

We use age ($A$) and BMI ($B$) as predictors---not hormone therapy ($H$)---because
LDL is measured at **baseline**, before treatment was administered.
:::

$$
\ba
Y|A,B &\sciid N(\mu(A,B), \sigma^2)\\
\mu(a,b) &= \beta_0 + \beta_A a + \beta_B b
\ea
$$ {#eq-hers-parallel}

::: {#tbl-hers-lm1}

```{r}
#| code-fold: true
hers_lm1 <- lm(
  formula = LDL ~ age + BMI,
  data = hers,
  na.action = na.exclude
)
hers_plot_data <- hers |>
  dplyr::filter(!is.na(age), !is.na(BMI), !is.na(LDL))

hers_lm1 |>
  parameters::parameters() |>
  parameters::print_md(
    select = "{estimate}"
  )
```

Regression parameter estimates for [Model @eq-hers-parallel]

:::

::::: {.panel-tabset}

### Statins: No

:::{#fig-hers-parallel-fit-no}

```{r}
#| code-fold: true
hers_no <- hers_plot_data |> dplyr::filter(statins == "no")

plotly::plot_ly(
  x = hers_no[["age"]],
  y = hers_no[["BMI"]],
  z = hers_no[["LDL"]],
  type = "scatter3d",
  mode = "markers",
  name = "No statins",
  marker = list(size = 3, opacity = 0.3, color = "steelblue")
) |>
  regress3d::add_3d_surface(
    model = hers_lm1,
    data = hers_plot_data,
    showlegend = TRUE
  ) |>
  plotly::layout(
    scene = list(
      xaxis = list(title = "Age (yr)"),
      yaxis = list(title = "BMI (kg/m\u00b2)"),
      zaxis = list(title = "LDL (mg/dL)")
    )
  )
```

Fitted regression surface from [Model @eq-hers-parallel]
(patients not taking statins)

:::

### Statins: Yes

:::{#fig-hers-parallel-fit-yes}

```{r}
#| code-fold: true
hers_yes <- hers_plot_data |> dplyr::filter(statins == "yes")

plotly::plot_ly(
  x = hers_yes[["age"]],
  y = hers_yes[["BMI"]],
  z = hers_yes[["LDL"]],
  type = "scatter3d",
  mode = "markers",
  name = "Yes statins",
  marker = list(size = 3, opacity = 0.3, color = "darkorange")
) |>
  regress3d::add_3d_surface(
    model = hers_lm1,
    data = hers_plot_data,
    showlegend = TRUE
  ) |>
  plotly::layout(
    scene = list(
      xaxis = list(title = "Age (yr)"),
      yaxis = list(title = "BMI (kg/m\u00b2)"),
      zaxis = list(title = "LDL (mg/dL)")
    )
  )
```

Fitted regression surface from [Model @eq-hers-parallel]
(patients taking statins)

:::

:::::

The coefficient estimates in @tbl-hers-lm1 quantify the shape of these surfaces.
Both tabs display the same fitted flat plane: the parallel-planes model constrains the age and BMI slopes to be the same regardless of statin use.
The statins-yes tab reveals that patients taking statins tend to have lower LDL values on average ---
a shift the parallel-planes model cannot capture, because it does not include statin use as a predictor.
