
:::{#def-predicted-value}
#### Predicted value

In a regression model $\p(y|\vx)$, 
the **predicted value** of $y$ given $\vx$ is the estimated mean of $Y$ given $\vX=\vx$:

$$\hat y \eqdef \hExpf{Y|\vX=\vx}$$
:::

---

For linear models, 
the predicted value can be straightforwardly calculated 
by multiplying each predictor value $x_j$ by its corresponding coefficient $\beta_j$ 
and adding up the results:

$$
\begin{aligned}
\hat y &= \hExpf{Y|\vX=\vx} \\
&= \tp{\vx}\hat\beta \\
&= \hat\beta_0\cdot 1 + \hat\beta_1 x_1 + ... + \hat\beta_p x_p
\end{aligned}
$$

---

:::{#def-fitted-value}
#### Fitted value

For a given dataset $(\vY, \mX)$ and corresponding fitted model $\p_{\hb}(\vy|\mx)$, 
the **fitted value** of $y_i$ is 
the @def-predicted-value of $y$ when $\vX=\vx_i$ using the estimated parameters $\hb$:

$$\hat y_i \eqdef \hExpf{Y_i|\vX=\vx_i} = \tp{\vxi}\hb$$
:::

---

#### Example: prediction for the `birthweight` data

```{r}
x <- c(1, 1, 40)
sum(x * coef(bw_lm1))
```

R has built-in functions for prediction:

```{r}
x <- tibble(age = 40, sex = "male")
bw_lm1 |> predict(newdata = x)
```

If you don't provide `newdata`, 
R will use the covariate values from the original dataset:

```{r}
predict(bw_lm1)
```

These special predictions are called the **fitted values** of the dataset
(see @def-fitted-value).

R has an extra function to get these values:

```{r}
fitted(bw_lm1)
```
