::: notes
In the parallel-planes model, the BMI slope is the same regardless of age.
Just as we added a sex-age interaction to the birthweight model to allow
the age slope to differ between males and females,
we now add an age-BMI interaction to allow the BMI slope to vary with age.
:::

$$
\ba
Y|A,B &\sciid N(\mu(A,B), \sigma^2)\\
\mu(a,b) &= \beta_0 + \beta_A a + \beta_B b + \beta_{AB}(a \cdot b)
\ea
$$ {#eq-hers-interact}

The slope of mean LDL with respect to BMI now depends on age:

$$
\ba
\deriv{b}\mu(A=\red{0}, B=b) &= \beta_B + \beta_{AB} \cdot \red{0} = \beta_B \\
\deriv{b}\mu(A=\red{a}, B=b) &= \beta_B + \beta_{AB} \red{a}
\ea
$$

::: notes
So the slope of LDL with respect to BMI changes by $\beta_{AB}$ for each one-year increase in age.
:::

::: {#tbl-hers-lm2}

```{r}
#| code-fold: true
hers_lm2 <- lm(
  LDL ~ age + BMI + age:BMI,
  data = hers,
  na.action = na.exclude
)

hers_lm2 |>
  parameters::parameters() |>
  parameters::print_md(
    select = "{estimate}"
  )
```

Regression parameter estimates for [Model @eq-hers-interact]

:::

::::: {.panel-tabset}

### Statins: No

:::{#fig-hers-interact-fit-no}

```{r}
#| code-fold: true
plotly::plot_ly(
  x = hers_no[["age"]],
  y = hers_no[["BMI"]],
  z = hers_no[["LDL"]],
  type = "scatter3d",
  mode = "markers",
  name = "No statins",
  marker = list(size = 3, opacity = 0.3, color = "steelblue")
) |>
  regress3d::add_3d_surface(
    model = hers_lm2,
    data = hers_plot_data,
    showlegend = TRUE
  ) |>
  plotly::layout(
    scene = list(
      xaxis = list(title = "Age (yr)"),
      yaxis = list(title = "BMI (kg/m\u00b2)"),
      zaxis = list(title = "LDL (mg/dL)")
    )
  )
```

Fitted regression surface from [Model @eq-hers-interact]
(patients not taking statins)

:::

### Statins: Yes

:::{#fig-hers-interact-fit-yes}

```{r}
#| code-fold: true
plotly::plot_ly(
  x = hers_yes[["age"]],
  y = hers_yes[["BMI"]],
  z = hers_yes[["LDL"]],
  type = "scatter3d",
  mode = "markers",
  name = "Yes statins",
  marker = list(size = 3, opacity = 0.3, color = "darkorange")
) |>
  regress3d::add_3d_surface(
    model = hers_lm2,
    data = hers_plot_data,
    showlegend = TRUE
  ) |>
  plotly::layout(
    scene = list(
      xaxis = list(title = "Age (yr)"),
      yaxis = list(title = "BMI (kg/m\u00b2)"),
      zaxis = list(title = "LDL (mg/dL)")
    )
  )
```

Fitted regression surface from [Model @eq-hers-interact]
(patients taking statins)

:::

:::::

Unlike the flat parallel-planes surface, the interaction model produces a twisted regression surface.
The BMI slope changes with age: at any given age, the BMI slope equals the base BMI coefficient plus the interaction coefficient times that age.
Both tabs plot the same interaction-model surface; statin use is used only to partition the observed data points for visual clarity.
