{{< include latex-macros/macros.qmd >}}

:::{#exr-ci-mean}

Given
a maximum likelihood estimate $\hvb$
and
a corresponding estimated covariance matrix
$\hS \eqdef \widehat{\Covt}(\hvb)$,
calculate a 95% confidence interval for the predicted mean
$\mu(\vx) = \E{Y|\vX=\vx}$.

:::

---

:::{#sol-ci-mean}

::: notes

By [the Gauss-Markov theorem](https://en.wikipedia.org/wiki/Gauss%E2%80%93Markov_theorem),
the OLS estimate $\hvb$ is the best linear unbiased estimator of $\vb$.
A 95% confidence interval for $\mu(\vx)$
can be constructed directly on the original scale:

$$\hat\mu(\vx) \pm t_{n-p}(0.975) \cdot \HSE{\hat\mu(\vx)}$$

where $\hat\mu(\vx) = \vx'\hvb$.

Unlike in logistic regression,
no transformation is needed;
the confidence interval is constructed directly on the mean scale.

:::

$$\mu(\vx) \in \paren{\hat\mu(\vx) \pm t_{n-p}(0.975) \cdot \HSE{\hat\mu(\vx)}}$$

:::

---

:::{#exr-se-mean}

::: notes

How can we estimate the standard error of
$\hat\mu(\vx)$?

:::

$$\blue{\HSE{\hat\mu(\vx)}} = \blue{?}$$

:::

---

:::{#sol-se-mean}

$$
\green{\SE{\hat\mu(\vx)}} =
\sqrt{\red{\Var{\hat\mu(\vx)}}}
$${#eq-se-sqrt-var-mean}

By the definition $\hat\mu(\vx) = \vx'\hvb$
and [the variance of a linear combination](probability.qmd#thm-var-lincom):

$$
\ba
\red{\Var{\hat\mu(\vx)}}
&= \Var{\vx'\hvb}
\\
&= \vx'\Cov{\hvb}\vx
\\
&= \red{\tp{\vx}\Sigma\vx}
\ea
$${#eq-var-mean}

where $\Sigma \eqdef \Covt(\hvb)$.

---

::: notes

Expanding @eq-var-mean out of matrix-vector notation, we have:

:::

$$
\ba
\red{\tp{\vx}\Sigma\vx}
&= \sumip\sumjp x_i \Sigma_{ij} x_j
\\
&= \red{\sumip\sumjp x_i \Covt(\hvb_i,\hvb_j) x_j}
\ea
$$

---

Combining @eq-var-mean and the Gauss-Markov theorem:

:::{#thm-est-se-mean}

#### Estimated variance and standard error of predicted mean

$$
\orange{\hVar{\hat\mu(\vx)}} = \orange{\tp{\vx}\hat{\Sigma}\vx}
$${#eq-est-var-mean}

$$
\blue{\HSE{\hat\mu(\vx)}} =
\blue{\sqrt{\tp{\vx}\hat{\Sigma}\vx}}
$${#eq-est-se-mean}

:::

::: notes

Note: on the RHS, we have plugged in $\hat{\Sigma}$,
our estimate of $\Sigma$.

:::

:::

---

#### In R

In R, `predict()` with `se.fit = TRUE`
computes the estimated mean $\hat\mu(\vx) = \vx'\hvb$
and its estimated standard error for each covariate pattern:

```{r}
#| label: tbl-bw-pred-mean
#| tbl-cap: "Predicted means and 95% CI for `birthweight` linear regression"
library(dplyr)
new_data <- tibble(
  age = c(36, 38, 40),
  sex = "male"
)

pred_mean <-
  bw_lm2 |>
  predict(
    newdata = new_data,
    se.fit = TRUE
  )

new_data |>
  mutate(
    mu_hat = pred_mean$fit,
    se = pred_mean$se.fit,
    ci_lower = mu_hat - qt(0.975, df = bw_lm2$df.residual) * se,
    ci_upper = mu_hat + qt(0.975, df = bw_lm2$df.residual) * se
  ) |>
  knitr::kable(digits = 3)
```

Alternatively, `predict()` with `interval = "confidence"` gives the same result:

```{r}
#| label: tbl-bw-pred-ci
#| tbl-cap: "Predicted means and 95% CI using `interval = 'confidence'`"
bw_lm2 |>
  predict(
    newdata = new_data,
    interval = "confidence"
  ) |>
  cbind(new_data) |>
  knitr::kable(digits = 3)
```

---

:::{#fig-pred-int}

```{r}
library(sjPlot)
bw_lm2 |>
  plot_model(type = "pred", terms = c("age", "sex"), show.data = TRUE) +
  theme_sjplot() +
  theme(legend.position = "bottom")
```

Predicted values and confidence bands for the `birthweight` model with interaction term
:::

### Why are confidence bands narrower near the center of the data? {#sec-why-ci-narrower}

The confidence bands in @fig-pred-int are visibly narrower
near the center of the gestational age range.
To understand why,
consider a simple linear regression with one predictor $a$ (gestational age)
and an intercept.
The covariate vector for a new observation at age $a$ is $\vx = \tp{(1, a)}$,
and the general variance formula for the predicted mean specializes to:

$$
\ba
\Var{\hat\mu(a)}
&= \sigma^2\, \tp{\vx}(\tp{\mX}\mX)^{-1}\vx \\
&= \red{\sigma^2 \left(\frac{1}{n} + \frac{(a - \bar{a})^2}{S_{AA}}\right)}
\ea
$${#eq-var-fitted-slr}

where $\bar{a} = \frac{1}{n}\sumin a_i$ is the mean gestational age
and $S_{AA} = \sumin (a_i - \bar{a})^2$ is the total variation in gestational age.

To derive @eq-var-fitted-slr,
first note that for $n$ observations with design matrix rows $(1, a_i)$:

$$
\tp{\mX}\mX
= \begin{pmatrix} n & n\bar{a} \\ n\bar{a} & \sumin a_i^2 \end{pmatrix}
$$

The determinant of $\tp{\mX}\mX$ is:

$$
\ba
\det(\tp{\mX}\mX)
&= n \cdot \sumin a_i^2 - (n\bar{a})^2 \\
&= n\left(\sumin a_i^2 - n\bar{a}^2\right) \\
&= n\, S_{AA}
\ea
$$

so its inverse is:

$$
(\tp{\mX}\mX)^{-1}
= \frac{1}{n\, S_{AA}}
  \begin{pmatrix} \sumin a_i^2 & -n\bar{a} \\ -n\bar{a} & n \end{pmatrix}
$$

Substituting $\vx = \tp{(1, a)}$ into the quadratic form:

$$
\ba
\tp{\vx}(\tp{\mX}\mX)^{-1}\vx
&= \frac{1}{n\, S_{AA}}
  (1,\; a)
  \begin{pmatrix} \sumin a_i^2 & -n\bar{a} \\ -n\bar{a} & n \end{pmatrix}
  \begin{pmatrix} 1 \\ a \end{pmatrix} \\
&= \frac{\sumin a_i^2 - 2n\bar{a}\,a + n\,a^2}{n\, S_{AA}} \\
&= \frac{\left(\sumin a_i^2 - n\bar{a}^2\right) + n(a - \bar{a})^2}{n\, S_{AA}} \\
&= \frac{S_{AA} + n(a - \bar{a})^2}{n\, S_{AA}} \\
&= \frac{1}{n} + \frac{(a - \bar{a})^2}{S_{AA}}
\ea
$$

Therefore the estimated standard error of the predicted mean at age $a$ is:

$$
\HSE{\hat\mu(a)}
= \hs \sqrt{\frac{1}{n} + \frac{(a - \bar{a})^2}{S_{AA}}}
$${#eq-se-fitted-slr}

Since $(a - \bar{a})^2 \geq 0$ and equals zero when $a = \bar{a}$,
the standard error is minimized at the mean gestational age $\bar{a}$
and increases as $a$ moves away from $\bar{a}$ in either direction.
Consequently, confidence intervals are narrowest at $\bar{a}$
and widen toward the extremes of the gestational age range.

Intuitively, the fitted line is "anchored" at the center of the data:
in simple linear regression with an intercept,
the OLS fitted line passes exactly through the sample mean point $(\bar{a}, \bar{y})$,
so the estimated mean at $\bar{a}$ is relatively stable across different samples.
Moving away from $\bar{a}$,
small changes in the estimated slope
cause the fitted line to "pivot" around that anchor,
producing larger changes in the predicted mean
the further $a$ is from $\bar{a}$.

Additionally, there is typically more nearby data near the center of the covariate range,
so we have more information about the true mean response there.
Near the edges of the covariate range,
there is less nearby data,
leaving us less confident in our estimate of the mean response at those values.
