
(c.f. @dobson4e §2.4.4)

---

::: notes
We can create ordinal variables in R using the `ordered()` function^[or equivalently, `factor(ordered = TRUE)`].
:::

:::{#exm-ordinal-variable}

The `hers` data [@vittinghoff2e] is available in the `rmb` package:

```{r}
hers <- rmb::hers |> haven::zap_labels()
```

::: {#tbl-hers-ordinal}

```{r}
#| code-fold: true
hers |> head()
```

HERS data
:::

The `physact` variable records each participant's self-reported physical activity,
relative to other women her age,
on an ordered five-level scale
(1 = much less active, …, 5 = much more active).
We make it an ordered factor so that `lm()` uses polynomial contrasts,
which respect the ordering,
instead of the default treatment contrasts:

```{r}
hers <- hers |>
  mutate(physact_ord = ordered(physact))

lm_physact <- lm(LDL ~ physact_ord, data = hers)
```

::: {#tbl-hers-ordinal-fit}

```{r}
#| code-fold: true
lm_physact |>
  parameters::model_parameters() |>
  parameters::print_md()
```

Mean LDL by physical-activity level, using polynomial contrasts
:::

With polynomial contrasts,
each coefficient summarizes a trend component across the ordered levels:
the linear term is the steady change in mean LDL per step up the scale,
and the quadratic and higher terms capture curvature.
Here the linear-trend coefficient is small and not statistically significant,
so these data show little evidence of a monotone association
between physical-activity level and mean LDL cholesterol.

:::

---

::: notes
For ordinal variables, 
we might want to use contrasts that respect the ordering of the categories.
The default treatment contrasts don't account for ordering.
:::

::: {.callout-note}
## Working with ordinal variables

When working with ordinal covariates in linear models:

1. Use `ordered()` or `factor(ordered = TRUE)` to create the variable
2. Consider using polynomial contrasts (`contr.poly`) which respect ordering
3. Alternatively, treat the ordinal variable as numeric if equal spacing is reasonable
4. Check `?codingMatrices::contr.diff` for additional contrast options

See @dobson4e §2.4.4 for more details on contrasts for ordinal variables.
:::
