
(adapted from @dobson4e §6.3.3; for more information on prediction, see @james2013introduction and @rms2e).

::: notes
If we have a lot of covariates in our dataset, we might want to choose a small subset to use in our model.

There are a few possible metrics to consider for choosing a "best" model.
:::

### DAGs for variable selection

For explanatory models,
variable inclusion should not rely on automated algorithms alone.
Directed acyclic graphs (DAGs)
can help us encode substantive assumptions
about which variables are confounders,
mediators,
or colliders.

In a Dobson-style workflow,
we use the DAG first
to decide a defensible adjustment set,
then compare candidate regression models
within that set using predictive or likelihood-based criteria.
This keeps model selection aligned with study design,
rather than only with numerical fit.

### Mean squared error

We might want to minimize the **mean squared error**,
$\E{(y-\hat y)^2}$,
for new observations
that weren't in our data set when we fit the model.

Unfortunately,
$$\frac{1}{n}\sum_{i=1}^n (y_i-\hat y_i)^2$$
gives a biased estimate of $\E{(y-\hat y)^2}$ for new data.
If we want an unbiased estimate,
we will have to be clever.

This is one reason that $R^2$ is not enough for model selection.
$R^2$ does not decrease as we add explanatory variables,
even when those variables do not improve out-of-sample prediction.
That can lead to overfitting.

With a training/test split,
we estimate coefficients in the training data
and compute prediction error in held-out data:

$$
\hat\b_{\text{train}} =
(\tp{X_{\text{train}}}X_{\text{train}})^{-1}
\tp{X_{\text{train}}}y_{\text{train}}
$$

$$
\hat e_{i,\text{test}} =
y_{\text{test},i} -
\cb{
\hat\b_{0,\text{train}} +
\sum_{j=1}^p \hat\b_{j,\text{train}}x_{\text{test},ij}
},
\quad i \in \text{test set}
$$

$$
\text{RMSE} = \sqrt{
\frac{1}{n_{\text{test}}}
\sum_{i \in \text{test set}} (\hat e_{i,\text{test}})^2
}
$$

#### Train/validation/test splits

In many applications,
we split the data into
training,
validation,
and test sets.
This extends the basic train/test split
by separating model tuning
from final model assessment.
This framework follows
[@james2021islr2e, pp. 198-201].

- Use the **training set** to estimate model parameters.
- Use the **validation set** to compare candidate models,
  choose transformations,
  or tune hyperparameters.
- Use the **test set** once at the end
  to estimate final out-of-sample performance.

Keeping the test set untouched during model building
helps avoid optimistic bias
from repeatedly trying many models.
If data are limited,
$k$-fold cross-validation
can replace a single validation split
for more stable tuning.

:::{#exm-train-validation-test-split}
#### Numerical example

The example below
uses R's built-in `mtcars` dataset
($n=32$ cars)
to predict fuel efficiency (`mpg`)
from vehicle weight (`wt`).
It uses one random split
to compare a linear model
(`mpg ~ wt`)
and a quadratic model
(`mpg ~ wt + I(wt^2)`)
on a validation set,
then reports the chosen model's test RMSE
on untouched test data
[@james2021islr2e, p. 213].

```{r}
#| code-fold: false
set.seed(108)
n <- nrow(mtcars)

n_train <- floor(0.6 * n)
n_valid <- floor(0.2 * n)

idx_train <- sample.int(n, size = n_train)
idx_remaining <- setdiff(seq_len(n), idx_train)
idx_valid <- sample(idx_remaining, size = n_valid)
idx_test <- setdiff(idx_remaining, idx_valid)

split_sizes <- tibble::tibble(
  split = c("training", "validation", "test"),
  n = c(length(idx_train), length(idx_valid), length(idx_test))
)
```

:::{#tbl-train-validation-test-split-sizes}
```{r}
split_sizes
```

Sizes of the training,
validation,
and test splits
:::

```{r}
train_dat <- mtcars[idx_train, ]
valid_dat <- mtcars[idx_valid, ]
test_dat <- mtcars[idx_test, ]

model_linear <- lm(mpg ~ wt, data = train_dat)
model_quadratic <- lm(mpg ~ wt + I(wt^2), data = train_dat)

compute_rmse <- function(model, data) {
  sqrt(mean((data$mpg - predict(model, newdata = data))^2))
}

validation_results <- tibble::tibble(
  model = c("linear", "quadratic"),
  validation_RMSE = c(
    compute_rmse(model_linear, valid_dat),
    compute_rmse(model_quadratic, valid_dat)
  )
)

chosen_model_name <-
  validation_results |>
  dplyr::arrange(validation_RMSE) |>
  dplyr::slice(1) |>
  dplyr::pull(model)

model_labels <- c(
  linear = "linear model",
  quadratic = "quadratic model"
)

chosen_model_label <- model_labels[[chosen_model_name]]

candidate_models <- list(
  linear = model_linear,
  quadratic = model_quadratic
)

chosen_model <- candidate_models[[chosen_model_name]]

performance_comparison <- tibble::tibble(
  split = c("validation", "test"),
  model = chosen_model_name,
  RMSE = c(
    min(validation_results$validation_RMSE),
    compute_rmse(chosen_model, test_dat)
  )
)
```

:::{#fig-linear-model-superimposed}
```{r}
#| fig-cap: ""
plot(
  train_dat$wt,
  train_dat$mpg,
  xlab = "wt",
  ylab = "mpg",
  main = "Linear model superimposed on training data",
  pch = 19,
  col = "grey40"
)
abline(model_linear, col = "blue", lwd = 2)
```

Linear model fit
superimposed on the training data
:::

:::{#fig-quadratic-model-superimposed}
```{r}
#| fig-cap: ""
plot(
  train_dat$wt,
  train_dat$mpg,
  xlab = "wt",
  ylab = "mpg",
  main = "Quadratic model superimposed on training data",
  pch = 19,
  col = "grey40"
)
wt_grid <- seq(min(train_dat$wt), max(train_dat$wt), length.out = 200)
quadratic_pred <- predict(
  model_quadratic,
  newdata = data.frame(wt = wt_grid)
)
lines(wt_grid, quadratic_pred, col = "blue", lwd = 2)
```

Quadratic model fit
superimposed on the training data
:::

:::{#tbl-train-validation-results}
```{r}
validation_results
```

Validation RMSE for candidate models
used to choose the final model
:::

The chosen model is
the `r chosen_model_label`,
because it has the lower validation RMSE.

:::{#tbl-train-validation-test-rmse}
```{r}
performance_comparison
```

Validation and test RMSE
for the chosen model
:::

:::

---

#### Cross-validation

Rather than one arbitrary split,
$k$-fold cross-validation repeatedly partitions the data into training and test folds.
For each split we compute prediction error,
then summarize errors across folds and replications.
The preferred model has lower prediction error,
with simpler models favored when errors are similar.

When the number of candidate explanatory variables is small,
we can compare all possible subsets ($2^p$ models)
using the same cross-validation metric.

```{r}
data("carbohydrate", package = "dobson")
library(cvTools)
full_model <- lm(carbohydrate ~ ., data = carbohydrate)
cv_full <-
  full_model |> cvFit(
    data = carbohydrate, K = 5, R = 10,
    y = carbohydrate$carbohydrate
  )

reduced_model <- full_model |> update(formula = ~ . - age)

cv_reduced <-
  reduced_model |> cvFit(
    data = carbohydrate, K = 5, R = 10,
    y = carbohydrate$carbohydrate
  )
```

---

```{r}
results_reduced <-
  tibble(
    model = "wgt+protein",
    errs = cv_reduced$reps[]
  )
results_full <-
  tibble(
    model = "wgt+age+protein",
    errs = cv_full$reps[]
  )

cv_results <-
  bind_rows(results_reduced, results_full)

cv_results |>
  ggplot(aes(y = model, x = errs)) +
  geom_boxplot()
```

---

##### comparing metrics

```{r}
compare_results <- tribble(
  ~model, ~cvRMSE, ~r.squared, ~adj.r.squared, ~trainRMSE, ~loglik,
  "full",
  cv_full$cv,
  summary(full_model)$r.squared,
  summary(full_model)$adj.r.squared,
  sigma(full_model),
  logLik(full_model) |> as.numeric(),
  "reduced",
  cv_reduced$cv,
  summary(reduced_model)$r.squared,
  summary(reduced_model)$adj.r.squared,
  sigma(reduced_model),
  logLik(reduced_model) |> as.numeric()
)

compare_results
```

---

```{r}
anova(full_model, reduced_model)
```

---

#### Best subset selection

When the number of candidate predictors is modest,
we can use **best subset selection**.
For each model size $k = 0, 1, \ldots, p$,
we fit all $\binom{p}{k}$ models
and keep the best model of size $k$
(e.g., by lowest RSS in the training data).

This gives at most $p+1$ candidate models to compare across model sizes,
using criteria such as
cross-validated prediction error,
$C_p$,
AIC/BIC,
or adjusted $R^2$.
In that sense,
best subset selection is more exhaustive
than one-path methods like forward or backward stepwise selection.

::: notes
For a full algorithmic treatment,
including validation and criterion-based comparison across model size,
see the ISLR treatment of best subset selection
[@james2013introduction].
:::

```{r}
hers_subset <- fs::path_package("rme", "extdata/hersdata.dta") |>
  haven::read_dta() |>
  dplyr::select(LDL, age, weight, BMI, HDL, TG, SBP) |>
  tidyr::drop_na()

hers_lm_subset <- lm(
  LDL ~ age + weight + BMI + HDL + TG + SBP,
  data = hers_subset
)

hers_best_subset <- olsrr::ols_step_best_subset(hers_lm_subset)

hers_best_subset$metrics |>
  dplyr::arrange(dplyr::desc(adjr)) |>
  dplyr::slice_head(n = 5)
```

---

#### Stepwise regression

Stepwise methods are another common approach.
Forward selection adds variables one at a time.
Backward selection starts with the full model
and removes variables sequentially.
Both approaches can select different models from the same data.

::: {.callout-warning}
## Caution about stepwise selection

Stepwise regression has several known problems:

- It tends to select too many variables (overfitting)
- P-values and confidence intervals are biased after selection
- It ignores model uncertainty
- Results can be unstable across different samples

Consider using cross-validation, penalized methods (like Lasso), 
or subject-matter knowledge instead.
See @rms2e and @heinze2018variable for more discussion.
:::

```{r}
library(olsrr)
olsrr:::ols_step_both_aic(full_model)
```

---

#### Lasso

Lasso is a penalized regression method
that shrinks coefficient estimates toward zero.
It adds an $L_1$ penalty term
to the objective function,
creating a trade-off between model fit and parsimony.
The intercept is not penalized.

As the tuning parameter $\lambda$ increases,
more coefficients are shrunken strongly,
and some become exactly zero.
So lasso both regularizes the model
and performs variable selection.
In practice,
$\lambda$ is usually chosen by cross-validation
to minimize prediction error.

For Gaussian linear models,
the penalized least-squares forms are:

$$
\hat\b^{\text{lasso}} =
\arg\min_{\b_0,\ldots,\b_p}
\cb{
\sum_{i=1}^n
\cb{
y_i - \b_0 - \sum_{j=1}^p \b_j x_{ij}
}^2
+
\lambda \sum_{j=1}^p |\b_j|
}
$$

$$
\hat\b^{\text{ridge}} =
\arg\min_{\b_0,\ldots,\b_p}
\cb{
\sum_{i=1}^n
\cb{
y_i - \b_0 - \sum_{j=1}^p \b_j x_{ij}
}^2
+
\lambda \sum_{j=1}^p \b_j^2
}
$$

$$
\hat\b^{\text{elastic-net}} =
\arg\min_{\b_0,\ldots,\b_p}
\cb{
\sum_{i=1}^n
\cb{
y_i - \b_0 - \sum_{j=1}^p \b_j x_{ij}
}^2
+
\lambda
\cb{
\a \sum_{j=1}^p |\b_j|
+
\frac{1-\a}{2}\sum_{j=1}^p \b_j^2
}
}
$$

```{r}
library(glmnet)
y <- carbohydrate$carbohydrate
x <- carbohydrate |>
  select(age, weight, protein) |>
  as.matrix()
fit <- glmnet(x, y)
```

---

```{r}
#| fig-cap: "Lasso selection"
#| label: fig-carbs-lasso
autoplot(fit, xvar = "lambda")
```

---

```{r}
cvfit <- cv.glmnet(x, y)
plot(cvfit)
```

---

```{r}
coef(cvfit, s = "lambda.1se")
```

---

### Likelihood ratio test for nested models {#sec-linreg-LRT}

For a general MLE-focused discussion of likelihood-ratio tests,
see [Likelihood ratio tests for MLEs](intro-MLEs.qmd#sec-lrt-mles).

A **likelihood ratio test (LRT)** compares two nested models
by computing twice the difference in their log-likelihoods:

$$2(\llik_1 - \llik_0) \dot \sim \chi^2_q \quad \text{(asymptotically under } H_0 \text{)}$$

where $q = p_2 - p_1$ is the number of extra parameters in the full model.

---

```{r}
logLik(bw_lm2)
logLik(bw_lm1)

log_LR <- (logLik(bw_lm2) - logLik(bw_lm1)) |> as.numeric()
delta_df <- (bw_lm1$df.residual - df.residual(bw_lm2))


x_max <- 1
```

---

```{r}
#| label: fig-chisq-plot
#| fig-cap: "Chi-square distribution"

d_log_LR <- function(x, df = delta_df) dchisq(x, df = df)

chisq_plot <-
  ggplot() +
  geom_function(fun = d_log_LR) +
  stat_function(
    fun = d_log_LR,
    xlim = c(2 * log_LR, x_max),
    geom = "area",
    fill = "gray"
  ) +
  geom_segment(
    aes(
      x = 2 * log_LR,
      xend = 2 * log_LR,
      y = 0,
      yend = d_log_LR(2 * log_LR)
    ),
    col = "red"
  ) +
  xlim(0.0001, x_max) +
  ylim(0, 4) +
  ylab("p(X=x)") +
  xlab("Likelihood-ratio test statistic [x] = 2 * log(likelihood ratio)") +
  theme_classic()
chisq_plot |> print()
```

---

Now we can get the p-value:

```{r}
#| label: LRT-pval
pchisq(
  q = 2 * log_LR,
  df = delta_df,
  lower = FALSE
) |>
  print()
```

---

In practice you don't have to do this by hand; there are functions to do it for you:

```{r}
# built in
lmtest::lrtest(bw_lm2, bw_lm1)
```

---

{{< include _subfiles/Linear-models-overview/_sec_linreg_Ftest.qmd >}}
