
::: notes
Maybe we can transform the data and model in some way 
to make it easier to inspect.
:::

{{< include _subfiles/Linear-models-overview/_def-residual-deviation.qmd >}}

{{< include _subfiles/Linear-models-overview/_def-signal.qmd >}}

---

:::{#thm-gaussian-resid-deviation}
#### Deviations in Gaussian models

If $Y$ has a Gaussian distribution, 
then $\devn(Y)$ also has a Gaussian distribution, and vice versa.
:::

:::{.proof}

By @def-deviation-subpop-mean, $\devn(Y \mid X = x) = Y - \E{Y \mid X = x}$.
Let $\mu(x) = \E{Y \mid X = x}$.
Since $\devn(Y \mid X = x) = Y - \mu(x)$ is an affine function of $Y$,
and Gaussian distributions are closed under affine transformations,
if $Y \mid X = x \sim \Normal(\mu(x), \ss)$, then:

$$
\devn(Y \mid X = x) = Y - \mu(x) \sim \Normal(\mu(x) - \mu(x), \ss) = \Normal(0, \ss)
$$

Conversely, since $Y = \E{Y \mid X = x} + \devn(Y \mid X = x) = \mu(x) + \devn(Y \mid X = x)$,
if $\devn(Y \mid X = x) \sim \Normal(0, \ss)$,
then $Y \mid X = x = \mu(x) + \devn(Y \mid X = x) \sim \Normal(\mu(x), \ss)$.

:::

---

:::{#def-resid-fitted}
#### Residuals

{{< include _subfiles/shared/_def-residual-observed-minus-fitted.qmd >}}
:::

---

:::{#exm-resid-bw}

#### Residuals in `birthweight` data

:::{#fig-bw-resids1}

```{r}
plot1_interact +
  facet_wrap(~sex) +
  geom_segment(
    aes(
      x = age,
      y = predlm2,
      xend = age,
      yend = weight,
      col = sex,
      group = id
    ),
    linetype = "dotted"
  )
```

Fitted values and residuals for interaction model for `birthweight` data

:::

:::

---

## Residuals of fitted values vs subpopulation-mean deviations

The residual is a plug-in estimate
of the deviation of an observation
from its subpopulation mean:

$$
\ba
\devn(y_i) &= y_i - \E{Y_i \mid X_i}
\\
\widehat{\devn(y_i)} &= y_i - \hat y_i
\\
&= \resid_i
\ea
$$

Different sources are not fully consistent
about these terms.
For terminology in this course,
we use:

- **deviation**
  for deviation of an observation from its population mean (typically conditional on covariates).
  This quantity is often called **error** in other sources,
  even though it doesn't directly involve an estimate;
  cf., <https://en.wikipedia.org/wiki/Errors_and_residuals>;
- **estimation error** for deviation of an estimate from its estimand;
- **residual** for deviation of an observed value from a fitted value.

---

## General characteristics of residuals

:::{#def-hat-matrix}
##### Hat matrix

For an ordinary least squares linear model
with design matrix $\mX$,
the **hat matrix** is

$$H = \mX(\tp{\mX}\mX)^{-1}\tp{\mX}$$

so the fitted values satisfy

$$\hat{\vY} = H\vY$$
:::

:::{#thm-resid-unbiased}
For an ordinary least squares linear model
with fitted values $\hat y_i = \dprodf{\vx_i}{\vb}$
(and fitted-value vector $\hat{\vY}$),
if the conditional mean is correctly specified so that:

$$ \E{\vY \mid \mX} = \mX\b$$

equivalently:

$$ \E{\hat{\vY} \mid \mX} = \E{\vY \mid \mX}$$

and if the errors are homoskedastic so that
$\Var{\vY \mid \mX} = \ss \idmat_n$ where $\idmat_n$ is the $n \times n$ identity matrix,
then,
treating $\mX$ as fixed in this derivation,
the residual moments are:

$$\E{\vec{\resid} \mid \mX} = 0$$ {#eq-mean-resid-unbiased}
$$\Var{\resid_i \mid \mX} = \ss(1-h_{ii})$$ {#eq-var-resid-leverage}

The conditional mean result (@eq-mean-resid-unbiased)
uses unbiasedness of fitted values.
The conditional variance result (@eq-var-resid-leverage)
uses the homoskedasticity condition
with the OLS hat-matrix representation.

When leverage is roughly uniform,
$h_{ii} \approx k/n$,
where $k = \mathrm{rank}(\mX)$
(so $k = p + 1$ for a full-rank model
with an intercept and $p$ predictors),
so $\Var{\resid_i} \approx \ss$
as $n$ grows.

:::

---

:::{.proof}

@eq-mean-resid-unbiased:

$$
\ba
\E{{\vec{\resid}} \mid \mX} &= \E{\vY - \hat{\vY} \mid \mX}
\\ &= \E{\vY \mid \mX} - \E{\hat{\vY} \mid \mX}
\\ &= \E{\vY \mid \mX} - \E{H\vY \mid \mX}
\\ &= \E{\vY \mid \mX} - H\E{\vY \mid \mX}
\\ &= \E{\vY \mid \mX} - H\mX\b
\\ &= \mX\b - H\mX\b
\\ &= \mX\b - \mX\b
\\ &= \vec{0}
\ea
$$

@eq-var-resid-leverage:

$$
\ba
{\vec{\resid}} &= (\idmat_n-H)\vY
\\ \Var{{\vec{\resid}} \mid \mX} &= \Var{(\idmat_n-H)\vY \mid \mX}
\\ &= (\idmat_n-H)\Var{\vY \mid \mX}\tp{(\idmat_n-H)}
\\ &= (\idmat_n-H)\ss \idmat_n\tp{(\idmat_n-H)}
\\ &= \ss(\idmat_n-H)\tp{(\idmat_n-H)}
\\ &= \ss(\idmat_n-H)^2
\\ &= \ss(\idmat_n - 2H + H^2)
\\ &= \ss(\idmat_n-H) \qquad (\text{since } H^2 = H)
\\ \Var{\resid_i \mid \mX} &= \left[\Var{{\vec{\resid}} \mid \mX}\right]_{ii}
\\ &= \left[\ss(\idmat_n-H)\right]_{ii}
\\ &= \ss(1-h_{ii})
\ea
$$

:::

---

:::{#thm-resid-gaussian}
#### Conditional distribution of residuals in Gaussian models

In a Gaussian linear regression model,
the residuals $\resid_i$,
conditional on the design matrix $\mX$,
are normally distributed:

$$\resid_i \mid \mX \sim \Normal\paren{0,\; \ss(1 - h_{ii})}$$ {#eq-resid-dist-exact}

where $h_{ii}$ is the $i$-th diagonal element of the hat matrix (@def-hat-matrix).
When leverage is roughly uniform,
$\ss(1-h_{ii}) \approx \ss$ for large $n$;
replacing the unknown $\ss$ with its estimate $\hs^2$ gives:

$$\resid_i \approx \Normal(0, \hs^2)$$ {#eq-resid-dist-approx}

:::

---

:::{.proof}

Using the hat matrix (@def-hat-matrix),
the residual vector is $\vec{\resid} = (I-H)\vY$,
so the $i$-th residual is a linear combination of the observations:

$$\resid_i = [(I-H)\vY]_i = \sum_{j=1}^n (\delta_{ij} - h_{ij})\, Y_j$$

where $\delta_{ij}$ is the Kronecker delta
($\delta_{ij} = 1$ if $i = j$, else $\delta_{ij} = 0$).

Given $\mX$,
the random variables $Y_1, \ldots, Y_n$ are conditionally independent Gaussians
by the Gaussian model assumption.
Therefore,
conditional on $\mX$,
the residual $\resid_i$ is a linear combination of independent Gaussians,
so $\resid_i$ is Gaussian.

The mean and variance follow from @thm-resid-unbiased:

$$
\ba
\Ef{\resid_i \mid \mX} &= 0
\\ \Var{\resid_i \mid \mX} &= \ss(1 - h_{ii})
\ea
$$

Therefore $\resid_i \mid \mX \sim \Normal(0, \ss(1 - h_{ii}))$.

:::

---

## Computing residuals in R

R provides a function for residuals:

```{r}
resid(bw_lm2)
```


---

:::{#exr-calc-resids}
Check R's output by computing the residuals directly.
:::

:::{.solution}
```{r}
bw$weight - fitted(bw_lm2)
```

This matches R's output!
:::

---

## Graphing the residuals

:::{#fig-resid-bw layout-ncol=2}

:::{#fig-bw-resids1-c2}

```{r}
plot1_interact +
  facet_wrap(~sex) +
  geom_segment(
    aes(
      x = age,
      y = predlm2,
      xend = age,
      yend = weight,
      col = sex,
      group = id
    ),
    linetype = "dotted"
  )
```

fitted values

:::

:::{#fig-resids-intxn-bw}
```{r}

bw <- bw |>
  mutate(
    resids_intxn =
      weight - fitted(bw_lm2)
  )

plot_bw_resid <-
  bw |>
  ggplot(aes(
    x = age,
    y = resids_intxn,
    linetype = sex,
    shape = sex,
    col = sex
  )) +
  theme_bw() +
  xlab("Gestational age (weeks)") +
  ylab("residuals (grams)") +
  theme(legend.position = "bottom") +
  geom_hline(aes(
    yintercept = 0,
    col = sex
  )) +
  geom_segment(
    aes(yend = 0),
    linetype = "dotted"
  ) +
  geom_point(alpha = .7)
# expand_limits(y = 0, x = 0)
print(plot_bw_resid + facet_wrap(~sex))
```

Residuals

:::

Fitted values and residuals for interaction model for `birthweight` data

:::

---

## Residuals versus predictors

```{r}
hers <- hers |>
  mutate(
    resids_no_intcpt =
      LDL - fitted(hers_lm_no_int),
    resids_with_intcpt =
      LDL - fitted(hers_lm_with_int)
  )
```

:::{#fig-resids-hers-1 layout-ncol=2}

:::{#fig-resid-hers-pred-no-intcpt} 

```{r}
hers |>
  arrange(age) |>
  ggplot() +
  aes(x = age, y = resids_no_intcpt, col = factor(smoking)) +
  geom_point() +
  geom_hline(aes(yintercept = 0, col = factor(smoking))) +
  facet_wrap(~smoking, labeller = "label_both") +
  theme(legend.position = "bottom") +
  geom_smooth(col = "blue")
```

no intercept
:::

:::{#fig-resid-hers-pred-with-intcpt}

```{r}
hers |>
  arrange(age) |>
  ggplot() +
  aes(x = age, y = resids_with_intcpt, col = factor(smoking)) +
  geom_point() +
  geom_hline(aes(yintercept = 0, col = factor(smoking))) +
  facet_wrap(~smoking, labeller = "label_both") +
  theme(legend.position = "bottom") +
  geom_smooth(col = "blue")
```


with intercept
:::

Residuals of `hers` data vs predictors

:::

---

## Residuals versus fitted values

::: notes
If the model contains multiple continuous covariates,
how do we check for errors in the mean structure assumption?
:::


```{r}
#| label: fig-resids-intxn-hers
#| fig-cap: "Residuals of interaction model for `hers` data"

library(ggfortify)
hers_lm_no_int |>
  update(na.action = na.omit) |>
  autoplot(
    which = 1,
    ncol = 1,
    smooth.colour = NA
  ) +
  geom_hline(yintercept = 0, col = "red")
```

---

We can add a LOESS smooth to visualize where the residual mean is nonzero:

```{r}
#| label: fig-resids-intxn-hers-smooth-no-int
#| fig-cap: "Residuals of interaction model for `hers` data, no intercept term"

library(ggfortify)
hers_lm_no_int |>
  update(na.action = na.omit) |>
  autoplot(
    which = 1,
    ncol = 1
  ) +
  geom_hline(yintercept = 0, col = "red")
```

---

:::{#fig-resid-hers-compare layout-ncol=2}

:::{#fig-resids-intxn-hers-smooth-no-int-2}

```{r}
library(ggfortify)
hers_lm_no_int |>
  update(na.action = na.omit) |>
  autoplot(
    which = 1,
    ncol = 1
  ) +
  geom_hline(yintercept = 0, col = "red")
```

no intercept term

:::

:::{#fig-resids-intxn-hers-smooth-with-int}

```{r}
hers_lm_with_int |>
  update(na.action = na.omit) |>
  autoplot(
    which = 1,
    ncol = 1
  ) +
  geom_hline(yintercept = 0, col = "red")
```

with intercept term

:::

Residuals of interaction model for `hers` data, with and without intercept term

:::

---

:::{#def-stred}

#### Standardized residuals

For an ordinary linear model,
the estimated standard deviation of the $i$th residual is

$$
\hSD{\resid_i} = \hat{\sigma}\sqrt{1-h_{ii}}
$$

so the standardized residual is

$$
\stdresid_i = \frac{\resid_i}{\hSD{\resid_i}}
= \frac{\resid_i}{\hat{\sigma}\sqrt{1-h_{ii}}}
$$
:::

Hence, with enough data and a correct model,
each standardized residual has an approximately standard Gaussian **marginal** distribution; that is,

$$
\stdresid_i \ddist N(0,1)
$$

but they are not independent.
If $H$ is the hat matrix with entries $h_{ij}$,
then

$$
\Cov{\resid_i,\resid_j \mid \mX}
=
\begin{cases}
\ss(1-h_{ii}), & i=j,\\
-\ss h_{ij}, & i\neq j,
\end{cases}
$$

and therefore

$$
\Cov{\stdresid_i,\stdresid_j \mid \mX}
\approx
\begin{cases}
1, & i=j,\\
\dfrac{-h_{ij}}{\sqrt{(1-h_{ii})(1-h_{jj})}}, & i\neq j,
\end{cases}
$$

In particular,
for $i \neq j$,
these covariances are generally nonzero.

