
(adapted from @dobson4e §6.3.3; for more information on prediction, see @james2013introduction and @rms2e).

::: notes
If we have a lot of covariates in our dataset, we might want to choose a small subset to use in our model.

There are a few possible metrics to consider for choosing a "best" model.
:::

### DAGs for variable selection

For explanatory models,
variable inclusion should not rely on automated algorithms alone.
Directed acyclic graphs (DAGs)
can help us encode substantive assumptions
about which variables are confounders,
mediators,
or colliders.

In a Dobson-style workflow,
we use the DAG first
to decide a defensible adjustment set,
then compare candidate regression models
within that set using predictive or likelihood-based criteria.
This keeps model selection aligned with study design,
rather than only with numerical fit.

In this workflow,
the DAG encodes hypothesized time ordering
and causal pathways.
That helps us decide
which variables belong in the candidate model set
before we run stepwise,
subset,
or penalized selection methods.
The key point
is that a variable can improve apparent fit
while still distorting the effect we care about
if it sits on a causal pathway
or induces collider bias
[@dobson4e, §6.3.3].

:::{#exm-dag-heat-deaths}

#### high temperatures and mortality

Suppose the research question is:
"What is the effect of high temperatures on deaths?"

A plausible DAG includes:
high temperatures $\rightarrow$ deaths,
high temperatures $\rightarrow$ blackouts,
high temperatures $\rightarrow$ forest fires,
blackouts $\rightarrow$ deaths,
forest fires $\rightarrow$ air pollution,
and both forest fires and air pollution $\rightarrow$ deaths
[@dobson4e, §6.3.3].

```{r}
#| label: fig-dag-heat-deaths
#| fig-cap: DAG for the high temperatures and mortality example.
#| fig-alt: DAG linking high temperatures, blackouts,
#|   forest fires, air pollution, and deaths.
#| echo: false
dag_heat_deaths <- ggdag::dagify(
  D ~ T + B + F + P,
  B ~ T,
  F ~ T,
  P ~ F,
  exposure = "T",
  outcome = "D",
  labels = c(
    T = "High temperatures",
    D = "Deaths",
    B = "Blackouts",
    F = "Forest fires",
    P = "Air pollution"
  )
)

dag_heat_deaths |>
  ggdag::ggdag(use_labels = "label") +
  ggdag::theme_dag()
```

If our estimand is the total effect
of high temperatures on deaths,
then blackouts,
fires,
and air pollution are mediators on downstream pathways.
Adjusting for them
would remove part of the effect
we are trying to estimate.

But if the estimand changes
to the effect of blackouts on deaths,
then high temperatures become a confounder
because they are a common cause
of both blackouts and deaths.
So the same variable
can be "adjust for" or "do not adjust for"
depending on the causal question.

:::

:::{#exm-dag-salmonellosis-weather}

#### salmonellosis and weather

For daily salmonellosis counts,
a plausible DAG has temperature and rainfall
as direct causes of risk,
with both variables also affecting humidity.
If humidity has no direct arrow to salmonellosis,
then humidity is not a causal driver;
it mainly reflects shared variation
from temperature and rainfall.
Including humidity in the regression
can therefore induce misleading associations
and bias weather-effect estimates
[@dobson4e, §6.3.3].

```{r}
#| label: fig-dag-salmonellosis-weather
#| fig-cap: DAG for the salmonellosis and weather example.
#| fig-alt: DAG linking season, temperature, rainfall,
#|   humidity, test change, and salmonellosis.
#| echo: false
dag_salmonellosis_weather <- ggdag::dagify(
  S ~ T + R + C,
  H ~ T + R,
  T ~ Se,
  R ~ Se,
  labels = c(
    S = "Salmonellosis",
    T = "Temperature",
    R = "Rainfall",
    H = "Humidity",
    C = "Test change",
    Se = "Season"
  )
)

dag_salmonellosis_weather |>
  ggdag::ggdag(use_labels = "label") +
  ggdag::theme_dag()
```

The same DAG can include
a "change in test" indicator
when surveillance switched
to a more sensitive diagnostic test.
That variable can improve predictive accuracy
and face validity,
but omitting it does not invalidate
causal interpretation
for temperature and rainfall effects
if it is not on their causal paths
[@dobson4e, §6.3.3].

Likewise,
season may cause both temperature and rainfall,
but if season has no direct path to salmonellosis
beyond those weather variables,
including season can obscure interpretation
of the proximal weather effects
[@dobson4e, §6.3.3].

Finally,
DAG-informed variable definitions matter.
In one intensive-care example,
a "nutrition score" looked highly predictive of mortality
until investigators clarified
that people near death were often coded as zero
because no assessment was completed.
So the recorded variable
combined nutrition with prognosis,
and naive selection would overstate
its scientific meaning
[@dobson4e, §6.3.3].

:::

### Mean squared error

We might want to minimize the **mean squared error**,
$\E{(y-\hat y)^2}$,
for new observations
that weren't in our data set when we fit the model.

Unfortunately,
$$\frac{1}{n}\sum_{i=1}^n (y_i-\hat y_i)^2$$
gives a biased estimate of $\E{(y-\hat y)^2}$ for new data.
If we want an unbiased estimate,
we will have to be clever.

This is one reason that $R^2$ is not enough for model selection.
$R^2$ does not decrease as we add explanatory variables,
even when those variables do not improve out-of-sample prediction.
That can lead to overfitting.

With a training/test split,
we estimate coefficients in the training data
and compute prediction error in held-out data:

$$
\hat\b_{\text{train}} =
(\tp{X_{\text{train}}}X_{\text{train}})^{-1}
\tp{X_{\text{train}}}y_{\text{train}}
$$

$$
\hat e_{i,\text{test}} =
y_{\text{test},i} -
\cb{
\hat\b_{0,\text{train}} +
\sum_{j=1}^p \hat\b_{j,\text{train}}x_{\text{test},ij}
},
\quad i \in \text{test set}
$$

$$
\text{RMSE} = \sqrt{
\frac{1}{n_{\text{test}}}
\sum_{i \in \text{test set}} (\hat e_{i,\text{test}})^2
}
$$

---

#### Cross-validation

Rather than one arbitrary split,
$k$-fold cross-validation repeatedly partitions the data into training and test folds.
For each split we compute prediction error,
then summarize errors across folds and replications.
The preferred model has lower prediction error,
with simpler models favored when errors are similar.

When the number of candidate explanatory variables is small,
we can compare all possible subsets ($2^p$ models)
using the same cross-validation metric.

```{r}
data("carbohydrate", package = "dobson")
library(cvTools)
full_model <- lm(carbohydrate ~ ., data = carbohydrate)
cv_full <-
  full_model |> cvFit(
    data = carbohydrate, K = 5, R = 10,
    y = carbohydrate$carbohydrate
  )

reduced_model <- full_model |> update(formula = ~ . - age)

cv_reduced <-
  reduced_model |> cvFit(
    data = carbohydrate, K = 5, R = 10,
    y = carbohydrate$carbohydrate
  )
```

---

```{r}
results_reduced <-
  tibble(
    model = "wgt+protein",
    errs = cv_reduced$reps[]
  )
results_full <-
  tibble(
    model = "wgt+age+protein",
    errs = cv_full$reps[]
  )

cv_results <-
  bind_rows(results_reduced, results_full)

cv_results |>
  ggplot(aes(y = model, x = errs)) +
  geom_boxplot()
```

---

##### comparing metrics

```{r}
compare_results <- tribble(
  ~model, ~cvRMSE, ~r.squared, ~adj.r.squared, ~trainRMSE, ~loglik,
  "full",
  cv_full$cv,
  summary(full_model)$r.squared,
  summary(full_model)$adj.r.squared,
  sigma(full_model),
  logLik(full_model) |> as.numeric(),
  "reduced",
  cv_reduced$cv,
  summary(reduced_model)$r.squared,
  summary(reduced_model)$adj.r.squared,
  sigma(reduced_model),
  logLik(reduced_model) |> as.numeric()
)

compare_results
```

---

```{r}
anova(full_model, reduced_model)
```

---

#### Best subset selection

When the number of candidate predictors is modest,
we can use **best subset selection**.
For each model size $k = 0, 1, \ldots, p$,
we fit all $\binom{p}{k}$ models
and keep the best model of size $k$
(e.g., by lowest RSS in the training data).

This gives at most $p+1$ candidate models to compare across model sizes,
using criteria such as
cross-validated prediction error,
$C_p$,
AIC/BIC,
or adjusted $R^2$.
In that sense,
best subset selection is more exhaustive
than one-path methods like forward or backward stepwise selection.

::: notes
For a full algorithmic treatment,
including validation and criterion-based comparison across model size,
see the ISLR treatment of best subset selection
[@james2013introduction].
:::

```{r}
hers_subset <- fs::path_package("rme", "extdata/hersdata.dta") |>
  haven::read_dta() |>
  dplyr::select(LDL, age, weight, BMI, HDL, TG, SBP) |>
  tidyr::drop_na()

hers_lm_subset <- lm(
  LDL ~ age + weight + BMI + HDL + TG + SBP,
  data = hers_subset
)

hers_best_subset <- olsrr::ols_step_best_subset(hers_lm_subset)

hers_best_subset$metrics |>
  dplyr::arrange(dplyr::desc(adjr)) |>
  dplyr::slice_head(n = 5)
```

---

#### Stepwise regression

Stepwise methods are another common approach.
Forward selection adds variables one at a time.
Backward selection starts with the full model
and removes variables sequentially.
Both approaches can select different models from the same data.

::: {.callout-warning}
## Caution about stepwise selection

Stepwise regression has several known problems:

- It tends to select too many variables (overfitting)
- P-values and confidence intervals are biased after selection
- It ignores model uncertainty
- Results can be unstable across different samples

Consider using cross-validation, penalized methods (like Lasso), 
or subject-matter knowledge instead.
See @rms2e and @heinze2018variable for more discussion.
:::

```{r}
library(olsrr)
olsrr:::ols_step_both_aic(full_model)
```

---

#### Lasso

Lasso is a penalized regression method
that shrinks coefficient estimates toward zero.
It adds an $L_1$ penalty term
to the objective function,
creating a trade-off between model fit and parsimony.
The intercept is not penalized.

As the tuning parameter $\lambda$ increases,
more coefficients are shrunken strongly,
and some become exactly zero.
So lasso both regularizes the model
and performs variable selection.
In practice,
$\lambda$ is usually chosen by cross-validation
to minimize prediction error.

For Gaussian linear models,
the penalized least-squares forms are:

$$
\hat\b^{\text{lasso}} =
\arg\min_{\b_0,\ldots,\b_p}
\cb{
\sum_{i=1}^n
\cb{
y_i - \b_0 - \sum_{j=1}^p \b_j x_{ij}
}^2
+
\lambda \sum_{j=1}^p |\b_j|
}
$$

$$
\hat\b^{\text{ridge}} =
\arg\min_{\b_0,\ldots,\b_p}
\cb{
\sum_{i=1}^n
\cb{
y_i - \b_0 - \sum_{j=1}^p \b_j x_{ij}
}^2
+
\lambda \sum_{j=1}^p \b_j^2
}
$$

$$
\hat\b^{\text{elastic-net}} =
\arg\min_{\b_0,\ldots,\b_p}
\cb{
\sum_{i=1}^n
\cb{
y_i - \b_0 - \sum_{j=1}^p \b_j x_{ij}
}^2
+
\lambda
\cb{
\a \sum_{j=1}^p |\b_j|
+
\frac{1-\a}{2}\sum_{j=1}^p \b_j^2
}
}
$$

```{r}
library(glmnet)
y <- carbohydrate$carbohydrate
x <- carbohydrate |>
  select(age, weight, protein) |>
  as.matrix()
fit <- glmnet(x, y)
```

---

```{r}
#| fig-cap: "Lasso selection"
#| label: fig-carbs-lasso
autoplot(fit, xvar = "lambda")
```

---

```{r}
cvfit <- cv.glmnet(x, y)
plot(cvfit)
```

---

```{r}
coef(cvfit, s = "lambda.1se")
```

---

### Likelihood ratio test for nested models {#sec-linreg-LRT}

For a general MLE-focused discussion of likelihood-ratio tests,
see [Likelihood ratio tests for MLEs](intro-MLEs.qmd#sec-lrt-mles).

A **likelihood ratio test (LRT)** compares two nested models
by computing twice the difference in their log-likelihoods:

$$2(\llik_1 - \llik_0) \dot \sim \chi^2_q \quad \text{(asymptotically under } H_0 \text{)}$$

where $q = p_2 - p_1$ is the number of extra parameters in the full model.

---

```{r}
logLik(bw_lm2)
logLik(bw_lm1)

log_LR <- (logLik(bw_lm2) - logLik(bw_lm1)) |> as.numeric()
delta_df <- (bw_lm1$df.residual - df.residual(bw_lm2))


x_max <- 1
```

---

```{r}
#| label: fig-chisq-plot
#| fig-cap: "Chi-square distribution"

d_log_LR <- function(x, df = delta_df) dchisq(x, df = df)

chisq_plot <-
  ggplot() +
  geom_function(fun = d_log_LR) +
  stat_function(
    fun = d_log_LR,
    xlim = c(2 * log_LR, x_max),
    geom = "area",
    fill = "gray"
  ) +
  geom_segment(
    aes(
      x = 2 * log_LR,
      xend = 2 * log_LR,
      y = 0,
      yend = d_log_LR(2 * log_LR)
    ),
    col = "red"
  ) +
  xlim(0.0001, x_max) +
  ylim(0, 4) +
  ylab("p(X=x)") +
  xlab("Likelihood-ratio test statistic [x] = 2 * log(likelihood ratio)") +
  theme_classic()
chisq_plot |> print()
```

---

Now we can get the p-value:

```{r}
#| label: LRT-pval
pchisq(
  q = 2 * log_LR,
  df = delta_df,
  lower = FALSE
) |>
  print()
```

---

In practice you don't have to do this by hand; there are functions to do it for you:

```{r}
# built in
lmtest::lrtest(bw_lm2, bw_lm1)
```

---

{{< include _subfiles/Linear-models-overview/_sec_linreg_Ftest.qmd >}}
