## Rescaling

::: notes
Centering covariates (subtracting a constant from them) 
can make coefficient interpretations more meaningful, 
particularly for the intercept and main effect terms in models with interactions.
:::

### Rescale age {.smaller}

::: notes
For example, 
the intercept $\beta_0$ in our interaction model represents 
the mean birthweight for females at age = 0 weeks, 
which is not a biologically plausible scenario.
If we center age at 36 weeks instead, 
$\beta_0$ will represent the mean birthweight for females at 36 weeks, 
which is more interpretable.
:::

---

{{< include _subfiles/Linear-models-overview/_exr-linreg-rescaling.qmd >}}

---

:::{#exr-rescaling-in-r}
Using R, fit the rescaled interaction model with $A^* = A - 36$ weeks
in place of $A$ in [Model @eq-BW-lm-interact].
Compare the coefficient estimates with those from the original model.
Which coefficients change, and which remain the same?
:::

---

:::{.solution}

```{r}
bw <-
  bw |>
  mutate(
    `age - mean` = age - mean(age),
    `age - 36wks` = age - 36
  )

lm1_c <- lm(weight ~ sex + `age - 36wks`, data = bw)

lm2_c <- lm(weight ~ sex + `age - 36wks` + sex:`age - 36wks`, data = bw)

parameters(lm2_c, ci_method = "wald") |> print_md()
```

Compare with what we got without rescaling:

```{r}
parameters(bw_lm2, ci_method = "wald") |> print_md()
```

::: notes
Notice that the slope coefficients (`age - 36wks` and `sexmale:age - 36wks`) 
remain the same, 
but the intercept and `sexmale` coefficient have changed 
to reflect means at age = 36 weeks rather than age = 0 weeks.
:::

:::

---

### Centering gestational age does not change predictions

Centering gestational age changes the coefficient parameterization,
but it does not change fitted values,
confidence bands,
or prediction bands.

The next output reports maximum absolute differences
between the uncentered and centered models.
All values should be near zero
(up to floating-point rounding),
which confirms that centering changes parameterization only.

```{r}
bw_centered <-
  bw |>
  dplyr::mutate(
    age_mean_centered = age - mean(age)
  )

bw_lm2_centered <-
  lm(
    weight ~ sex + age_mean_centered + sex:age_mean_centered,
    data = bw_centered
  )

pred_uncentered_ci <-
  predict(
    bw_lm2,
    newdata = bw_centered,
    interval = "confidence"
  ) |>
  tibble::as_tibble()

pred_centered_ci <-
  predict(
    bw_lm2_centered,
    newdata = bw_centered,
    interval = "confidence"
  ) |>
  tibble::as_tibble()

pred_uncentered_pi <-
  predict(
    bw_lm2,
    newdata = bw_centered,
    interval = "predict"
  ) |>
  tibble::as_tibble()

pred_centered_pi <-
  predict(
    bw_lm2_centered,
    newdata = bw_centered,
    interval = "predict"
  ) |>
  tibble::as_tibble()

tibble::tibble(
  fitted_max_abs_diff = max(abs(pred_uncentered_ci$fit - pred_centered_ci$fit)),
  ci_lwr_max_abs_diff = max(abs(pred_uncentered_ci$lwr - pred_centered_ci$lwr)),
  ci_upr_max_abs_diff = max(abs(pred_uncentered_ci$upr - pred_centered_ci$upr)),
  pi_lwr_max_abs_diff = max(abs(pred_uncentered_pi$lwr - pred_centered_pi$lwr)),
  pi_upr_max_abs_diff = max(abs(pred_uncentered_pi$upr - pred_centered_pi$upr))
)
```

:::{#fig-pred-int-centered-comparison}
```{r}
comparison_plot_data <-
  dplyr::bind_rows(
    bw_centered |>
      dplyr::bind_cols(
        pred_uncentered_ci |>
          dplyr::transmute(
            fit,
            ci_lwr = lwr,
            ci_upr = upr
          ),
        pred_uncentered_pi |>
          dplyr::transmute(
            pred_lwr = lwr,
            pred_upr = upr
          )
      ) |>
      dplyr::transmute(
        age,
        weight,
        sex,
        model = "Without centering age",
        fit,
        ci_lwr,
        ci_upr,
        pred_lwr,
        pred_upr
      ),
    bw_centered |>
      dplyr::bind_cols(
        pred_centered_ci |>
          dplyr::transmute(
            fit,
            ci_lwr = lwr,
            ci_upr = upr
          ),
        pred_centered_pi |>
          dplyr::transmute(
            pred_lwr = lwr,
            pred_upr = upr
          )
      ) |>
      dplyr::transmute(
        age,
        weight,
        sex,
        model = "With centered age",
        fit,
        ci_lwr,
        ci_upr,
        pred_lwr,
        pred_upr
      )
  ) |>
  dplyr::mutate(
    model = factor(
      model,
      levels = c("Without centering age", "With centered age")
    )
  )

comparison_plot_data |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = age,
      y = weight,
      col = sex,
      fill = sex,
      group = sex
    )
  ) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::geom_ribbon(
    ggplot2::aes(
      ymin = pred_lwr,
      ymax = pred_upr
    ),
    alpha = 0.15,
    col = NA
  ) +
  ggplot2::geom_ribbon(
    ggplot2::aes(
      ymin = ci_lwr,
      ymax = ci_upr
    ),
    alpha = 0.3,
    col = NA
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = fit)
  ) +
  ggplot2::facet_wrap(~model) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(
    x = "Gestational age (weeks)",
    y = "Birthweight (grams)",
    col = "Sex",
    fill = "Sex"
  )
```

Two-panel comparison of the fitted `birthweight` model `bw_lm2`
with and without centering gestational age,
showing identical fitted values,
confidence bands,
and prediction bands in both panels.
:::
