::: notes
Until now, statin use ($U$) has appeared only as the color grouping in our HERS plots;
here it finally enters the model, as the stratifying variable.

As an alternative to the age–BMI models above,
we can model LDL as a function of BMI,
stratifying by statin use
(i.e., fitting a separate intercept and BMI slope
for each statin group).
We omit age from this stratified specification
to keep it simple and focus on how the BMI–LDL relationship differs by statin use;
age could be added back as an additional covariate if desired.
:::

$$
\E{LDL|B=b, V=v} =
\beta_{no} (1-v) + \beta_{no,B}\, b(1-v) +
\beta_{yes} v + \beta_{yes,B}\, bv
$$ {#eq-hers-strat}

Equivalently, we can fit the same model using reference (indicator) coding,
with an overall intercept, a main effect for statin use,
a BMI slope, and a statin–BMI interaction:

$$
\E{LDL|B=b, V=v} =
\beta_0 + \beta_V\, v + \beta_B\, b + \beta_{VB}\, (v \cdot b)
$$ {#eq-hers-strat-interact}

:::: {.columns}

::: {.column width=45%}

::: {#tbl-hers-lm-strat}

```{r}
#| code-fold: true
hers_lm_strat <-
  hers |>
  lm(
    formula = LDL ~ statins + statins:BMI - 1,
    data = _,
    na.action = na.exclude
  )

hers_lm_strat |>
  parameters::parameters() |>
  parameters::print_md(
    select = "{estimate}"
  )
```

Stratified regression estimates for `hers` data ([Model @eq-hers-strat])

:::

:::

:::{.column width=10%}
:::

:::{.column width=45%}

::: {#tbl-hers-lm-statin-interact}

```{r}
#| code-fold: true
hers_lm_statins_interact <- lm(
  LDL ~ statins + BMI + statins:BMI,
  data = hers,
  na.action = na.exclude
)

hers_lm_statins_interact |>
  parameters::parameters() |>
  parameters::print_md(
    select = "{estimate}"
  )
```

Regression parameter estimates for [Model @eq-hers-strat-interact]

:::

:::

::::
