::: notes
We compare the parallel-planes model (`hers_lm1`: `LDL ~ age + BMI`)
and the interaction model (`hers_lm2`: `LDL ~ age + BMI + age:BMI`)
using AIC, BIC, and deviance.
Lower AIC and BIC values indicate a better trade-off between fit and complexity;
lower deviance indicates a better-fitting model (at the cost of more parameters).
:::

#### AIC and BIC for `hers` models

::: {#tbl-hers-aic-bic}

```{r}
#| code-fold: true
AIC(hers_lm1, hers_lm2)
BIC(hers_lm1, hers_lm2)
```

AIC and BIC for the `hers` parallel-planes (`hers_lm1`) and interaction (`hers_lm2`) models

:::

The model with the lower AIC or BIC value is preferred as the better balance of fit and parsimony.
A smaller value for the interaction model indicates the added age-BMI interaction term is worthwhile;
a smaller value for the parallel-planes model would favor the simpler specification.


#### Deviance for `hers` models

::: {#tbl-hers-deviance}

```{r}
#| code-fold: true
deviance(hers_lm1)
deviance(hers_lm2)
```

Deviance for the `hers` parallel-planes (`hers_lm1`) and interaction (`hers_lm2`) models

:::

Lower deviance indicates a closer fit to the data; the interaction model uses one more parameter.
The likelihood ratio test below provides a formal comparison of whether the improvement justifies the added complexity.

