
If our Gaussian linear regression model is correct, the residuals $\resid_i$ and standardized residuals $\stdresid_i$ should have:

-   an approximately Gaussian distribution, with:
-   a mean of 0
-   a constant variance

This should be true **for every** value of $x$.

---

If we didn't correctly guess the functional form of the linear component of the mean,
for covariate vector $x = (x_1,\ldots,x_p)$,
$$\E{Y \mid X=x} = \b_0 + \b_1 x_1 + \cdots + \b_p x_p$$

Then the residuals might have nonzero mean.

Regardless of whether we guessed the mean function correctly, the variance of the residuals might differ between values of $x$.

---

#### Residuals versus fitted values

::: notes
To look for these issues, we can plot the residuals $\resid_i$ against the fitted values $\hat y_i$ (@fig-bw_lm2-resid-vs-fitted).
:::

:::{#fig-bw_lm2-resid-vs-fitted}

```{r}
autoplot(bw_lm2, which = 1, ncol = 1) |> print()
```

`birthweight` model  (@eq-BW-lm-interact): residuals versus fitted values
:::

::: notes
If the model is correct, the blue line should stay flat and close to 0, and the cloud of dots should have the same vertical spread regardless of the fitted value.

If not,
we probably need to change the functional form
of the linear component of the mean,
$$\E{Y \mid X=x} = \b_0 + \b_1 x_1 + \cdots + \b_p x_p$$

:::

---

{{< include exm-plot-fits.qmd >}}
---

#### Scale-location plot

::: notes
We can also plot the square roots of the absolute values of the standardized residuals against the fitted values (@fig-bw-scale-loc).
:::

```{r}
#| label: fig-bw-scale-loc
#| fig-cap: "Scale-location plot of `birthweight` data"
autoplot(bw_lm2, which = 3, ncol = 1) |> print()
```

::: notes
Here, the blue line doesn't need to be near 0, 
but it should be flat. 
If not, the residual variance $\sigma^2$ might not be constant, 
and we might need to transform our outcome $Y$ 
(or use a model that allows non-constant variance).
:::

---

#### Residuals versus leverage

::: notes

We can also plot our standardized residuals against "leverage", which roughly speaking is a measure of how unusual each $x_i$ value is. Very unusual $x_i$ values can have extreme effects on the model fit, so we might want to remove those observations as outliers, particularly if they have large residuals.

:::

::: {#fig-bw_lm2_resid-vs-leverage}

```{r}
autoplot(bw_lm2, which = 5, ncol = 1) |> print()
```

`birthweight` model with interactions (@eq-BW-lm-interact): residuals versus leverage
:::

::: notes
The blue line should be relatively flat and close to 0 here.
The curved dashed contour lines show where Cook's distance equals 0.5 and 1;
observations near or beyond those contours have substantial influence on the fit
(see the Cook's distance section below).
:::

---

