::: notes
We check whether the assumptions of the parallel-planes model
(`hers_lm1`, @eq-hers-parallel, fitted in @tbl-hers-lm1) hold
by examining residuals vs. fitted values (to assess linearity and constant variance)
and a QQ plot (to assess normality of residuals).
Plots are faceted by statin use to check for group-level residual patterns.
:::

#### Residuals vs fitted for `hers_lm1`

:::{#fig-hers-resid-fitted}

```{r}
#| code-fold: true
hers_diag <- hers_lm_data |>
  dplyr::mutate(
    .fitted = fitted(hers_lm1),
    .resid = residuals(hers_lm1)
  )

ggplot(hers_diag, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~statins, labeller = label_both) +
  xlab("Fitted values") +
  ylab("Residuals") +
  theme_bw()
```

Residuals vs fitted values for `hers_lm1` (parallel planes model)

:::

#### QQ plot for `hers_lm1`

:::{#fig-hers-qq}

```{r}
#| code-fold: true
ggplot(hers_diag, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~statins, labeller = label_both) +
  theme_bw()
```

QQ plot of residuals for `hers_lm1` (parallel planes model)

:::

::: notes
The residuals scatter around zero with no clear curve (@fig-hers-resid-fitted),
so the linearity assumption looks reasonable;
their spread grows only slightly at higher fitted values,
indicating at most mild non-constant variance.
The QQ plot (@fig-hers-qq) bends above the reference line in the upper tail,
reflecting the right skew of LDL --- a mild departure from normality.
With several thousand observations,
the central limit theorem makes inference about the model coefficients robust to this mild non-normality.
Both statin groups show the same pattern.
:::
