#### Example diagnostics with code and output

The example below simulates ordered data with positively autocorrelated errors.
That creates a setting where independence should fail.
We generate the error term from an AR(1) process with coefficient 0.7,
matching the notation in the mathematical details section.

```{r}
#| label: exm-independence-sim-data
#| code-fold: false
set.seed(204)

n_obs <- 120
x <- seq(-1, 1, length.out = n_obs)
ar_errors <- as.numeric(arima.sim(model = list(ar = 0.7), n = n_obs, sd = 1))
y <- 2 + 1.5 * x + ar_errors

serial_dep_exm <- tibble::tibble(
  time_index = seq_len(n_obs),
  x = x,
  y = y
)

serial_dep_exm_lm <- lm(y ~ x, data = serial_dep_exm)

summary(serial_dep_exm_lm)
```

Residuals versus order:

```{r}
#| label: fig-independence-resid-order
#| fig-cap: "Residuals versus observation order for simulated data"
serial_dep_exm |>
  dplyr::mutate(resid = resid(serial_dep_exm_lm)) |>
  ggplot(aes(x = time_index, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(color = "steelblue") +
  theme_classic() +
  labs(
    x = "Observation order",
    y = "Residual"
  )
```

Correlogram diagnostics:

```{r}
#| label: fig-independence-correlogram
#| fig-cap: "Residual ACF and PACF for simulated data"
op <- par(no.readonly = TRUE)
on.exit(par(op), add = TRUE)
par(mfrow = c(1, 2))
acf(
  resid(serial_dep_exm_lm),
  main = "Residual ACF"
)
pacf(
  resid(serial_dep_exm_lm),
  main = "Residual PACF"
)
```

Durbin-Watson and Breusch-Godfrey tests:

```{r}
#| label: exm-independence-formal-tests
#| code-fold: false
lmtest::dwtest(serial_dep_exm_lm, alternative = "two.sided")
lmtest::bgtest(serial_dep_exm_lm, order = 4)
```

In this example,
the residual-order plot and correlogram suggest serial dependence,
and both formal tests reject independence.

#### Real-data examples

We can also run the same diagnostics on real ordered data sets.

##### Nile annual river flow

```{r}
#| label: exm-independence-real-nile
#| code-fold: false
nile_df <- tibble::tibble(
  year = as.numeric(time(datasets::Nile)),
  flow = as.numeric(datasets::Nile)
)

nile_lm <- lm(flow ~ year, data = nile_df)

summary(nile_lm)
lmtest::dwtest(nile_lm, alternative = "two.sided")
lmtest::bgtest(nile_lm, order = 4)
```

```{r}
#| label: fig-independence-real-nile-resid-order
#| fig-cap: "Residuals versus year for Nile flow model"
nile_df |>
  dplyr::mutate(resid = resid(nile_lm)) |>
  ggplot(aes(x = year, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(color = "steelblue") +
  theme_classic() +
  labs(
    x = "Year",
    y = "Residual"
  )
```

##### Mauna Loa atmospheric carbon dioxide (`co2`)

```{r}
#| label: exm-independence-real-co2
#| code-fold: false
co2_df <- tibble::tibble(
  decimal_year = as.numeric(time(datasets::co2)),
  co2_ppm = as.numeric(datasets::co2)
)

co2_lm <- lm(co2_ppm ~ decimal_year, data = co2_df)

summary(co2_lm)
lmtest::dwtest(co2_lm, alternative = "two.sided")
lmtest::bgtest(co2_lm, order = 12)
```

For both real-data examples,
the diagnostics indicate residual dependence,
consistent with their time-ordered structure.
