{{< include latex-macros/macros.qmd >}}

:::{#exr-ci-mean}

Given
a maximum likelihood estimate $\hvb$
and
a corresponding estimated covariance matrix
$\hS \eqdef \widehat{\Covt}(\hvb)$,
calculate a 95% confidence interval for the predicted mean
$\mu(\vx) = \E{Y|\vX=\vx}$.

:::

---

:::{#sol-ci-mean}

::: notes

By [the Gauss-Markov theorem](https://en.wikipedia.org/wiki/Gauss%E2%80%93Markov_theorem),
the OLS estimate $\hvb$ is the best linear unbiased estimator of $\vb$.
A 95% confidence interval for $\mu(\vx)$
can be constructed directly on the original scale:

$$\hat\mu(\vx) \pm t_{n-p}(0.975) \cdot \HSE{\hat\mu(\vx)}$$

where $\hat\mu(\vx) = \vx'\hvb$.

Unlike in logistic regression,
no transformation is needed;
the confidence interval is constructed directly on the mean scale.

:::

$$\mu(\vx) \in \paren{\hat\mu(\vx) \pm t_{n-p}(0.975) \cdot \HSE{\hat\mu(\vx)}}$$

:::

---

:::{#exr-se-mean}

::: notes

How can we estimate the standard error of
$\hat\mu(\vx)$?

:::

$$\blue{\HSE{\hat\mu(\vx)}} = \blue{?}$$

:::

---

:::{#sol-se-mean}

$$
\green{\SE{\hat\mu(\vx)}} =
\sqrt{\red{\Var{\hat\mu(\vx)}}}
$${#eq-se-sqrt-var-mean}

By the definition $\hat\mu(\vx) = \vx'\hvb$
and [the variance of a linear combination](probability.qmd#thm-var-lincom):

$$
\ba
\red{\Var{\hat\mu(\vx)}}
&= \Var{\vx'\hvb}
\\
&= \vx'\Cov{\hvb}\vx
\\
&= \red{\tp{\vx}\Sigma\vx}
\ea
$${#eq-var-mean}

where $\Sigma \eqdef \Covt(\hvb)$.

---

::: notes

Expanding @eq-var-mean out of matrix-vector notation, we have:

:::

$$
\ba
\red{\tp{\vx}\Sigma\vx}
&= \sumip\sumjp x_i \Sigma_{ij} x_j
\\
&= \red{\sumip\sumjp x_i \Covt(\hvb_i,\hvb_j) x_j}
\ea
$$

---

Combining @eq-var-mean and the Gauss-Markov theorem:

:::{#thm-est-se-mean}

#### Estimated variance and standard error of predicted mean

$$
\orange{\hVar{\hat\mu(\vx)}} = \orange{\tp{\vx}\hat{\Sigma}\vx}
$${#eq-est-var-mean}

$$
\blue{\HSE{\hat\mu(\vx)}} =
\blue{\sqrt{\tp{\vx}\hat{\Sigma}\vx}}
$${#eq-est-se-mean}

:::

::: notes

Note: on the RHS, we have plugged in $\hat{\Sigma}$,
our estimate of $\Sigma$.

:::

:::

---

#### In R

In R, `predict()` with `se.fit = TRUE`
computes the estimated mean $\hat\mu(\vx) = \vx'\hvb$
and its estimated standard error for each covariate pattern:

```{r}
#| label: tbl-bw-pred-mean
#| tbl-cap: "Predicted means and 95% CI for `birthweight` linear regression"
library(dplyr)
new_data <- tibble(
  age = c(36, 38, 40),
  sex = "male"
)

pred_mean <-
  bw_lm2 |>
  predict(
    newdata = new_data,
    se.fit = TRUE
  )

new_data |>
  mutate(
    mu_hat = pred_mean$fit,
    se = pred_mean$se.fit,
    ci_lower = mu_hat - qt(0.975, df = bw_lm2$df.residual) * se,
    ci_upper = mu_hat + qt(0.975, df = bw_lm2$df.residual) * se
  ) |>
  knitr::kable(digits = 3)
```

Alternatively, `predict()` with `interval = "confidence"` gives the same result:

```{r}
#| label: tbl-bw-pred-ci
#| tbl-cap: "Predicted means and 95% CI using `interval = 'confidence'`"
bw_lm2 |>
  predict(
    newdata = new_data,
    interval = "confidence"
  ) |>
  cbind(new_data) |>
  knitr::kable(digits = 3)
```
