
::: notes
The most direct way to examine the fit of a model is 
to compare it to the raw observed data.
:::

```{r}
#| label: fig-bw-interaction2
#| fig-cap: "Birthweight model with interaction term"
bw <-
  bw |>
  mutate(
    predlm2 = predict(bw_lm2)
  ) |>
  arrange(sex, age)

plot1_interact <-
  plot1 %+% bw +
  geom_line(aes(y = predlm2))

print(plot1_interact)
```

::: notes
It's not easy to assess these assumptions from this model.
If there are multiple continuous covariates, 
it becomes even harder to visualize the raw data.
:::

---

#### Fitted model for `hers` data {.smaller}

Consider the `hers` data from @vittinghoff2e.

{{< include _subfiles/shared/_sec_hers_intro.qmd >}}

::: {.landscape}

:::{#tbl-hers-data}

```{r}
hers <- rmb::hers |> haven::zap_labels()
hers
```

`hers` data
:::

:::

---

::: notes
Suppose we consider models with and without intercept terms (i.e.,
possibly forcing the intercept to go through 0):
:::

```{r}
hers_lm_with_int <- lm(
  na.action = na.exclude,
  LDL ~ smoking * age, data = hers
)


library(equatiomatic)
equatiomatic::extract_eq(hers_lm_with_int)
```

```{r}
hers_lm_no_int <- lm(
  na.action = na.exclude,
  LDL ~ age + smoking:age - 1, data = hers
)

library(equatiomatic)
equatiomatic::extract_eq(hers_lm_no_int)
```


::: {.landscape}

:::{#tbl-hers-maybe-int layout-ncol=2}

:::{#tbl-hers-with-int}


```{r}
library(gtsummary)
hers_lm_with_int |> 
  tbl_regression(intercept = TRUE)
```
With intercept
:::

:::{#tbl-hers-no-int}

```{r}
hers_lm_no_int |> 
  tbl_regression(intercept = TRUE)
```

No intercept
:::

`hers` data models with and without intercepts

:::

:::

---

:::{#fig-hers-maybe-int layout-ncol=2}


:::{#fig-hers-no-int}

```{r}
library(sjPlot)

hers_plot1 <- hers_lm_no_int |>
  sjPlot::plot_model(
    type = "pred",
    terms = c("age", "smoking"),
    show.data = TRUE
  ) +
  facet_wrap(~group_col, ncol = 1) +
  expand_limits(y = 0) +
  theme(legend.position = "bottom")

hers_plot1
```

No intercept

:::

:::{#fig-hers-with-intercept}

```{r}
library(sjPlot)

hers_plot2 <- hers_lm_with_int |>
  sjPlot::plot_model(
    type = "pred",
    terms = c("age", "smoking"),
    show.data = TRUE
  ) +
  facet_wrap(~group_col, ncol = 1) +
  expand_limits(y = 0) +
  theme(legend.position = "bottom")

hers_plot2
```

With intercept
:::

`hers` data models with and without intercepts

:::

