:::{.callout-note}
This section is based on [@vittinghoff2e, Chapter 4].
:::

::: notes

{{< include _subfiles/shared/_sec_hers_intro.qmd >}}

:::

```{r}
#| eval: false
#| code-fold: show
library(haven)
hers <- haven::read_dta(
  paste0(
    "https://regression.ucsf.edu/sites/g/files",
    "/tkssra6706/f/wysiwyg/home/data/hersdata.dta"
  )
)
```

```{r}
#| include: false
library(haven)
hers <-
  fs::path_package("rme", "extdata/hersdata.dta") |>
  read_dta() |>
  dplyr::mutate(
    HT = as_factor(HT) |>
      relevel(ref = "placebo"),
    statins = as_factor(statins) |>
      relevel(ref = "no")
  )

# Snapshot the factor-coded `hers` so the diagnostics sections below use
# exactly the data the models were fit on, regardless of any later
# reassignment of `hers`.
hers_lm_data <- hers
```

#### Data as table

:::{#tbl-hers-ch4}

```{r}
#| code-fold: true
hers |> head()
```

`hers` data

:::

{{< slidebreak >}}

#### Data as graph

:::{#fig-hers-scatter}

```{r}
#| code-fold: true
hers_scatter_data <- hers |>
  dplyr::filter(!is.na(age), !is.na(BMI), !is.na(LDL), !is.na(statins))
hers_scatter_no <- hers_scatter_data |> dplyr::filter(statins == "no")
hers_scatter_yes <- hers_scatter_data |> dplyr::filter(statins == "yes")

hers_scatter_axes <- list(
  xaxis = list(title = "Age (yr)"),
  yaxis = list(title = "BMI (kg/m²)"),
  zaxis = list(title = "LDL (mg/dL)")
)

plotly::plot_ly() |>
  plotly::add_trace(
    data = hers_scatter_no,
    x = ~age, y = ~BMI, z = ~LDL,
    type = "scatter3d", mode = "markers",
    name = "No statins", scene = "scene",
    marker = list(size = 3, opacity = 0.5, color = "steelblue")
  ) |>
  plotly::add_trace(
    data = hers_scatter_yes,
    x = ~age, y = ~BMI, z = ~LDL,
    type = "scatter3d", mode = "markers",
    name = "Yes statins", scene = "scene2",
    marker = list(size = 3, opacity = 0.5, color = "darkorange")
  ) |>
  plotly::layout(
    annotations = list(
      list(
        text = "No statins", x = 0.22, y = 0.95,
        xref = "paper", yref = "paper", showarrow = FALSE
      ),
      list(
        text = "Yes statins", x = 0.78, y = 0.95,
        xref = "paper", yref = "paper", showarrow = FALSE
      )
    ),
    scene = c(
      hers_scatter_axes,
      list(domain = list(x = c(0, 0.48), y = c(0, 1)))
    ),
    scene2 = c(
      hers_scatter_axes,
      list(domain = list(x = c(0.52, 1), y = c(0, 1)))
    )
  )
```

`hers` data (@vittinghoff2e):
age (years) and BMI (kg/m²) vs. baseline LDL (mg/dL),
split by statin use (left: no statins; right: statins).

:::

{{< slidebreak >}}

#### Key variables

:::{#fig-hers-key-vars}

```{r}
#| code-fold: true
#| fig-height: 7
#| fig-width: 8
library(GGally)
hers |>
  dplyr::select(LDL, HT, BMI, statins, age) |>
  ggpairs(
    mapping = aes(col = statins),
    lower = list(continuous = GGally::wrap("points", alpha = 0.3)),
    columnLabels = c(
      "LDL (mg/dL)",
      "HT",
      "BMI (kg/m²)",
      "Statins",
      "Age (yr)"
    )
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Key variables in `hers` data: outcome (LDL),
treatment (HT), and covariates (BMI, statins, age)

:::

{{< slidebreak >}}

#### Data notation {.smaller}

::: notes
Let's define some notation to represent this data:
:::

-   $Y$: LDL cholesterol (mg/dL)
-   $A$: age (years)
-   $B$: BMI (kg/m²)
-   $T$: hormone therapy treatment assignment
    ("placebo" or "hormone therapy")
-   $H$: indicator variable for $T$ = "hormone therapy"
    -   $H = 0$ if $T$ = "placebo"
    -   $H = 1$ if $T$ = "hormone therapy"
-   $U$: statin use ("no" or "yes")
-   $V$: indicator variable for $U$ = "yes"
    -   $V = 0$ if $U$ = "no"
    -   $V = 1$ if $U$ = "yes"

::: notes
"Placebo" is the **reference level** for the categorical variable $T$,
and "no" is the **reference level** for statin use $U$.
The choice of reference level is arbitrary;
it only affects the interpretation of the intercept and corresponding indicator coefficients.

Since LDL is measured at **baseline** (before the hormone therapy was administered),
$H$ is not included as a predictor in our regression models for LDL.
We instead focus on statin use $U$ (and its indicator $V$) as the key grouping variable.
:::
