::: notes
We compare the parallel-planes model (`hers_lm1`: `LDL ~ age + BMI`)
and the interaction model (`hers_lm2`: `LDL ~ age + BMI + age:BMI`)
using AIC, BIC, and deviance.
Lower AIC and BIC values indicate a better trade-off between fit and complexity;
lower deviance indicates a better-fitting model (at the cost of more parameters).
:::

#### AIC and BIC for `hers` models

::: {#tbl-hers-aic-bic}

```{r}
#| code-fold: true
aic <- AIC(hers_lm1, hers_lm2)
bic <- BIC(hers_lm1, hers_lm2)
data.frame(
  df  = aic$df,
  AIC = aic$AIC,
  BIC = bic$BIC,
  row.names = rownames(aic)
) |>
  knitr::kable()
```

AIC and BIC for the `hers` parallel-planes (`hers_lm1`) and interaction (`hers_lm2`) models

:::

::: notes
Both AIC and BIC are slightly lower for the parallel-planes model (@tbl-hers-aic-bic):
adding the age-BMI interaction does not improve either criterion,
so both favor the simpler parallel-planes specification ---
BIC, which penalizes extra parameters more heavily, by the wider margin.
:::


#### Deviance for `hers` models

::: {#tbl-hers-deviance}

```{r}
#| code-fold: true
data.frame(
  deviance = c(deviance(hers_lm1), deviance(hers_lm2)),
  row.names = c("hers_lm1", "hers_lm2")
) |>
  knitr::kable()
```

Deviance for the `hers` parallel-planes (`hers_lm1`) and interaction (`hers_lm2`) models

:::

::: notes
The interaction model has marginally lower deviance (@tbl-hers-deviance) ---
as it must, since adding a parameter can only reduce the residual sum of squares ---
but the reduction is tiny relative to the total.
The likelihood ratio test in @tbl-hers-lrt formally tests
whether that improvement is larger than we would expect by chance from one extra parameter.
:::

