### Partial F-test for nested linear models {#sec-partial-F-test}

::: notes
See @vittinghoff2e [§4.3] and @dobson4e [§6.3] for further discussion.
:::

#### Setup {#sec-partial-F-test-setup}

Suppose we have two **nested** linear regression models:

::: {#def-nested-lm}
#### Nested linear models

$$
\E{Y \mid \vx}
= \tp{\vx}\vb
= \beta_0 + \beta_1 x_1 + \cdots + \beta_{p_1 - 1} x_{p_1 - 1}
$${#eq-lm-reduced}

$$
\E{Y \mid \vx, \vec{z}}
= \tp{\vx}\vb + \tp{\vec{z}}\vec{\g}
= \beta_0 + \beta_1 x_1 + \cdots + \beta_{p_1 - 1} x_{p_1 - 1}
  + \gamma_1 z_1 + \cdots + \gamma_q z_q
$${#eq-lm-full}

The **reduced model** (@eq-lm-reduced) has $p_1$ parameters.
The **full model** (@eq-lm-full) adds $q$ extra predictors
and has $p_2 = p_1 + q$ parameters.
The reduced model is a special case of the full model
with the constraint $\vec{\g} = \vec 0$.
:::

---

The null and alternative hypotheses are:

$$H_0: \vec{\g} = \vec 0 \qquad \text{(reduced model)}$$
$$H_A: \vec{\g} \neq \vec 0 \qquad \text{(full model)}$$

---

::: {#exm-nested-lm-bw}
#### Nested models: `birthweight` data

In the `birthweight` example, the **reduced model** has $p_1 = 3$ parameters:

$$\E{\text{weight} \mid \text{sex}, \text{age}} = \beta_0 + \beta_{\text{sex}} \cdot \text{sex} + \beta_A \cdot \text{age}$$

The **full model** adds an interaction term ($q = 1$ extra parameter, $p_2 = 4$):

$$\E{\text{weight} \mid \text{sex}, \text{age}} = \beta_0 + \beta_{\text{sex}} \cdot \text{sex} + \beta_A \cdot \text{age} + \beta_{AM} \cdot \text{sex} \cdot \text{age}$$

$H_0: \beta_{AM} = 0$ vs. $H_A: \beta_{AM} \neq 0$.
:::

---

#### F-statistic {#sec-F-statistic}

::: {#def-partial-F-statistic}
#### Partial F-statistic

Let $\text{RSS}_0$ and $\text{RSS}_1$ denote the residual sums of squares
from the reduced and full models, respectively.
The **partial F-statistic** is:

$$
F
= \frac{(\text{RSS}_0 - \text{RSS}_1) \,/\, q}
       {\text{RSS}_1 \,/\, (n - p_2)}
$${#eq-partial-F}

where:

-   $\text{RSS}_0 = \sumin (y_i - \hat{y}_i^{(0)})^2$ is the residual SS under $H_0$
-   $\text{RSS}_1 = \sumin (y_i - \hat{y}_i^{(1)})^2$ is the residual SS under $H_A$
-   $q = p_2 - p_1$ is the number of constraints (extra parameters in the full model)
-   $n - p_2$ is the residual degrees of freedom of the full model
:::

---

::: {#thm-partial-F-null-dist}
#### Null distribution of the partial F-statistic

Under $H_0$ and the Gaussian linear regression assumptions,

$$
F \dist F_{q,\; n - p_2}
$${#eq-F-null-dist}

This is an **exact** result (not an asymptotic approximation):
it holds for any sample size $n$ when the errors $\epsilon_i \siid N(0, \sigma^2)$.
:::

---

::: proof

Under $H_0$, the extra predictors $\vec{z}$ contribute nothing,
so the numerator $\text{RSS}_0 - \text{RSS}_1 \dist \sigma^2 \chi^2_q$
and the denominator $\text{RSS}_1 \dist \sigma^2 \chi^2_{n-p_2}$
are independent chi-squared random variables
(this follows from the Gauss-Markov theorem and the properties
of projections in the column spaces of the design matrices).
Therefore,

$$
F
= \frac{(\text{RSS}_0 - \text{RSS}_1)/q}{\text{RSS}_1/(n - p_2)}
= \frac{\chi^2_q / q}{\chi^2_{n-p_2} / (n-p_2)}
\dist F_{q,\; n-p_2}.
$$

:::

---

#### Connection to deviance {#sec-F-test-deviance}

From @sec-compare-lm-glm-deviance,
the Gaussian deviance of model $k$ is
$D_k = \text{RSS}_k / \hat\sigma^2$,
where $\hat\sigma^2$ is an estimate of $\sigma^2$.
Because $\hat\sigma^2$ appears in both the numerator and denominator
of @eq-partial-F, it cancels:

$$
F
= \frac{(D_0 - D_1) \,/\, q}{D_1 \,/\, (n - p_2)}
= \frac{(\text{RSS}_0 - \text{RSS}_1) \,/\, q}{\text{RSS}_1 \,/\, (n - p_2)}
$${#eq-F-from-deviance}

The denominator $s^2 \eqdef \text{RSS}_1/(n - p_2)$
is an unbiased estimator of $\sigma^2$ under both $H_0$ and $H_A$.

---

#### Connection to the likelihood ratio test {#sec-F-test-LRT}

The **approximate likelihood ratio test** (LRT) for MLEs
(see the [table of Gaussian vs. MLE-based tests](intro-MLEs.qmd#tbl-gaussian-vs-mle-tests)
and @sec-compare-lm-glm-deviance) uses the statistic:

$$
\lambda = 2(\llik_1 - \llik_0) \dot \sim \chi^2_q
\quad \text{under } H_0 \text{ (asymptotically)}
$${#eq-LRT-stat}

For Gaussian linear regression,
the MLE of $\sigma^2$ under model $k$ is $\hat\sigma^2_k = \text{RSS}_k / n$,
so the log-likelihood at the MLE is:

$$
\ba
\llik_k
&= -\frac{n}{2}\log(2\pi\hat\sigma^2_k) - \frac{n}{2}
\\
&= -\frac{n}{2}\log\!\paren{\frac{2\pi\, \text{RSS}_k}{n}} - \frac{n}{2}
\ea
$${#eq-gaussian-loglik-at-MLE}

Substituting into @eq-LRT-stat:

$$
\ba
\lambda
&= 2(\llik_1 - \llik_0)
\\
&= 2\left[
   -\frac{n}{2}\log\!\paren{\frac{\text{RSS}_1}{n}}
   +\frac{n}{2}\log\!\paren{\frac{\text{RSS}_0}{n}}
   \right]
\\
&= n \log\!\paren{\frac{\text{RSS}_0}{\text{RSS}_1}}
\ea
$${#eq-gaussian-LRT}

---

##### Asymptotic equivalence of F-test and LRT

For large $n$, $F$ and $\lambda$ are approximately related by:

$$
\lambda \approx q \cdot F
$${#eq-F-LRT-approx}

This follows because when $H_0$ is true and $n$ is large,
$(\text{RSS}_0 - \text{RSS}_1) / \text{RSS}_1$ is small, and:

$$
\ba
\lambda
&= n \log\!\paren{\frac{\text{RSS}_0}{\text{RSS}_1}}
\\
&= n \log\!\paren{1 + \frac{\text{RSS}_0 - \text{RSS}_1}{\text{RSS}_1}}
\\
&\approx n \cdot \frac{\text{RSS}_0 - \text{RSS}_1}{\text{RSS}_1}
\ea
$$

$$
\ba
q \cdot F
&= q \cdot \frac{(\text{RSS}_0 - \text{RSS}_1)/q}{\text{RSS}_1/(n-p_2)}
\\
&= (n - p_2) \cdot \frac{\text{RSS}_0 - \text{RSS}_1}{\text{RSS}_1}
\\
&\approx n \cdot \frac{\text{RSS}_0 - \text{RSS}_1}{\text{RSS}_1}
\ea
$$

So $\lambda \approx q \cdot F$ for large $n$,
and both statistics have the same asymptotic null distribution $\chi^2_q$
(since $q \cdot F_{q, n-p_2} \dot \sim \chi^2_q$ as $n \to \infty$).

---

::: {.callout-note}
#### Why use the F-test instead of the LRT?

For Gaussian linear regression:

- The **F-test** is **exact** — it has the correct $F_{q,n-p_2}$ null distribution
  for any sample size $n$,
  provided the errors are Gaussian.
  It accounts for the estimation of $\sigma^2$ via the residual degrees of freedom.

- The **LRT** is **approximate** — it relies on the asymptotic $\chi^2_q$ distribution,
  which requires large $n$ and treats $\sigma^2$ as known (fixed at its MLE).

Both tests give similar p-values for large $n$.
For small to moderate $n$, the F-test is preferred because it is exact.

For non-Gaussian GLMs (Poisson, Binomial),
the F-test is not applicable;
the LRT (or Wald test) is the standard approach.
:::

---

#### In R {#sec-partial-F-test-R}

In R, the partial F-test for two nested `lm` models
is performed with `anova()`:

```{r}
#| label: tbl-partial-F-bw
#| code-fold: false
anova(bw_lm1, bw_lm2)
```

::: notes
`anova(bw_lm1, bw_lm2)` performs the partial F-test comparing
the reduced model `bw_lm1` (parallel slopes, no interaction)
against the full model `bw_lm2` (with sex-age interaction).
The output shows the RSS for each model, the difference, the F-statistic, and the p-value.
:::

---

For comparison, the approximate likelihood ratio test using the `lmtest` package:

```{r}
#| label: tbl-LRT-bw-compare
#| code-fold: false
lmtest::lrtest(bw_lm1, bw_lm2)
```

::: notes
Compare the p-values from `anova()` (exact F-test) and `lmtest::lrtest()` (approximate LRT).
They are similar but not identical:
the F-test uses the $F_{1, n-4}$ distribution,
while the LRT uses the asymptotic $\chi^2_1$ distribution.
For large $n$, these p-values converge.
:::

---

##### Extracting F-test components by hand

To understand the computation, we can replicate the F-statistic manually:

```{r}
#| label: tbl-F-manual
#| code-fold: false
rss0 <- deviance(bw_lm1)   # RSS of reduced model
rss1 <- deviance(bw_lm2)   # RSS of full model
n    <- nobs(bw_lm2)
p2   <- length(coef(bw_lm2))
q    <- length(coef(bw_lm2)) - length(coef(bw_lm1))
s2   <- rss1 / (n - p2)    # residual variance estimate from full model

F_stat <- ((rss0 - rss1) / q) / s2
p_val  <- pf(F_stat, df1 = q, df2 = n - p2, lower.tail = FALSE)

cat("RSS_0 =", rss0, "\n")
cat("RSS_1 =", rss1, "\n")
cat("q     =", q,    "\n")
cat("s^2   =", s2,   "\n")
cat("F     =", F_stat, "\n")
cat("p-val =", p_val,  "\n")
```

::: notes
The LRT statistic $\lambda = n \log(\text{RSS}_0 / \text{RSS}_1) \approx q \cdot F$
for large $n$.
:::

```{r}
#| label: tbl-LRT-manual
#| code-fold: false
lambda <- n * log(rss0 / rss1)
cat("LRT statistic lambda =", lambda, "\n")
cat("q * F               =", q * F_stat, "\n")
cat("LRT p-val (chi^2)   =", pchisq(lambda, df = q, lower.tail = FALSE), "\n")
```

::: notes
The LRT statistic $\lambda$ and $q \cdot F$ are close in value,
illustrating the asymptotic equivalence @eq-F-LRT-approx.
:::
