:::{.callout-tip}
This section is adapted from @dobson4e [§6.3].
:::

Residuals and leverage each capture one aspect of an observation's influence on the model fit.
Cook's distance [@cook1977detection] combines them into a single diagnostic.

{{< slidebreak >}}

:::{#def-cooks-distance}
#### Cook's distance

For ordinary least squares with $p$ parameters,
the **Cook's distance** for observation $i$ is

$$
D_i
= \frac{\tp{(\hvb_{(-i)} - \hvb)}
\tp{\mX}\mX
(\hvb_{(-i)} - \hvb)}{p\,\hs^2}
$$ {#eq-cooks-distance-def}

where $\hvb_{(-i)}$ is the least-squares estimate obtained after deleting observation $i$.
:::

{{< slidebreak >}}

Because refitting the model $n$ times is expensive,
@eq-cooks-distance-def has an equivalent closed-form expression
that uses only quantities from the full-data fit:

$$
D_i = \frac{(\stdresid_i)^2}{p} \cdot \frac{h_{ii}}{1-h_{ii}}
$$ {#eq-cooks-distance-formula}

where $\stdresid_i$ is the standardized residual (@def-stred)
and $h_{ii}$ is the leverage (diagonal of the hat matrix, @def-hat-matrix).

{{< slidebreak >}}

:::{#exm-cooks-distance}
Consider a regression with $p = 2$ parameters.
Suppose observation $i$ has standardized residual $\stdresid_i = 2$ and leverage $h_{ii} = 0.3$.
Applying @eq-cooks-distance-formula,

$$
D_i = \frac{(2)^2}{2} \cdot \frac{0.3}{1 - 0.3} = 2 \cdot 0.43 \approx 0.86.
$$
:::

{{< slidebreak >}}

@eq-cooks-distance-formula shows that influence has two ingredients:

- **Residual size** ($(\stdresid_i)^2$):
  a poorly fitted observation pushes $D_i$ up regardless of its leverage.
- **Leverage** ($h_{ii}/(1-h_{ii})$):
  high leverage amplifies the effect of even a moderate residual.

An observation must be both unusual in its predictor values *and* poorly fitted to have a large Cook's distance.

{{< slidebreak >}}

#### Cook's distance plot

::: notes
We can visualize Cook's distances for all observations using `autoplot` with `which = 4` (@fig-bw-cooks-distance).
:::

:::{#fig-bw-cooks-distance}

```{r}
#| code-fold: true
autoplot(bw_lm2, which = 4, ncol = 1) |> print()
```

Cook's distance for `birthweight` model with interactions (@eq-BW-lm-interact)
:::

{{< slidebreak >}}

::: notes
A commonly cited rule of thumb [@dobson4e, §6.3]
flags observation $i$ as influential if

$$D_i > 1$$

A more conservative threshold is $D_i > 4/n$ [@bollen1990regression].

Observations above either threshold warrant closer inspection:
check whether they represent genuine data or recording errors,
and consider refitting the model with and without them to assess sensitivity.
:::

{{< slidebreak >}}

::: notes
The residuals-versus-leverage plot (@fig-bw_lm2_resid-vs-leverage)
also displays Cook's distance contours as curved dashed lines.
An observation lying near or beyond a contour line has substantial influence on the fit.
:::

{{< slidebreak >}}

Computing Cook's distances in R:

```{r}
cooks.distance(bw_lm2)
```

{{< slidebreak >}}

:::{#exr-cooks-d-formula}
Verify @eq-cooks-distance-formula for the observations in `bw_lm2`:
compute $D_i$ using the standardized residuals and leverage values,
and check that the result matches `cooks.distance(bw_lm2)`.
:::

:::{.solution}
```{r}
r_prime <- rstandard(bw_lm2)
h <- hatvalues(bw_lm2)
p <- length(coef(bw_lm2))

D_manual <- (r_prime^2 / p) * (h / (1 - h))
cbind(manual = D_manual, builtin = cooks.distance(bw_lm2)) |> head(3)
```
:::
