
::: notes
Another way to assess whether the standardized residuals follow a standard normal distribution
is a **quantile-quantile (QQ) plot**.

A QQ plot compares the empirical distribution of a dataset to a theoretical reference distribution.
Each point corresponds to one observation:

- The **x-axis** (Theoretical Quantiles) shows the $N(0,1)$ quantile
  corresponding to the plotting position $p_i$
  used for the $i$th ordered residual.
  In R, `qqnorm()` obtains these plotting positions
  from `ppoints(n)`.
- The **y-axis** (Standardized Residuals) shows the observed standardized residuals,
  plotted in increasing order as $r_{(1)}\le\cdots\le r_{(n)}$.
  In QQ-plot terminology,
  these ordered values are [sample quantiles](nonparametric-models.qmd#def-sample-quantile).
  The QQ plot uses points
  $$\left(\Phi^{-1}(p_i),\, r_{(i)}\right), \quad i = 1, \ldots, n,$$
  where $p_i$ denotes the plotting positions returned by `ppoints(n)`.
  For the common midpoint choice $p_i=(i-0.5)/n$,
  the sample quantile at $p_i$ equals $r_{(i)}$.
  More generally,
  `qqnorm()` pairs the $i$th ordered residual
  with the corresponding theoretical quantile
  computed from its plotting position.
  So both labels are correct:
  `qqnorm()` emphasizes the quantile role ("Sample Quantiles"),
  while `autoplot.lm()` emphasizes the underlying ordered values
  ("Standardized residuals").

If the standardized residuals approximately follow $N(0,1)$,
the points should fall approximately on a straight line.
:::

```{r}
library(ggfortify)
# needed to make ggplot2::autoplot() work for `lm` objects

qqplot_lm2_auto <-
  bw_lm2 |>
  autoplot(
    which = 2, # options are 1:6; can do multiple at once
    ncol = 1
  ) +
  theme_classic()

print(qqplot_lm2_auto)
```

::: notes
If the Gaussian model were correct, the points should fall approximately on the reference line.

The **reference line** passes through the pairs of theoretical and empirical quantiles
at probabilities 0.25 and 0.75:
it connects $(\Phi^{-1}(0.25),\, \hat{Q}_{0.25})$ and $(\Phi^{-1}(0.75),\, \hat{Q}_{0.75})$,
where $\Phi^{-1}$ is the standard normal quantile function
and $\hat{Q}_p$ is the $p$-th sample quantile of the standardized residuals.
If the standardized residuals are approximately $N(0,1)$,
this line should be close to the identity line $y = x$,
and the points should fall roughly along a straight line.

Systematic deviations from the line suggest departures from normality:

- **S-shaped curves**: heavier or lighter tails than the normal distribution
- **Concave or convex curves**: skewness
- **Individual points far from the line**: potential outliers

Fig 2.4 panel (c) in @dobson4e is a little different; they didn't specify how they produced it, but other statistical analysis systems do things differently from R.

See also @dunn2018generalized [§3.5.4](https://link.springer.com/chapter/10.1007/978-1-4419-0118-7_3#Sec14:~:text=3.5.4%20Q%E2%80%93Q%20Plots%20and%20Normality).

:::

---

## QQ plot: typical patterns

:::{#fig-qqplot-patterns}

```{r}
#| fig-height: 5
#| fig-width: 10
set.seed(36) # seed chosen for Fig. 3.6 reproduction
n <- 150     # sample size as in @dunn2018generalized Fig. 3.6

sims <- list(
  "Right skewed"    = rchisq(n, df = 3),
  "Left skewed"     = -rchisq(n, df = 3),
  "Tails too heavy" = rt(n, df = 3),
  "Tails too light" = runif(n, -sqrt(3), sqrt(3)),
  "True normal"     = rnorm(n),
  "Bimodal"         = c(rnorm(n / 2, -2.5), rnorm(n / 2, 2.5))
)

old_par <- par(
  mfcol = c(2, 6),
  mar = c(3, 3, 2, 0.5),
  oma = c(0, 0, 0, 0)
)

for (nm in names(sims)) {
  qqnorm(
    sims[[nm]],
    main = nm,
    xlab = "Theoretical Quantiles",
    ylab = "Sample Quantiles",
    pch = 1,
    cex = 0.6
  )
  qqline(sims[[nm]])
  hist(
    sims[[nm]],
    main = "",
    xlab = "Residuals",
    ylab = "Frequency",
    col = "grey70",
    border = "white",
    breaks = 15
  )
}

par(old_par)
```

Typical QQ-plot patterns for six types of residual distribution.
Adapted from @dunn2018generalized [Fig. 3.6], with thanks to the authors;
reproduced here using simulated data ($n = 150$).
Each column shows one distribution scenario:
a QQ plot (top) paired with a histogram of the simulated data (bottom).

:::

---

### QQ plot - how it's built

::: notes
To construct a QQ plot by hand:

1. Assign each standardized residual a **plotting position** $p_i$
   based on its rank among the $n$ observations.
   R's default formula for $n > 10$ is:
   $$p_i = \frac{\text{rank}_i - 0.5}{n}$$
2. Compute the **theoretical quantile**: $q_i = \Phi^{-1}(p_i)$,
   the value from $N(0,1)$ that corresponds to probability $p_i$.
3. Plot $(q_i, \stdresid_i)$ — theoretical quantile $q_i$ on the x-axis,
   observed standardized residual $\stdresid_i$ (see @def-stred) on the y-axis.
4. Add the **reference line** through the pairs of theoretical and empirical quantiles
   at probabilities 0.25 and 0.75.
:::

```{r}
bw <- bw |>
  mutate(
    # plotting position: (rank - 0.5) / n
    p = (rank(std_resid_lm2) - 1 / 2) / n(),
    expected_quantiles_lm2 = qnorm(p)
  )

qqplot_lm2 <-
  bw |>
  ggplot(
    aes(
      x = expected_quantiles_lm2,
      y = std_resid_lm2,
      col = sex,
      shape = sex
    )
  ) +
  geom_point() +
  theme_classic() +
  theme(legend.position = "none") + # removing the plot legend
  ggtitle("Normal Q-Q") +
  xlab("Theoretical Quantiles") +
  ylab("Standardized residuals")

# find the reference line through the (theoretical, empirical) quantile pairs
# at probabilities 0.25 and 0.75:

ps <- c(.25, .75) # reference probabilities
a <- quantile(rstandard(bw_lm2), ps) # empirical quantiles
b <- qnorm(ps) # theoretical quantiles

qq_slope <- diff(a) / diff(b)
qq_intcpt <- a[1] - b[1] * qq_slope

qqplot_lm2 <-
  qqplot_lm2 +
  geom_abline(slope = qq_slope, intercept = qq_intcpt)
```

:::{#fig-qqplot-comparison layout-ncol=3}

:::{#fig-qqplot-manual}

```{r}
print(qqplot_lm2)
```

`ggplot2` (manual construction)

:::

:::{#fig-qqplot-baser}

```{r}
std_resid_lm2 <- rstandard(bw_lm2)

qqnorm(
  std_resid_lm2,
  main = "Normal Q-Q",
  xlab = "Theoretical Quantiles",
  ylab = "Standardized residuals"
)
qqline(std_resid_lm2)
```

base R (`qqnorm` + `qqline`)

:::

:::{#fig-qqplot-autoplot}

```{r}
library(ggfortify)
autoplot(
  bw_lm2,
  which = 2,
  ncol = 1
) +
  theme_classic()
```

`ggfortify::autoplot()`

:::

Three equivalent ways to produce a QQ plot of the standardized residuals
for the `birthweight` model (@eq-BW-lm-interact).
All three plots show the same data and reference line.

:::

---

