::: notes
We compare the parallel-planes model (`hers_lm1`: `LDL ~ age + BMI`)
and the interaction model (`hers_lm2`: `LDL ~ age + BMI + age:BMI`)
using AIC, BIC, and deviance.
Lower AIC and BIC values indicate a better trade-off between fit and complexity;
lower deviance indicates a better-fitting model (at the cost of more parameters).
:::

#### AIC and BIC for `hers` models

::: {#tbl-hers-aic-bic}

```{r}
#| code-fold: true
aic <- AIC(hers_lm1, hers_lm2)
bic <- BIC(hers_lm1, hers_lm2)
data.frame(
  df  = aic$df,
  AIC = aic$AIC,
  BIC = bic$BIC,
  row.names = rownames(aic)
) |>
  knitr::kable()
```

AIC and BIC for the `hers` parallel-planes (`hers_lm1`) and interaction (`hers_lm2`) models

:::

The model with the lower AIC or BIC value is preferred as the better balance of fit and parsimony.
A smaller value for the interaction model indicates the added age-BMI interaction term is worthwhile;
a smaller value for the parallel-planes model would favor the simpler specification.


#### Deviance for `hers` models

::: {#tbl-hers-deviance}

```{r}
#| code-fold: true
data.frame(
  deviance = c(deviance(hers_lm1), deviance(hers_lm2)),
  row.names = c("hers_lm1", "hers_lm2")
) |>
  knitr::kable()
```

Deviance for the `hers` parallel-planes (`hers_lm1`) and interaction (`hers_lm2`) models

:::

Lower deviance indicates a closer fit to the data; the interaction model uses one more parameter.
The likelihood ratio test in @tbl-hers-lrt provides a formal comparison of whether the improvement justifies the added complexity.

