{{< include latex-macros/macros.qmd >}}

These paper-and-pencil exercises
use actual datasets.
No numerical computation is required.

## Exercise 1: Main-effects model interpretation {#exr-lm-eoc-main-effects}

Adapted from
@vittinghoff2e [Chapter 4].

Use the
`ToothGrowth`
dataset.
Consider the additive model

$$
\E{\texttt{len}|\texttt{dose}=x,\texttt{supp}=z}
= \b_0 + \b_{\texttt{dose}}x + \b_{\texttt{supp}}z,
$$

where
$z=1$
for
`VC`
and
$z=0$
for
`OJ`.

State the interpretation of
$\b_0$,
$\b_{\texttt{dose}}$,
and
$\b_{\texttt{supp}}$.
Which interpretations depend on the reference level
`OJ`?

```{r}
#| label: fig-eoc-lm-main-effects
#| fig-cap: "ToothGrowth data with additive (parallel-lines) fit"
eoc_tg <-
  ToothGrowth |>
  mutate(supp = relevel(supp, ref = "OJ"))

eoc_tg_add <-
  lm(len ~ dose + supp, data = eoc_tg)

eoc_tg_grid <-
  expand_grid(
    dose = seq(min(eoc_tg$dose), max(eoc_tg$dose), length.out = 100),
    supp = levels(eoc_tg$supp)
  ) |>
  mutate(len_hat = predict(eoc_tg_add, newdata = cur_data()))

ggplot(eoc_tg, aes(x = dose, y = len, color = supp)) +
  geom_point(alpha = 0.7) +
  geom_line(
    data = eoc_tg_grid,
    aes(y = len_hat)
  ) +
  labs(color = "supplement")
```

::: {.callout-note collapse="true"}
### Solution

$\b_0$
is the mean tooth length
when
`dose = 0`
and
`supp = OJ`.
So it depends on the reference level
`OJ`.

$\b_{\texttt{dose}}$
is the change in mean tooth length
for a one-unit increase in dose,
holding supplement type fixed.
Because the model is additive,
this slope is the same for both supplement groups.

$\b_{\texttt{supp}}$
is the mean difference
(`VC` minus `OJ`)
at a fixed dose.
Its interpretation depends on
`OJ`
being the reference level.
:::

---

## Exercise 2: Interaction interpretation {#exr-lm-eoc-interaction}

Adapted from
@dobson4e [Section 6.7.1].

Use the same
`ToothGrowth`
dataset,
but now consider the interaction model

$$
\E{\texttt{len}|\texttt{dose}=x,\texttt{supp}=z}
= \b_0 + \b_{\texttt{dose}}x + \b_{\texttt{supp}}z
+ \b_{\texttt{dose:supp}}(x\cdot z).
$$

Using the plot,
write the slope with respect to
`dose`
for
`OJ`
and for
`VC`.
Then interpret
$\b_{\texttt{dose:supp}}$.

```{r}
#| label: fig-eoc-lm-interaction
#| fig-cap: "ToothGrowth data with interaction (non-parallel-lines) fit"
eoc_tg_int <-
  lm(len ~ dose * supp, data = eoc_tg)

eoc_tg_grid2 <-
  expand_grid(
    dose = seq(min(eoc_tg$dose), max(eoc_tg$dose), length.out = 100),
    supp = levels(eoc_tg$supp)
  ) |>
  mutate(len_hat = predict(eoc_tg_int, newdata = cur_data()))

ggplot(eoc_tg, aes(x = dose, y = len, color = supp)) +
  geom_point(alpha = 0.7) +
  geom_line(
    data = eoc_tg_grid2,
    aes(y = len_hat)
  ) +
  labs(color = "supplement")
```

::: {.callout-note collapse="true"}
### Solution

For
`OJ`
(`z=0`),
the slope is
$\b_{\texttt{dose}}$.

For
`VC`
(`z=1`),
the slope is
$\b_{\texttt{dose}} + \b_{\texttt{dose:supp}}$.

Therefore,
$\b_{\texttt{dose:supp}}$
is the difference in dose slopes
between
`VC`
and
`OJ`.
:::

---

## Exercise 3: Centering and coefficient changes {#exr-lm-eoc-centering}

Adapted from end-of-chapter exercises in
@kleinbaum2014applied [Chapter 14].

Use the
`PLOS`
dataset from
`dobson`.
Define
$X = \texttt{authors}$
and
$X^* = X - 10$.
Start from

$$
\E{\texttt{nchar}|X=x}
= \b_0 + \b_Xx.
$$

Rewrite the model in terms of
$X^*$.
Express
$\g_0$
and
$\g_X$
in terms of
$\b_0$
and
$\b_X$.

```{r}
#| label: fig-eoc-lm-centering
#| fig-cap: "PLOS data with original and centered parameterizations"
data(PLOS, package = "dobson")

eoc_plos <-
  PLOS |>
  mutate(authors_centered = authors - 10)

eoc_plos_orig <-
  lm(nchar ~ authors, data = eoc_plos)

eoc_plos_cent <-
  lm(nchar ~ authors_centered, data = eoc_plos)

eoc_plos_grid <-
  tibble(
    authors = seq(
      min(eoc_plos$authors),
      max(eoc_plos$authors),
      length.out = 100
    )
  ) |>
  mutate(
    authors_centered = authors - 10,
    fit_original = predict(eoc_plos_orig, newdata = cur_data()),
    fit_centered = predict(eoc_plos_cent, newdata = cur_data())
  )

ggplot(eoc_plos, aes(x = authors, y = nchar)) +
  geom_point(alpha = 0.7) +
  geom_line(
    data = eoc_plos_grid,
    aes(y = fit_original, color = "original parameterization")
  ) +
  geom_line(
    data = eoc_plos_grid,
    aes(y = fit_centered, color = "centered parameterization"),
    linetype = 2
  ) +
  labs(color = "")
```

::: {.callout-note collapse="true"}
### Solution

Since
$x = x^* + 10$:

$$
\ba
\E{\texttt{nchar}|X^*=x^*}
&= \b_0 + \b_X(x^* + 10)\\
&= (\b_0 + 10\b_X) + \b_Xx^*.
\ea
$$

So

$$
\g_0 = \b_0 + 10\b_X,
\quad
\g_X = \b_X.
$$

Centering shifts the intercept,
but does not change the slope.
:::

---

## Exercise 4: Diagnostics from observed residual patterns {#exr-lm-eoc-diagnostics}

Adapted from
@dunn2018generalized [Chapter 3].

Use the
`mtcars`
dataset and the model

$$
\E{\texttt{mpg}|\texttt{wt}=x,\texttt{hp}=h}
= \b_0 + \b_{\texttt{wt}}x + \b_{\texttt{hp}}h.
$$

Based on the residual-vs-fitted
and Q-Q plots,
which assumptions look most questionable?
State one practical follow-up step.

```{r}
#| label: fig-eoc-lm-diagnostics
#| fig-cap: "Diagnostic plots for linear model fit to mtcars data"
#| layout-ncol: 2
#| fig-subcap:
#| - "Residuals versus fitted"
#| - "Normal Q-Q"
eoc_mtcars <-
  mtcars |>
  as_tibble()

eoc_mtcars_fit <-
  lm(mpg ~ wt + hp, data = eoc_mtcars)

autoplot(eoc_mtcars_fit, which = 1, ncol = 1)
autoplot(eoc_mtcars_fit, which = 2, ncol = 1)
```

::: {.callout-note collapse="true"}
### Solution

The residual-vs-fitted plot should be checked
for curvature and changing spread.
The Q-Q plot should be checked
for tail departures.

If either pattern is substantial,
a reasonable next step is to compare
an alternative specification
(for example,
nonlinear terms or transformation)
and then re-check diagnostics.
:::
