{{< include latex-macros/macros.qmd >}}

These paper-and-pencil exercises
use actual datasets.
No numerical computation is required.

::: {#exr-lm-eoc-main-effects}
Main-effects model interpretation.

Adapted from
the main-effects interpretation exercise in
@vittinghoff2e [Chapter 4].

Dataset summary:
The
`ToothGrowth`
dataset contains one row per guinea pig.
Variable definitions:
- $Y$:
  tooth length (`len`)
- $X$:
  dose (`dose`)
- $Z$:
  supplement indicator (`supp`)
  with
  $Z=1$
  for
  `VC`
  and
  $Z=0$
  for
  `OJ`

Consider the additive model

$$
\E{Y|X=x,Z=z}
= \b_0 + \b_X x + \b_Z z.
$$

State the interpretation of
$\b_0$,
$\b_X$,
and
$\b_Z$.
Which interpretations depend on the reference level
`OJ`?
:::

```{r}
eoc_tg <-
  ToothGrowth |>
  as_tibble() |>
  mutate(supp = relevel(supp, ref = "OJ"))

eoc_tg
glimpse(eoc_tg)
```

```{r}
#| label: fig-eoc-lm-main-effects
#| fig-cap: "ToothGrowth data with additive (parallel-lines) fit"
#| code-fold: true

eoc_tg_add <-
  lm(len ~ dose + supp, data = eoc_tg)

eoc_tg_grid <-
  expand_grid(
    dose = seq(min(eoc_tg$dose), max(eoc_tg$dose), length.out = 100),
    supp = levels(eoc_tg$supp)
  )

eoc_tg_grid <-
  eoc_tg_grid |>
  mutate(len_hat = predict(eoc_tg_add, newdata = eoc_tg_grid))

ggplot(eoc_tg, aes(x = dose, y = len, color = supp)) +
  geom_point(alpha = 0.7) +
  geom_line(
    data = eoc_tg_grid,
    aes(y = len_hat)
  ) +
  labs(color = "supplement")
```

::: {.solution}
\ 

$\b_0$
is the mean tooth length
for guinea pigs given
`OJ`
at dose
0.

$\b_X$
is the change in mean tooth length
for each one-unit increase in dose.
In this additive model,
that dose effect is the same
for
`OJ`
and
`VC`.

$\b_Z$
is the mean tooth-length difference
between
`VC`
and
`OJ`
at the same dose.
The interpretations that depend on
`OJ`
as the reference level are
$\b_0$
and
$\b_Z$.
:::

---

::: {#exr-lm-eoc-interaction}
Interaction interpretation.

Adapted from
the interaction-slope interpretation example in
@dobson4e [Section 6.7.1].

Use
`ToothGrowth`
as introduced in
[Exercise @exr-lm-eoc-main-effects].
Variable definitions:
- $Y$:
  tooth length (`len`)
- $X$:
  dose (`dose`)
- $Z$:
  supplement indicator (`supp`)
  with
  $Z=1$
  for
  `VC`
  and
  $Z=0$
  for
  `OJ`

Consider the interaction model

$$
\E{Y|X=x,Z=z}
= \b_0 + \b_X x + \b_Z z + \b_{XZ}(x\cdot z).
$$

Using the plot,
write the slope with respect to
$X$
for
`OJ`
and for
`VC`.
Then interpret
$\b_{XZ}$.
:::

```{r}
#| label: fig-eoc-lm-interaction
#| fig-cap: "ToothGrowth data with interaction (non-parallel-lines) fit"
#| code-fold: true
eoc_tg_int <-
  lm(len ~ dose * supp, data = eoc_tg)

eoc_tg_grid2 <-
  expand_grid(
    dose = seq(min(eoc_tg$dose), max(eoc_tg$dose), length.out = 100),
    supp = levels(eoc_tg$supp)
  )

eoc_tg_grid2 <-
  eoc_tg_grid2 |>
  mutate(len_hat = predict(eoc_tg_int, newdata = eoc_tg_grid2))

ggplot(eoc_tg, aes(x = dose, y = len, color = supp)) +
  geom_point(alpha = 0.7) +
  geom_line(
    data = eoc_tg_grid2,
    aes(y = len_hat)
  ) +
  labs(color = "supplement")
```

::: {.solution}
\ 

For
`OJ`,
the dose slope is
$\b_X$.

For
`VC`,
the dose slope is
$\b_X + \b_{XZ}$.

Therefore,
$\b_{XZ}$
is the difference in dose slopes
between
`VC`
and
`OJ`.
:::

---

::: {#exr-lm-eoc-centering}
Centering and coefficient changes.

Adapted from
the centering/reparameterization exercise in
@kleinbaum2014applied [Chapter 14].

Dataset summary:
The
`PLOS`
dataset contains one row per paper.
Variable definitions:
- $Y$:
  title length (`nchar`)
- $X$:
  number of authors (`authors`)
- $X^*$:
  centered author count (`authors_centered`),
  constructed from `authors`,
  defined as
  $X^* = X - 10$

Start from

$$
\E{Y|X=x}
= \b_0 + \b_Xx.
$$

Rewrite the model in terms of
$X^*$.
Express
$\g_0$
and
$\g_X$
in terms of
$\b_0$
and
$\b_X$.
:::

```{r}
data(PLOS, package = "dobson")

eoc_plos <-
  PLOS |>
  as_tibble() |>
  mutate(authors_centered = authors - 10)

eoc_plos
glimpse(eoc_plos)
```

```{r}
#| label: fig-eoc-lm-centering
#| fig-cap: "PLOS data with original and centered parameterizations"
#| code-fold: true

eoc_plos_orig <-
  lm(nchar ~ authors, data = eoc_plos)

eoc_plos_cent <-
  lm(nchar ~ authors_centered, data = eoc_plos)

eoc_plos_grid <-
  tibble(
    authors = seq(
      min(eoc_plos$authors),
      max(eoc_plos$authors),
      length.out = 100
    )
  ) |>
  mutate(authors_centered = authors - 10)

eoc_plos_grid <-
  eoc_plos_grid |>
  mutate(
    fit_original = predict(eoc_plos_orig, newdata = eoc_plos_grid),
    fit_centered = predict(eoc_plos_cent, newdata = eoc_plos_grid)
  )

ggplot(eoc_plos, aes(x = authors, y = nchar)) +
  geom_point(alpha = 0.7) +
  geom_line(
    data = eoc_plos_grid,
    aes(y = fit_original, color = "original parameterization")
  ) +
  geom_line(
    data = eoc_plos_grid,
    aes(y = fit_centered, color = "centered parameterization"),
    linetype = 2
  ) +
  labs(color = "")
```

::: {.solution}
\ 

Since
$x = x^* + 10$:

$$
\ba
\E{Y|X^*=x^*}
&= \b_0 + \b_X(x^* + 10)\\
&= (\b_0 + 10\b_X) + \b_Xx^*.
\ea
$$

So

$$
\g_0 = \b_0 + 10\b_X,
\quad
\g_X = \b_X.
$$

Centering shifts the intercept,
but does not change the slope.
:::

---

::: {#exr-lm-eoc-diagnostics}
Diagnostics from observed residual patterns.

Adapted from
the residual-diagnostics example in
@dunn2018generalized [Chapter 3].

Dataset summary:
The
`mtcars`
dataset contains one row per car model.
Variable definitions:
- $Y$:
  fuel economy (`mpg`)
- $X$:
  vehicle weight (`wt`)
- $H$:
  horsepower (`hp`)

Consider the model

$$
\E{Y|X=x,H=h}
= \b_0 + \b_Xx + \b_Hh.
$$

Based on the residual-vs-fitted
and Q-Q plots,
which assumptions look most questionable?
State one practical follow-up step.
:::

```{r}
eoc_mtcars <-
  mtcars |>
  as_tibble()

eoc_mtcars
glimpse(eoc_mtcars)
```

```{r}
#| label: fig-eoc-lm-diagnostics
#| fig-cap: "Diagnostic plots for linear model fit to mtcars data"
#| code-fold: true
#| layout-ncol: 2
#| fig-subcap:
#| - "Residuals versus fitted"
#| - "Normal Q-Q"
eoc_mtcars_fit <-
  lm(mpg ~ wt + hp, data = eoc_mtcars)

autoplot(eoc_mtcars_fit, which = 1, ncol = 1)
autoplot(eoc_mtcars_fit, which = 2, ncol = 1)
```

::: {.solution}
\ 

The residual-vs-fitted plot shows
nonlinearity and non-constant spread.
The Q-Q plot shows
mild tail departures from normality.

The correct conclusion is that
linearity and constant-variance assumptions
are the most questionable here.
A practical next step is to fit
a model with a nonlinear weight term
(for example,
a spline or polynomial in
`wt`)
and then re-check diagnostics.
:::
