{{< include macros.qmd >}}

## Acknowledgements{.unnumbered}

This content is adapted from:

-   Dobson & Barnett 2018, "An Introduction to Generalized Linear Models", 4th edition, Chapter 9
-   Vittinghoff et al 2012, "Regression Methods in Biostatistics", 2nd edition, Chapter 8
-   Lecture slides and videos from the 2021 edition of this course (http://dmrocke.ucdavis.edu/Class/EPI204-Spring-2021/EPI204-Spring-2021.html)

::: {.content-hidden when-format="revealjs"}
## Configuring R{.unnumbered}

Functions from these packages will be used throughout this document:

```{r packages, message = FALSE}
library(pander) # format tables for markdown
library(ggplot2) # graphics
library(dplyr) # manipulate data
library(haven) # import Stata files
library(knitr) # format R output for markdown
library(tidyr) # Tools to help to create tidy data
library(plotly) # interactive graphics
library(dobson) # datasets from Dobson and Barnett 2018
library(parameters) # format model output tables for markdown
library(conflicted) # check for conflicting function definitions
```

Here are some R settings I use in this document:

```{r options, message=FALSE}
rm(list = ls()) # delete any data that's already loaded into R
knitr::opts_chunk$set(message = FALSE)
pander::panderOptions("table.emphasize.rownames", FALSE)
options("digits" = 4)
```
:::

# Introduction

There are three intersecting modeling challenges in this section.

1. censoring

3. time-to-event modeling

4. semiparametric models

## Time to Event Data
* Survival Analysis is a term for analyzing time-to-event data.
* This is used in clinical trials, where the event is often death or
recurrence of disease.
\item It is used in engineering reliability analysis, where the event is failure
of a device or system.
\item It is used in insurance, particularly life insurance, where the event is death.
\ei
\end{frame}

# The exponential distribution

## PDF

## Mean

## Variance

## MLE

## Information

# some new quantities

## survival function

## hazard function

# Censoring

## noninformative censoring

C and T are independent, at least conditional on the covariates, and don't share any parameters.

So the likelihood factors, and we can ignore the C part.

# Truncation

We have to define the starting point:

* disease onset? 
* disease diagnosis?
* birth?
* study enrollment date?

What's the earliest time point when the event could have occurred for each individual? that's typically time 0.

In some of these cases, some participants don't enter the data set until sometime after time 0. If the event had occurred between time 0 and when they entered the study.

We need to account for that missing time.

