## Gaussian Deviance vs. GLM Deviance {#sec-compare-lm-glm-deviance}

The residual deviance is a general concept that applies to all GLMs,
including Gaussian linear regression.
For any GLM, the residual deviance is:

$$D = 2(\ell_{\text{sat}} - \ell(\hat\beta))$$ {#eq-deviance-general}

where $\ell_{\text{sat}}$ is the log-likelihood of the saturated model
and $\ell(\hat\beta)$ is the log-likelihood of the fitted model.
However, this formula simplifies differently depending on the distribution family,
and the resulting test statistics have different null distributions.

### Gaussian linear regression deviance

From @eq-linreg-loglik, the Gaussian log-likelihood is:

$$
\ell(\hat\beta, \hat\sigma^2)
= -\frac{n}{2}\log(2\pi\hat\sigma^2)
- \frac{1}{2\hat\sigma^2}
  \sumin (y_i - \hat{y}_i)^2
$$

The saturated model has one free mean parameter per distinct covariate pattern.
When there are no replicates (each covariate pattern appears exactly once),
it sets $\hat\mu_i = y_i$ for every observation,
so its residual sum of squares is zero:

$$\ell_{\text{sat}}(\hat\sigma^2) = -\frac{n}{2}\log(2\pi\hat\sigma^2)$$

Substituting into @eq-deviance-general, the Gaussian deviance simplifies to
the **residual sum of squares** (RSS), scaled by $\hat\sigma^2$:

$$
D
= 2(\ell_{\text{sat}} - \ell(\hat\beta))
= \frac{\sumin (y_i - \hat{y}_i)^2}{\hat\sigma^2}
= \frac{\text{RSS}}{\hat\sigma^2}
$$

Because $\sigma^2$ is unknown and must be estimated,
comparing deviances between two nested Gaussian models uses an **F-test**,
which is exact under Gaussian assumptions.
Substituting $D = \text{RSS}/\hat\sigma^2$, the $\hat\sigma^2$ factors cancel:

$$
F = \frac{(D_1 - D_2) / (p_2 - p_1)}{D_2 / (n - p_2)}
= \frac{(\text{RSS}_1 - \text{RSS}_2) / (p_2 - p_1)}{\text{RSS}_2 / (n - p_2)}
\dist F_{p_2 - p_1,\; n - p_2}
$$

In R, `deviance()` applied to an `lm` object returns **RSS**
(the unscaled deviance):

```{r}
#| code-fold: false
deviance(bw_lm2)
sum(residuals(bw_lm2)^2)
```

::: notes
The two expressions above return the same value.
:::

The Gaussian linear regression model can equivalently be fit as a GLM
with `family = gaussian`, which also returns RSS as the deviance:

```{r}
#| code-fold: false
bw_glm2 <- glm(
  weight ~ sex + age + sex:age,
  data = bw,
  family = gaussian
)

deviance(bw_lm2)
deviance(bw_glm2)
```

::: notes
`deviance(bw_lm2)` and `deviance(bw_glm2)` return the same value,
confirming that Gaussian linear regression is a special case of the GLM framework
where the deviance equals RSS.
:::

### Non-Gaussian GLM deviance

For non-Gaussian GLMs,
the deviance does **not** simplify to RSS.
The formula depends on the distribution family:

**Poisson:**

$$
D = 2\sumin
\left[y_i \log\!\left(\frac{y_i}{\hat\mu_i}\right) - (y_i - \hat\mu_i)\right]
$$

**Binomial:**

$$
D = 2\sumin
\left[y_i \log\!\left(\frac{y_i}{\hat\pi_i}\right)
+ (n_i - y_i)\log\!\left(\frac{n_i - y_i}{n_i - \hat\pi_i}\right)\right]
$$

For Poisson and Binomial models, the dispersion parameter $\phi = 1$ is **fixed**
rather than estimated.
Consequently, the difference in deviances between two nested models
follows an approximate $\chi^2$ distribution:

$$D_1 - D_2 \;\dot \sim\; \chi^2_{p_2 - p_1}$$

This asymptotic result replaces the exact F-test used for Gaussian models.

### Saturated vs. fully parametrized models with replicates {#sec-sat-vs-full}

When some covariate patterns appear more than once (i.e., when there are **replicates**),
it is important to distinguish between two special models:

- The **saturated model** has one free mean parameter per **distinct covariate pattern**
  ($q$ parameters, where $q \leq n$ is the number of unique patterns).
- The **fully parametrized model** has one free mean parameter per **observation**
  ($n$ parameters).

When there are no replicates ($q = n$), these two coincide.
When there are replicates ($q < n$), the saturated model constrains all
observations sharing a covariate pattern to have the same mean,
but places no constraint on means across different patterns.
See @kleinbaum2010logistic for further discussion of this distinction.

Deviance is always measured relative to the **saturated model**,
not the fully parametrized model.

#### Gaussian deviance with replicates

When covariate pattern $k$ has $n_k$ replicates with sample mean $\bar{y}_k$,
the saturated model fits $\hat\mu_k = \bar{y}_k$ for each pattern $k$,
so its residual sum of squares equals the **within-group (pure error) SS**:

$$
\text{RSS}_{\text{sat}}
= \sum_{k=1}^q \sum_{i: \vx_i = \vx_k} (y_i - \bar{y}_k)^2
$$

This is nonzero whenever any covariate pattern has replicates with
different response values.
The Gaussian deviance relative to the saturated model is therefore:

$$
D
= 2(\ell_{\text{sat}} - \ell(\hat\beta))
= \frac{\text{RSS} - \text{RSS}_{\text{sat}}}{\hat\sigma^2}
$$

::: callout-note
R's `deviance()` applied to an `lm` object returns the **total fitted-model RSS**,
not the deviance relative to the saturated model.
To compute the deviance relative to the saturated model,
subtract the saturated model's RSS:
`deviance(lm_fit) - deviance(lm_saturated)`.
:::

For the `birthweight` data, we can verify this directly using
`bw_lm2` (the interaction model) and `lm_max` (the saturated model):

```{r}
#| code-fold: false
deviance(bw_lm2)                      # total RSS of fitted model
deviance(lm_max)                      # within-group (pure error) SS
deviance(bw_lm2) - deviance(lm_max)   # lack-of-fit SS (deviance vs. saturated)
```

::: notes
The lack-of-fit SS (`deviance(bw_lm2) - deviance(lm_max)`) measures
how much additional unexplained variation remains after accounting for
pure measurement error within covariate patterns.
:::

#### GLM deviance with replicates

For a Binomial GLM fit to data **grouped by covariate pattern**
(with $y_k$ events and $n_k$ observations for pattern $k$),
the saturated model sets $\hat\pi_k^{\text{sat}} = y_k / n_k$ for each pattern $k$.
R's `deviance()` correctly computes $2(\ell_{\text{sat}} - \ell(\hat\beta))$
using this grouping:

$$
D = 2\sum_{k=1}^q
\left[
  y_k \log\!\left(\frac{y_k / n_k}{\hat\pi_k}\right)
  + (n_k - y_k)\log\!\left(\frac{1 - y_k/n_k}{1 - \hat\pi_k}\right)
\right]
$$

::: notes
By convention, terms with $y_k = 0$ or $y_k = n_k$ contribute zero to the sum,
since $0\log(0) = 0$ in the limit.
:::

When binomial data is **ungrouped** (individual Bernoulli observations, $n_i = 1$),
R uses the **fully parametrized** model as its reference —
assigning $\hat\pi_i = y_i \in \{0, 1\}$ to each individual observation.
Since each predicted probability is exactly 0 or 1,
every Bernoulli likelihood contribution equals 1,
and hence $\ell_{\text{fp}} = 0$.
Thus R's `deviance()` returns $-2\ell(\hat\beta)$.

We can verify this directly:
observations are ungrouped Bernoulli with repeated covariate patterns
($x \in \{0, 1\}$ appears three times each).

```{r}
#| code-fold: false
set.seed(42)
x_ug <- rep(0:1, each = 3)
y_ug <- c(0L, 1L, 0L, 1L, 1L, 0L)
fit_ug <- glm(y_ug ~ x_ug, family = binomial)
deviance(fit_ug)                     # R's deviance
-2 * as.numeric(logLik(fit_ug))      # -2 * log-likelihood (using ell_fp = 0)
```

::: notes
The two values are equal, confirming that R uses $\ell_{\text{fp}} = 0$ as its reference
when data are ungrouped.
:::

::: callout-note
This reference model is the **fully parametrized** model (one parameter per observation),
not the **saturated** model (one parameter per distinct covariate pattern).
They coincide only when there are no repeated covariate patterns ($q = n$).
When patterns do repeat ($q < n$), the saturated model sets $\hat\pi_k = y_k/n_k$ per pattern,
giving $\ell_{\text{sat}} < 0$.

`deviance()` for ungrouped data cannot be used as a goodness-of-fit test
against the $\chi^2$ distribution,
because the asymptotic $\chi^2(q - p)$ approximation
requires the number of observations per covariate pattern to grow ($n_k \to \infty$).
With ungrouped Bernoulli data ($n_i = 1$), this condition never holds,
so the approximation is unreliable regardless of total sample size $n$.
:::

### Summary

| Feature | Gaussian linear regression | Non-Gaussian GLMs (e.g., Poisson, Binomial) |
|---|---|---|
| Deviance formula | $(\text{RSS} - \text{RSS}_{\text{sat}})/\hat\sigma^2$ | $2(\ell_{\text{sat}} - \ell(\hat\beta))$ |
| `deviance()` in R | Returns total RSS | Returns $2(\ell_{\text{sat}} - \ell(\hat\beta))$ |
| Saturated model reference | Per covariate pattern | Per covariate pattern (grouped) or per observation (ungrouped) |
| Dispersion $\phi$ | Unknown; estimated as $s^2 = \text{RSS}/(n-p)$ | Fixed ($\phi = 1$) |
| Test for nested models | F-test (exact) | $\chi^2$-test (asymptotic) |
