### Risk differences

::: notes
The simplest comparison of two probabilities, $\pi_1$, and $\pi_2$, is the difference of their values:
:::

:::{#def-RD}
#### Risk difference
The **risk difference** of two probabilities, $\pi_1$, and $\pi_2$, is the difference of their values:
$$\delta(\pi_1,\pi_2) \eqdef \pi_1 - \pi_2$$
:::

---

:::{#exm-RD}

#### Difference in MI risk

In @exm-oc-mi, the maximum likelihood estimate of the difference in MI risk between OC users and OC non-users is:
```{r, include = FALSE}
#| label: compute-risk-diff
rd_OC <- p_MI_OC - p_MI_nOC # nolint: object_name_linter
```

$$
\begin{aligned}
\hat\delta(\pi(OC), \pi(\neg OC))
&= \delta(\hat\pi(OC), \hat\pi(\neg OC))\\
&= \hat\pi(OC) - \hat\pi(\neg OC)\\
&= `r p_MI_OC` - `r p_MI_nOC`\\
&= `r rd_OC`
\end{aligned}
$$

:::

---

:::{#exr-interpret-rr}
#### interpreting risk differences

How can we interpret the preceding relative risk estimate in prose?
:::

---

:::{#sol-interpret-rd}

#### interpreting risk differences

"The difference in risk of MI between OC users and non-users was `r rd_OC`."

or

"The difference in risk of MI between OC users and non-users was `r rd_OC*100`
[percentage points](https://en.wikipedia.org/wiki/Percentage_point)."

:::

::: notes
See the note about working with percentages in the [Appendix](notation.qmd#percent-sign).
:::

---

### Risk ratios

:::{#exr-name-RR}

If $\pi_1$ and $\pi_2$ are two probabilities,
what do we call the following ratio?

$$\rho(\pi_1,\pi_2) = \frac{\pi_1}{\pi_2}$$
:::

---

:::{#sol-exr-name-RR}

:::{#def-RR}
### Relative risk ratios

The ratio of two probabilities $\pi_1$ and $\pi_2$,

$$\rho(\pi_1,\pi_2) = \frac{\pi_1}{\pi_2}$$

is called the:

- **risk ratio**,
- **relative risk ratio**,
- **probability ratio**,
- or **rate ratio**

of $\pi_1$ compared to $\pi_2$.

:::

:::


---

:::{#exr-RR}

:::: notes
Above, we estimated that:
::::

$$\ph(MI|OC) = `r 13/5000`$$

$$\ph(MI|\neg OC) = `r 7/10000`$$

Now, estimate the *relative risk* of MI for OC versus non-OC.

:::

---

:::{#sol-exr-RR}

:::: notes
The *relative risk* of MI for OC versus non-OC is:
::::

```{r}
#| label: compute-risk-ratio
rr <- (13 / 5000) / (7 / 10000)
```

$$
\begin{aligned}
\hat\rho(OC, \neg OC)
&=\frac{\ph(MI|OC)}{\ph(MI|\neg OC)}\\
&= \frac{`r 13/5000`}{`r 7/10000`}\\
&= `r rr`
\end{aligned}
$$

:::

---

:::{#exr-interpret-rr}
How can we interpret the preceding relative risk estimate in prose?
:::

---

:::{#sol-interpret-rr}

:::: notes
We might summarize this result by saying that "the estimated probability of MI among OC users was `r rr` [times as high as]{.underline}, or `r rr - 1` [times higher than]{.underline}, the estimated probability among OC non-users.
::::

:::

---

### Relative risk differences

:::{#def-RRD}

#### Relative risk difference

:::: notes
Sometimes, we divide the risk difference by the comparison probability; the result is called the **relative risk difference**:
::::

$$\xi(\pi_1,\pi_2) \eqdef \frac{\delta(\pi_1,\pi_2)}{\pi_2}$$

:::

---

:::{#thm-rrd-vs-rr}
#### Relative risk difference equals risk ratio minus 1

$$\xi(\pi_1,\pi_2) = \rho(\pi_1,\pi_2) - 1$$
:::

---

::: proof
$$
\begin{aligned}
\xi(\pi_1,\pi_2)
&\eqdef \frac{\delta(\pi_1,\pi_2)}{\pi_2}
\\&= \frac{\pi_1-\pi_2}{\pi_2}
\\&= \frac{\pi_1}{\pi_2} - 1
\\&= \rho(\pi_1,\pi_2) - 1
\end{aligned}
$$
:::

---

### Changing reference groups in risk comparisons

:::: notes
Risk differences, risk ratios, and relative risk differences are defined by two probabilities, plus a choice of which probability is the **baseline** or **reference** probability (i.e., which probability is the subtrahend of the risk difference or the denominator of the risk ratio).
:::

$$\delta(\pi_2,\pi_1) = -\delta(\pi_1, \pi_2)$$

$$\rho(\pi_2,\pi_1) = \inv{\rho(\pi_1,\pi_2)}$$
$$\xi(\pi_2,\pi_1) = \inv{\xi(\pi_1,\pi_2) + 1} - 1$$

:::{#exr-change-ref-group}

Prove the relationships above.

:::

---

:::{#exm-ref}

#### Switching the reference group in a risk ratio

Above, we estimated that the risk ratio of OC versus non-OC is:

$$
\begin{aligned}
\rho(OC, \neg OC)
&= `r (13/5000)/(7/10000)`
\end{aligned}
$$

In comparison, the risk ratio for non-OC versus OC is:

$$
\begin{aligned}
\rho(\neg OC, OC)
&=\frac{\ph(MI|\neg OC)}{\ph(MI|OC)}\\
&= \frac{`r 7/10000`}{`r 13/5000`}\\
&= `r (7/10000)/(13/5000)`\\
&= \frac{1}{\rho(OC, \neg OC)}
\end{aligned}
$$

:::

