
::: notes

* In @exm-oc-mi, we estimated the risk and the odds of MI for two groups,
defined by oral contraceptive use.

* If the predictor is quantitative (dose) or there is more than one predictor, the task becomes more difficult.

* In this case, we will use logistic regression, which is a generalization of the linear regression models you have been using that can account for a binary response instead of a continuous one.

:::

### Binary outcomes models - one group, no covariates {.smaller}

$$
\begin{aligned}
\P(Y=1) &= \pi\\
\P(Y=0) &= 1-\pi\\
\P(Y=y) &= \pi^y (1-\pi)^{1-y}\\
\mathbf y  &= (y_1, ..., y_n)\\
\Lik(\pi;\mathbf y) &= \pi^{\sum y_i} (1-\pi)^{n - \sum y_i}\\
\ell(\pi, \mathbf y) &= \left({\sum y_i}\right) \logf{\pi} + \left(n - \sum y_i\right) \logf{1-\pi}\\
&= \left({\sum y_i}\right) \left(\logf{\pi} - \logf{1-\pi}\right) + n \cdot \logf{1-\pi}\\
&= \left({\sum y_i}\right) \logf{\frac{\pi}{ 1-\pi}} + n \cdot \logf{1-\pi}
\\ &= \left({\sum y_i}\right) \logit(\pi) + n \cdot \logf{1-\pi}
\end{aligned}
$$

### Binary outcomes - general  {.smaller}

$$
\ba
\P(Y_i=1) &= \pi_i
\\ \P(Y_i=0) &= 1-\pi_i
\ea
$$

$$\P(Y_i=y_i) = (\pi_i)^y_i (1-\pi_i)^{1-y_i}$$

$$\Lik_i(\pi_i) = \P(Y_i=y_i)$$

$$
\ba
\ell_i(\pi_i)
   &= \logf{\Lik_i(\pi_i)}
\\ &= y_i \logf{\pi_i} + (1-y_i) \logf{1-\pi_i}
\ea
$$

---

For $\iid$ data $\vec y  = (y_1, ..., y_n)$:

$$
\ba
\Lik(\pi;\vec y) &= \P(\dsvn{Y}{y})
\\ &= \prod_{i=1}^n \P(Y_i=y_i)
\\ &= \prod_{i=1}^n \Lik_i(\pi_i)
\ea
$$

### Modeling $\pi_i$ as a function of $X_i$

If there are only a few distinct $X_i$ values, we can model $\pi_i$
separately for each value of $X_i$.

Otherwise, we need regression.

$$
\begin{aligned}
\pi(x) &\equiv \text{E}(Y=1|X=x)\\
&= f(x^\top\beta)
\end{aligned}
$$

Typically, we use the $\expit$ inverse-link:

$$\pi(\vec x) = \expit(\vx'\beta)$${#eq-logistic-link}


### Meet the beetles

:::{#tbl-beetles-data}

```{r}

library(glmx)

data(BeetleMortality, package = "glmx")
beetles <- BeetleMortality |>
  mutate(
    pct = died / n,
    survived = n - died,
    dose_c = dose - mean(dose)
  )
beetles
```

Mortality rates of adult flour beetles after five hours’ exposure to gaseous carbon disulphide [@bliss1935beetles]

:::

---

:::{#fig-beetles_1a}

```{r}

plot1 <-
  beetles |>
  ggplot(aes(x = dose, y = pct)) +
  geom_point(aes(size = n)) +
  xlab("Dose (log mg/L)") +
  ylab("Mortality rate (%)") +
  scale_y_continuous(labels = scales::percent) +
  scale_size(range = c(1, 2)) +
  theme_bw(base_size = 18)

print(plot1)
```

Mortality rates of adult flour beetles after five hours' exposure to gaseous carbon disulphide [@bliss1935beetles]

:::

### Why don't we use linear regression?

:::{#fig-beetles_2}

```{r}
beetles_long <- beetles |>
  reframe(
    .by = everything(),
    outcome = c(
      rep(1, times = died),
      rep(0, times = survived)
    )
  ) |>
  as_tibble()

lm1 <- beetles_long |> lm(formula = outcome ~ dose)
f_linear <- function(x) predict(lm1, newdata = data.frame(dose = x))

range1 <- range(beetles$dose) + c(-.2, .2)

plot2 <-
  plot1 +
  geom_function(
    fun = f_linear,
    aes(col = "Straight line")
  ) +
  labs(colour = "Model", size = "")

plot2 |> print()
```

Mortality rates of adult flour beetles after five hours' exposure to gaseous carbon disulphide [@bliss1935beetles]

:::

### Zoom out

:::{#fig-beetles_3}

```{r}
(plot2 + expand_limits(x = c(1.6, 2))) |> print()
```

Mortality rates of adult flour beetles after five hours' exposure to gaseous carbon disulphide [@bliss1935beetles]

:::

### log transformation of dose?

:::{#fig-beetles_4}

```{r}
lm2 <- beetles_long |> lm(formula = outcome ~ log(dose))
f_linearlog <- function(x) predict(lm2, newdata = data.frame(dose = x))

plot3 <- plot2 +
  expand_limits(x = c(1.6, 2)) +
  geom_function(fun = f_linearlog, aes(col = "Log-transform dose"))
(plot3 + expand_limits(x = c(1.6, 2))) |> print()
```

Mortality rates of adult flour beetles after five hours' exposure to gaseous carbon disulphide [@bliss1935beetles]

:::

### Logistic regression

:::{#fig-beetles_5}

```{r}
beetles_glm_grouped <- beetles |>
  glm(formula = cbind(died, survived) ~ dose, family = "binomial")
f <- function(x) {
  beetles_glm_grouped |>
    predict(newdata = data.frame(dose = x), type = "response")
}

plot4 <- plot3 + geom_function(fun = f, aes(col = "Logistic regression"))
plot4 |> print()
```

Mortality rates of adult flour beetles
after five hours' exposure to gaseous carbon disulphide [@bliss1935beetles]

:::

---

### Three parts to regression models

-   What distribution does the outcome have for a specific subpopulation
defined by covariates? (outcome model)

-   How does the combination of covariates relate to the mean? (link
function)

-   How do the covariates combine? (linear predictor, interactions)

---

{{< include _sec_logistic_slope_mean.qmd >}}

---

{{< include _sec_derivs_MLE.qmd >}}

---

### Logistic regression in R

```{r}
beetles_glm_grouped <-
  beetles |>
  glm(
    formula = cbind(died, survived) ~ dose,
    family = "binomial"
  )

library(parameters)
beetles_glm_grouped |>
  parameters() |>
  print_md()
```

---

Fitted values:

```{r}
fitted.values(beetles_glm_grouped)
predict(beetles_glm_grouped, type = "response")
predict(beetles_glm_grouped, type = "link")

fit_y <- beetles$n * fitted.values(beetles_glm_grouped)
```

---

### Individual observations

```{r}
#| tbl-cap: "`beetles` data in long format"
#| label: tbl-beetles-long

beetles_long
```

---

:::{#tbl-beetles-model-ungrouped}

```{r}
beetles_glm_ungrouped <-
  beetles_long |>
  glm(
    formula = outcome ~ dose,
    family = "binomial"
  )

beetles_glm_ungrouped |>
  parameters() |>
  print_md()
```

logistic regression model for beetles data with individual Bernoulli data

:::

::: notes
Here's the previous version again:
:::

:::{#tbl-beetles-model-grouped}

```{r}
beetles_glm_grouped |>
  parameters() |>
  print_md()
```

logistic regression model for beetles data with grouped (binomial) data

:::

---

::: notes
They seem the same! But not quite:
:::

```{r}
logLik(beetles_glm_grouped)
logLik(beetles_glm_ungrouped)
```

::: notes
The difference is due to the binomial coefficient
$\left(n\atop x \right)$ which isn't included in the
individual-observations (Bernoulli) version of the model.
:::
