
We don't have to do these calculations by hand every time; the
`survival` package and several others have functions available to
automate many of these tasks (full list:
<https://cran.r-project.org/web/views/Survival.html>).

### The `Surv` function

To use the `survival` package, the first step is telling R how to
combine the exit time and exit reason (censoring versus event) columns.
The `Surv()` function accomplishes this task.

#### Example: `Surv()` with `drug6mp` data

```{r}
#| code-fold: show
#| code-line-numbers: "5-7"
library(survival)
drug6mp.v3 <-
  drug6mp.v2 |>
  mutate(
    surv2 = Surv(
      time = t2,
      event = (outcome == "relapsed")
    )
  )

print(drug6mp.v3)
```

The output of `Surv()` is a vector of objects with class `Surv`. When we
print this vector:

-   observations where the event was observed are printed as the event
time (for example, `surv2 = 10` on line 1)

-   observations where the event was right-censored are printed as the
censoring time with a plus sign (`+`; for example, `surv2 = 32+` on
line 3).

### The `survfit` function

Once we have constructed our `Surv` variable, we can calculate the
Kaplan-Meier estimate of the survival curve using the `survfit()`
function.

::: callout-note
The documentation for `?survfit` isn't too helpful; the
`survfit.formula` documentation is better.

```{r, printr.help.sections = c("description", "usage")}
#| include: false
?survfit.formula
```
:::

---

#### Example: `survfit()` with `drug6mp` data

Here we use `survfit()` to create a `survfit` object, which contains the
Kaplan-Meier estimate:

```{r}
#| code-fold: show
drug6mp.km_model <- survfit(
  formula = surv2 ~ 1,
  data = drug6mp.v3
)
```

`print.survfit()` just gives some summary statistics:

```{r}
#| code-fold: show
print(drug6mp.km_model)
```

`summary.survfit()` shows us the underlying Kaplan-Meier table:

```{r}
#| code-fold: show
summary(drug6mp.km_model)
```

---

We can specify which time points we want using the `times` argument:

```{r}
#| code-fold: show
summary(
  drug6mp.km_model,
  times = c(0, drug6mp.v3$t2)
)
```

---

```{r, printr.help.sections = c("description", "usage", "arguments")}
#| code-fold: show
?summary.survfit
```

### Plotting estimated survival functions

We can plot `survfit` objects with `plot()`, `autoplot()`, or
`ggsurvplot()`:

```{r}
#| code-fold: show
#| fig-cap: "Kaplan-Meier Survival Curve for 6-MP Patients"

library(ggfortify)
autoplot(drug6mp.km_model)

# not shown:
# plot(drug6mp.km_model)

# library(survminer)
# ggsurvplot(drug6mp.km_model)
```

---

#### quantiles of survival curve

We can extract quantiles with `quantile()`:

```{r}
#| code-line-numbers: "2"
drug6mp.km_model |>
  quantile(p = c(.25, .5)) |>
  as_tibble() |>
  mutate(p = c(.25, .5)) |>
  relocate(p, .before = everything())
```

### R Packages Ecosystem for Survival Analysis

R provides a rich ecosystem of packages for survival analysis,
ranging from core functionality to specialized visualization tools and advanced modeling capabilities.

#### Visualization Packages

##### ggsurvfit

**[`ggsurvfit`](https://danieldsjoberg.com/ggsurvfit/)** is a modern package for creating publication-ready survival plots using ggplot2.

- Developed by [Daniel D. Sjoberg](https://www.danieldsjoberg.com/) and collaborators
- Seamlessly integrates with ggplot2,
allowing full customization using familiar ggplot2 syntax
- Key features:
  - Create Kaplan-Meier plots with `ggsurvfit()`
  - Add confidence intervals with `add_confidence_interval()`
  - Add risk tables with `add_risktable()`
  - Add quantile annotations with `add_quantile()`
  - Support for competing risks with `ggcuminc()`
- Produces publication-ready legends without showing raw variable names
- All plot elements are native ggplot2 layers,
making customization straightforward
- Available on [CRAN](https://cran.r-project.org/web/packages/ggsurvfit/index.html) and [GitHub](https://github.com/pharmaverse/ggsurvfit)

##### survminer

**[`survminer`](https://cran.r-project.org/web/packages/survminer/index.html)** is another popular visualization package for survival analysis.

- Creates ggplot2-based survival plots using `ggsurvplot()`
- Provides functions for visualizing Cox model results
- Includes tools for testing proportional hazards assumptions
- Offers built-in themes and color palettes for publication-quality plots
- Can display risk tables, cumulative events, and censoring counts

##### ggfortify

**[`ggfortify`](https://cran.r-project.org/web/packages/ggfortify/index.html)** extends ggplot2 with automatic plotting capabilities (as shown above).

- Provides `autoplot()` method for `survfit` objects
- One-line code to create survival curves with confidence intervals
- Automatically handles grouping variables and creates legends
- Shows censoring marks on plots by default
- Returns ggplot2 objects for further customization
- Also supports many other statistical objects beyond survival analysis

#### Additional Modeling Packages

Several packages extend the modeling capabilities beyond the core `survival` package:

- **[`flexsurv`](https://cran.r-project.org/web/packages/flexsurv/index.html)**: Flexible parametric survival models with a wide range of distributions
- **[`rms`](https://cran.r-project.org/web/packages/rms/index.html)** (Regression Modeling Strategies): 
Advanced regression modeling including survival models with tools for model validation and calibration
- **[`cmprsk`](https://cran.r-project.org/web/packages/cmprsk/index.html)**: Competing risks regression using the Fine-Gray model
- **[`mstate`](https://cran.r-project.org/web/packages/mstate/index.html)**: Multi-state survival models for complex event histories
- **[`timereg`](https://cran.r-project.org/web/packages/timereg/index.html)**: Flexible regression models for survival data,
including additive hazards models
- **[`frailtypack`](https://cran.r-project.org/web/packages/frailtypack/index.html)**: Frailty models for clustered or recurrent event data
- **[`riskRegression`](https://cran.r-project.org/web/packages/riskRegression/index.html)**: Risk prediction models and performance evaluation
- **[`pec`](https://cran.r-project.org/web/packages/pec/index.html)**: Prediction error curves for survival model validation
- **[`prodlim`](https://cran.r-project.org/web/packages/prodlim/index.html)**: Fast estimation of survival models including competing risks

#### Package Selection Guidance

For most standard survival analyses:

- Use **`survival`** for core functionality (Kaplan-Meier, Cox models, parametric models)
- Choose **`ggsurvfit`** for modern,
customizable survival plots if you're familiar with ggplot2
- Consider **`survminer`** as an alternative visualization package with many built-in features
- Use **`ggfortify`** if you want quick,
automatic plots with minimal code

For specialized analyses:

- **Competing risks**: `cmprsk` or `riskRegression`
- **Multi-state models**: `mstate`
- **Flexible parametric models**: `flexsurv`
- **Clustered/recurrent events**: `frailtypack`
- **Model validation**: `rms` or `pec`

The [CRAN Survival Analysis Task View](https://cran.r-project.org/web/views/Survival.html) provides the most comprehensive and up-to-date list of available packages.
