For Gaussian linear regression,
the ordinary least squares (OLS) estimates $\hat\beta_k$ are exactly Gaussian
when the error terms $\epsilon_i$ are Gaussian,
for any sample size.
See also the [table of Gaussian vs. MLE-based tests](intro-MLEs.qmd#tbl-gaussian-vs-mle-tests).

#### Wald test statistic

To test $H_0: \beta_k = \beta_{k,0}$ (typically $\beta_{k,0} = 0$):

$$t_k = \frac{\hat\beta_k - \beta_{k,0}}{\widehat{SE}(\hat\beta_k)}$$

Under $H_0$, $t_k \sim t_{n-p}$ exactly when errors are Gaussian.

#### Confidence intervals for regression coefficients

A 95% confidence interval for $\beta_k$ is:

$$\hat\beta_k \pm t_{n-p}(0.975) \cdot \widehat{SE}(\hat\beta_k)$$

#### In R

In R, `parameters()` from the `parameters` package automatically computes
Wald tests and confidence intervals for linear regression model coefficients:

```{r}
#| label: tbl-bw-wald-tests
#| tbl-cap: "Wald tests and 95% CIs for `birthweight` linear regression"
bw_lm2 |>
  parameters() |>
  print_md(
    include_reference = TRUE
  )
```

To understand what's happening,
let's replicate these results by hand for the interaction term.

#### P-values {.smaller}

```{r}
bw_lm2 |>
  parameters(keep = "sexmale:age") |>
  print_md(
    include_reference = TRUE
  )
```

```{r}
beta_hat <- coef(summary(bw_lm2))["sexmale:age", "Estimate"]
se_hat <- coef(summary(bw_lm2))["sexmale:age", "Std. Error"]
dfresid <- bw_lm2$df.residual
t_stat <- abs(beta_hat) / se_hat
pval_t <-
  pt(-t_stat, df = dfresid, lower.tail = TRUE) +
  pt(t_stat, df = dfresid, lower.tail = FALSE)
```

$$
\begin{aligned}
&P\paren{
| \hat \beta_{AM}  | >
| `r coef(bw_lm2)["sexmale:age"]`| \middle| H_0
}
\\
&= \Pr \paren{
\abs{ \frac{\hat\beta_{AM}}{\hat{SE}(\hat\beta_{AM})} } >
\abs{ \frac{`r beta_hat`}{`r se_hat`} } \middle| H_0
}\\
&= \Pr \paren{
\abs{ T_{`r dfresid`} }  >  `r t_stat` | H_0
}\\
&= `r pval_t`
\end{aligned}
$$

::: notes
This matches the result in the table above.
:::

#### Confidence intervals {.smaller}

```{r}
#| label: confint-t
q_t_upper <- qt(
  p = 0.975,
  df = dfresid,
  lower.tail = TRUE
)

q_t_lower <- qt(
  p = 0.025,
  df = dfresid,
  lower.tail = TRUE
)

confint_radius_t <-
  se_hat * q_t_upper

confint_t <- beta_hat + c(-1, 1) * confint_radius_t

print(confint_t)
```

::: notes
This also matches.
:::

#### Gaussian approximations {.smaller}

For large samples, the t-distribution is well-approximated by the standard Gaussian:

```{r}
pval_z <- pnorm(abs(t_stat), lower = FALSE) * 2

print(pval_z)
```

```{r}
confint_radius_z <- se_hat * qnorm(0.975, lower = TRUE)
confint_z <-
  beta_hat + c(-1, 1) * confint_radius_z
print(confint_z)
```
