{{< include latex-macros/macros.qmd >}}

## Evaluating a predictor of primary interest {#sec-pred-sel-eval-primary}

::: notes

When the goal is to estimate the causal effect of
a specific exposure on an outcome,
the central challenge in observational data is **confounding**.
Relatively inclusive models — those including more potential confounders —
are better at minimizing confounding bias.

:::

The following principles guide predictor selection for this goal.

## Including predictors for face validity {#sec-pred-sel-face-validity}

Some predictors are such well-established causal antecedents of the outcome
that they should be included regardless of their statistical significance
in the current data.
This ensures the **face validity** of the model:
readers and reviewers can be confident that obvious confounders
have been accounted for.

For example, age and smoking are established CHD risk factors
and would be included in any model for CHD outcomes,
even if they are not statistically significant in a small sample.

## Selecting confounders on statistical grounds {#sec-pred-sel-confounders}

In addition to variables included for face validity,
we may consider other potential confounders
identified in previous studies or on substantive grounds.

A practical approach is **backward selection with a liberal criterion**:

1. Start with a full model including all pre-specified candidates.
2. At each step, remove the predictor with the largest p-value,
   provided it exceeds a liberal threshold (e.g., $p > 0.2$).
3. Stop when all remaining predictors meet the retention criterion.

The liberal threshold ($p > 0.2$ rather than $p > 0.05$)
is important for confounding control:
in small samples, even important confounders may have large p-values,
and removing them could introduce bias.

An alternative retention criterion is the
**10% change-in-estimate rule**
[@kleinbaum2010logistic; @vittinghoff2e]:

Retain a candidate confounder $C$ in the model if
dropping it changes the estimated coefficient for the primary predictor
by more than 10–15%.

More precisely,
let $\hat\beta$ be the estimated coefficient for the exposure
from the **full model** (including $C$),
and let $\hat\beta_0$ be the estimate from the **reduced model** (dropping $C$).
Then $C$ is judged a meaningful confounder if:

$$\left| \frac{\hat\beta - \hat\beta_0}{\hat\beta} \right| > 0.10$$

The threshold is often taken as 10–20%, depending on the context.
A smaller threshold (e.g., 10%) retains more variables;
a larger threshold (e.g., 20%) allows more aggressive pruning.

::: {.callout-note}

The 10% change-in-estimate rule focuses on **bias** rather than **statistical significance**.
A variable can be a meaningful confounder even if it is not statistically significant
(because of small sample size or measurement error),
and conversely, a statistically significant variable may cause negligible confounding.
The p-value criterion and the change-in-estimate rule can disagree,
and @vittinghoff2e recommends using both as complementary checks.

:::

### Numerical example: 10% rule in WCGS {#sec-pred-sel-10pct-wcgs}

::: notes

The WCGS (Western Collaborative Group Study) is an observational
prospective cohort of 3,154 men followed for CHD events.
Unlike HERS (a randomized trial),
the WCGS exposure — Type A behavior pattern — was not randomly assigned,
so confounding is a genuine concern.

:::

In the WCGS study,
the exposure of interest is Type A behavior pattern (`dibpat`),
and the outcome is CHD incidence (`chd69`).
We fit logistic regression models
and apply the 10% change-in-estimate rule
to assess whether age, cholesterol, SBP, smoking, and BMI
are meaningful confounders of the Type A → CHD relationship.

::: notes

Adapted from @kleinbaum2010logistic [Chapter 8],
which describes the change-in-estimate procedure for logistic regression models.

:::

```{r}
#| label: wcgs-data-setup

wcgs <- readRDS(fs::path_package("rme", "extdata/wcgs.rds"))
```

```{r}
#| label: wcgs-10pct-rule

# Full model (all candidate confounders)
m_full <- glm(
  chd69 ~ dibpat + age + chol + sbp + smoke + bmi,
  data = wcgs,
  family = binomial
)

# Unadjusted model (Type A only)
m_unadj <- glm(
  chd69 ~ dibpat,
  data = wcgs,
  family = binomial
)

# Estimate of Type A effect from full model
beta_full <- coef(m_full)["dibpatType A"]
beta_unadj <- coef(m_unadj)["dibpatType A"]

# Models dropping one confounder at a time
confounders <- c("age", "chol", "sbp", "smoke", "bmi")

results <- lapply(confounders, function(v) {
  f <- as.formula(
    paste("chd69 ~ dibpat +",
          paste(setdiff(confounders, v), collapse = " + "))
  )
  m_red <- glm(f, data = wcgs, family = binomial)
  beta_red <- coef(m_red)["dibpatType A"]
  pct_change <- abs((beta_full - beta_red) / beta_full) * 100
  data.frame(
    Confounder = v,
    Beta_full = round(beta_full, 4),
    Beta_reduced = round(beta_red, 4),
    Pct_change = round(pct_change, 1),
    Meaningful = ifelse(pct_change > 10, "Yes (>10%)", "No (<=10%)")
  )
})

results_df <- do.call(rbind, results)
rownames(results_df) <- NULL
```

::: {#tbl-wcgs-10pct}

```{r}
knitr::kable(
  results_df,
  col.names = c(
    "Dropped variable",
    "β (full model)",
    "β (reduced model)",
    "% change",
    "Meaningful confounder?"
  )
)
```

10% change-in-estimate rule applied to the WCGS data.
The Type A coefficient from the full model
is compared with coefficients from models that drop one confounder at a time.
A percentage change exceeding 10% indicates meaningful confounding.

:::

::: notes

In this example,
none of the individual confounders causes a >10% change when dropped,
suggesting they are not strong individual confounders of the
Type A → CHD relationship in the WCGS.
However, the unadjusted log-OR for Type A ($\approx$ `r round(beta_unadj, 2)`)
is somewhat different from the adjusted log-OR ($\approx$ `r round(beta_full, 2)`),
suggesting that the overall set of covariates does provide some adjustment.
In practice, covariates with established biological relationships to the outcome
(such as age, smoking, and blood pressure for CHD)
are typically retained regardless of the change-in-estimate result,
for reasons of face validity.

:::

::: {.callout-note}
## Collinearity between confounders is not a problem

Unlike collinearity between a predictor of primary interest
and other covariates (which inflates its standard error),
collinearity *among* adjustment variables does not affect
the precision of the primary predictor's estimate.
Thus, two collinear confounders can both be retained
if both are judged necessary on substantive grounds.
:::

## Example: backward selection for CHD confounders {#sec-pred-sel-backward-hers}

In the HERS data, we can use backward selection
to identify which baseline covariates confound
the effect of hormone therapy on LDL.
Since HT was randomized in HERS,
adjustment is needed for precision, not confounding;
but this illustrates the procedure.

```{r}
#| label: backward-sel-ldl-hers

full_model <- lm(
  LDL ~ HT + age + BMI + diabetes + smoking + statins + SBP,
  data = hers_ldl
)

# Backward selection using AIC
back_model <- MASS::stepAIC(full_model, direction = "backward", trace = FALSE)
summary(back_model)
```

::: notes

The backward selection procedure retains the predictors that
most improve model fit according to AIC.
In this setting, since HT was randomized,
we interpret the selected set as prognostic covariates
that improve precision of the HT estimate,
not as confounders.

:::

## Interactions with the primary predictor {#sec-pred-sel-interactions}

A useful check on the validity of the selected model is to test
for **interactions** between the primary predictor and
key covariates.
Particularly for novel findings,
showing that the association is consistent across subgroups
strengthens credibility.

::: notes

If a substantial and biologically plausible interaction is found,
the analysis must account for it.
However, exploratory interaction tests are susceptible to
false-positive findings,
and unexpected interactions should be interpreted cautiously.

:::

## Randomized experiments {#sec-pred-sel-rct}

In randomized trials, participants are allocated to treatment randomly,
so the treatment indicator is by design uncorrelated
with pre-treatment covariates in expectation.
Confounding adjustment is therefore **not required** for causal inference.

However, there are several reasons to include covariates
in trial analyses:

1. **Improving precision** (continuous outcomes).
   Adjusting for strong predictors of the outcome
   reduces residual variance and tightens the treatment effect estimate.
   Because treatment is uncorrelated with the covariates by randomization,
   variance inflation is negligible.

2. **De-attenuation** (binary and survival outcomes).
   In logistic and Cox models,
   omitting important predictors attenuates the treatment effect estimate.
   Adjusting for them moves the estimate closer to the true subject-specific effect.

3. **Accounting for stratified designs**.
   In stratified or cluster-randomized trials,
   including stratification variables in the model
   yields correct standard errors and p-values.

4. **Correcting for chance imbalances**.
   In small trials, randomization may leave residual imbalance
   in important prognostic covariates.
   Adjusted analyses demonstrate that the main inference
   is not materially affected by any such imbalance.

::: {.callout-note}
## Pre-specification is important in trials

Adjusted analyses in clinical trials should ideally be
**pre-specified in the study protocol**,
to prevent post-hoc selection of covariates
that minimize the treatment effect p-value.
:::
