
::: notes
The Nelson-Aalen estimate of the cumulative hazard is usually used for estimates of the hazard and often the cumulative hazard.

If the hazards of the three groups are proportional, that means that the ratio of the hazards is constant over $t$. We can test this using the ratios of the estimated cumulative hazards, which also would be
proportional, as shown above.
:::

```{r}
#| label: load-bmt
library(KMsurv)
library(survival)
library(dplyr)
data(bmt)

bmt =
  bmt |>
  as_tibble() |>
  mutate(
    group =
      group |>
      factor(
        labels = c("ALL","Low Risk AML","High Risk AML")))

nafit = survfit(
  formula = Surv(t2,d3) ~ group,
  type = "fleming-harrington",
  data = bmt)

bmt_curves = tibble(timevec = 1:1000)
sf1 <- with(nafit[1], stepfun(time,c(1,surv)))
sf2 <- with(nafit[2], stepfun(time,c(1,surv)))
sf3 <- with(nafit[3], stepfun(time,c(1,surv)))

bmt_curves =
  bmt_curves |>
  mutate(
    cumhaz1 = -log(sf1(timevec)),
    cumhaz2 = -log(sf2(timevec)),
    cumhaz3 = -log(sf3(timevec)))
```

```{r}
#| fig-cap: "Hazard Ratios by Disease Group for `bmt` data"
#| label: fig-HR-bmt
library(ggplot2)
bmt_rel_hazard_plot =
  bmt_curves |>
  ggplot(
    aes(
      x = timevec,
      y = cumhaz1/cumhaz2)
  ) +
  geom_line(aes(col = "ALL/Low Risk AML")) +
  ylab("Hazard Ratio") +
  xlab("Time") +
  ylim(0,6) +
  geom_line(aes(y = cumhaz3/cumhaz1, col = "High Risk AML/ALL")) +
  geom_line(aes(y = cumhaz3/cumhaz2, col = "High Risk AML/Low Risk AML")) +
  theme_bw() +
  labs(colour = "Comparison") +
  theme(legend.position="bottom")

print(bmt_rel_hazard_plot)
```

---

We can zoom in on the first 300 days to take a closer look:

```{r}
#| fig-cap: "Hazard Ratios by Disease Group (0-300 Days)"
#| label: fig-HR-bmt-inset
bmt_rel_hazard_plot + xlim(c(0,300))
```

---

The cumulative hazard curves should also be proportional

```{r}
#| fig-cap: "Disease-Free Cumulative Hazard by Disease Group"
#| label: fig-cuhaz-bmt
library(ggfortify)
plot_cuhaz_bmt =
  bmt |>
  survfit(formula = Surv(t2, d3) ~ group) |>
  autoplot(fun = "cumhaz",
           mark.time = TRUE) +
  ylab("Cumulative hazard")

plot_cuhaz_bmt |> print()

```

---

```{r}
#| fig-cap: "Disease-Free Cumulative Hazard by Disease Group (log-scale)"
#| label: fig-cuhaz-bmt-loglog
plot_cuhaz_bmt +
  scale_y_log10() +
  scale_x_log10()
```

### Smoothed hazard functions

::: notes
The Nelson-Aalen estimate of the cumulative hazard is usually used for
estimates of the hazard. Since the hazard is the derivative of the
cumulative hazard, we need a smooth estimate of the cumulative hazard,
which is provided by smoothing the step-function cumulative hazard.

The R package `muhaz` handles this for us. What we are looking for is
whether the hazard function is more or less the same shape, increasing,
decreasing, constant, etc. Are the hazards "proportional"?
:::


```{r}
#| fig-cap: "Smoothed Hazard Rate Estimates by Disease Group"
#| label: fig-smoothed-hazard-bmt
library(muhaz)

muhaz(bmt$t2,bmt$d3,bmt$group=="High Risk AML") |> plot(lwd=2,col=3)
muhaz(bmt$t2,bmt$d3,bmt$group=="ALL") |> lines(lwd=2,col=1)
muhaz(bmt$t2,bmt$d3,bmt$group=="Low Risk AML") |> lines(lwd=2,col=2)
legend("topright",c("ALL","Low Risk AML","High Risk AML"),col=1:3,lwd=2)
```

::: notes
Group 3 was plotted first because it has the highest hazard.

Except for an initial blip in the high risk AML group,
the hazards look roughly proportional.
They are all strongly decreasing.
:::
