{{< include latex-macros/macros.qmd >}}

## Directed Acyclic Graphs (DAGs) {#sec-pred-sel-dag}

A **Directed Acyclic Graph (DAG)** is a graphical model
that encodes assumed causal relationships among variables.

In a DAG:

- **Nodes** represent variables.
- **Directed edges** (arrows) represent direct causal effects.
- The graph is **acyclic**: there are no directed cycles
(a variable cannot cause itself, even indirectly).

DAGs are used in causal inference to:

- Identify **confounders** that must be adjusted for.
- Identify **mediators** that should generally *not* be adjusted for
when estimating the total causal effect.
- Identify **colliders** that should *not* be adjusted for
(adjusting for a collider opens a non-causal path).

## Key concepts from DAG theory {#sec-pred-sel-dag-concepts}

:::{#def-confounder}
### Confounder

A **confounder** of the relationship between an exposure $X$ and
outcome $Y$ is a variable $C$
that is a common cause of both $X$ and $Y$.
By contrast,
a mediator lies on the causal path
from $X$ to $Y$
and is not a common cause of $X$ and $Y$.

Adjusting for confounders removes confounding bias.
:::

---

:::{#def-mediator}
### Mediator

A **mediator** $M$ is a variable that lies on the causal pathway
from the exposure $X$ to the outcome $Y$:
$$X \to M \to Y$$

Adjusting for a mediator blocks the indirect causal pathway
through $M$,
so the estimated coefficient for $X$ reflects only
the *direct* effect of $X$ on $Y$ not through $M$.

Whether to adjust for mediators depends on the research question:

- If you want the **total causal effect** of $X$ on $Y$,
do **not** adjust for $M$.
- If you want the **direct effect** of $X$ on $Y$ not through $M$,
adjust for $M$.
:::

---

:::{#def-collider}
### Collider

A **collider** is a variable $L$ that is caused by two or more
variables on a path:
$$X \to L \leftarrow Y$$

Adjusting for a collider opens a non-causal (backdoor) path
between $X$ and $Y$,
introducing **collider bias**.
:::

## HERS study DAG {#sec-pred-sel-hers-dag}

{{< include _subfiles/shared/_sec_hers_intro.qmd >}}

We now examine the assumed causal structure
among key baseline variables, treatment, and recurrent CHD events.
<!-- {{< include _subfiles/shared/_sec_hers_conclusions.qmd >}} -->

@fig-hers-dag shows a DAG for the HERS study,
representing the assumed causal structure among key variables.

The main goal of the HERS study was to estimate
the causal effect of hormone therapy (HT)
on recurrent CHD event risk.

::: {#fig-hers-dag}

```{r}
#| fig-width: 7
#| fig-height: 5

library(dagitty)

hers_dag <- dagitty(
  'dag {
    HT [exposure, pos="0,0"]
    CHD [outcome, pos="4,0"]
    age [pos="2,2"]
    smoking [pos="-1,-2"]
    diabetes [pos="1,-1.5"]
    BMI [pos="-1,1"]
    LDL [pos="2,1"]
    HDL [pos="2,-1"]
    statins [pos="3,2"]
    SBP [pos="3,-2"]

    HT -> CHD
    HT -> HDL
    HT -> LDL
    age -> CHD
    smoking -> CHD
    smoking -> HDL
    BMI -> CHD
    BMI -> diabetes
    BMI -> SBP
    BMI -> LDL
    diabetes -> CHD
    diabetes -> LDL
    LDL -> CHD
    HDL -> CHD
    statins -> LDL
    SBP -> CHD
  }'
)

plot(hers_dag)
```

Directed Acyclic Graph (DAG) for the HERS study.
`HT` = hormone therapy (exposure);
`CHD` = recurrent coronary heart disease events (outcome);
`HDL` = high-density lipoprotein;
`LDL` = low-density lipoprotein;
`SBP` = systolic blood pressure.

:::

In @fig-hers-dag:

- **`age`** is prognostic for CHD:
older women are at higher risk of CHD events.
Since HT was randomized in HERS,
age is not a confounder of the HT → CHD relationship,
but adjusting for age can improve precision.
- **`LDL`** and **`HDL`** are *mediators*:
hormone therapy affects lipid levels,
which in turn affect CHD risk.
- **`BMI`** is a common cause of diabetes, SBP, and LDL.
- **`statins`** (statin use) affects LDL independently of HT
(statins are medications prescribed to lower LDL).
- This DAG is simplified; a more complete model might include
additional variables such as anti-hypertensive medication use.

## What variables should we adjust for? {#sec-pred-sel-hers-adjust}

Using DAG theory,
we can identify the minimum adjustment set
needed to estimate the total causal effect of HT on CHD.

```{r}
#| label: tbl-hers-adj
adjustmentSets(
  hers_dag,
  exposure = "HT",
  outcome = "CHD",
  type = "minimal"
)
```

::: notes

Since HT was randomized in HERS,
there are no backdoor paths from HT to CHD.
The `adjustmentSets()` function confirms that no adjustment is needed
to estimate the total causal effect of HT on CHD.

The variables `LDL` and `HDL` are mediators on the path
HT → LDL/HDL → CHD,
so they should **not** be included
when estimating the *total* causal effect of HT on CHD.

Including `LDL` and `HDL` in the model would block
a part of the causal effect of HT,
giving us only the *direct* effect of HT on CHD
(not through lipids).

:::

## Analyzing LDL in the HERS study {#sec-pred-sel-hers-ldl}

::: notes

The HERS baseline dataset includes LDL and HDL cholesterol,
allowing us to examine the effect of hormone therapy on lipid levels.

Here we build models to predict LDL cholesterol,
illustrating how DAG thinking guides predictor selection.

:::

```{r}
#| label: hers-data-setup

library(haven)
library(dplyr)

hers <- read_stata(fs::path_package("rme", "extdata/hersdata.dta"))

hers_ldl <-
  hers |>
  filter(!is.na(LDL)) |>
  mutate(
    HT = factor(HT, levels = c(0, 1), labels = c("Placebo", "HT")),
    statins = factor(statins, levels = c(0, 1), labels = c("No", "Yes")),
    diabetes = factor(diabetes, levels = c(0, 1), labels = c("No", "Yes")),
    smoking = factor(smoking, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

::: {#tbl-hers-ldl-full}

```{r}

library(gtsummary)

hers_ldl |>
  dplyr::select(LDL, HT, age, BMI, diabetes, smoking, statins) |>
  tbl_summary(by = HT) |>
  add_overall() |>
  bold_labels()
```

Summary table for LDL analysis variables in HERS

:::

### Unadjusted model

::: {#tbl-hers-ldl-unadj}

```{r}

model_ldl_unadj <-
  lm(LDL ~ HT, data = hers_ldl)

model_ldl_unadj |>
  tbl_regression() |>
  bold_labels()
```

Unadjusted linear regression for LDL in HERS

:::

### Adjusted model

Based on the DAG in @fig-hers-dag,
BMI is a common cause of LDL and other outcomes
(but age does not confound the HT → LDL relationship,
since HT was randomized).

In a randomized trial,
treatment assignment is independent of pre-treatment covariates,
so confounding adjustment is not required to estimate the causal effect.
However, adjusting for strong predictors of the outcome
can improve precision.

::: {#tbl-hers-ldl-adj}

```{r}

model_ldl_adj <-
  lm(LDL ~ HT + age + BMI + diabetes + statins, data = hers_ldl)

model_ldl_adj |>
  tbl_regression() |>
  bold_labels()
```

Adjusted linear regression for LDL in HERS

:::

Note that `statins` is a strong predictor of LDL
(statin medications are prescribed specifically to lower LDL).
Including `statins` substantially reduces residual variability,
improving precision of the HT coefficient.

## An observational study example: WCGS {#sec-pred-sel-wcgs-dag}

The HERS study is a randomized trial,
so confounding of the treatment-outcome relationship is not a concern.
To illustrate DAG-guided confounder selection in an **observational study**,
we use the Western Collaborative Group Study (WCGS).

The WCGS was a prospective cohort study of 3,154 men
employed by California companies,
followed from 1960 to 1969 [@rosenman1975coronary].
The exposure of primary interest is **Type A behavior pattern**
(a personality trait characterized by competitive drive,
time urgency, and hostility),
hypothesized to increase CHD risk
independently of classic cardiovascular risk factors.

::: {#fig-wcgs-dag}

```{r}
#| fig-width: 7
#| fig-height: 5

wcgs_dag <- dagitty(
  'dag {
    TypeA [exposure, pos="0,0"]
    CHD [outcome, pos="4,0"]
    age [pos="2,2"]
    smoking [pos="-1,-2"]
    chol [pos="2,-1"]
    sbp [pos="3,-2"]
    bmi [pos="-1,1"]

    TypeA -> CHD
    age -> CHD
    age -> chol
    age -> sbp
    age -> bmi
    age -> TypeA
    TypeA -> smoking
    smoking -> CHD
    chol -> CHD
    sbp -> CHD
    bmi -> CHD
    bmi -> chol
    bmi -> sbp
  }'
)

plot(wcgs_dag)
```

Directed Acyclic Graph (DAG) for the WCGS study.
`TypeA` = Type A behavior pattern (exposure);
`CHD` = coronary heart disease events (outcome);
`chol` = serum cholesterol;
`sbp` = systolic blood pressure.

:::

In @fig-wcgs-dag,
`age` is a **confounder**:
it has direct causal arrows to both the exposure (`TypeA`)
and the outcome (`CHD`).
We assume `TypeA -> smoking`
rather than `smoking -> TypeA`
as an explicit modeling assumption
for this teaching example:
we treat Type A behavior pattern
as a relatively stable trait
that can influence smoking behavior.
Under that assumption,
`smoking` is a mediator on the path
`TypeA -> smoking -> CHD`,
not a confounder.
BMI is causally affected by age.
Cholesterol and SBP are causally affected
by both age and BMI.
All three have direct effects on CHD.
Unlike in HERS,
where HT was randomized
to estimate its causal effect on CHD event risk,
we cannot assume the exposure is unconfounded here.

```{r}
#| label: wcgs-adj-sets

wcgs_adj <- adjustmentSets(
  wcgs_dag,
  exposure = "TypeA",
  outcome = "CHD",
  type = "minimal"
)
wcgs_adj
```

::: notes

The `adjustmentSets()` function returns the minimal set(s) of variables
that block all backdoor paths from `TypeA` to `CHD`.
Adjusting for any of these minimal sets is sufficient to
estimate the total causal effect of Type A behavior on CHD
(under the assumptions encoded in the DAG).

:::
