
## AIC and BIC

::: notes
This section is adapted from @vittinghoff2e [§4.5] and @kleinbaum2014applied [§15].
:::

When we use likelihood ratio tests, we are comparing how well different models fit the data.

Likelihood ratio tests require **nested** models: one must be a special case of the other.

AIC and BIC can be used to compare both nested and non-nested models.

---

### Formulas

:::{#def-aic}
### Akaike Information Criterion (AIC)

$$\text{AIC} = -2 \ell(\hat\theta) + 2 p$$

where $\ell(\hat\theta)$ is the log-likelihood evaluated at the
maximum-likelihood estimates $\hat\theta$,
$p$ is the number of estimated parameters
(including $\hat\sigma^2$ for Gaussian models),
and $n$ is the number of observations.
:::

---

:::{#def-bic}
### Bayesian Information Criterion (BIC)

$$\text{BIC} = -2 \ell(\hat\theta) + p \log(n)$$

where $\ell(\hat\theta)$, $p$, and $n$ are defined as in @def-aic.
:::

---

### Conceptual basis

Each criterion has two components:

-   **Fit term** ($-2\ell(\hat\theta)$):
    measures lack of fit — lower is better.
    A model with more parameters will always achieve a higher (or equal)
    log-likelihood on the observed data.

-   **Penalty term** ($2 p$ for AIC; $p \log(n)$ for BIC):
    penalizes model complexity to guard against overfitting.

Together, they balance **goodness of fit** against **parsimony**.

::: notes
The AIC was introduced by @akaike1974new and is grounded in
information theory (specifically, the Kullback-Leibler divergence).
The BIC was introduced by @schwarz1978estimating as an approximation
to the Bayes factor for model comparison.
:::

---

### AIC vs. BIC

| Criterion | Penalty per parameter | Tends to select |
|-----------|----------------------|-----------------|
| AIC       | $2$                  | larger models   |
| BIC       | $\log(n)$            | smaller models  |

The BIC penalty exceeds the AIC penalty whenever $\log(n) > 2$,
i.e., when $n > e^2 \approx 7.4$.
In practice, BIC almost always penalizes additional parameters more
heavily than AIC and therefore tends to select simpler (more parsimonious)
models [@vittinghoff2e; @kleinbaum2014applied].

---

### AIC in R

```{r}
-2 * logLik(bw_lm2) |> as.numeric() +
  2 * (length(coef(bw_lm2)) + 1) # sigma counts as a parameter here

AIC(bw_lm2)
```

### BIC in R

```{r}
-2 * logLik(bw_lm2) |> as.numeric() +
  (length(coef(bw_lm2)) + 1) * log(nobs(bw_lm2))

BIC(bw_lm2)
```

---

### Interpretation

-   **Lower values are better.**
    There are no hypothesis tests or p-values associated with these
    criteria.

-   To compare models, calculate the criterion for each model and
    choose the model with the **smallest value**.

-   Differences of less than 2 units are generally considered negligible;
    differences greater than 10 are considered strong evidence in favor
    of the lower-criterion model.

-   BIC tends to favor simpler models than AIC,
    especially when the sample size is large.

## (Residual) Deviance

Let $q$ be the number of distinct covariate combinations in a data set.

```{r}
bw_X_unique <-
  bw |>
  count(sex, age)

n_unique_bw <- nrow(bw_X_unique)
```

For example, in the `birthweight` data, 
there are $q = `r n_unique_bw`$ unique patterns (@tbl-bw-x-combos).

::: {#tbl-bw-x-combos}

```{r}
bw_X_unique
```

Unique covariate combinations in the `birthweight` data, with replicate counts

:::

---

::: {#def-replicates}
### Replicates
If a given covariate pattern has more than one observation in a dataset, 
those observations are called **replicates**.
:::

---

::: {#exm-replicate-bw}

### Replicates in the `birthweight` data

In the `birthweight` dataset, there are 2 replicates of the combination "female, age 36" (@tbl-bw-x-combos).

:::

---

::: {#exr-replicate-bw}

### Replicates in the `birthweight` data

Which covariate pattern(s) in the `birthweight` data has the most replicates?

:::

---

::: {#sol-replicate-bw}

### Replicates in the `birthweight` data

Two covariate patterns are tied for most replicates: males at age 40 weeks 
and females at age 40 weeks.
40 weeks is the usual length for human pregnancy (@polin2011fetal), so this result makes sense.

```{r}
bw_X_unique |> dplyr::filter(n == max(n))
```

:::

---

### Saturated models {.smaller}

The most complicated model we could fit would have one parameter (a mean) for each covariate pattern, plus a variance parameter:

```{r}
#| label: tbl-bw-model-sat
#| tbl-cap: "Saturated model for the `birthweight` data"
lm_max <-
  bw |>
  mutate(age = factor(age)) |>
  lm(
    formula = weight ~ sex:age - 1,
    data = _
  )

lm_max |>
  parameters() |>
  print_md()
```

We call this model the **full**, **maximal**, or **saturated** model for this dataset.

---

```{r}
library(rlang) # defines the `.data` pronoun
plot_PIs_and_CIs <- function(model, data) {
  cis <- model |>
    predict(interval = "confidence") |>
    suppressWarnings() |>
    tibble::as_tibble()
  names(cis) <- paste("ci", names(cis), sep = "_")

  preds <- model |>
    predict(interval = "predict") |>
    suppressWarnings() |>
    tibble::as_tibble()
  names(preds) <- paste("pred", names(preds), sep = "_")
  dplyr::bind_cols(bw, cis, preds) |>
    ggplot2::ggplot() +
    ggplot2::aes(
      x = .data$age, 
      y = .data$weight, 
      col = .data$sex
    ) +
    ggplot2::geom_point() +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::geom_line(ggplot2::aes(y = .data$ci_fit)) +
    ggplot2::geom_ribbon(
      ggplot2::aes(
        ymin = .data$pred_lwr,
        ymax = .data$pred_upr
      ),
      alpha = 0.2
    ) +
    ggplot2::geom_ribbon(
      ggplot2::aes(
        ymin = .data$ci_lwr,
        ymax = .data$ci_upr
      ),
      alpha = 0.5
    ) +
    ggplot2::facet_wrap(~sex)
}
```

:::{#fig-max-vs-intxn-fits layout-ncol=2}

:::{#fig-fit-intxn}

```{r}
plot_PIs_and_CIs(bw_lm2, bw)
```
Model [-@eq-BW-lm-interact] (linear with age:sex interaction)
:::

:::{#fig-fit-max}

```{r}
plot_PIs_and_CIs(lm_max, bw)
```
Saturated model
:::

Model [-@eq-BW-lm-interact] and saturated model for `birthweight` data, 
with confidence and prediction intervals

:::


---

We can calculate the log-likelihood of this model as usual:

```{r}
logLik(lm_max)
```

We can compare this model to our other models using chi-square tests, as usual:

```{r}
lrtest(lm_max, bw_lm2)
```

The likelihood ratio statistic for this test is 
$$\lambda = 2 * (\ell_{\text{full}} - \ell) = `r 2 * (logLik(lm_max) - logLik(bw_lm2))`$$ 
where:

-   $\ell_{\text{full}}$ is the log-likelihood of the full model: `r logLik(lm_max)`
-   $\ell$ is the log-likelihood of our comparison model (two slopes, two intercepts): `r logLik(bw_lm2)`

This statistic is called the **deviance** or **residual deviance** for our two-slopes and two-intercepts model; it tells us how much the likelihood of that model deviates from the likelihood of the maximal model.

The corresponding p-value tells us whether there we have enough evidence to detect that our two-slopes, two-intercepts model is a worse fit for the data than the maximal model; in other words, it tells us if there's evidence that we missed any important patterns. (Remember, a nonsignificant p-value could mean that we didn't miss anything and a more complicated model is unnecessary, or it could mean we just don't have enough data to tell the difference between these models.)

## Null Deviance

Similarly, the *least* complicated model we could fit would have only one mean parameter, an intercept:

$$\text E[Y|X=x] = \beta_0$$ We can fit this model in R like so:

```{r}
lm0 <- lm(weight ~ 1, data = bw)

lm0 |>
  parameters() |>
  print_md()
```

---

:::{#fig-null-model}
```{r}
lm0 |> plot_PIs_and_CIs()
```

Null model for birthweight data, with 95% confidence and prediction intervals.
:::


---

This model also has a likelihood:

```{r}
logLik(lm0)
```

And we can compare it to more complicated models using a likelihood ratio test:

```{r}
lrtest(bw_lm2, lm0)
```

The likelihood ratio statistic for the test comparing the null model to the maximal model is $$\lambda = 2 * (\ell_{\text{full}} - \ell_{0}) = `r 2 * (logLik(lm_max) - logLik(lm0))`$$ where:

-   $\ell_{\text{0}}$ is the log-likelihood of the null model: `r logLik(lm0)`
-   $\ell_{\text{full}}$ is the log-likelihood of the maximal model: `r logLik(lm_max)`

In R, this test is:

```{r}
lrtest(lm_max, lm0)
```

This log-likelihood ratio statistic is called the **null deviance**. It tells us whether we have enough data to detect a difference between the null and full models.

{{< include _sec_compare_lm_glm_deviance.qmd >}}

