
## Motivating example: `birthweight` data

Research question: is there really an interaction between sex and age?

$H_0: \beta_{AM} = 0$

$H_A: \beta_{AM} \neq 0$

$P(|\hat\beta_{AM}| > |`r coef(bw_lm2)["sexmale:age"]`| \mid H_0)$ = ?

## Inference for individual predictor coefficients

{{< include _sec_linreg_coef_sampling_dist.qmd >}}

---

### Estimated covariance matrix and standard errors

{{< include _sec_bw_vcov.qmd >}}

### Wald tests and confidence intervals

{{< include _sec_linreg_wald.qmd >}}

## Inference for predicted means

{{< include _sec_linreg_pred_inference.qmd >}}

## Inference for differences in means

{{< include _sec_linreg_diff_means_inference.qmd >}}

## Likelihood ratio statistics

```{r}
logLik(bw_lm2)
logLik(bw_lm1)

log_LR <- (logLik(bw_lm2) - logLik(bw_lm1)) |> as.numeric()
delta_df <- (bw_lm1$df.residual - df.residual(bw_lm2))


x_max <- 1
```

---

```{r}
#| label: fig-chisq-plot
#| fig-cap: "Chi-square distribution"

d_log_LR <- function(x, df = delta_df) dchisq(x, df = df)

chisq_plot <-
  ggplot() +
  geom_function(fun = d_log_LR) +
  stat_function(
    fun = d_log_LR,
    xlim = c(log_LR, x_max),
    geom = "area",
    fill = "gray"
  ) +
  geom_segment(
    aes(
      x = log_LR,
      xend = log_LR,
      y = 0,
      yend = d_log_LR(log_LR)
    ),
    col = "red"
  ) +
  xlim(0.0001, x_max) +
  ylim(0, 4) +
  ylab("p(X=x)") +
  xlab("log(likelihood ratio) statistic [x]") +
  theme_classic()
chisq_plot |> print()
```

---

Now we can get the p-value:

```{r}
#| label: LRT-pval
pchisq(
  q = 2 * log_LR,
  df = delta_df,
  lower = FALSE
) |>
  print()
```


---

In practice you don't have to do this by hand; there are functions to do it for you:

```{r}
# built in
library(lmtest)
lrtest(bw_lm2, bw_lm1)
```

---

{{< include _sec_linreg_Ftest.qmd >}}

