In model [-@eq-BW-lm-interact] above, $\heinf(\beta)$ is:

:::{#tbl-vcov}

```{r}
#| code-fold: show
bw_lm2 |> vcov()
```

Covariance matrix of $\hvb$ for `birthweight` model [-@eq-BW-lm-interact] (with interaction term)

:::

---

If we take the square roots of the diagonals, we get the standard errors listed in the model output:

```{r}
bw_lm2 |>
  vcov() |>
  diag() |>
  sqrt()
```

```{r}
#| label: tbl-mod-intx
#| tbl-cap: "Estimated model for `birthweight` data with interaction term"
bw_lm2 |>
  parameters() |>
  print_md()
```

So we can do confidence intervals, hypothesis tests, and p-values exactly as in the one-variable case we looked at previously.

### Residual Standard Deviation

::: notes
$\hs$ represents an *estimate* of the *Residual Standard Deviation* parameter, $\s$. 
We can extract $\hs$ from the fitted model, using the `sigma()` function:
:::

```{r}
#| code-fold: show
sigma(bw_lm2)
```

---

#### $\s$ is NOT "Residual standard error"

::: notes
In the `summary.lm()` output, this estimate is labeled as `"Residual standard error"`:
:::

```{r}
#| code-fold: show
summary(bw_lm2)
```

::: notes
However, this is a misnomer: see note in `?stats::sigma`
:::
