
## Example: `birthweight`

In the birthweight example, the variable `sex` had only two observed values:

```{r}
unique(bw$sex)
```

If there are more than two observed values, we can't just use a single variable with 0s and 1s.

## 

:::{.notes}
For example, @tbl-iris-data shows the 
[(in)famous](https://www.meganstodel.com/posts/no-to-iris/) 
`iris` data (@anderson1935irises), 
and @tbl-iris-summary provides summary statistics. 
The data include three species: "setosa", "versicolor", and "virginica".
:::

```{r}
#| tbl-cap: "The `iris` data"
#| label: tbl-iris-data
head(iris)
```

```{r}
#| tbl-cap: "Summary statistics for the `iris` data"
#| label: tbl-iris-summary

library(table1)
table1(
  x = ~ . | Species,
  data = iris,
  overall = FALSE
)
```

---

If we want to model `Sepal.Length` by species, we could create a variable $X$ that represents "setosa" as $X=1$, "virginica" as $X=2$, and "versicolor" as $X=3$.

```{r}
#| label: tbl-numeric-coding
#| tbl-cap: "`iris` data with numeric coding of species"
data(iris) # this step is not always necessary, but ensures you're starting
# from the original version of a dataset stored in a loaded package

iris <-
  iris |>
  tibble() |>
  mutate(
    X = case_when(
      Species == "setosa" ~ 1,
      Species == "virginica" ~ 2,
      Species == "versicolor" ~ 3
    )
  )

iris |>
  distinct(Species, X)
```

Then we could fit a model like:

```{r}
#| label: tbl-iris-numeric-species
#| tbl-cap: "Model of `iris` data with numeric coding of `Species`"
iris_lm1 <- lm(Sepal.Length ~ X, data = iris)
iris_lm1 |>
  parameters() |>
  print_md()
```

## Let's see how that model looks:

```{r}
#| label: fig-iris-numeric-species-model
#| fig-cap: "Model of `iris` data with numeric coding of `Species`"
iris_plot1 <- iris |>
  ggplot(
    aes(
      x = X,
      y = Sepal.Length
    )
  ) +
  geom_point(alpha = .1) +
  geom_abline(
    intercept = coef(iris_lm1)[1],
    slope = coef(iris_lm1)[2]
  ) +
  theme_bw(base_size = 18)
print(iris_plot1)
```

We have forced the model to use a straight line for the three estimated means. Maybe not a good idea?

## Let's see what R does with categorical variables by default:

```{r}
#| label: tbl-iris-model-factor1
#| tbl-cap: "Model of `iris` data with `Species` as a categorical variable"
iris_lm2 <- lm(Sepal.Length ~ Species, data = iris)
iris_lm2 |>
  parameters() |>
  print_md()
```

## Re-parametrize with no intercept

If you don't want the default and offset option, you can use "-1" like we've seen previously:

```{r}
#| label: tbl-iris-no-intcpt
iris_lm2_no_int <- lm(Sepal.Length ~ Species - 1, data = iris)
iris_lm2_no_int |>
  parameters() |>
  print_md()
```

## Let's see what these new models look like:

```{r}
#| label: fig-iris-no-intcpt
iris_plot2 <-
  iris |>
  mutate(
    predlm2 = predict(iris_lm2)
  ) |>
  arrange(X) |>
  ggplot(aes(x = X, y = Sepal.Length)) +
  geom_point(alpha = .1) +
  geom_line(aes(y = predlm2), col = "red") +
  geom_abline(
    intercept = coef(iris_lm1)[1],
    slope = coef(iris_lm1)[2]
  ) +
  theme_bw(base_size = 18)

print(iris_plot2)
```

## Let's see how R did that:

```{r}
#| label: tbl-iris-model-matrix-factor
formula(iris_lm2)
model.matrix(iris_lm2) |>
  as_tibble() |>
  unique()
```

This format is called a "corner point parametrization" 
(e.g., in @dobson4e) or "treatment coding" (e.g., in @dunn2018generalized).

The default contrasts are controlled by `options("contrasts")`:

```{r}
options("contrasts")
```

See `?options` for more details.

---

```{r}
#| label: tbl-iris-group-point-parameterization
formula(iris_lm2_no_int)
model.matrix(iris_lm2_no_int) |>
  as_tibble() |>
  unique()
```

This format is called a "group point parametrization" (e.g., in @dobson4e).

There are more options; see @dobson4e §6.4.1 and the 
[`codingMatrices` package](https://CRAN.R-project.org/package=codingMatrices) 
[vignette](https://cran.r-project.org/web/packages/codingMatrices/vignettes/codingMatrices.pdf) 
(@venablescodingMatrices).

