
We don't have to do these calculations by hand every time; the
`survival` package and several others have functions available to
automate many of these tasks (full list:
<https://cran.r-project.org/web/views/Survival.html>).

### The `Surv` function

To use the `survival` package, the first step is telling R how to
combine the exit time and exit reason (censoring versus event) columns.
The `Surv()` function accomplishes this task.

#### Example: `Surv()` with `drug6mp` data

```{r}
#| code-fold: show
#| code-line-numbers: "5-7"
library(survival)
drug6mp.v3 <-
  drug6mp.v2 |>
  mutate(
    surv2 = Surv(
      time = t2,
      event = (outcome == "relapsed")
    )
  )

print(drug6mp.v3)
```

The output of `Surv()` is a vector of objects with class `Surv`. When we
print this vector:

-   observations where the event was observed are printed as the event
time (for example, `surv2 = 10` on line 1)

-   observations where the event was right-censored are printed as the
censoring time with a plus sign (`+`; for example, `surv2 = 32+` on
line 3).

### The `survfit` function

Once we have constructed our `Surv` variable, we can calculate the
Kaplan-Meier estimate of the survival curve using the `survfit()`
function.

::: callout-note
The documentation for `?survfit` isn't too helpful; the
`survfit.formula` documentation is better.

```{r, printr.help.sections = c("description", "usage")}
#| include: false
?survfit.formula
```
:::

---

#### Example: `survfit()` with `drug6mp` data

Here we use `survfit()` to create a `survfit` object, which contains the
Kaplan-Meier estimate:

```{r}
#| code-fold: show
drug6mp.km_model <- survfit(
  formula = surv2 ~ 1,
  data = drug6mp.v3
)
```

`print.survfit()` just gives some summary statistics:

```{r}
#| code-fold: show
print(drug6mp.km_model)
```

`summary.survfit()` shows us the underlying Kaplan-Meier table:

```{r}
#| code-fold: show
summary(drug6mp.km_model)
```

---

We can specify which time points we want using the `times` argument:

```{r}
#| code-fold: show
summary(
  drug6mp.km_model,
  times = c(0, drug6mp.v3$t2)
)
```

---

```{r, printr.help.sections = c("description", "usage", "arguments")}
#| code-fold: show
?summary.survfit
```

### Plotting estimated survival functions

We can plot `survfit` objects with `plot()`, `autoplot()`, or
`ggsurvplot()`:

```{r}
#| code-fold: show
#| fig-cap: "Kaplan-Meier Survival Curve for 6-MP Patients"

library(ggfortify)
autoplot(drug6mp.km_model)

# not shown:
# plot(drug6mp.km_model)

# library(survminer)
# ggsurvplot(drug6mp.km_model)
```

---

#### quantiles of survival curve

We can extract quantiles with `quantile()`:

```{r}
#| code-line-numbers: "2"
drug6mp.km_model |>
  quantile(p = c(.25, .5)) |>
  as_tibble() |>
  mutate(p = c(.25, .5)) |>
  relocate(p, .before = everything())
```

### R Packages Ecosystem for Survival Analysis

{{< include _sec-surv-r-packages-ecosystem.qmd >}}
