## Rescaling

::: notes
Centering covariates (subtracting a constant from them) 
can make coefficient interpretations more meaningful, 
particularly for the intercept and main effect terms in models with interactions.
:::

### Rescale age {.smaller}

::: notes
For example, 
the intercept $\beta_0$ in our interaction model represents 
the mean birthweight for females at age = 0 weeks, 
which is not a biologically plausible scenario.
If we center age at 36 weeks instead, 
$\beta_0$ will represent the mean birthweight for females at 36 weeks, 
which is more interpretable.
:::

```{r}
bw <-
  bw |>
  mutate(
    `age - mean` = age - mean(age),
    `age - 36wks` = age - 36
  )

lm1_c <- lm(weight ~ sex + `age - 36wks`, data = bw)

lm2_c <- lm(weight ~ sex + `age - 36wks` + sex:`age - 36wks`, data = bw)

parameters(lm2_c, ci_method = "wald") |> print_md()
```

Compare with what we got without rescaling:

```{r}
parameters(bw_lm2, ci_method = "wald") |> print_md()
```

::: notes
Notice that the slope coefficients (`age - 36wks` and `sexmale:age - 36wks`) 
remain the same, 
but the intercept and `sexmale` coefficient have changed 
to reflect means at age = 36 weeks rather than age = 0 weeks.
:::

---

{{< include _exr-linreg-rescaling.qmd >}}
