
### Estimating survival in datasets without censoring

::: notes
In the `drug6mp` dataset, the estimated survival function for the placebo patients is easy to compute. 
For any time $t$ in months, $\surv(t)$ is the fraction of patients with times greater than $t$:
:::

```{r}
#| include: false
#| code-summary: example-km-curve

y <- c(1, 2, 3, 3, 3, 5, 9, 13, 23, 100)
d <- c(1, 1, 1, 0, 1, 0, 1, 1, 1, 1)
library(survival)
temp1 <- survfit(Surv(y, d) ~ 1)
```


### Estimating survival in datasets with censoring

-   For the 6-MP patients, we cannot ignore the censored data because we
know that the time to relapse is greater than the censoring time.

-   For any time $t$ in months, we know that 6-MP patients with times
greater than $t$ have not relapsed, and those with relapse time less
than $t$ have relapsed, but we don't know if patients with censored
time less than $t$ have relapsed or not.

-   The procedure we usually use is the Kaplan-Meier product-limit
estimator of the survival function.

-   The Kaplan-Meier estimator is a step function (like the empirical
cdf), which changes value only at the event times, not at the
censoring times.

-   At each event time $t$, we compute the at-risk group size $Y$, which
is all those observations whose event time or censoring time is at
least $t$.

-   If $d$ of the observations have an event time (not a censoring time)
of $t$, then the group of survivors immediately following time $t$
is reduced by the fraction $$\frac{Y-d}{Y}=1-\frac{d}{Y}$$

---

{{< include _def-km-estimator.qmd >}}

---

:::{#thm-KM-est-no-cens}
#### Kaplan-Meier Estimate with No Censored Observations

If there are no censored data, and there are $n$ data points, then just
after (say) the third event time 

$$
\begin{aligned}
\hsurv(t) 
&= \prod_{t_i < t}\sb{1-\frac{d_i}{Y_i}}
\\ &= \sb{\frac{n-d_1}{n}} \sb{\frac{n-d_1-d_2}{n-d_1}} \sb{\frac{n-d_1-d_2-d_3}{n-d_1-d_2}}
\\ &= \frac{n-d_1-d_2-d_3}{n}
\\ &=1-\frac{d_1+d_2+d_3}{n}
\\ &=1-\hat F(t)
\end{aligned}
$$

where $\hat F(t)$ is the usual empirical CDF estimate.

:::

### Kaplan-Meier curve for `drug6mp` data

Here is the Kaplan-Meier estimated survival curve for the patients who
received 6-MP in the `drug6mp` dataset (we will see code to produce
figures like this one shortly):

```{r}
#| fig-cap: "Kaplan-Meier Survival Curve for 6-MP Patients"
#| label: fig-KM-mp6
# | echo: false

require(KMsurv)
data(drug6mp)
library(dplyr)
library(survival)

drug6mp_km_model1 <-
  drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, data = _)

library(ggfortify)
drug6mp_km_model1 |>
  autoplot(
    mark.time = TRUE,
    conf.int = FALSE
  ) +
  expand_limits(y = 0) +
  xlab("Time since diagnosis (months)") +
  ylab("KM Survival Curve")
```

### Kaplan-Meier calculations {.smaller}

Let's compute these estimates and build the chart by hand:

```{r}
library(KMsurv)
library(dplyr)
data(drug6mp)

drug6mp.v2 <-
  drug6mp |>
  as_tibble() |>
  mutate(
    remstat = remstat |>
      case_match(
        1 ~ "partial",
        2 ~ "complete"
      ),
    # renaming to "outcome" while relabeling is just a style choice:
    outcome = relapse |>
      case_match(
        0 ~ "censored",
        1 ~ "relapsed"
      )
  )

km.6mp <-
  drug6mp.v2 |>
  summarize(
    .by = t2,
    Relapses = sum(outcome == "relapsed"),
    Censored = sum(outcome == "censored")
  ) |>
  # here we add a start time row, so the graph starts at time 0:
  bind_rows(
    tibble(
      t2 = 0,
      Relapses = 0,
      Censored = 0
    )
  ) |>
  # sort in time order:
  arrange(t2) |>
  mutate(
    Exiting = Relapses + Censored,
    `Study Size` = sum(Exiting),
    Exited = cumsum(Exiting) |> dplyr::lag(default = 0),
    `At Risk` = `Study Size` - Exited,
    Hazard = Relapses / `At Risk`,
    `KM Factor` = 1 - Hazard,
    `Cumulative Hazard` = cumsum(`Hazard`),
    `KM Survival Curve` = cumprod(`KM Factor`)
  )

library(pander)
pander(km.6mp)
```

---

#### Summary

For the 6-MP patients at time 6 months, there are 21 patients at risk.
At $t=6$ there are 3 relapses and 1 censored observations.

The Kaplan-Meier factor is $(21-3)/21 = 0.857$. The number at risk for
the next time ($t=7$) is $21-3-1=17$.

At time 7 months, there are 17 patients at risk. At $t=7$ there is 1
relapse and 0 censored observations. The Kaplan-Meier factor is
$(17-1)/17 = 0.941$. The Kaplan Meier estimate is
$0.857\times0.941=0.807$. The number at risk for the next time ($t=9$)
is $17-1=16$.

---

Now, let's graph this estimated survival curve using `ggplot()`:

```{r}
#| label: fig-km-by-hand
#| fig-cap: "KM curve for 6MP patients, calculated by hand"
library(ggplot2)
conflicts_prefer(dplyr::filter)
km.6mp |>
  ggplot(aes(x = t2, y = `KM Survival Curve`)) +
  geom_step() +
  geom_point(data = km.6mp |> filter(Censored > 0), shape = 3) +
  expand_limits(y = c(0, 1), x = 0) +
  xlab("Time since diagnosis (months)") +
  ylab("KM Survival Curve") +
  scale_y_continuous(labels = scales::percent)
```
