
## Prediction for linear models

:::{#def-predicted-value}
### Predicted value

In a regression model $\p(y|\vx)$, 
the **predicted value** of $y$ given $\vx$ is the estimated mean of $Y$ given $\vX=\vx$:

$$\hat y \eqdef \hE{Y|\vX=\vx}$$
:::

---

For linear models, 
the predicted value can be straightforwardly calculated 
by multiplying each predictor value $x_j$ by its corresponding coefficient $\beta_j$ 
and adding up the results:

$$
\begin{aligned}
\hat y &= \hE{Y|\vX=\vx} \\
&= \vx'\hat\beta \\
&= \hat\beta_0\cdot 1 + \hat\beta_1 x_1 + ... + \hat\beta_p x_p
\end{aligned}
$$

---

## Example: prediction for the `birthweight` data

```{r}
x <- c(1, 1, 40)
sum(x * coef(bw_lm1))
```

R has built-in functions for prediction:

```{r}
x <- tibble(age = 40, sex = "male")
bw_lm1 |> predict(newdata = x)
```

If you don't provide `newdata`, 
R will use the covariate values from the original dataset:

```{r}
predict(bw_lm1)
```

These special predictions are called the *fitted values* of the dataset:

:::{#def-fitted-value}

For a given dataset $(\vY, \mX)$ and corresponding fitted model $\p_{\hb}(\vy|\mx)$, 
the **fitted value** of $y_i$ is 
the predicted value of $y$ when $\vX=\vx_i$ using the estimate parameters $\hb$.

:::

R has an extra function to get these values:

```{r}
fitted(bw_lm1)
```

## Confidence intervals {#sec-se-fitted}

Use `predict(se.fit = TRUE)` to compute SEs for predicted values:

```{r}
bw_lm1 |>
  predict(
    newdata = x,
    se.fit = TRUE
  )
```

The output of `predict.lm(se.fit = TRUE)` is a `list()`; 
you can extract the elements with `$` or `magrittr::use_series()`:

```{r}
library(magrittr)
bw_lm1 |>
  predict(
    newdata = x,
    se.fit = TRUE
  ) |>
  use_series(se.fit)
```

---

We can construct **confidence intervals** for $\E{Y|X=x}$ 
using the usual formula:


$$\mu(\vx) \in 
\paren{\red{\hat{\mu}(\vx)} \pm \blue{\zeta_\alpha}}$$

$$
\blue{\zeta_\alpha} = 
t_{n-p}\paren{1-\frac{\alpha}{2}} * \hse{\hat{\mu}(\vx)}
$$

$$\red{\hat{\mu}(\vx)} = \vx \cdot \hb$$

$$\se{\hat{\mu}(\vx)} = \sqrt{\Var{\hat{\mu}(\vx)}}$$
$$
\ba
\Var{\hat{\mu}(\vx)} &= \Var{x'\hat{\beta}}
\\&= x' \Var{\hat{\beta}} x
\\&= x' \sigma^2(\mX'\mX)^{-1} x
\\&= \sigma^2 x' (\mX'\mX)^{-1} x
\\&= \sumin \sumjn x_i x_j \Covs{\hb_i}{\hb_j}
\ea
$$
$$\hVar{\hat{\mu}(\vx)} = \hs^2 x' (\mX'\mX)^{-1} x$$

```{R}
bw_lm2 |> predict(
  newdata = x,
  interval = "confidence"
)
```

---

:::{#fig-pred-int}

```{r}
library(sjPlot)
bw_lm2 |>
  plot_model(type = "pred", terms = c("age", "sex"), show.data = TRUE) +
  theme_sjplot() +
  theme(legend.position = "bottom")
```

Predicted values and confidence bands for the `birthweight` model with interaction term
:::


## Prediction intervals {#sec-se-pred}

We can also construct **prediction intervals** for 
the value of a new observation $Y^*$,
given a covariate pattern $\vx^*$:

$$
\ba
\hat{Y^*} &= \hat\mu({\vx}^*) + \hat{\epsilon}^*
\\ \Var{\hat{Y}} &= \Var{\hat\mu} + \Var{\hat\epsilon}
\\ \Var{\hat{Y}} &= \Var{x'\hb} + \Var{\hat\epsilon}
\\ &= {{\vx}^{*\top}} \Var{\hb} {{\vx}^*} + \sigma^2
\\ &=  {{\vx}^{*\top}} \paren{\sigma^2(\mX'\mX)^{-1}} {{\vx}^*} + \sigma^2
\\ &= \sigma^2 {{\vx}^{*\top}} (\mX'\mX)^{-1} {{\vx}^*} + \sigma^2
\ea
$${#eq-prediction-var}

::: notes
See @hoggtanis2015 §7.6 (p. 340) for a longer version.
:::

```{r}
#| code-fold: show
bw_lm2 |>
  predict(newdata = x, interval = "predict")
```

---

If you don't specify `newdata`, you get a warning:

```{r}
#| code-fold: show
#| warning: true
bw_lm2 |>
  predict(interval = "predict") |>
  head()
```

The warning from the last command is: "predictions on current data refer to *future* responses" (since you already know what happened to the current data, and thus don't need to predict it).

See `?predict.lm` for more.

---

```{r}
library(rlang) # defines the `.data` pronoun
plot_PIs_and_CIs <- function(data, model = NULL) {
  if (is.null(model)) {
    rlang::abort("`model` must be supplied to `plot_PIs_and_CIs()`.")
  }

  if (!is.data.frame(data) && is.data.frame(model)) {
    rlang::warn(
      paste0(
        "`plot_PIs_and_CIs()` received `model` as the first argument ",
        "and `data` as the second. ",
        "Please update the call to use `data` first and `model` second."
      )
    )
    tmp <- data
    data <- model
    model <- tmp
  }

  cis <- model |>
    predict(newdata = data, interval = "confidence") |>
    suppressWarnings() |>
    tibble::as_tibble()
  names(cis) <- paste("ci", names(cis), sep = "_")

  preds <- model |>
    predict(newdata = data, interval = "predict") |>
    suppressWarnings() |>
    tibble::as_tibble()
  names(preds) <- paste("pred", names(preds), sep = "_")
  dplyr::bind_cols(data, cis, preds) |>
    ggplot2::ggplot() +
    ggplot2::aes(
      x = .data$age,
      y = .data$weight,
      col = .data$sex
    ) +
    ggplot2::geom_point() +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::geom_line(ggplot2::aes(y = .data$ci_fit)) +
    ggplot2::geom_ribbon(
      ggplot2::aes(
        ymin = .data$pred_lwr,
        ymax = .data$pred_upr
      ),
      alpha = 0.2
    ) +
    ggplot2::geom_ribbon(
      ggplot2::aes(
        ymin = .data$ci_lwr,
        ymax = .data$ci_upr
      ),
      alpha = 0.5
    ) +
    ggplot2::facet_wrap(~sex)
}
```

:::{#fig-pred-int-bw}
```{r}
plot_PIs_and_CIs(bw_lm2, bw)
```

Confidence and prediction intervals for `birthweight` model [-@eq-BW-lm-interact]

:::

