
## Prediction for linear models

:::{#def-predicted-value}
### Predicted value

In a regression model $\p(y|\vx)$, 
the **predicted value** of $y$ given $\vx$ is the estimated mean of $Y$ given $\vX=\vx$:

$$\hat y \eqdef \hE{Y|\vX=\vx}$$
:::

---

For linear models, 
the predicted value can be straightforwardly calculated 
by multiplying each predictor value $x_j$ by its corresponding coefficient $\beta_j$ 
and adding up the results:

$$
\begin{aligned}
\hat y &= \hE{Y|\vX=\vx} \\
&= \vx'\hat\beta \\
&= \hat\beta_0\cdot 1 + \hat\beta_1 x_1 + ... + \hat\beta_p x_p
\end{aligned}
$$

---

## Example: prediction for the `birthweight` data

```{r}
x <- c(1, 1, 40)
sum(x * coef(bw_lm1))
```

R has built-in functions for prediction:

```{r}
x <- tibble(age = 40, sex = "male")
bw_lm1 |> predict(newdata = x)
```

If you don't provide `newdata`, 
R will use the covariate values from the original dataset:

```{r}
predict(bw_lm1)
```

These special predictions are called the *fitted values* of the dataset:

:::{#def-fitted-value}

For a given dataset $(\vY, \mX)$ and corresponding fitted model $\p_{\hb}(\vy|\mx)$, 
the **fitted value** of $y_i$ is 
the predicted value of $y$ when $\vX=\vx_i$ using the estimate parameters $\hb$.

:::

R has an extra function to get these values:

```{r}
fitted(bw_lm1)
```

## Confidence intervals {#sec-se-fitted}

Use `predict(se.fit = TRUE)` to compute SEs for predicted values:

```{r}
bw_lm1 |>
  predict(
    newdata = x,
    se.fit = TRUE
  )
```

The output of `predict.lm(se.fit = TRUE)` is a `list()`; 
you can extract the elements with `$` or `magrittr::use_series()`:

```{r}
library(magrittr)
bw_lm1 |>
  predict(
    newdata = x,
    se.fit = TRUE
  ) |>
  use_series(se.fit)
```

---

We can construct **confidence intervals** for $\E{Y|X=x}$ 
using the usual formula:


$$\mu(\vx) \in 
\paren{\red{\hat{\mu}(\vx)} \pm \blue{\zeta_\alpha}}$$

$$
\blue{\zeta_\alpha} = 
t_{n-p}\paren{1-\frac{\alpha}{2}} * \hse{\hat{\mu}(\vx)}
$$

$$\red{\hat{\mu}(\vx)} = \vx \cdot \hb$$

$$\se{\hat{\mu}(\vx)} = \sqrt{\Var{\hat{\mu}(\vx)}}$$
$$
\ba
\Var{\hat{\mu}(\vx)} &= \Var{x'\hat{\beta}}
\\&= x' \Var{\hat{\beta}} x
\\&= x' \sigma^2(\mX'\mX)^{-1} x
\\&= \sigma^2 x' (\mX'\mX)^{-1} x
\\&= \sumin \sumjn x_i x_j \Covs{\hb_i}{\hb_j}
\ea
$$
$$\hVar{\hat{\mu}(\vx)} = \hs^2 x' (\mX'\mX)^{-1} x$$

```{R}
bw_lm2 |> predict(
  newdata = x,
  interval = "confidence"
)
```

---

:::{#fig-pred-int}

```{r}
library(sjPlot)
bw_lm2 |>
  plot_model(type = "pred", terms = c("age", "sex"), show.data = TRUE) +
  theme_sjplot() +
  theme(legend.position = "bottom")
```

Predicted values and confidence bands for the `birthweight` model with interaction term
:::

### Why are confidence bands narrower near the center of the data? {#sec-why-ci-narrower}

The confidence bands in @fig-pred-int are visibly narrower
near the center of the gestational age range.
To understand why,
consider a simple linear regression with one predictor $a$ (gestational age)
and an intercept.
The covariate vector for a new observation at age $a$ is $\vx = \tp{(1, a)}$,
and the general variance formula for the predicted mean specializes to:

$$
\ba
\Var{\hat\mu(a)}
&= \sigma^2\, \tp{\vx}(\mX'\mX)^{-1}\vx \\
&= \red{\sigma^2 \left(\frac{1}{n} + \frac{(a - \bar{a})^2}{S_{AA}}\right)}
\ea
$${#eq-var-fitted-slr}

where $\bar{a} = \frac{1}{n}\sumin a_i$ is the mean gestational age
and $S_{AA} = \sumin (a_i - \bar{a})^2$ is the total variation in gestational age.

To derive @eq-var-fitted-slr,
first note that for $n$ observations with design matrix rows $(1, a_i)$:

$$
\mX'\mX
= \begin{pmatrix} n & n\bar{a} \\ n\bar{a} & \sumin a_i^2 \end{pmatrix}
$$

The determinant of $\mX'\mX$ is:

$$
\ba
\det(\mX'\mX)
&= n \cdot \sumin a_i^2 - (n\bar{a})^2 \\
&= n\left(\sumin a_i^2 - n\bar{a}^2\right) \\
&= n\, S_{AA}
\ea
$$

so its inverse is:

$$
(\mX'\mX)^{-1}
= \frac{1}{n\, S_{AA}}
  \begin{pmatrix} \sumin a_i^2 & -n\bar{a} \\ -n\bar{a} & n \end{pmatrix}
$$

Substituting $\vx = \tp{(1, a)}$ into the quadratic form:

$$
\ba
\tp{\vx}(\mX'\mX)^{-1}\vx
&= \frac{1}{n\, S_{AA}}
  (1,\; a)
  \begin{pmatrix} \sumin a_i^2 & -n\bar{a} \\ -n\bar{a} & n \end{pmatrix}
  \begin{pmatrix} 1 \\ a \end{pmatrix} \\
&= \frac{\sumin a_i^2 - 2n\bar{a}\,a + n\,a^2}{n\, S_{AA}} \\
&= \frac{\left(\sumin a_i^2 - n\bar{a}^2\right) + n(a - \bar{a})^2}{n\, S_{AA}} \\
&= \frac{S_{AA} + n(a - \bar{a})^2}{n\, S_{AA}} \\
&= \frac{1}{n} + \frac{(a - \bar{a})^2}{S_{AA}}
\ea
$$

Therefore the estimated standard error of the predicted mean at age $a$ is:

$$
\HSE{\hat\mu(a)}
= \hs \sqrt{\frac{1}{n} + \frac{(a - \bar{a})^2}{S_{AA}}}
$${#eq-se-fitted-slr}

Since $(a - \bar{a})^2 \geq 0$ and equals zero when $a = \bar{a}$,
the standard error is minimized at the mean gestational age $\bar{a}$
and increases as $a$ moves away from $\bar{a}$ in either direction.
Consequently, confidence intervals are narrowest at $\bar{a}$
and widen toward the extremes of the gestational age range.

Intuitively, the fitted line is "anchored" at the center of the data:
any fitted line must pass near the mean point $(\bar{a}, \bar{y})$,
so the estimated mean at $\bar{a}$ is relatively stable across different samples.
Moving away from $\bar{a}$,
small changes in the estimated slope
cause the fitted line to "pivot" around that anchor,
producing larger changes in the predicted mean
the further $a$ is from $\bar{a}$.

Additionally, there is typically more nearby data near the center of the covariate range,
so we have more information about the true mean response there.
Near the edges of the covariate range,
there is less nearby data,
leaving us less confident in our estimate of the mean response at those values.

## Prediction intervals {#sec-se-pred}

We can also construct **prediction intervals** for 
the value of a new observation $Y^*$,
given a covariate pattern $\vxs$.

A new observation $Y^*$ at $\vxs$ follows the same linear model as the training data:

$$Y^* = \tp{(\vxs)}\vb + \eps^*$$

where $\eps^* \sim \ndist{0, \sigma^2}$ is independent of the training data
(and therefore independent of $\hb$).

The predicted value of $Y^*$ is:

$$\hat{Y}^* \eqdef \hat\mu(\vxs) = \tp{(\vxs)}\hb$$

We base the prediction interval on the **prediction error** $Y^* - \hat{Y}^*$:

$$
\ba
Y^* - \hat{Y}^*
&= \tp{(\vxs)}\vb + \eps^* - \tp{(\vxs)}\hb
\\
&= \eps^* - \tp{(\vxs)}(\hb - \vb)
\ea
$$

The prediction error is unbiased:

$$
\ba
\E{Y^* - \hat{Y}^*}
&= \E{\eps^*} - \tp{(\vxs)}\E{\hb - \vb}
\\
&= 0 - \tp{(\vxs)} \cdot 0
\\
&= 0
\ea
$$

The variance of the prediction error is:

$$
\ba
\Var{Y^* - \hat{Y}^*}
&= \Var{\eps^* - \tp{(\vxs)}(\hb - \vb)}
\\
&= \Var{\eps^*} + \Var{\tp{(\vxs)}\hb}
\quad \text{(since } \eps^* \ind \hb\text{)}
\\
&= \sigma^2 + \tp{(\vxs)}\Var{\hb}\vxs
\\
&= \sigma^2 + \tp{(\vxs)}\paren{\sigma^2(\mX'\mX)^{-1}}\vxs
\\
&= \sigma^2 + \sigma^2\tp{(\vxs)}(\mX'\mX)^{-1}\vxs
\\
&= \sigma^2\paren{1 + \tp{(\vxs)}(\mX'\mX)^{-1}\vxs}
\ea
$${#eq-prediction-var}

::: notes
See @hoggtanis2015 §7.6 (p. 340).
:::

The variance (@eq-prediction-var) has two components:

- $\Var{\eps^*} = \sigma^2$:
the variance of the future observation's own noise,
which is irreducible.

- $\Var{\tp{(\vxs)}\hb} = \tp{(\vxs)}\Var{\hb}\vxs$:
the variance due to estimating the mean $\tp{(\vxs)}\vb$ from the data.

The first component is not present in the confidence interval for $\hat\mu(\vxs)$,
which only accounts for estimation uncertainty.
This is why prediction intervals are always wider than the corresponding confidence intervals:
a prediction interval must additionally account for
the irreducible noise $\eps^*$ in a new observation.

The **standard error** of the prediction error is the square root of its variance:

$$
\SE{Y^* - \hat{Y}^*}
= \sqrt{\Var{Y^* - \hat{Y}^*}}
= \sigma\sqrt{1 + \tp{(\vxs)}(\mX'\mX)^{-1}\vxs}
$$

Replacing $\sigma$ with the residual standard error $\hs$ gives the estimated standard error:

$$
\HSE{Y^* - \hat{Y}^*}
= \hs\sqrt{1 + \tp{(\vxs)}(\mX'\mX)^{-1}\vxs}
$${#eq-prediction-se}

Since $\eps^* \sim \ndist{0, \sigma^2}$ and $\hb \sim \ndist{\vb, \sigma^2(\mX'\mX)^{-1}}$
are both normally distributed,
and since $\eps^* \ind \hb$ (as noted above in the variance derivation),
their difference $Y^* - \hat{Y}^* = \eps^* - \tp{(\vxs)}(\hb - \vb)$
is also normally distributed,
with mean 0 (as shown above) and variance given by @eq-prediction-var.
Standardizing by dividing by $\sigma\sqrt{1 + \tp{(\vxs)}(\mX'\mX)^{-1}\vxs}$
yields a standard normal random variable.
Replacing $\sigma$ with $\hs$,
which is estimated with $n-p$ degrees of freedom and is independent of $Y^* - \hat{Y}^*$,
the standardized prediction error follows a $t$ distribution:

$$
\frac{Y^* - \hat{Y}^*}{\HSE{Y^* - \hat{Y}^*}} \sim t_{n-p}
$$

Therefore, a $(1-\alpha)$ **prediction interval** for $Y^*$ is:

$$
Y^* \in
\paren{
\hat{Y}^*
\pm
t_{n-p}\paren{1 - \frac{\alpha}{2}}
\cdot
\HSE{Y^* - \hat{Y}^*}
}
$${#eq-prediction-interval}

```{r}
#| code-fold: show
bw_lm2 |>
  predict(newdata = x, interval = "predict")
```

---

If you don't specify `newdata`, you get a warning:

```{r}
#| code-fold: show
#| warning: true
bw_lm2 |>
  predict(interval = "predict") |>
  head()
```

The warning from the last command is: "predictions on current data refer to *future* responses" (since you already know what happened to the current data, and thus don't need to predict it).

See `?predict.lm` for more.

---

```{r}
library(rlang) # defines the `.data` pronoun
plot_PIs_and_CIs <- function(data, model = NULL) {
  if (is.null(model)) {
    rlang::abort("`model` must be supplied to `plot_PIs_and_CIs()`.")
  }

  if (!is.data.frame(data) && is.data.frame(model)) {
    rlang::warn(
      paste0(
        "`plot_PIs_and_CIs()` received `model` as the first argument ",
        "and `data` as the second. ",
        "Please update the call to use `data` first and `model` second."
      )
    )
    tmp <- data
    data <- model
    model <- tmp
  }

  cis <- model |>
    predict(newdata = data, interval = "confidence") |>
    suppressWarnings() |>
    tibble::as_tibble()
  names(cis) <- paste("ci", names(cis), sep = "_")

  preds <- model |>
    predict(newdata = data, interval = "predict") |>
    suppressWarnings() |>
    tibble::as_tibble()
  names(preds) <- paste("pred", names(preds), sep = "_")
  dplyr::bind_cols(data, cis, preds) |>
    ggplot2::ggplot() +
    ggplot2::aes(
      x = .data$age,
      y = .data$weight,
      col = .data$sex
    ) +
    ggplot2::geom_point() +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::geom_line(ggplot2::aes(y = .data$ci_fit, group = .data$sex)) +
    ggplot2::geom_ribbon(
      ggplot2::aes(
        ymin = .data$pred_lwr,
        ymax = .data$pred_upr,
        group = .data$sex
      ),
      alpha = 0.2
    ) +
    ggplot2::geom_ribbon(
      ggplot2::aes(
        ymin = .data$ci_lwr,
        ymax = .data$ci_upr,
        group = .data$sex
      ),
      alpha = 0.5
    ) +
    ggplot2::facet_wrap(~sex)
}
```

:::{#fig-pred-int-bw}
```{r}
plot_PIs_and_CIs(bw, bw_lm2)
```

Confidence and prediction intervals for `birthweight` model [-@eq-BW-lm-interact]

:::

