By [the central limit theorem for MLEs](intro-MLEs.qmd#thm-dist-mle),
the maximum likelihood estimates $\hat\beta_k$ are approximately Gaussian
for large sample sizes;
see also the [table of Gaussian vs. MLE-based tests](intro-MLEs.qmd#tbl-gaussian-vs-mle-tests).

#### Wald test statistic

To test $H_0: \beta_k = \beta_{k,0}$ (typically $\beta_{k,0} = 0$):

$$z_k = \frac{\hat\beta_k - \beta_{k,0}}{\widehat{SE}(\hat\beta_k)}$$

Under $H_0$, $z_k \overset{\cdot}{\sim} N(0,1)$ for large $n$.

#### Confidence intervals for regression coefficients

A 95% confidence interval for $\beta_k$ is:

$$\hat\beta_k \pm 1.96 \cdot \widehat{SE}(\hat\beta_k)$$

#### Confidence intervals for odds ratios (exponentiated coefficients)

By the invariance property of MLEs,
the MLE of $e^{\beta_k}$ is $e^{\hat\beta_k}$.

A 95% confidence interval for $e^{\beta_k}$ is obtained by
exponentiating the endpoints of the CI for $\beta_k$:

$$e^{\beta_k} \in \left(e^{\hat\beta_k - 1.96 \cdot \widehat{SE}(\hat\beta_k)},\; e^{\hat\beta_k + 1.96 \cdot \widehat{SE}(\hat\beta_k)}\right)$$

#### In R

In R, `parameters()` from the `parameters` package automatically computes
Wald tests and confidence intervals for logistic regression model coefficients:

```{r}
#| label: tbl-beetles-wald-tests
#| tbl-cap: "Wald tests and 95% CIs for `beetles` logistic regression"
beetles_glm_ungrouped |>
  parameters() |>
  print_md()
```

Pass `exponentiate = TRUE` to `parameters()` to report results
on the odds ratio scale:

```{r}
#| label: tbl-beetles-wald-tests-OR
#| tbl-cap: "Odds ratios and 95% CIs for `beetles` logistic regression"
beetles_glm_ungrouped |>
  parameters(exponentiate = TRUE) |>
  print_md()
```
