### Fitting and manipulating logistic regression models in R

:::{#tbl-beetles-model-grouped}

```{r}
#| label: beetles-model-grouped-
beetles_glm_grouped <-
  beetles |>
  glm(
    formula = cbind(died, survived) ~ dose,
    family = "binomial"
  )

library(parameters)
beetles_glm_grouped |>
  parameters() |>
  print_md()
```

logistic regression model for beetles data with grouped (binomial) data

:::

---

Fitted values:

```{r}
#| label: beetles-model-fitted-values
fitted.values(beetles_glm_grouped)
predict(beetles_glm_grouped, type = "response")
predict(beetles_glm_grouped, type = "link")

fit_y <- beetles$n * fitted.values(beetles_glm_grouped)
```

---

### Individual observations

```{r}
#| tbl-cap: "`beetles` data in long format"
#| label: tbl-beetles-long

beetles_long
```

---

:::{#tbl-beetles-model-ungrouped}

```{r}
#| label: beetles-ungrouped-param-ests
beetles_glm_ungrouped <-
  beetles_long |>
  glm(
    formula = outcome ~ dose,
    family = "binomial"
  )

beetles_glm_ungrouped |>
  parameters() |>
  print_md()
```

logistic regression model for beetles data with individual Bernoulli data

:::

:::{#exr-compare-grouped-ungrouped}
Compare this model with the grouped-observations model (@tbl-beetles-model-grouped).
:::

---

:::{#sol-compare-grouped-ungrouped}

::: notes
They seem the same! But not quite:
:::

```{r}
#| code-fold: show
logLik(beetles_glm_grouped)
logLik(beetles_glm_ungrouped)
```

::: notes
The difference is due to the binomial coefficient
$\left(n\atop x \right)$ which isn't included in the
individual-observations (Bernoulli) version of the model.
:::

:::
