## Gaussian Deviance vs. GLM Deviance {#sec-compare-lm-glm-deviance}

The residual deviance is a general concept that applies to all GLMs,
including Gaussian linear regression.
For any GLM, the residual deviance is:

$$D = 2(\ell_{\text{sat}} - \ell(\hat\beta))$$ {#eq-deviance-general}

where $\ell_{\text{sat}}$ is the log-likelihood of the saturated model
and $\ell(\hat\beta)$ is the log-likelihood of the fitted model.
However, this formula simplifies differently depending on the distribution family,
and the resulting test statistics have different null distributions.

### Gaussian linear regression deviance

From @eq-linreg-loglik, the Gaussian log-likelihood is:

$$
\ell(\hat\beta, \hat\sigma^2)
= -\frac{n}{2}\log(2\pi\hat\sigma^2)
- \frac{1}{2\hat\sigma^2}
  \sumin (y_i - \hat{y}_i)^2
$$

The saturated model sets $\hat\mu_i = y_i$ for each observation
(one free parameter per observation),
so its residual sum of squares is zero:

$$\ell_{\text{sat}}(\hat\sigma^2) = -\frac{n}{2}\log(2\pi\hat\sigma^2)$$

Substituting into @eq-deviance-general, the Gaussian deviance simplifies to
the **residual sum of squares** (RSS), scaled by $\hat\sigma^2$:

$$
D
= 2(\ell_{\text{sat}} - \ell(\hat\beta))
= \frac{\sumin (y_i - \hat{y}_i)^2}{\hat\sigma^2}
= \frac{\text{RSS}}{\hat\sigma^2}
$$

Because $\sigma^2$ is unknown and must be estimated,
comparing deviances between two nested Gaussian models uses an **F-test**,
which is exact under Gaussian assumptions:

$$
F = \frac{(D_1 - D_2) / (p_2 - p_1)}{D_2 / (n - p_2)} \dist F_{p_2 - p_1,\; n - p_2}
$$

In R, `deviance()` applied to an `lm` object returns **RSS**
(the unscaled deviance):

```{r}
deviance(bw_lm2)
sum(residuals(bw_lm2)^2)
```

::: notes
The two expressions above return the same value.
:::

The Gaussian linear regression model can equivalently be fit as a GLM
with `family = gaussian`, which also returns RSS as the deviance:

```{r}
bw_glm2 <- glm(
  weight ~ sex + age + sex:age,
  data = bw,
  family = gaussian
)

deviance(bw_lm2)
deviance(bw_glm2)
```

::: notes
`deviance(bw_lm2)` and `deviance(bw_glm2)` return the same value,
confirming that Gaussian linear regression is a special case of the GLM framework
where the deviance equals RSS.
:::

### Non-Gaussian GLM deviance

For non-Gaussian GLMs,
the deviance does **not** simplify to RSS.
The formula depends on the distribution family:

**Poisson:**

$$
D = 2\sumin
\left[y_i \log\!\left(\frac{y_i}{\hat\mu_i}\right) - (y_i - \hat\mu_i)\right]
$$

**Binomial:**

$$
D = 2\sumin
\left[y_i \log\!\left(\frac{y_i}{\hat\pi_i}\right)
+ (n_i - y_i)\log\!\left(\frac{n_i - y_i}{n_i - \hat\pi_i}\right)\right]
$$

For Poisson and Binomial models, the dispersion parameter $\phi = 1$ is **fixed**
rather than estimated.
Consequently, the difference in deviances between two nested models
follows an approximate $\chi^2$ distribution:

$$D_1 - D_2 \;\dot \sim\; \chi^2_{p_2 - p_1}$$

This asymptotic result replaces the exact F-test used for Gaussian models.

### Summary

| Feature | Gaussian linear regression | Non-Gaussian GLMs (e.g., Poisson, Binomial) |
|---|---|---|
| Deviance formula | $\text{RSS}/\hat\sigma^2$ | $2(\ell_{\text{sat}} - \ell(\hat\beta))$ |
| `deviance()` in R | Returns RSS | Returns $2(\ell_{\text{sat}} - \ell(\hat\beta))$ |
| Dispersion $\phi$ | Unknown; estimated as $s^2 = \text{RSS}/(n-p)$ | Fixed ($\phi = 1$) |
| Test for nested models | F-test (exact) | $\chi^2$-test (asymptotic) |
