
(adapted from @dobson4e ┬¦6.3.3; for more information on prediction, see @james2013introduction and @rms2e).

::: notes
If we have a lot of covariates in our dataset, we might want to choose a small subset to use in our model.

There are a few possible metrics to consider for choosing a "best" model.
:::

### Mean squared error

We might want to minimize the **mean squared error**, $\text E[(y-\hat y)^2]$, for new observations that weren't in our data set when we fit the model.

Unfortunately, $$\frac{1}{n}\sum_{i=1}^n (y_i-\hat y_i)^2$$ gives a biased estimate of $\text E[(y-\hat y)^2]$ for new data. If we want an unbiased estimate, we will have to be clever.

---

#### Cross-validation

```{r}
data("carbohydrate", package = "dobson")
library(cvTools)
full_model <- lm(carbohydrate ~ ., data = carbohydrate)
cv_full <-
  full_model |> cvFit(
    data = carbohydrate, K = 5, R = 10,
    y = carbohydrate$carbohydrate
  )

reduced_model <- full_model |> update(formula = ~ . - age)

cv_reduced <-
  reduced_model |> cvFit(
    data = carbohydrate, K = 5, R = 10,
    y = carbohydrate$carbohydrate
  )
```

---

```{r}
results_reduced <-
  tibble(
    model = "wgt+protein",
    errs = cv_reduced$reps[]
  )
results_full <-
  tibble(
    model = "wgt+age+protein",
    errs = cv_full$reps[]
  )

cv_results <-
  bind_rows(results_reduced, results_full)

cv_results |>
  ggplot(aes(y = model, x = errs)) +
  geom_boxplot()
```

---

##### comparing metrics

```{r}
compare_results <- tribble(
  ~model, ~cvRMSE, ~r.squared, ~adj.r.squared, ~trainRMSE, ~loglik,
  "full",
  cv_full$cv,
  summary(full_model)$r.squared,
  summary(full_model)$adj.r.squared,
  sigma(full_model),
  logLik(full_model) |> as.numeric(),
  "reduced",
  cv_reduced$cv,
  summary(reduced_model)$r.squared,
  summary(reduced_model)$adj.r.squared,
  sigma(reduced_model),
  logLik(reduced_model) |> as.numeric()
)

compare_results
```

---

```{r}
anova(full_model, reduced_model)
```

---

#### stepwise regression

::: {.callout-warning}
## Caution about stepwise selection

Stepwise regression has several known problems:

- It tends to select too many variables (overfitting)
- P-values and confidence intervals are biased after selection
- It ignores model uncertainty
- Results can be unstable across different samples

Consider using cross-validation, penalized methods (like Lasso), 
or subject-matter knowledge instead.
See @rms2e and @heinze2018variable for more discussion.
:::

```{r}
library(olsrr)
olsrr:::ols_step_both_aic(full_model)
```

---

#### Lasso

$$\am_{\theta} \cb{\llik(\th) - \lambda \sum_{j=1}^p|\beta_j|}$$

```{r}
library(glmnet)
y <- carbohydrate$carbohydrate
x <- carbohydrate |>
  select(age, weight, protein) |>
  as.matrix()
fit <- glmnet(x, y)
```

---

```{r}
#| fig-cap: "Lasso selection"
#| label: fig-carbs-lasso
autoplot(fit, xvar = "lambda")
```

---

```{r}
cvfit <- cv.glmnet(x, y)
plot(cvfit)
```

---

```{r}
coef(cvfit, s = "lambda.1se")
```


