---
title: "posterior credible interval"
output: pdf_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}

library(dplyr)
library(ggplot2)
library(pander)
bayes = tibble(
  pi = seq(0, 1, by = 0.1),
  `P(Pi=pi)` = 1/length(pi)
) 

p_x_given_pi = function(pi, x = 8, n = 10)
{
  # translate your mathematical formula here
  # you can use the `choose()` function for the binomial constant
  choose(n,x) * (pi^x) * ((1-pi)^(n-x))
  # it's ok to use `dbinom()` if you prefer
  # dbinom(x, n, pi)
}


x = 8
n = 10

bayes = 
  bayes |>
  mutate(
    `P(X|pi)` = p_x_given_pi(pi=pi)
  ) 

p_x_given_pi_plot = ggplot(bayes, aes(x = pi, y = `P(X|pi)`)) + 
  geom_col() +
  theme_bw()

bayes = 
  bayes |> 
  mutate(
    `P(x,pi)` = `P(Pi=pi)` * `P(X|pi)`
  )

bayes = 
  bayes |> 
  mutate(
    `P(X=8)` = sum(`P(x,pi)`),
    `P(Pi=pi|X=8)` =  `P(x,pi)` / `P(X=8)`
  )

```

If everything went well up through 2.5.1, you should have the following posterior distribution (note that I've deleted some of the intermediate columns that should be in this tibble):

```{r}

bayes = 
  bayes |> 
  select(
    pi, 
    `P(Pi=pi|X=8)`
  )

bayes |> pander()

```

Now, let's add two more columns:

```{r}
bayes = 
  bayes |> 
  mutate(
    `P(Pi<=pi|X=8)` = cumsum(`P(Pi=pi|X=8)`), # `cumsum()` gives cumulative sums
    `P(Pi>pi|X=8)` = 1 - `P(Pi<=pi|X=8)` # because P(A) + P(!A) = 1
  )

pander(bayes)

```

By inspecting this table, we can see that the largest $\pi$ for which $P(\Pi > \pi |X=8) > 0.95$ is $\pi = 0.4$.
So we can summarize our findings as a "posterior interval", similar to a confidence interval, and say that 
$P(\Pi \in (0.4, 1]|X=8) > 0.95$. This is the answer we are looking for in 2.5.4.

If you want to extract the answer $\pi = 0.4$ programmatically, you can do it like this:

```{r}
 
library(conflicted) # this package checks for conflicting definitions of functions
# from different packages. It will alert us that `filter()` has two definitions; 
# one from `stats` and one from `dplyr`; we want the one from dplyr:
conflicts_prefer(dplyr::filter)

max.pi = 
  bayes |>
  filter(`P(Pi>pi|X=8)` >= 0.95) |> # this step gives us all values of $\pi$ that 
  # have this condition
  arrange(pi) |>  # `arrange()` re-orders a tibble() using the provided columns 
  # (in ascending order, by default)
  slice_tail() # get the last row (i.e., the largest pi in our subset, since we 
  # just sorted on pi)

pander(max.pi)
```

