---
title: "Predictor Selection"
subtitle: "Choosing which variables to include in a regression model"
format:
  html: default
  revealjs:
    output-file: predictor-selection-slides.html
  pdf:
    output-file: predictor-selection-handout.pdf
---

{{< include shared-config.qmd >}}

# Acknowledgements {.unnumbered}

This content is adapted from:

-   @vittinghoff2e, Chapter 10
-   @heinze2018variable

# Introduction {#sec-pred-sel-intro}

{{< include _sec-pred-sel-overview.qmd >}}

# Causal Thinking and DAGs {#sec-pred-sel-causal}

{{< include _sec-pred-sel-dag.qmd >}}

# Testing-based selection {#sec-pred-sel-stepwise}

::: notes

Testing-based selection methods use statistical tests
(such as F-tests or likelihood ratio tests)
to decide which predictors to include in a model.

:::

## Stepwise regression

::: {.callout-warning}
# Caution about stepwise selection

Stepwise regression has well-known problems:

- It tends to select too many variables (overfitting).
- P-values and confidence intervals are biased after selection.
- It ignores model uncertainty.
- Results can be unstable across different samples.

Consider using subject-matter knowledge (DAGs),
cross-validation, or penalized methods (such as LASSO) instead.
See @heinze2018variable for a thorough review.
:::

The basic idea of stepwise selection:

- **Forward selection**: Start with no predictors;
add the predictor that most improves model fit at each step.
- **Backward elimination**: Start with all predictors;
remove the predictor that least worsens model fit at each step.
- **Bidirectional stepwise**: Combine forward and backward steps.

## Example: stepwise selection for LDL

```{r}
#| label: stepwise-ldl

# Full model with all candidate predictors
full_model_ldl <- lm(
  LDL ~ HT + age + BMI + diabetes + smoking + statins + SBP,
  data = hers_ldl
)

# Backward elimination using AIC
step_model_ldl <- step(
  full_model_ldl,
  direction = "backward",
  trace = 0
)

summary(step_model_ldl)
```

# Criterion-based selection {#sec-pred-sel-aic-bic}

## AIC and BIC

When comparing non-nested models,
we can use information criteria:

- **AIC** (Akaike Information Criterion):
$$\text{AIC} = -2\,\hat\ell + 2p$$
- **BIC** (Bayesian Information Criterion):
$$\text{BIC} = -2\,\hat\ell + p\log n$$

where $\hat\ell$ is the maximized log-likelihood,
$p$ is the number of model parameters,
and $n$ is the sample size.

Lower values of AIC or BIC indicate a better model.
BIC applies a larger penalty for model complexity,
especially for large $n$.

## Comparing models for LDL

```{r}
#| label: tbl-model-comparison
#| tbl-cap: "Comparison of LDL models by AIC and BIC"

library(tibble)

models_ldl <- list(
  "HT only" = lm(LDL ~ HT, data = hers_ldl),
  "HT + age" = lm(LDL ~ HT + age, data = hers_ldl),
  "HT + statins" = lm(LDL ~ HT + statins, data = hers_ldl),
  "HT + age + BMI + diabetes + statins" = lm(
    LDL ~ HT + age + BMI + diabetes + statins,
    data = hers_ldl
  ),
  "Full model" = lm(
    LDL ~ HT + age + BMI + diabetes + smoking + statins + SBP,
    data = hers_ldl
  )
)

tibble(
  Model = names(models_ldl),
  AIC = sapply(models_ldl, AIC),
  BIC = sapply(models_ldl, BIC),
  Adj_R2 = sapply(
    models_ldl,
    function(m) summary(m)$adj.r.squared
  )
)
```

# Penalized regression {#sec-pred-sel-lasso}

::: notes

When there are many candidate predictors,
penalized regression methods simultaneously
estimate model coefficients and perform variable selection
by shrinking small coefficients toward zero.

:::

## The LASSO

The **LASSO** (Least Absolute Shrinkage and Selection Operator)
adds an $\ell_1$ penalty to the log-likelihood:

$$\hat\beta_\lambda = \arg\max_\beta \left\{
\hat\ell(\beta) - \lambda \sum_{j=1}^p |\beta_j|
\right\}$$

The tuning parameter $\lambda$ controls the strength of the penalty:

- $\lambda = 0$: no penalty (ordinary least squares).
- $\lambda \to \infty$: all coefficients shrink to zero.

For intermediate values of $\lambda$,
some coefficients are exactly zero (excluded),
while others are shrunk but non-zero.

The optimal $\lambda$ is usually chosen by cross-validation.

## Example: LASSO for LDL

```{r}
#| label: fig-lasso-ldl
#| fig-cap: "LASSO coefficient paths for LDL predictors in HERS"

library(glmnet)
library(ggfortify)

hers_ldl_complete <-
  hers_ldl |>
  select(LDL, HT, age, BMI, diabetes, smoking, statins, SBP) |>
  tidyr::drop_na()

y_ldl <- hers_ldl_complete$LDL
x_ldl <-
  hers_ldl_complete |>
  select(-LDL) |>
  mutate(
    HT = as.integer(HT == "HT"),
    statins = as.integer(statins == "Yes"),
    diabetes = as.integer(diabetes == "Yes"),
    smoking = as.integer(smoking == "Yes")
  ) |>
  as.matrix()

lasso_fit <- glmnet(x_ldl, y_ldl)
autoplot(lasso_fit, xvar = "lambda") +
  theme_bw()
```

---

```{r}
#| label: fig-lasso-cv
#| fig-cap: "Cross-validated LASSO for LDL: MSE vs. log(lambda)"

cv_lasso <- cv.glmnet(x_ldl, y_ldl)
plot(cv_lasso)
```

---

```{r}
#| label: tbl-lasso-coefs
#| tbl-cap: "LASSO coefficients at two values of the regularization parameter:
#|   `lambda.min` minimizes cross-validation error;
#|   `lambda.1se` uses a more conservative (larger) lambda,
#|   within one standard error of the minimum,
#|   producing a sparser model."

data.frame(
  lambda.min = as.vector(coef(cv_lasso, s = "lambda.min")),
  lambda.1se = as.vector(coef(cv_lasso, s = "lambda.1se")),
  row.names = rownames(coef(cv_lasso))
)
```

::: notes

The LASSO identifies `statins` and `HT` as the strongest predictors
of LDL cholesterol in the HERS data,
consistent with the biological knowledge encoded in the DAG
in @fig-hers-dag.

:::

# Summary {#sec-pred-sel-summary}

Key principles for predictor selection:

1. **Start with subject-matter knowledge**.
Draw a DAG to identify confounders, mediators, and colliders.

2. **Adjust for confounders** to remove bias in causal effect estimates.

3. **Do not adjust for mediators** if you want the total causal effect.

4. **Do not adjust for colliders** to avoid introducing bias.

5. **Use AIC/BIC** to compare models of similar types.

6. **Be cautious with stepwise selection**.
It tends to overfit and produces biased inference.

7. **Consider penalized regression** (LASSO, ridge)
when there are many candidate predictors.

8. **In randomized trials** (like HERS),
confounding adjustment is not required for causal inference,
but adjusting for strong predictors of the outcome
can improve precision.
