---
title: "Predictor Selection"
subtitle: "Choosing which variables to include in a regression model"
format:
  html: default
  revealjs:
    output-file: predictor-selection-slides.html
  pdf:
    output-file: predictor-selection-handout.pdf
---

{{< include shared-config.qmd >}}

# Acknowledgements {.unnumbered}

This content is adapted from:

-   @vittinghoff2e, Chapter 10
-   @heinze2018variable

# Introduction {#sec-pred-sel-intro}

{{< include _sec-pred-sel-overview.qmd >}}

# Inferential Goals {#sec-pred-sel-goals-chapter}

{{< include _sec-pred-sel-goals.qmd >}}

# Causal Thinking and DAGs {#sec-pred-sel-causal}

{{< include _sec-pred-sel-dag.qmd >}}

# Prediction {#sec-pred-sel-prediction}

{{< include _sec-pred-sel-prediction.qmd >}}

# Evaluating a Predictor of Primary Interest {#sec-pred-sel-eval}

{{< include _sec-pred-sel-primary.qmd >}}

# Identifying Multiple Important Predictors {#sec-pred-sel-multiple-chapter}

{{< include _sec-pred-sel-multiple.qmd >}}

# Testing-based selection {#sec-pred-sel-stepwise}

::: notes

Testing-based selection methods use statistical tests
(such as F-tests or likelihood ratio tests)
to decide which predictors to include in a model.

:::

## Stepwise regression

::: {.callout-warning}
# Caution about stepwise selection

Stepwise regression has well-known problems:

- It tends to select too many variables (overfitting).
- P-values and confidence intervals are biased after selection.
- It ignores model uncertainty.
- Results can be unstable across different samples.

Consider using subject-matter knowledge (DAGs),
cross-validation, or penalized methods (such as LASSO) instead.
See @heinze2018variable for a thorough review.
:::

The basic idea of stepwise selection:

- **Forward selection**: Start with no predictors;
add the predictor that most improves model fit at each step.
- **Backward elimination**: Start with all predictors;
remove the predictor that least worsens model fit at each step.
- **Bidirectional stepwise**: Combine forward and backward steps.

## Example: stepwise selection for LDL

```{r}
#| label: stepwise-ldl

# Full model with all candidate predictors
full_model_ldl <- lm(
  LDL ~ HT + age + BMI + diabetes + smoking + statins + SBP,
  data = hers_ldl
)

# Backward elimination using AIC
step_model_ldl <- step(
  full_model_ldl,
  direction = "backward",
  trace = 0
)

summary(step_model_ldl)
```

# Criterion-based selection {#sec-pred-sel-aic-bic}

## AIC and BIC

When comparing non-nested models,
we can use information criteria:

- **AIC** (Akaike Information Criterion):
$$\text{AIC} = -2\,\hat\ell + 2p$$
- **BIC** (Bayesian Information Criterion):
$$\text{BIC} = -2\,\hat\ell + p\log n$$

where $\hat\ell$ is the maximized log-likelihood,
$p$ is the number of model parameters,
and $n$ is the sample size.

Lower values of AIC or BIC indicate a better model.
BIC applies a larger penalty for model complexity,
especially for large $n$.

## Comparing models for LDL

::: {#tbl-model-comparison}

```{r}

library(tibble)

models_ldl <- list(
  "HT only" = lm(LDL ~ HT, data = hers_ldl),
  "HT + age" = lm(LDL ~ HT + age, data = hers_ldl),
  "HT + statins" = lm(LDL ~ HT + statins, data = hers_ldl),
  "HT + age + BMI + diabetes + statins" = lm(
    LDL ~ HT + age + BMI + diabetes + statins,
    data = hers_ldl
  ),
  "Full model" = lm(
    LDL ~ HT + age + BMI + diabetes + smoking + statins + SBP,
    data = hers_ldl
  )
)

tibble(
  Model = names(models_ldl),
  AIC = sapply(models_ldl, AIC),
  BIC = sapply(models_ldl, BIC),
  Adj_R2 = sapply(
    models_ldl,
    function(m) summary(m)$adj.r.squared
  )
)
```

Comparison of LDL models by AIC, BIC, and adjusted $R^2$.

:::

# Penalized regression {#sec-pred-sel-lasso}

::: notes

When there are many candidate predictors,
penalized regression methods simultaneously
estimate model coefficients and perform variable selection
by shrinking small coefficients toward zero.

:::

## The LASSO

The **LASSO** (Least Absolute Shrinkage and Selection Operator)
adds an $\ell_1$ penalty to the log-likelihood:

$$\hat\beta_\lambda = \arg\max_\beta \left\{
\hat\ell(\beta) - \lambda \sum_{j=1}^p |\beta_j|
\right\}$$

The tuning parameter $\lambda$ controls the strength of the penalty:

- $\lambda = 0$: no penalty (ordinary least squares).
- $\lambda \to \infty$: all coefficients shrink to zero.

For intermediate values of $\lambda$,
some coefficients are exactly zero (excluded),
while others are shrunk but non-zero.

The optimal $\lambda$ is usually chosen by cross-validation.

## Example: LASSO for LDL

::: {#fig-lasso-ldl}

```{r}

library(glmnet)
library(ggfortify)

hers_ldl_complete <-
  hers_ldl |>
  dplyr::select(LDL, HT, age, BMI, diabetes, smoking, statins, SBP) |>
  tidyr::drop_na()

y_ldl <- hers_ldl_complete$LDL
x_ldl <-
  hers_ldl_complete |>
  dplyr::select(-LDL) |>
  mutate(
    HT = as.integer(HT == "HT"),
    statins = as.integer(statins == "Yes"),
    diabetes = as.integer(diabetes == "Yes"),
    smoking = as.integer(smoking == "Yes")
  ) |>
  as.matrix()

lasso_fit <- glmnet(x_ldl, y_ldl)
autoplot(lasso_fit, xvar = "lambda") +
  theme_bw()
```

LASSO coefficient paths for LDL predictors in HERS.
Each line traces one predictor's coefficient as $\lambda$ increases.
Coefficients shrink toward zero as the penalty grows;
some reach exactly zero (variables excluded from the model).

:::


---

::: {#fig-lasso-cv}

```{r}

cv_lasso <- cv.glmnet(x_ldl, y_ldl)
plot(cv_lasso)
```

Cross-validated LASSO for LDL: mean squared error vs. $\log(\lambda)$.
The left dashed line marks `lambda.min` (minimum CV error);
the right dashed line marks `lambda.1se` (largest $\lambda$ within 1 SE of the minimum).

:::

---

::: {#tbl-lasso-coefs}

```{r}

data.frame(
  lambda.min = as.vector(coef(cv_lasso, s = "lambda.min")),
  lambda.1se = as.vector(coef(cv_lasso, s = "lambda.1se")),
  row.names = rownames(coef(cv_lasso))
)
```

LASSO coefficients at two values of the regularization parameter.
`lambda.min` minimizes cross-validation error;
`lambda.1se` uses a more conservative (larger) $\lambda$,
within one standard error of the minimum,
producing a sparser model.

:::

::: notes

The LASSO identifies `statins` and `HT` as the strongest predictors
of LDL cholesterol in the HERS data,
consistent with the biological knowledge encoded in the DAG
in @fig-hers-dag.

:::

# Some Details {#sec-pred-sel-details}

{{< include _sec-pred-sel-details.qmd >}}

# Summary {#sec-pred-sel-summary}

Predictor selection is the process of choosing appropriate variables
for inclusion in a multivariable regression model.
The right approach depends on the **inferential goal**:

**For prediction** (Goal 1):

- Pre-specify well-motivated candidate predictors.
- Use cross-validation of a target PE measure
  to select among models without overfitting.
- Shrinkage methods (LASSO, ridge) are effective
  when there are many candidates.

**For evaluating a predictor of primary interest** (Goal 2):

- Use a DAG to identify confounders, mediators, and colliders.
- Include all well-established confounders for face validity.
- Use backward selection with a liberal criterion ($p < 0.2$)
  to remove apparent non-confounders.
- In randomized trials, covariates are not required for confounding control
  but improve precision (and de-attenuate estimates in logistic/Cox models).
- Pre-specify adjusted analyses in trial protocols.

**For identifying multiple important predictors** (Goal 3):

- Apply the same inclusive strategy as Goal 2.
- Consider the Allen–Cady modified backward selection procedure
  to limit false-positive findings.
- Interpret novel, weak, and borderline significant associations
  cautiously; report all candidate predictors considered.

**Across all goals**:

- Collinearity between a primary predictor and a confounder
  can inflate standard errors; between adjustment variables, it does not.
- The EPV rule (10 events per predictor) flags potential problems
  but is not a hard limit.
- Model selection complicates inference:
  nominal p-values and confidence intervals are anticonservative
  when predictors have been selected from the data.

# Learning Objectives {#sec-pred-sel-objectives .unnumbered}

After completing this chapter, students should be able to:

1. **Identify the inferential goal** of a predictor selection problem
   (prediction, evaluating a primary predictor,
   or identifying multiple important predictors)
   and choose an appropriate strategy.

2. **Use a DAG** to identify confounders, mediators, and colliders,
   and determine which covariates to include or exclude.

3. **Apply backward selection with a liberal criterion**
   to select confounders in observational studies.

4. **Use AIC and BIC** to compare alternative models.

5. **Apply and interpret the LASSO**,
   including cross-validated selection of $\lambda$.

6. **Explain why stepwise selection is problematic**
   and describe its known pitfalls.

7. **Describe the bias–variance trade-off** and explain
   why cross-validation provides a less optimistic estimate of PE
   than naive within-sample measures.

8. **Identify and handle collinearity**,
   including using the VIF and understanding its implications
   for different inferential goals.

9. **Articulate how model selection complicates inference**
   and describe best practices for reporting in each inferential context.

