:::{#exm-marginal-risks}

#### Estimated marginal risks

Using the WCGS data with `dibpat` (behavior pattern) as the exposure
and `chd69` (coronary heart disease by 1969) as the outcome,
adjusted for `age`:

```{r}
#| label: wcgs-marginal-setup
#| code-fold: true
#| message: false
#| warning: false
library(dplyr)
library(haven)

wcgs_path <- here::here("Data", "wcgs.rda")
if (!file.exists(wcgs_path)) {
  stop(
    "Could not find WCGS data file at: ", wcgs_path,
    ". This file should be in the repository's Data directory. ",
    paste0(
      "Run this chapter from the repository root and verify ",
      "that Data/wcgs.rda is present."
    )
  )
}
wcgs_env <- new.env(parent = emptyenv())
invisible(load(wcgs_path, envir = wcgs_env))
wcgs <- wcgs_env$wcgs

wcgs_clean <- wcgs |>
  mutate(
    dibpat = dibpat |>
      haven::as_factor() |>
      relevel(ref = "Type B"),
    chd69_binary = as.numeric(chd69 == "Yes"),
    across(where(is.labelled), haven::as_factor)
  ) |>
  filter(!is.na(chd69_binary), !is.na(dibpat), !is.na(age))

fit <- glm(
  chd69_binary ~ dibpat + age,
  data = wcgs_clean,
  family = binomial(link = "logit")
)
```

The **observed marginal risks**
$\pi(a) = \Pr(Y=1 \mid A=a)$
can be estimated in two equivalent ways.

Directly, as the sample proportion of events within each exposure group:

```{r}
#| label: wcgs-obs-marginal-risks-direct
pi1_obs <- mean(wcgs_clean$chd69_binary[wcgs_clean$dibpat == "Type A"])
pi0_obs <- mean(wcgs_clean$chd69_binary[wcgs_clean$dibpat == "Type B"])
```

$$
\hat\pi(1) = `r round(pi1_obs, 3)`, \quad
\hat\pi(0) = `r round(pi0_obs, 3)`
$$

Or via **predictive margins** (@def-predictive-margins):
average the model's fitted risks over the **subgroup** actually exposed to $A = a$,
keeping each man's own covariates and observed exposure:

```{r}
#| label: wcgs-obs-marginal-risks-pm
fitted_risk <- predict(fit, type = "response")
pi1_obs_pm <- mean(fitted_risk[wcgs_clean$dibpat == "Type A"])
pi0_obs_pm <- mean(fitted_risk[wcgs_clean$dibpat == "Type B"])
```

$$
\hat\pi(1) = `r round(pi1_obs_pm, 3)`, \quad
\hat\pi(0) = `r round(pi0_obs_pm, 3)`
$$

The two estimates agree exactly (to displayed precision).
This is not a coincidence:
because the model includes `dibpat` as a covariate,
the maximum-likelihood score equations force
the mean fitted risk within each `dibpat` group
to equal that group's observed event proportion.
Predictive margins over the exposure subgroup therefore reproduce
the raw proportions here.

The **causal marginal risks**
$\hat\pi_a = \frac{1}{n}\sumin \hat\pi(a, z_i)$
standardized over the full covariate distribution
(@def-g-computation):

```{r}
#| label: wcgs-causal-marginal-risks
dibpat_levels <- levels(wcgs_clean$dibpat)
wcgs_counterfactual_a <- wcgs_clean |>
  mutate(
    dibpat = factor("Type A", levels = dibpat_levels)
  )
wcgs_counterfactual_b <- wcgs_clean |>
  mutate(
    dibpat = factor("Type B", levels = dibpat_levels)
  )

pi1_hat <- mean(
  predict(fit, newdata = wcgs_counterfactual_a, type = "response")
)
pi0_hat <- mean(
  predict(fit, newdata = wcgs_counterfactual_b, type = "response")
)
```

$$
\hat\pi_1 = `r round(pi1_hat, 3)`, \quad
\hat\pi_0 = `r round(pi0_hat, 3)`
$$

Note that $\hat\pi(1) \neq \hat\pi_1$ and $\hat\pi(0) \neq \hat\pi_0$:
because age is associated with behavioral pattern (Type A and Type B men
have different age distributions), the observed marginal and causal marginal
risks differ.
Standardization via g-computation adjusts for this difference.

:::
