:::{#exm-wcgs-observed-marginal-risks}

#### Estimated observed marginal risks (WCGS)

Continuing the WCGS example (@exm-wcgs-causal-marginal-risks),
the **observed marginal risks**
$\pi(a) = \Pr(Y=1 \mid A=a)$
can be estimated in two equivalent ways.

Directly, as the sample proportion of events within each exposure group:

```{r}
#| label: wcgs-obs-marginal-risks-direct
pi1_obs <- mean(wcgs_clean$chd69_binary[wcgs_clean$dibpat == "Type A"])
pi0_obs <- mean(wcgs_clean$chd69_binary[wcgs_clean$dibpat == "Type B"])
```

$$
\hat\pi(1) = `r round(pi1_obs, 3)`, \quad
\hat\pi(0) = `r round(pi0_obs, 3)`
$$

Or via **predictive margins** (@def-predictive-margins):
average the model's fitted risks over the **subgroup** actually exposed to $A = a$,
keeping each man's own covariates and observed exposure:

```{r}
#| label: wcgs-obs-marginal-risks-pm
fitted_risk <- predict(fit, type = "response")
pi1_obs_pm <- mean(fitted_risk[wcgs_clean$dibpat == "Type A"])
pi0_obs_pm <- mean(fitted_risk[wcgs_clean$dibpat == "Type B"])
```

$$
\hat\pi_1(1) = `r round(pi1_obs_pm, 3)`, \quad
\hat\pi_0(0) = `r round(pi0_obs_pm, 3)`
$$

The two estimates agree exactly (to displayed precision),
as @lem-subgroup-pm-proportion guarantees:
because the model includes `dibpat` as a covariate,
the exposed-subgroup predictive margin reproduces
the raw event proportion in each `dibpat` group.

Note that $\hat\pi(1) \neq \hat\pi_1$ and $\hat\pi(0) \neq \hat\pi_0$
(@exm-wcgs-causal-marginal-risks):
because age is associated with behavioral pattern (Type A and Type B men
have different age distributions), the observed marginal and causal marginal
risks differ.
Standardization via g-computation adjusts for this difference.

:::
