
If you want to report risk differences or risk ratios instead of odds ratios,
you can obtain estimates from logistic regression models,
as long as you didn't stratify sampling by the outcome;
in other words, not in case-control studies
(see @sec-CC-studies).

To compute risk ratios from logistic regression models:

- Apply the expit function to the linear predictor for each covariate pattern
to compute the (estimated) risks,
- Then take the differences or ratios of the risks, as needed.

To quantify uncertainty for risk difference or risk ratio estimates
derived from logistic regression models
(e.g., to calculate SEs, CIs, and p-values),
you will need to use the
bootstrap,
the multivariate delta method,
or some other special technique.

::: notes
Adapted from
[@vittinghoff2e, Section 3.6, p. 62],
[@hastie2009esl2e, Section 7.11, p. 249],
and
[@james2021islr2e, Chapter 5, Section 5.2, p. 209].
:::

#### Bootstrap workflow for derived risk measures

When using nonparametric bootstrap inference
for a derived quantity
like a marginal risk difference or risk ratio:

1. Fit the logistic model on the observed data.
2. Compute the target estimand
   (for example, a g-computation risk difference or risk ratio).
3. Repeat for
   $b = 1, \ldots, B$:
   draw a bootstrap sample of size $n$
   with replacement,
   refit the model,
   and recompute the target estimand.
4. Use the empirical standard deviation of the
   $B$ bootstrap estimates
   as the bootstrap standard error.
5. Build a confidence interval from the bootstrap distribution
   (percentile or BCa intervals are common when the bootstrap distribution is skewed).

In practice,
larger $B$ is usually needed for interval estimation than for standard errors;
for percentile or BCa intervals,
it is common to use at least 1,000 resamples,
and often 2,000 or more for stability.

::: {#exm-wcgs-bootstrap-rd}

#### Bootstrap CI for marginal risk difference and risk ratio (WCGS)

We estimate the marginal risk difference and risk ratio
comparing Type A vs. Type B personality patterns in the WCGS study,
using g-computation to derive the estimands
and the nonparametric bootstrap to quantify uncertainty.

**Steps 1–2: Point estimates via g-computation**

```{r}
#| label: exm-boot-gcomp-point
#| code-fold: true
# Fit logistic model: CHD ~ behavioral pattern + age
fit_wcgs <- glm(
  chd69 ~ dibpat + age,
  family = binomial,
  data = wcgs
)

# G-computation: set everyone to Type A, then to Type B,
# and average the predicted risks
dibpat_levels <- levels(wcgs$dibpat)  # c("Type B", "Type A")

newdata_A <- wcgs |> mutate(dibpat = factor("Type A", levels = dibpat_levels))
newdata_B <- wcgs |> mutate(dibpat = factor("Type B", levels = dibpat_levels))

risk_A <- mean(predict(fit_wcgs, newdata = newdata_A, type = "response"))
risk_B <- mean(predict(fit_wcgs, newdata = newdata_B, type = "response"))

tibble::tibble(
  measure   = c("Risk (Type A)", "Risk (Type B)", "Risk difference", "Risk ratio"),
  estimate  = c(risk_A, risk_B, risk_A - risk_B, risk_A / risk_B)
)
```

**Steps 3–5: Bootstrap SE and BCa CIs**

```{r}
#| label: exm-boot-gcomp-ci
#| code-fold: true
library(boot)

boot_gcomp <- function(data, indices) {
  d <- data[indices, ]
  fit <- glm(chd69 ~ dibpat + age, family = binomial, data = d)
  lev <- levels(d$dibpat)
  risk_A <- mean(
    predict(fit, newdata = mutate(d, dibpat = factor("Type A", levels = lev)),
            type = "response")
  )
  risk_B <- mean(
    predict(fit, newdata = mutate(d, dibpat = factor("Type B", levels = lev)),
            type = "response")
  )
  c(RD = risk_A - risk_B, RR = risk_A / risk_B)
}

set.seed(42)
boot_wcgs <- boot(data = wcgs, statistic = boot_gcomp, R = 1000)

boot_wcgs
```

```{r}
#| label: exm-boot-gcomp-ci-rd
#| code-fold: true
# BCa CI for risk difference (index = 1)
boot.ci(boot_wcgs, type = "bca", index = 1)
```

```{r}
#| label: exm-boot-gcomp-ci-rr
#| code-fold: true
# BCa CI for risk ratio (index = 2)
boot.ci(boot_wcgs, type = "bca", index = 2)
```

The bootstrap standard errors and BCa intervals
provide valid inference for both the risk difference and risk ratio
without requiring any closed-form delta-method derivation.

:::
