Contrasting marginal risks by ratio rather than by difference
gives the corresponding marginal risk ratios:

{{< slidebreak >}}

:::{#def-observed-marginal-rr}

#### Observed marginal risk ratio

The **observed marginal risk ratio** (**OMRR**) is:

$$
\text{OMRR} \eqdef \frac{\pi(1)}{\pi(0)}
$$

:::

{{< slidebreak >}}

:::{#exm-observed-marginal-rr}

#### Estimated observed marginal risk ratio

Continuing from @exm-wcgs-observed-marginal-risks,
the observed marginal risk ratio estimate is:

```{r}
#| label: wcgs-observed-marginal-rr
rr_obs <- pi1_obs / pi0_obs
```

$$
\ba
\widehat{\text{OMRR}}
&= \frac{\hat\pi(1)}{\hat\pi(0)}
&& \text{[definition of OMRR]}
\\
&= \frac{`r round(pi1_obs, 3)`}{`r round(pi0_obs, 3)`}
&& \text{[substitute estimates]}
\\
&= `r round(rr_obs, 2)`
&& \text{[arithmetic, computed from full-precision } \hat\pi(1) / \hat\pi(0) \text{]}
\ea
$$

:::

{{< slidebreak >}}

:::{#def-causal-marginal-rr}

#### Causal marginal risk ratio

The **causal marginal risk ratio** (**CMRR**) is:

$$
\text{CMRR} \eqdef \frac{\pi_1}{\pi_0}
$$

:::

{{< slidebreak >}}

:::{#exm-causal-marginal-rr}

#### Estimated causal marginal risk ratio

Continuing from @exm-wcgs-causal-marginal-risks,
the causal marginal risk ratio estimate is:

```{r}
#| label: wcgs-causal-marginal-rr
rr_hat <- pi1_hat / pi0_hat
```

$$
\ba
\widehat{\text{CMRR}}
&= \frac{\hat\pi_1}{\hat\pi_0}
&& \text{[definition of CMRR]}
\\
&= \frac{`r round(pi1_hat, 3)`}{`r round(pi0_hat, 3)`}
&& \text{[substitute estimates]}
\\
&= `r round(rr_hat, 2)`
&& \text{[arithmetic, computed from full-precision } \hat\pi_1 / \hat\pi_0 \text{]}
\ea
$$

:::

{{< slidebreak >}}

The point estimates in @exm-observed-marginal-rr and @exm-causal-marginal-rr
use the same fitted model as @exm-causal-marginal-rd.
To quantify their uncertainty, we bootstrap the marginal risk ratio,
modifying the statistic function to compute ratios instead of differences:

```{r}
#| label: marginal-rr-example
#| code-fold: show
#| eval: false

library(boot)

compute_marginal_rr <- function(model,
                                data,
                                exposure_var,
                                exposed_level,
                                unexposed_level) {
  counterfactual_data <- make_counterfactual_datasets(
    data = data,
    exposure_var = exposure_var,
    exposed_level = exposed_level,
    unexposed_level = unexposed_level
  )
  data_exposed <- counterfactual_data$data_exposed
  data_unexposed <- counterfactual_data$data_unexposed

  risk_exposed <- predict(
    model,
    newdata = data_exposed,
    type = "response"
  )

  risk_unexposed <- predict(
    model,
    newdata = data_unexposed,
    type = "response"
  )

  mean(risk_exposed) / mean(risk_unexposed)
}

statistic_fn_rr <- function(data, indices) {
  boot_data <- data[indices, ]
  boot_fit <- glm(
    chd69_binary ~ dibpat + age,
    data = boot_data,
    family = binomial(link = "logit")
  )
  compute_marginal_rr(
    model = boot_fit,
    data = boot_data,
    exposure_var = "dibpat",
    exposed_level = "Type A",
    unexposed_level = "Type B"
  )
}

set.seed(20260512)
boot_results_rr <- boot(
  data = wcgs_clean,
  statistic = statistic_fn_rr,
  # Keep the rendered example fast; increase to 2000+ for final analyses.
  R = 300
)

# BCa intervals can be unstable at this small R;
# increase to R = 2000+ before relying on bca results.
boot_ci_rr <- boot.ci(boot_results_rr, type = c("perc", "bca"))

cat("Marginal Risk Ratio (Type A vs Type B):",
    round(boot_results_rr$t0, 3), "\n")
cat("95% Percentile CI: (",
    round(boot_ci_rr$percent[4], 3), ",",
    round(boot_ci_rr$percent[5], 3), ")\n")
```
