
:::: notes
Odds ratios have a special property: we can swap a covariate with the outcome, and the odds ratio remains the same.
::::

:::{#thm-or-swap}
##### Odds ratios are reversible


For any two events $A$, $B$:

$$\ror(A|B) = \ror(B|A)$$

:::

{{< slidebreak >}}

::: {.proof}

{{< include prf-OR-reversible.qmd >}}

:::

{{< slidebreak >}}

:::{#exm-or-inv-MI}

#### Reversibility of the OC-MI odds ratio

@thm-or-swap says $\ror(MI|OC) = \ror(OC|MI)$:
we can swap the roles of the outcome (MI) and the covariate (OC use)
without changing the odds ratio.
We can confirm this equality numerically
using the counts from @exm-oc-mi.

```{r}
#| label: or-swap-numeric
#| include: false
n_MI_OC <- 13 # OC users with MI # nolint: object_name_linter
n_nMI_OC <- 4987 # OC users without MI # nolint: object_name_linter
n_MI_nOC <- 7 # OC non-users with MI # nolint: object_name_linter
n_nMI_nOC <- 9993 # OC non-users without MI # nolint: object_name_linter

# disease odds ratio: odds of MI, comparing OC users to non-users
odds_MI_OC <- n_MI_OC / n_nMI_OC # nolint: object_name_linter
odds_MI_nOC <- n_MI_nOC / n_nMI_nOC # nolint: object_name_linter
ror_MI_given_OC <- odds_MI_OC / odds_MI_nOC # nolint: object_name_linter

# exposure odds ratio: odds of OC use, comparing MI cases to non-cases
odds_OC_MI <- n_MI_OC / n_MI_nOC # nolint: object_name_linter
odds_OC_nMI <- n_nMI_OC / n_nMI_nOC # nolint: object_name_linter
ror_OC_given_MI <- odds_OC_MI / odds_OC_nMI # nolint: object_name_linter
```

The **disease** odds ratio
(the odds of MI, comparing OC users to non-users) is:

$$
\ba
\ror(MI|OC)
&\eqdef \frac{\odds(MI|OC)}{\odds(MI|\neg OC)}
&& \text{(@def-OR)}
\\ &\eqdef \frac{13/4987}{7/9993}
&& \text{(@def-c-odds, with counts from @exm-oc-mi)}
\\ &= \frac{`r odds_MI_OC`}{`r odds_MI_nOC`}
&& \text{(dividing within each OC-use group)}
\\ &= `r ror_MI_given_OC`
&& \text{(dividing)}
\ea
$$

The **exposure** odds ratio
(the odds of OC use, comparing MI cases to non-cases) is:

$$
\ba
\ror(OC|MI)
&\eqdef \frac{\odds(OC|MI)}{\odds(OC|\neg MI)}
&& \text{(@def-OR)}
\\ &\eqdef \frac{13/7}{4987/9993}
&& \text{(@def-c-odds, with counts from @exm-oc-mi)}
\\ &= \frac{`r odds_OC_MI`}{`r odds_OC_nMI`}
&& \text{(dividing within each MI-status group)}
\\ &= `r ror_OC_given_MI`
&& \text{(dividing)}
\ea
$$

Both equal `r ror_MI_given_OC`,
confirming @thm-or-swap.
:::

{{< slidebreak >}}

:::{#exr-2x2-generic}
For @tbl-2x2-generic, show that $\hror(Exposed, Unexposed) = \hror(Event, \neg Event)$.
:::

{{< slidebreak >}}

:::: notes
Conditional odds ratios have the same reversibility property:
::::

:::{#thm-conditional-OR-swap}
##### Conditional odds ratios are reversible


For any three events $A$, $B$, $C$:

$$\ror(A|B,C) = \ror(B|A,C)$$
:::

{{< slidebreak >}}

:::{.proof}
Apply the same steps as for @thm-or-swap,
inserting $C$ into the conditions (RHS of $|$) of every expression.
:::

{{< slidebreak >}}

:::{#exm-conditional-or-swap}

#### Reversibility within age strata in the WCGS data

@thm-conditional-OR-swap says a conditional odds ratio is also reversible:
$\ror(A|B,C) = \ror(B|A,C)$.
We can illustrate this equality using the `wcgs` data,
taking the covariate $B$ to be behavioral pattern (Type A versus Type B),
the outcome $A$ to be coronary heart disease (CHD) by 1969,
and the conditioning event $C$ to be an age stratum.

```{r}
#| label: cond-or-swap-data
#| include: false
library(here)
library(fs)
here::here() |>
  fs::path("Data/wcgs.rda") |>
  load()

library(dplyr)
# count CHD cases and non-cases by age stratum and behavioral pattern
or_by_age <-
  wcgs |>
  mutate(age_group = if_else(age >= 50, "50 or older", "under 50")) |>
  summarize(
    .by = age_group,
    a_e = sum(dibpat == "Type A" & chd69 == "Yes"),
    a_n = sum(dibpat == "Type A" & chd69 == "No"),
    b_e = sum(dibpat == "Type B" & chd69 == "Yes"),
    b_n = sum(dibpat == "Type B" & chd69 == "No")
  ) |>
  mutate(
    # disease OR: odds of CHD, comparing Type A to Type B
    disease_or = (a_e / a_n) / (b_e / b_n),
    # exposure OR: odds of Type A, comparing CHD cases to non-cases
    exposure_or = (a_e / b_e) / (a_n / b_n)
  )

d_young <- or_by_age |> filter(age_group == "under 50") |> pull(disease_or)
e_young <- or_by_age |> filter(age_group == "under 50") |> pull(exposure_or)
d_old <- or_by_age |> filter(age_group == "50 or older") |> pull(disease_or)
e_old <- or_by_age |> filter(age_group == "50 or older") |> pull(exposure_or)
```

Among participants under 50,
the disease odds ratio
(the odds of CHD, comparing Type A to Type B) is `r d_young`,
and the exposure odds ratio
(the odds of Type A, comparing CHD cases to non-cases) is `r e_young`;
the two agree, as @thm-conditional-OR-swap requires.

Among participants aged 50 or older,
the disease and exposure odds ratios are
`r d_old` and `r e_old`,
which likewise agree.

:::
