### Risk differences

::: notes
The simplest comparison of two probabilities, $\pi_1$, and $\pi_2$, is the difference of their values:
:::

:::{#def-RD}
#### Risk difference
The **risk difference** of two probabilities, $\pi_1$, and $\pi_2$, is the difference of their values:
$$\delta(\pi_1,\pi_2) \eqdef \pi_1 - \pi_2$$
:::

{{< slidebreak >}}

:::{#exm-RD}

#### Difference in MI risk

Here, we introduce the shorthand notation

$$\pi(OC) \eqdef \Pr(MI|OC), \quad \pi(\neg OC) \eqdef \Pr(MI|\neg OC)$$

for the probability of MI given OC use and given OC non-use,
respectively.

In @exm-oc-mi, the maximum likelihood estimate of the difference in MI risk between OC users and OC non-users is:
```{r, include = FALSE}
#| label: compute-risk-diff
rd_OC <- p_MI_OC - p_MI_nOC # nolint: object_name_linter
```

$$
\begin{aligned}
\hat\delta(\pi(OC), \pi(\neg OC))
&= \delta(\hat\pi(OC), \hat\pi(\neg OC))\\
&= \hat\pi(OC) - \hat\pi(\neg OC)\\
&= `r p_MI_OC` - `r p_MI_nOC`\\
&= `r rd_OC`
\end{aligned}
$$

:::

{{< slidebreak >}}

:::{#exr-interpret-rd}
#### interpreting risk differences

How can we interpret the preceding risk difference estimate in prose?
:::

{{< slidebreak >}}

:::{#sol-interpret-rd}

#### interpreting risk differences

"The difference in risk of MI between OC users and non-users was `r rd_OC`."

or

"The difference in risk of MI between OC users and non-users was `r rd_OC*100`
[percentage points](https://en.wikipedia.org/wiki/Percentage_point)."

:::

::: notes
See the note about working with percentages in the [Appendix](notation.qmd#percent-sign).
:::

{{< slidebreak >}}

### Risk ratios

:::{#exr-name-RR}

If $\pi_1$ and $\pi_2$ are two probabilities,
what do we call the following ratio?

$$\rho(\pi_1,\pi_2) = \frac{\pi_1}{\pi_2}$$
:::

{{< slidebreak >}}

:::{#sol-exr-name-RR}

:::{#def-RR}
### Relative risk ratios

The ratio of two probabilities $\pi_1$ and $\pi_2$,

$$\rho(\pi_1,\pi_2) = \frac{\pi_1}{\pi_2}$$

is called the:

- **risk ratio**,
- **relative risk ratio**,
- **probability ratio**,
- or **rate ratio**

of $\pi_1$ compared to $\pi_2$.

:::

:::


{{< slidebreak >}}

:::{#exr-RR}

:::: notes
In @sol-compute-probs, we estimated that:
::::

$$\ph(MI|OC) = `r 13/5000`$$

$$\ph(MI|\neg OC) = `r 7/10000`$$

Now, estimate the *relative risk* of MI for OC versus non-OC.

:::

{{< slidebreak >}}

:::{#sol-exr-RR}

:::: notes
The *relative risk* of MI for OC versus non-OC is:
::::

```{r}
#| label: compute-risk-ratio
rr <- (13 / 5000) / (7 / 10000)
```

$$
\begin{aligned}
\hat\rho(OC, \neg OC)
&=\frac{\ph(MI|OC)}{\ph(MI|\neg OC)}\\
&= \frac{`r 13/5000`}{`r 7/10000`}\\
&= `r rr`
\end{aligned}
$$

:::

{{< slidebreak >}}

:::{#exr-interpret-rr}
How can we interpret the preceding relative risk estimate in prose?
:::

{{< slidebreak >}}

:::{#sol-interpret-rr}

::: notes
We might summarize this result by saying that:
:::

- "The estimated probability of MI among OC users was 
`r rr` [times as high as]{.underline} 
the estimated probability among OC non-users."

or

- "The estimated probability of MI among OC users was 
`r rr - 1` [times higher than]{.underline}, 
the estimated probability among OC non-users."

::: notes
see also @vittinghoff2e [Section 8.1.4], which uses similar phrasing.
:::

:::

{{< slidebreak >}}

### Relative risk differences

::: notes
The second approach in @sol-interpret-rr,
where we subtract 1 from the risk ratio,
is actually reporting a slightly different metric:
:::

:::{#def-RRD}

#### Relative risk difference

:::: notes
Sometimes, we divide the risk difference by the comparison probability; the result is called the **relative risk difference**:
::::

$$\xi(\pi_1,\pi_2) \eqdef \frac{\delta(\pi_1,\pi_2)}{\pi_2}$$

:::

{{< slidebreak >}}

:::{#thm-rrd-vs-rr}
#### Relative risk difference equals risk ratio minus 1

$$\xi(\pi_1,\pi_2) = \rho(\pi_1,\pi_2) - 1$$
:::

{{< slidebreak >}}

::: proof
$$
\begin{aligned}
\xi(\pi_1,\pi_2)
&\eqdef \frac{\delta(\pi_1,\pi_2)}{\pi_2}
\\&= \frac{\pi_1-\pi_2}{\pi_2}
\\&= \frac{\pi_1}{\pi_2} - 1
\\&= \rho(\pi_1,\pi_2) - 1
\end{aligned}
$$
:::

{{< slidebreak >}}

:::{#exm-RRD}

#### Relative difference in MI risk

In @exm-oc-mi,
the maximum likelihood estimate of
the relative risk difference in MI risk
between OC users and OC non-users is:

```{r, include = FALSE}
#| label: compute-rrd
rrd_OC <- rd_OC / p_MI_nOC # nolint: object_name_linter
```

$$
\ba
\hat\xi(\pi(OC), \pi(\neg OC))
&= \xi(\hat\pi(OC), \hat\pi(\neg OC))
&& \text{(invariance property of MLEs)}
\\ &\eqdef \frac{\delta(\hat\pi(OC), \hat\pi(\neg OC))}{\hat\pi(\neg OC)}
&& \text{(definition of relative risk difference)}
\\ &= \frac{`r rd_OC`}{`r p_MI_nOC`}
&& \text{(substituting the estimates)}
\\ &= `r rrd_OC`
&& \text{(dividing)}
\ea
$$

::: notes
The numerator `r rd_OC` is the risk difference estimate from @exm-RD,
and the denominator `r p_MI_nOC` is the estimate $\hat\pi(\neg OC)$ from @sol-compute-probs.
:::

We can check this estimate using @thm-rrd-vs-rr
and the risk ratio estimate from @sol-exr-RR:

$$
\ba
\hat\xi(\pi(OC), \pi(\neg OC))
&= \hat\rho(OC, \neg OC) - 1
&& \text{(relative risk difference as risk ratio minus 1)}
\\ &= `r rr` - 1
&& \text{(substituting the risk ratio estimate)}
\\ &= `r rr - 1`
&& \text{(subtracting)}
\ea
$$

The two calculations agree.

:::

{{< slidebreak >}}

### Changing reference groups in risk comparisons

:::: notes
Risk differences, risk ratios, and relative risk differences are defined by two probabilities, plus a choice of which probability is the **baseline** or **reference** probability (i.e., which probability is the subtrahend of the risk difference or the denominator of the risk ratio).
:::

$$\delta(\pi_2,\pi_1) = -\delta(\pi_1, \pi_2)$$ {#eq-rd-switch-ref}

$$\rho(\pi_2,\pi_1) = \inv{\rho(\pi_1,\pi_2)}$$ {#eq-rr-switch-ref}

$$\xi(\pi_2,\pi_1) = \inv{\xi(\pi_1,\pi_2) + 1} - 1$$ {#eq-rrd-switch-ref}

:::{#exr-change-ref-group}

Prove @eq-rd-switch-ref, @eq-rr-switch-ref, and @eq-rrd-switch-ref.

:::

{{< slidebreak >}}

:::{#sol-change-ref-group}

To prove @eq-rd-switch-ref:

$$
\ba
\delta(\pi_2,\pi_1)
&\eqdef \pi_2 - \pi_1
&& \text{(definition of risk difference)}
\\ &= -(\pi_1 - \pi_2)
&& \text{(factoring out } -1 \text{)}
\\ &= -\delta(\pi_1, \pi_2)
&& \text{(definition of risk difference)}
\ea
$$

To prove @eq-rr-switch-ref:

$$
\ba
\rho(\pi_2,\pi_1)
&\eqdef \frac{\pi_2}{\pi_1}
&& \text{(definition of risk ratio)}
\\ &= \inv{\frac{\pi_1}{\pi_2}}
&& \text{(inverse of the inverted fraction)}
\\ &= \inv{\rho(\pi_1,\pi_2)}
&& \text{(definition of risk ratio)}
\ea
$$

To prove @eq-rrd-switch-ref,
first rearrange @thm-rrd-vs-rr to express the risk ratio
in terms of the relative risk difference:

$$
\ba
\xi(\pi_1,\pi_2)
&= \rho(\pi_1,\pi_2) - 1
&& \text{(relative risk difference as risk ratio minus 1)}
\\ \xi(\pi_1,\pi_2) + 1
&= \rho(\pi_1,\pi_2)
&& \text{(adding 1 to both sides)}
\ea
$$

Then apply @thm-rrd-vs-rr to the switched-reference pair $(\pi_2, \pi_1)$,
substitute @eq-rr-switch-ref,
and substitute the rearranged form of @thm-rrd-vs-rr:

$$
\ba
\xi(\pi_2,\pi_1)
&= \rho(\pi_2,\pi_1) - 1
&& \text{(relative risk difference as risk ratio minus 1)}
\\ &= \inv{\rho(\pi_1,\pi_2)} - 1
&& \text{(risk ratio with switched reference group)}
\\ &= \inv{\xi(\pi_1,\pi_2) + 1} - 1
&& \text{(substituting } \rho(\pi_1,\pi_2) = \xi(\pi_1,\pi_2) + 1 \text{)}
\ea
$$

:::

{{< slidebreak >}}

:::{#exm-ref-rd}

#### Switching the reference group in a risk difference

In @exm-RD, we estimated that the risk difference of MI for OC versus non-OC is:

$$
\begin{aligned}
\delta(OC, \neg OC)
&= `r rd_OC`
\end{aligned}
$$

In comparison, the risk difference for non-OC versus OC is:

$$
\begin{aligned}
\delta(\neg OC, OC)
&= \ph(MI|\neg OC) - \ph(MI|OC)\\
&= `r p_MI_nOC` - `r p_MI_OC`\\
&= `r p_MI_nOC - p_MI_OC`\\
&= -\delta(OC, \neg OC)
\end{aligned}
$$

as @eq-rd-switch-ref states.

:::

{{< slidebreak >}}

:::{#exm-ref}

#### Switching the reference group in a risk ratio

In @sol-exr-RR, we estimated that the risk ratio of OC versus non-OC is:

$$
\begin{aligned}
\rho(OC, \neg OC)
&= `r (13/5000)/(7/10000)`
\end{aligned}
$$

In comparison, the risk ratio for non-OC versus OC is:

$$
\begin{aligned}
\rho(\neg OC, OC)
&=\frac{\ph(MI|\neg OC)}{\ph(MI|OC)}\\
&= \frac{`r 7/10000`}{`r 13/5000`}\\
&= `r (7/10000)/(13/5000)`\\
&= \frac{1}{\rho(OC, \neg OC)}
\end{aligned}
$$

as @eq-rr-switch-ref states.

:::

{{< slidebreak >}}

:::{#exm-ref-rrd}

#### Switching the reference group in a relative risk difference

In @exm-RRD, we estimated that the relative risk difference of MI for OC versus non-OC is:

$$
\begin{aligned}
\xi(OC, \neg OC)
&= `r rrd_OC`
\end{aligned}
$$

In comparison, the relative risk difference for non-OC versus OC is:

$$
\begin{aligned}
\xi(\neg OC, OC)
&= \frac{\ph(MI|\neg OC) - \ph(MI|OC)}{\ph(MI|OC)}\\
&= \frac{`r p_MI_nOC - p_MI_OC`}{`r p_MI_OC`}\\
&= `r (p_MI_nOC - p_MI_OC) / p_MI_OC`
\end{aligned}
$$

We can check this value against the switched-reference identity @eq-rrd-switch-ref:

$$
\begin{aligned}
\frac{1}{\xi(OC, \neg OC) + 1} - 1
&= \frac{1}{`r rrd_OC` + 1} - 1\\
&= `r 1 / (rrd_OC + 1) - 1`
\end{aligned}
$$

The two calculations agree.

:::

