{{< include latex-macros/macros.qmd >}}

{{< include _subfiles/logistic-regression/_sec_intro_wcgs.qmd >}}

### Load the data

Here, I load the data:

```{r}
#| code-fold: show
#| eval: false

### load the data directly from a UCSF website:
library(haven)
url <- paste0(
  # I'm breaking up the url into two chunks for readability
  "https://regression.ucsf.edu/sites/g/files/",
  "tkssra6706/f/wysiwyg/home/data/wcgs.dta"
)
wcgs <- haven::read_dta(url)
```

```{r}
#| include: false
# I presaved the data in my project's `data` folder
library(here) # provides the `here()` function
library(fs) # provides the `path()` function
here::here() |>
  fs::path("Data/wcgs.rda") |>
  load()
```

A copy is also available on
[Kaggle](https://www.kaggle.com/datasets/nmd2104/wcgs-data).

```{r}
#| label: tbl-wcgs
#| tbl-cap: "`wcgs` data"
wcgs
```

{{< slidebreak >}}

### Data cleaning

::: notes
Now let's do some data cleaning
:::

```{r}
#| code-fold: show
library(arsenal) # provides `set_labels()`
library(forcats) # provides `as_factor()`
library(haven)
library(plotly)
wcgs <- wcgs |>
  mutate(
    age = age |>
      arsenal::set_labels("Age (years)"),
    arcus = arcus |>
      as.logical() |>
      arsenal::set_labels("Arcus Senilis"),
    time169 = time169 |>
      as.numeric() |>
      arsenal::set_labels("Observation (follow up) time (days)"),
    dibpat = dibpat |>
      as_factor() |>
      relevel(ref = "Type B") |>
      arsenal::set_labels("Behavioral Pattern"),
    typchd69 = typchd69 |>
      labelled(
        label = "Type of CHD Event",
        labels =
          c(
            "None" = 0,
            "infdeath" = 1,
            "silent" = 2,
            "angina" = 3
          )
      ),

    # turn stata-style labelled variables in to R-style factors:
    across(
      where(is.labelled),
      haven::as_factor
    )
  )
```

### What's in the data {.smaller}

::: notes
@tbl-wcgs-demos summarizes the data.
:::

:::{#tbl-wcgs-demos}

```{r}
library(gtsummary)
wcgs |>
  dplyr::select(
    -dplyr::all_of(c("id", "uni", "t1"))
  ) |>
  gtsummary::tbl_summary(
    by = "chd69",
    missing_text = "Missing"
  ) |>
  gtsummary::add_p() |>
  gtsummary::add_overall() |>
  gtsummary::bold_labels() |> 
  gtsummary::separate_p_footnotes()
```

Baseline characteristics by CHD status at end of follow-up

:::

### Data by age and personality type

For now, we will look at the interaction between age and personality
type (`dibpat`). 
To make it easier to visualize the data, 
we summarize the event rates for each combination of age:

```{r}
#| label: "event rates by age"
library(dplyr)
odds <- function(pi) pi / (1 - pi)
chd_grouped_data <-
  wcgs |>
  summarize(
    .by = c(age, dibpat),
    n = sum(chd69 %in% c("Yes", "No")),
    x = sum(chd69 == "Yes")
  ) |>
  mutate(
    `n - x` = n - x,
    `p(chd)` = (x / n) |>
      labelled(label = "CHD Event by 1969"),
    `odds(chd)` = `p(chd)` / (1 - `p(chd)`),
    `logit(chd)` = log(`odds(chd)`)
  )

chd_grouped_data
```

### Graphical exploration

```{r}
#| label: fig-chd-graph-prob-scale
#| fig-cap: "CHD rates by age group, probability scale"

library(ggplot2)
library(scales)
chd_plot_probs <-
  chd_grouped_data |>
  ggplot() +
  aes(
    x = age,
    y = `p(chd)`,
    col = dibpat
  ) +
  geom_point(aes(size = n), alpha = .7) +
  scale_size(range = c(1, 4)) +
  geom_line() +
  theme_bw() +
  ylab("P(CHD Event by 1969)") +
  scale_y_continuous(
    labels = scales::label_percent(),
    sec.axis = sec_axis(
      ~ odds(.),
      name = "odds(CHD Event by 1969)"
    )
  ) +
  theme(legend.position = "bottom")

print(chd_plot_probs)
```

{{< slidebreak >}}

## Odds scale

```{r}
#| label: fig-chd-odds-scale
#| fig-cap: "CHD rates by age group, odds spacing"

odds_inv <- function(omega) omega / (1 + omega)
trans_odds <- trans_new(
  name = "odds",
  transform = odds,
  inverse = odds_inv
)

chd_plot_odds <- chd_plot_probs +
  scale_y_continuous(
    trans = trans_odds, # this line changes the vertical spacing
    name = chd_plot_probs$labels$y,
    sec.axis = sec_axis(
      ~ odds(.),
      name = "odds(CHD Event by 1969)"
    )
  )

print(chd_plot_odds)
```

{{< slidebreak >}}

## Log-odds (logit) scale

```{r}
#| label: fig-chd_plot_logit
#| fig-cap: "CHD data (logit-scale)"

logit <- function(pi) log(odds(pi))
expit <- function(eta) odds_inv(exp(eta))
trans_logit <- trans_new(
  name = "logit",
  transform = logit,
  inverse = expit
)

chd_plot_logit <-
  chd_plot_probs +
  scale_y_continuous(
    trans = trans_logit, # this line changes the vertical spacing
    name = chd_plot_probs$labels$y,
    breaks = c(seq(.01, .1, by = .01), .15, .2),
    minor_breaks = NULL,
    sec.axis = sec_axis(
      ~ logit(.),
      name = "log(odds(CHD Event by 1969))"
    )
  )

print(chd_plot_logit)
```

### Logistic regression models for CHD data {.smaller}

{{< include _subfiles/logistic-regression/_sec_wcgs_notation.qmd >}}

```{r}
#| results: asis
chd_glm_contrasts <-
  wcgs |>
  glm(
    "data" = _,
    "formula" = chd69 == "Yes" ~ dibpat * age,
    "family" = binomial(link = "logit")
  )

library(equatiomatic)
equatiomatic::extract_eq(chd_glm_contrasts)
```

Or in more formal notation:

{{< include _subfiles/logistic-regression/_eq_wcgs_model_1.qmd >}}

{{< include _subfiles/logistic-regression/_sec_wcgs_model_graphs.qmd >}}

### Interpreting the model parameters

{{< include _subfiles/logistic-regression/_exm_wcgs_coef_interp.qmd >}}

### Interpreting the model parameter estimates

@tbl-model-corner-point shows the fitted model.

```{r}
#| label: tbl-model-corner-point
#| tbl-cap: "CHD model (corner-point parametrization)"

library(parameters)
chd_glm_contrasts |>
  parameters() |>
  print_md()
```

{{< slidebreak >}}

We can get the corresponding odds ratio estimates ($e^{\hat{\beta}}$s) 
by passing `exponentiate = TRUE` to `parameters()`:

```{r}
#| tbl-cap: "Odds ratio estimates for CHD model"
#| label: tbl-chd-glm-contrasts-OR
chd_glm_contrasts |>
  parameters(exponentiate = TRUE) |>
  print_md()
```


### Stratified parametrization

We could instead use a stratified parametrization:

```{r}
#| results: asis
chd_glm_strat <- glm(
  "formula" = chd69 == "Yes" ~ dibpat + dibpat:age - 1,
  "data" = wcgs,
  "family" = binomial(link = "logit")
)
equatiomatic::extract_eq(chd_glm_strat)
```

```{r}
#| tbl-cap: "CHD model, stratified parametrization"
#| label: tbl-chd-strat
chd_glm_strat |>
  parameters() |>
  print_md()
```

{{< slidebreak >}}

Again, we can get the corresponding odds ratios ($e^{\beta}$s) by passing
`exponentiate = TRUE` to `parameters()`:

```{r}
#| tbl-cap: "Odds ratio estimates for CHD model"
#| label: tbl-chd-glm-strat-OR
chd_glm_strat |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

Compare with @tbl-model-corner-point.

{{< slidebreak >}}

:::{#exr-strat-to-contrast}
If I give you model 1, how would you get the coefficients of model 2?
:::

{{< slidebreak >}}

:::{#sol-strat-to-contrast}

Model 1 is the corner-point (reference/contrast) parametrization
fitted in `chd_glm_contrasts`,
with linear predictor (@eq-wcgs-model-1):

$$\eta(p, a) = \beta_0 + \beta_P p + \beta_A a + \beta_{PA} pa$$

Model 2 is the stratified parametrization
fitted in `chd_glm_strat`,
which gives each personality type
its own intercept and its own age slope.
Writing $\gamma_B, \gamma_A$ for the stratum-specific intercepts
and $\delta_B, \delta_A$ for the stratum-specific age slopes:

$$
\eta(p, a) \eqdef
\gamma_B (1 - p) + \gamma_A p + \delta_B (1 - p) a + \delta_A p a
$$ {#eq-wcgs-model-strat}

The two parametrizations span the same set of linear-predictor functions,
so both models have the same fitted linear predictor $\eta(p, a)$.
We can therefore recover the model 2 coefficients
by equating the two expressions for $\eta(p, a)$
within each personality-type stratum.

{{< slidebreak >}}

**Type B stratum ($p = 0$):**

In model 1 (@eq-wcgs-model-1):

$$
\ba
\eta(0, a)
&= \beta_0 + \beta_P \cdot 0 + \beta_A a + \beta_{PA} \cdot 0 \cdot a
&& \text{(substitute } p = 0\text{)}
\\
&= \beta_0 + 0 + \beta_A a + 0
&& \text{(multiply by 0)}
\\
&= \beta_0 + \beta_A a
&& \text{(drop the zero terms)}
\ea
$$

In model 2 (@eq-wcgs-model-strat):

$$
\ba
\eta(0, a)
&= \gamma_B (1 - 0) + \gamma_A \cdot 0 + \delta_B (1 - 0) a + \delta_A \cdot 0 \cdot a
&& \text{(substitute } p = 0\text{)}
\\
&= \gamma_B \cdot 1 + \gamma_A \cdot 0 + \delta_B \cdot 1 \cdot a + \delta_A \cdot 0 \cdot a
&& \text{(} 1 - 0 = 1 \text{)}
\\
&= \gamma_B + 0 + \delta_B a + 0
&& \text{(multiply by 1 and by 0)}
\\
&= \gamma_B + \delta_B a
&& \text{(drop the zero terms)}
\ea
$$

{{< slidebreak >}}

Equating the two expressions for $\eta(0, a)$, for all ages $a$:

$$
\gamma_B + \delta_B a = \beta_0 + \beta_A a
$$ {#eq-wcgs-strat-match-b}

Setting $a = 0$ in @eq-wcgs-strat-match-b:

$$
\ba
\gamma_B + \delta_B \cdot 0
&= \beta_0 + \beta_A \cdot 0
&& \text{(substitute } a = 0\text{)}
\\
\gamma_B + 0
&= \beta_0 + 0
&& \text{(multiply by 0)}
\\
\gamma_B
&= \beta_0
&& \text{(drop the zero terms)}
\ea
$$

Subtracting $\gamma_B = \beta_0$ from both sides of @eq-wcgs-strat-match-b
and then setting $a = 1$:

$$
\ba
\delta_B a
&= \beta_A a
&& \text{(subtract } \gamma_B = \beta_0 \text{ from both sides)}
\\
\delta_B \cdot 1
&= \beta_A \cdot 1
&& \text{(substitute } a = 1\text{)}
\\
\delta_B
&= \beta_A
&& \text{(multiply by 1)}
\ea
$$

{{< slidebreak >}}

**Type A stratum ($p = 1$):**

In model 1 (@eq-wcgs-model-1):

$$
\ba
\eta(1, a)
&= \beta_0 + \beta_P \cdot 1 + \beta_A a + \beta_{PA} \cdot 1 \cdot a
&& \text{(substitute } p = 1\text{)}
\\
&= \beta_0 + \beta_P + \beta_A a + \beta_{PA} a
&& \text{(multiply by 1)}
\\
&= (\beta_0 + \beta_P) + (\beta_A + \beta_{PA}) a
&& \text{(group the constant terms; factor out } a \text{)}
\ea
$$

In model 2 (@eq-wcgs-model-strat):

$$
\ba
\eta(1, a)
&= \gamma_B (1 - 1) + \gamma_A \cdot 1 + \delta_B (1 - 1) a + \delta_A \cdot 1 \cdot a
&& \text{(substitute } p = 1\text{)}
\\
&= \gamma_B \cdot 0 + \gamma_A \cdot 1 + \delta_B \cdot 0 \cdot a + \delta_A \cdot 1 \cdot a
&& \text{(} 1 - 1 = 0 \text{)}
\\
&= 0 + \gamma_A + 0 + \delta_A a
&& \text{(multiply by 0 and by 1)}
\\
&= \gamma_A + \delta_A a
&& \text{(drop the zero terms)}
\ea
$$

Equating the two expressions for $\eta(1, a)$
and applying the same two steps as in the Type B stratum
(set $a = 0$; then subtract the intercepts and set $a = 1$):

$$
\ba
\gamma_A &= \beta_0 + \beta_P
\\
\delta_A &= \beta_A + \beta_{PA}
\ea
$$

{{< slidebreak >}}

**Summary of the conversion:**

$$
\ba
\gamma_B &= \beta_0
\\
\gamma_A &= \beta_0 + \beta_P
\\
\delta_B &= \beta_A
\\
\delta_A &= \beta_A + \beta_{PA}
\ea
$$ {#eq-wcgs-strat-from-contrast}

Solving @eq-wcgs-strat-from-contrast for the model 1 coefficients
inverts the conversion,
so either parametrization determines the other:

$$
\ba
\beta_0 &= \gamma_B
&& \text{(first equation of the summary)}
\\
\beta_P &= \gamma_A - \gamma_B
&& \text{(subtract the first equation from the second)}
\\
\beta_A &= \delta_B
&& \text{(third equation)}
\\
\beta_{PA} &= \delta_A - \delta_B
&& \text{(subtract the third equation from the fourth)}
\ea
$$

{{< slidebreak >}}

**Numerical check:**
applying @eq-wcgs-strat-from-contrast to the fitted model 1 coefficients
(@tbl-model-corner-point)
reproduces the fitted model 2 coefficients (@tbl-chd-strat):

```{r}
#| label: strat-to-contrast-check
#| code-fold: true

b <- coef(chd_glm_contrasts)

gamma_delta <- c(
  "dibpatType B" = unname(b["(Intercept)"]),
  "dibpatType A" = unname(b["(Intercept)"] + b["dibpatType A"]),
  "dibpatType B:age" = unname(b["age"]),
  "dibpatType A:age" = unname(b["age"] + b["dibpatType A:age"])
)

cbind(
  "converted from model 1" = gamma_delta,
  "model 2 (fitted directly)" = coef(chd_glm_strat)[names(gamma_delta)]
)
```

:::
