{{< include latex-macros/macros.qmd >}}

:::{#exr-ci-prob}

Given
a maximum likelihood estimate $\hb$
and
a corresponding estimated covariance matrix
$\hS \eqdef \widehat{\Covt}(\hb)$,
calculate a 95% confidence interval for the predicted probability
$\pi(\vx) = \Pr(Y = 1 | \vX = \vx)$.

:::

{{< slidebreak >}}

:::{#sol-ci-prob}

::: notes

By [the central limit theorem for MLEs](intro-MLEs.qmd#thm-dist-mle),
a 95% confidence interval for $\pi(\vx)$
can be constructed as:

$$\hat\pi(\vx) \pm 1.96 * \HSE{\hat\pi(\vx)}$$

However, $\HSE{\hat\pi(\vx)}$ seems difficult to compute;
doing so would require using the
[delta method](https://en.wikipedia.org/wiki/Delta_method).

Instead, using the invariance property of MLEs,
we can first calculate a confidence interval for the log-odds,

$$\logodds(\vx) = \vx'\vb \in (L, R)$$

and then apply $\expit$ to the endpoints of that log-odds-scale confidence interval:

:::

$$\pi(\vx) \in (\expit(L), \expit(R))$${#eq-ci-prob-exp}

:::

{{< slidebreak >}}

:::{#exr-logodds-ci}

Find a 95% confidence interval for the log-odds $\logodds(\vx) = \vx'\vb$.

:::

{{< slidebreak >}}

:::{#sol-logodds-ci}

By [the central limit theorem for MLEs](intro-MLEs.qmd#thm-dist-mle),
a 95% confidence interval for $\logodds(\vx)$
can be constructed as:

$$\hat\logodds(\vx) \pm 1.96 * \teal{\HSE{\hat\logodds(\vx)}}$$

where $\hat\logodds(\vx) = \vx'\hb$.

:::

{{< slidebreak >}}

:::{#exr-se-logodds}

::: notes

How can we estimate the standard error of
$\hat\logodds(\vx)$?

:::

$$\teal{\HSE{\hat\logodds(\vx)}} = \teal{?}$$

:::

{{< slidebreak >}}

:::{#sol-se-logodds}

$$
\green{\SE{\hat\logodds(\vx)}} =
\sqrt{\red{\Var{\hat\logodds(\vx)}}}
$${#eq-se-sqrt-var-logodds}

By the definition $\hat\logodds(\vx) = \vx'\hb$
and [the variance of a linear combination](probability.qmd#thm-var-lincom):

$$
\ba
\red{\Var{\hat\logodds(\vx)}}
&= \Var{\tp{\vx}\hb}
\\
&= \tp{\vx}\Cov{\hb}\vx
\\
&= \red{\tp{\vx}\Sigma\vx}
\ea
$${#eq-var-logodds}

where $\Sigma \eqdef \Covt(\hb)$.

{{< slidebreak >}}

::: notes

Expanding @eq-var-logodds out of matrix-vector notation, we have:

:::

$$
\ba
\red{\tp{\vx}\Sigma\vx}
&= \sumip\sumjp x_i \Sigma_{ij} x_j
\\
&= \red{\sumip\sumjp x_i \Covt(\hb_i,\hb_j) x_j}
\ea
$$

:::

{{< slidebreak >}}

Combining @eq-var-logodds with the invariance property of MLEs
gives an estimator for the variance and standard error:

:::{#thm-est-se-logodds}

#### Estimated variance and standard error of log-odds

$$
\orange{\hVar{\hat\logodds(\vx)}} = \orange{\tp{\vx}\hS\vx}
$${#eq-est-var-logodds}

$$
\teal{\HSE{\hat\logodds(\vx)}} =
\teal{\sqrt{\tp{\vx}\hS\vx}}
$${#eq-est-se-logodds}

:::

::: proof
By @sol-se-logodds,
$\Var{\hat\logodds(\vx)} = \tp{\vx}\Sigma\vx$ (@eq-var-logodds).
By the invariance property of MLEs,
we estimate $\Sigma$ with $\hS \eqdef \widehat{\Covt}(\hb)$,
which yields @eq-est-var-logodds;
taking the square root gives @eq-est-se-logodds.
:::

{{< slidebreak >}}

:::{#exm-est-se-logodds-beetles}

#### Standard error and CI for a beetle log-odds

::: notes
We instantiate @thm-est-se-logodds
for the `beetles` logistic regression model
(@tbl-beetles-model-ungrouped).
:::

Consider the covariate pattern $\vx = (1, 1.8)\'$,
representing the intercept and a dose of 1.8.

```{r}
#| label: exm-se-logodds-beetles-setup
#| code-fold: show
x_vec <- c(1, 1.8)
sigma_hat <- vcov(beetles_glm_ungrouped)
logodds_hat <- as.numeric(x_vec %*% coef(beetles_glm_ungrouped))
se_logodds <- sqrt(as.numeric(t(x_vec) %*% sigma_hat %*% x_vec))
ci_logodds <- logodds_hat + c(-1, 1) * 1.96 * se_logodds
```

By @eq-est-se-logodds,
the estimated standard error of the log-odds is
$\HSE{\hat\logodds(\vx)} = \sqrt{\tp{\vx}\hS\vx} = `r se_logodds`$,
while the estimated log-odds itself is
$\hat\logodds(\vx) = \vx'\hb = `r logodds_hat`$.

The resulting 95% confidence interval for the log-odds is
$(`r ci_logodds[1]`, `r ci_logodds[2]`)$,
and applying $\expit$ to its endpoints (@eq-ci-prob-exp)
gives a 95% confidence interval for $\pi(\vx)$ of
$(`r plogis(ci_logodds[1])`, `r plogis(ci_logodds[2])`)$.

These values match the `dose = 1.8` row of @tbl-beetles-pred-logodds,
which uses `predict()` with `type = "link"` and `se.fit = TRUE`
to compute the estimated log-odds $\hat\logodds(\vx) = \vx'\hb$
and its estimated standard error for each covariate pattern:

```{r}
#| label: tbl-beetles-pred-logodds
#| tbl-cap: "Predicted log-odds and 95% CI for `beetles` logistic regression"
library(dplyr)
new_doses <- tibble(dose = c(1.7, 1.8, 1.9))

pred_logodds <-
  beetles_glm_ungrouped |>
  predict(
    newdata = new_doses,
    type = "link",
    se.fit = TRUE
  )

new_doses |>
  mutate(
    logodds_hat = pred_logodds$fit,
    se = pred_logodds$se.fit,
    ci_lower_logodds = logodds_hat - 1.96 * se,
    ci_upper_logodds = logodds_hat + 1.96 * se,
    pi_hat = plogis(logodds_hat),
    ci_lower_prob = plogis(ci_lower_logodds),
    ci_upper_prob = plogis(ci_upper_logodds)
  ) |>
  knitr::kable(digits = 3)
```

:::
