If you want to report risk differences or risk ratios instead of odds ratios,
you can obtain estimates from logistic regression models,
as long as you didn't stratify sampling by the outcome;
in other words, not in case-control studies
(see @sec-CC-studies).

### Computing marginal risk differences

::: {.callout-note}
This section is **optional for 2026** and will not be tested on exams.
:::

#### Conceptual approach

To compute risk ratios or risk differences from logistic regression models:

- Apply the expit function (@def-expit)
  to the linear predictor
  for each covariate pattern
  to compute the (estimated) risks,
- Then take the differences or ratios of the risks, as needed.

For example,
suppose we have a logistic regression model
with a binary exposure $A$ and a covariate $Z$:

$$
\logitf{\Pr(Y = 1 | A, Z)} = \b_0 + \b_A A + \b_Z Z
$$

:::{#def-cond-predicted-risk}

#### Conditional predicted risk

The **conditional predicted risk** (or **conditional risk**)
for an individual with exposure $A = a$ and covariate $Z = z$
is the population probability

$$
\pi(a, z) \eqdef \Pr(Y = 1 \mid A = a, Z = z).
$$

:::

Under the logistic model above, this population risk equals

$$
\pi(a, z) = \expitf{\b_0 + \b_A a + \b_Z z}.
$$

:::{#def-cond-rd}

#### Conditional risk difference

The **conditional risk difference** comparing exposure levels $a = 1$ vs. $a = 0$
at a fixed covariate value $z$ is, using the conditional predicted risk
(@def-cond-predicted-risk):

$$
\text{RD}(z) \eqdef \pi(1, z) - \pi(0, z)
$$

:::

:::{#def-cond-rrr}

#### Conditional risk ratio

The **conditional risk ratio** comparing exposure levels $a = 1$ vs. $a = 0$
at a fixed covariate value $z$ is, using the conditional predicted risk
(@def-cond-predicted-risk):

$$
\text{RR}(z) \eqdef \frac{\pi(1, z)}{\pi(0, z)}
$$

:::

{{< slidebreak >}}

:::{#exm-cond-rd-rr}

#### Numerical example: conditional RD and RR at a fixed covariate value

Using the WCGS data with `dibpat` (behavior pattern) as the exposure
and `chd69` (coronary heart disease by 1969) as the outcome,
adjusted for `age`:

```{r}
#| label: wcgs-marginal-setup
#| code-fold: true
#| message: false
#| warning: false
library(dplyr)
library(haven)

wcgs_clean <- rmb::wcgs |>
  mutate(
    across(where(is.labelled), haven::as_factor),
    dibpat = dibpat |> relevel(ref = "Type B"),
    chd69_binary = as.numeric(chd69 == "Yes")
  ) |>
  filter(!is.na(chd69_binary), !is.na(dibpat), !is.na(age))

fit <- glm(
  chd69_binary ~ dibpat + age,
  data = wcgs_clean,
  family = binomial(link = "logit")
)
```

At age $z = 50$ (within the WCGS sample's observed age range):

```{r}
#| label: cond-rd-rr-example
#| code-fold: true

z <- 50
dibpat_levels <- levels(wcgs_clean$dibpat)
wcgs_cf_a <- data.frame(dibpat = factor("Type A", levels = dibpat_levels), age = z)
wcgs_cf_b <- data.frame(dibpat = factor("Type B", levels = dibpat_levels), age = z)

pi_hat_1 <- predict(fit, newdata = wcgs_cf_a, type = "response")
pi_hat_0 <- predict(fit, newdata = wcgs_cf_b, type = "response")

rd_hat_z <- pi_hat_1 - pi_hat_0
rr_hat_z <- pi_hat_1 / pi_hat_0

tibble::tibble(
  quantity = c(
    "$\\hat\\pi(1, 50)$", "$\\hat\\pi(0, 50)$",
    "$\\widehat{\\text{RD}}(50)$", "$\\widehat{\\text{RR}}(50)$"
  ),
  value = c(pi_hat_1, pi_hat_0, rd_hat_z, rr_hat_z)
) |>
  knitr::kable(digits = 3)
```

These are *conditional* estimates: both hold age fixed at 50
and compare only the two exposure levels at that one age.

:::

#### Marginal vs. conditional effects

{{< include _subfiles/logistic-regression/_sec-marginal-vs-conditional.qmd >}}

{{< slidebreak >}}

#### Standard errors and confidence intervals

To quantify uncertainty for risk difference or risk ratio estimates
derived from logistic regression models
(e.g., to calculate SEs, CIs, and p-values),
you will need to use the
bootstrap,
the multivariate delta method,
or some other special technique.

The **bootstrap** is generally the most straightforward approach
and does not require deriving complex formulas.

{{< slidebreak >}}

### Bootstrap inference

{{< include _subfiles/logistic-regression/_sec-bootstrap-inference.qmd >}}

{{< slidebreak >}}

### Computing marginal risk ratios

{{< include _subfiles/logistic-regression/_sec-marginal-rr.qmd >}}

{{< slidebreak >}}

### Notes and caveats

<!-- Reviewer note (intentional, do not re-flag): no slidebreak between this
heading and the include. The slidebreak above already opens a fresh slide; the
heading and the first div (#def-collapsibility) intentionally share that slide
to avoid a title-only slide. -->
{{< include _subfiles/logistic-regression/_sec-marginal-notes.qmd >}}
