By [the central limit theorem for MLEs](intro-MLEs.qmd#thm-dist-mle),
the maximum likelihood estimates $\hb_k$ are approximately Gaussian
for large sample sizes;
see also the [table of Gaussian vs. MLE-based tests](intro-MLEs.qmd#tbl-gaussian-vs-mle-tests).

#### Wald test statistic

To test $H_0: \beta_k = \beta_{k,0}$ (typically $\beta_{k,0} = 0$):

$$z_k = \frac{\hb_k - \beta_{k,0}}{\HSE{\hb_k}}$$

Under $H_0$, $z_k \overset{\cdot}{\sim} N(0,1)$ for large $n$.

#### Confidence intervals for regression coefficients

A 95% confidence interval for $\beta_k$ is:

$$\hb_k \pm 1.96 \cdot \HSE{\hb_k}$$

#### Confidence intervals for exponentiated coefficients

By the invariance property of MLEs,
the MLE of $e^{\beta_k}$ is $e^{\hb_k}$.

A 95% confidence interval for $e^{\beta_k}$ is obtained by
exponentiating the endpoints of the CI for $\beta_k$:

$$e^{\beta_k} \in \left(e^{\hb_k - 1.96 \cdot \HSE{\hb_k}},\; e^{\hb_k + 1.96 \cdot \HSE{\hb_k}}\right)$$

For covariate coefficients ($k \neq 0$), $e^{\beta_k}$ is an odds ratio
(the multiplicative change in odds per 1-unit increase in $x_k$,
holding all other covariates fixed).
For the intercept ($k = 0$), $e^{\beta_0}$ is the baseline odds
(the odds when all covariates equal zero), not an odds ratio.

#### In R

In R, `parameters()` from the `parameters` package automatically computes
Wald tests and confidence intervals for logistic regression model coefficients:

```{r}
#| label: tbl-beetles-wald-tests
#| tbl-cap: "Wald tests and 95% CIs for `beetles` logistic regression"
beetles_glm_ungrouped |>
  parameters() |>
  print_md()
```

Pass `exponentiate = TRUE` to `parameters()` to report
exponentiated coefficients.
The intercept is then interpreted as baseline odds,
while the non-intercept coefficients are odds ratios:

```{r}
#| label: tbl-beetles-wald-tests-OR
#| tbl-cap: "Exponentiated coefficients and 95% CIs for `beetles`"
beetles_glm_ungrouped |>
  parameters(exponentiate = TRUE) |>
  print_md()
```
