Now we use the bootstrap to estimate
the standard error and construct a 95% confidence interval.
In the non-parametric bootstrap,
each iteration applies both the model-fitting step
and the covariate standardization to the same bootstrap resample.
This joint refit-and-restandardize approach mirrors the original world in the bootstrap world
and is standard practice for this type of estimator.
(An alternative would fix the covariates at the observed values,
but is less common for g-computation estimators.)

```{r}
#| label: bootstrap-ci
#| code-fold: show
#| cache: true
#| message: false
#| warning: false

# Set seed for reproducibility
set.seed(20260512)

# Number of bootstrap iterations in this rendered example
# (illustrative only; use >= 2000 for final CIs)
B <- 300

boot_estimates <- numeric(B)

for (b in 1:B) {

  boot_indices <- sample(
    seq_len(nrow(wcgs_clean)),
    size = nrow(wcgs_clean),
    replace = TRUE
  )

  boot_data <- wcgs_clean[boot_indices, ]

  boot_fit <- glm(
    chd69_binary ~ dibpat + age,
    data = boot_data,
    family = binomial(link = "logit")
  )

  boot_estimates[b] <- compute_marginal_rd(
    model = boot_fit,
    data = boot_data,
    exposure_var = "dibpat",
    exposed_level = "Type A",
    unexposed_level = "Type B"
  )
}

boot_se <- sd(boot_estimates)

# percentile method: quantiles of the bootstrap distribution
ci_lower <- as.numeric(quantile(boot_estimates, 0.025))
ci_upper <- as.numeric(quantile(boot_estimates, 0.975))
```

::: {#tbl-marginal-rd-results}

```{r}
#| code-fold: true

tibble(
  Estimate = marginal_rd_est,
  `Std. Error` = boot_se,
  `95% CI Lower` = ci_lower,
  `95% CI Upper` = ci_upper
) |>
  knitr::kable(digits = 4)
```

Causal marginal risk difference (CMRD) for Type A vs Type B behavioral pattern
(adjusted for age) via g-computation,
with bootstrap standard error and 95% confidence interval

:::
