Note that these risk differences and risk ratios are **conditional** on the covariate value $z$.
In many applications,
we want to summarize the effect across the population,
which requires computing a **marginal** risk difference or risk ratio.

{{< slidebreak >}}

:::{#def-observed-marginal-risk}

#### Observed marginal risk

The **observed marginal risk** under exposure level $A = a$ is:

$$\pi(a) \eqdef \Pr(Y = 1 \mid A = a)$$

:::

{{< slidebreak >}}

:::{#exm-observed-marginal-risk-direct}

#### Illustration: observed marginal risk

Continuing the WCGS example (@exm-cond-rd-rr):

```{r}
#| label: wcgs-obs-marginal-risk-direct-illustration
#| code-fold: true

pi1_direct <- mean(wcgs_clean$chd69_binary[wcgs_clean$dibpat == "Type A"])
```

The observed marginal risk among the Type A men ($A = 1$)
is simply the proportion of them who developed CHD by 1969:

$$
\pi(1) = \Pr(Y = 1 \mid A = 1) = `r round(pi1_direct, 3)`
$$

:::

This definition collapses the outcome over everything else in one step.
When the model also conditions on covariates $Z$,
we can unpack $\pi(a)$ as the covariate-specific risk $\pi(a, z)$
averaged over the distribution of $Z$ among those with $A = a$:

{{< slidebreak >}}

{{< include _subfiles/logistic-regression/_thm-observed-marginal-risk.qmd >}}

{{< slidebreak >}}

:::{#exm-observed-marginal-risk}

#### Observed marginal risk by standardization

Take $Z$ to be an indicator for age 50 or older.
Among the WCGS men with $A = 1$ (Type A):

```{r}
#| label: wcgs-age-strata-setup
#| code-fold: true

wcgs_clean$age_ge50 <- as.numeric(wcgs_clean$age >= 50)
wcgs_exposed <- wcgs_clean[wcgs_clean$dibpat == "Type A", ]
pi1_z0 <- mean(wcgs_exposed$chd69_binary[wcgs_exposed$age_ge50 == 0])
pi1_z1 <- mean(wcgs_exposed$chd69_binary[wcgs_exposed$age_ge50 == 1])
p_z1_a1 <- mean(wcgs_exposed$age_ge50)
```

the age-specific risks are $\pi(1, 0) = `r round(pi1_z0, 3)`$
(under 50) and $\pi(1, 1) = `r round(pi1_z1, 3)`$ (50 or older),
and $\Pr(Z = 1 \mid A = 1) = `r round(p_z1_a1, 3)`$
of the Type A men are 50 or older.
Then @thm-observed-marginal-risk gives:

$$
\ba
\pi(1)
&= \E{\pi(1, Z) \mid A = 1}
&& \text{[observed marginal risk theorem]}
\\
&= (1 - `r round(p_z1_a1, 3)`)(`r round(pi1_z0, 3)`)
+ (`r round(p_z1_a1, 3)`)(`r round(pi1_z1, 3)`)
&& \text{[substitute stratum risks and weights]}
\\
&= `r round((1 - p_z1_a1) * pi1_z0 + p_z1_a1 * pi1_z1, 3)`
&& \text{[arithmetic]}
\ea
$$

matching the direct estimate $\pi(1) = `r round(pi1_direct, 3)`$
from @exm-observed-marginal-risk-direct
(up to rounding).

:::

In causal inference,
[potential outcomes](causal-inference.qmd#def-potential-outcomes) $Y^a$
represent the outcome each unit would have under exposure level $a$.
Population-level causal estimands summarize these potential outcomes:

{{< slidebreak >}}

{{< include _subfiles/logistic-regression/_sec-causal-marginal-risk.qmd >}}

Two related procedures estimate these marginal risks from a fitted regression model:

{{< slidebreak >}}

{{< include _subfiles/logistic-regression/_sec-g-computation.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/logistic-regression/_sec-predictive-margins.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/logistic-regression/_sec-marginal-risk-differences.qmd >}}
