
::: notes
When the outcome is rare (i.e., its probability is small) for both groups being compared in an odds ratio, the odds of the outcome will be similar to the probability of the outcome, and thus the risk ratio will be similar to the odds ratio.
:::

##### Case 1: rare events

For rare events, odds ratios and probability (a.k.a. risk, a.k.a.
prevalence) ratios will be close:

$$\pi_1 = .01$$
$$\pi_2 = .02$$

```{r "rare events or-rr"}
#| code-fold: show
pi1 <- .01
pi2 <- .02
pi2 / pi1
odds(pi2) / odds(pi1)
```

{{< slidebreak >}}

:::{#exm-or-rr-OC-MI}
In @exm-oc-mi, the outcome is rare for both OC and non-OC participants, so the odds for both groups are similar to the corresponding probabilities, and the odds ratio is similar the risk ratio.
:::

{{< slidebreak >}}

##### Case 2: frequent events

$$\pi_1 = .4$$

$$\pi_2 = .5$$

For more frequently-occurring outcomes, this won't be the case:

```{r}
#| code-fold: show
pi1 <- .4
pi2 <- .5
pi2 / pi1
odds(pi2) / odds(pi1)
```

{{< slidebreak >}}

::: notes
The relationship between the two spot-checks
is captured by a simple algebraic identity:
:::

{{< slidebreak >}}

:::{#thm-rare-disease-or-rr}
#### Odds ratio as risk ratio times a correction factor

For two probabilities $\pi_1$ and $\pi_2$,
the odds ratio equals the risk ratio
times a correction factor:

$$\ratio(\odds_1, \odds_2)
= \rho(\pi_1, \pi_2) \cdot \frac{1-\pi_2}{1-\pi_1}$$ {#eq-or-rr-correction}

When $\pi_1$ and $\pi_2$ are both small,
the correction factor $\frac{1-\pi_2}{1-\pi_1}$ is close to 1,
so the odds ratio is close to the risk ratio.
:::

{{< slidebreak >}}

::: proof
$$
\ba
\ratio(\odds_1, \odds_2)
&\eqdef \frac{\odds_1}{\odds_2}
&& \text{(@def-OR)}
\\ &= \frac{\paren{\frac{\pi_1}{1-\pi_1}}}{\paren{\frac{\pi_2}{1-\pi_2}}}
&& \text{(@thm-prob-to-odds, applied to each odds)}
\\ &= \frac{\pi_1}{\pi_2} \cdot \frac{1-\pi_2}{1-\pi_1}
&& \text{(rearranging the compound fraction)}
\\ &= \rho(\pi_1, \pi_2) \cdot \frac{1-\pi_2}{1-\pi_1}
&& \text{(@def-RR)}
\ea
$$
:::

{{< slidebreak >}}

:::{#exm-or-rr-correction}

#### The correction factor in the two spot-checks

@thm-rare-disease-or-rr explains the two spot-checks preceding it.

```{r}
#| label: or-rr-correction
#| include: false
odds <- function(pi) pi / (1 - pi)
# Case 1 (rare): higher-risk group pi_1 = 0.02, reference pi_2 = 0.01
rr_rare <- 0.02 / 0.01
or_rare <- odds(0.02) / odds(0.01)
corr_rare <- (1 - 0.01) / (1 - 0.02)
# Case 2 (frequent): higher-risk group pi_1 = 0.5, reference pi_2 = 0.4
rr_freq <- 0.5 / 0.4
or_freq <- odds(0.5) / odds(0.4)
corr_freq <- (1 - 0.4) / (1 - 0.5)
```

In **Case 1** (rare events),
take $\pi_1 = 0.02$ and $\pi_2 = 0.01$.
The correction factor $\frac{1-\pi_2}{1-\pi_1} = `r corr_rare`$ is close to 1,
so the odds ratio (`r or_rare`) is close to the risk ratio (`r rr_rare`).

In **Case 2** (frequent events),
take $\pi_1 = 0.5$ and $\pi_2 = 0.4$.
The correction factor $\frac{1-\pi_2}{1-\pi_1} = `r corr_freq`$ is farther from 1,
so the odds ratio (`r or_freq`) and risk ratio (`r rr_freq`) diverge.

:::

{{< slidebreak >}}

{{< include _subfiles/logistic-regression/_fig-rd-rr-or.qmd >}}

{{< slidebreak >}}

Testing whether an odds ratio = 1 is equivalent to
testing whether the corresponding risk ratio = 1,
and also equivalent to
testing whether the risk difference = 0.
Therefore,
in **hypothesis testing**,
if the null hypothesis is no effect,
then we can shift between RD, RR, and OR.
But when we're talking about **point estimates** and **CIs**,
we need to limit our conclusions to the effect measure(s) that we actually estimated,
because the *sizes* of RDs, RRs, and ORs don't have a simple relationship to each other,
except when $\pi_1=\pi_2$ (as shown by @fig-rd-rr-or).
