:::::{#exm-oc-mi}
### Oral Contraceptive Use and Heart Attack

* Research question:
how does oral contraceptive (OC) use
affect the risk of myocardial infarction (MI; a.k.a. heart attack)?

:::{.notes}
This was an issue when oral contraceptives were first developed,
because the original formulations used
higher concentrations of hormones.
Modern OCs don't have this issue.

@tbl-oc-mi contains simulated data for
an imaginary follow-up (a.k.a. *prospective*) study
in which two groups are identified,
one using OCs and another not using OCs,
and both groups are tracked for three years
to determine how many in each groups have MIs.

:::

:::{#tbl-oc-mi}

```{r}
#| message: false
#| code-fold: true
library(dplyr)
oc_mi <-
  tribble(
    ~OC, ~MI, ~Total,
    "OC use", 13, 5000,
    "No OC use", 7, 10000
  ) |>
  mutate(`No MI` = Total - MI) |>
  relocate(`No MI`, .after = MI)

totals <-
  oc_mi |>
  summarize(across(c(MI, `No MI`, Total), sum)) |>
  mutate(OC = "Total")

tbl_oc_mi <- bind_rows(oc_mi, totals)

tbl_oc_mi |> pander::pander()
```

Simulated data from study of oral contraceptive use and heart attack risk

:::

:::::

{{< slidebreak >}}

::::{#exr-probs}

Estimate the probabilities of MI for OC users and non-OC users in @exm-oc-mi.
::::

{{< slidebreak >}}

:::{#sol-compute-probs}

```{r, include = FALSE}
#| label: calc-prs
p_MI_OC <- 13 / 5000 # nolint: object_name_linter
p_MI_nOC <- 7 / 10000 # nolint: object_name_linter
```

$$\ph(MI|OC) = \frac{13}{5000} = `r p_MI_OC`$$

$$\ph(MI|\neg OC) = \frac{7}{10000} = `r p_MI_nOC`$$

:::

{{< slidebreak >}}

:::{#exr-def-controls}

What does the term "controls" mean in the context of study design?

:::

{{< slidebreak >}}

:::{#sol-def-controls}
::::{#def-controls}
##### Two meanings of "controls"

Depending on context, "controls" can mean either:

- individuals who don't experience an *exposure* of interest,
- or individuals who don't experience an *outcome* of interest.

::::
:::

{{< slidebreak >}}

:::{#exr-controls-use-cases}
What types of studies do the two definitions of controls correspond to?
:::

{{< slidebreak >}}

:::{#sol-controls-use-cases}
:::{#def-cases-retrospective}
##### cases and controls in retrospective studies
In *retrospective case-control studies*,
participants who experience the outcome of interest are called **cases**,
while participants who don't experience that outcome are called **controls**.
:::

:::{#def-cases-prospective}
##### treatment groups and control groups in prospective studies
In *prospective studies*,
the group of participants who experience the treatment or exposure of interest
is called the **treatment group**,
while the participants who receive the baseline or comparison treatment
(for example, clinical trial participants who receive a placebo
or a standard-of-care treatment rather than an experimental treatment)
are called **controls**.

:::
:::

{{< slidebreak >}}

:::{#exm-cases-controls}

#### Cases and controls in the OC-MI study

@exm-oc-mi describes a *prospective* (follow-up) study:
participants were enrolled based on their exposure status
(OC use versus no OC use)
and then followed for three years to observe MI outcomes.

Therefore, @def-cases-prospective applies:
the 5000 OC users form the treatment (exposed) group,
and the 10,000 OC non-users are the **controls**.

@def-cases-retrospective does not apply to this study design,
because participants were not selected based on their outcomes.
If we had instead recruited participants by outcome status —
enrolling the 20 participants who experienced MIs,
along with a sample of participants who did not —
the study would have been a *retrospective case-control study*:
the 20 participants with MIs would be the **cases**,
and the sampled MI-free participants would be the **controls**.

:::
