Alternatively,
if you want to estimate risk ratios more directly from the model,
you can sometimes change the link function from $\logitf{}$ to $\logf{}$;
then you can obtain risk ratios by exponentiating coefficients
^[or linear combinations of coefficients,
depending on what covariate patterns you are contrasting],
just like we did for odds ratios with the logit link:

```{r}
data(anthers, package = "dobson")
anthers_sum <- aggregate(
  anthers[c("n", "y")],
  by = anthers[c("storage")], FUN = sum
)

anthers_glm_log <- glm(
  formula = cbind(y, n - y) ~ storage,
  data = anthers_sum,
  family = binomial(link = "log")
)

anthers_glm_log |>
  parameters() |>
  print_md()
```

{{< slidebreak >}}

Now $\exp{\beta}$ gives us risk ratios instead of odds ratios:

```{r}
anthers_glm_log |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

{{< slidebreak >}}

Let's compare this model with a logistic model:

```{r}
anthers_glm_logit <- glm(
  formula = cbind(y, n - y) ~ storage,
  data = anthers_sum,
  family = binomial(link = "logit")
)

anthers_glm_logit |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

{{< slidebreak >}}

We can compare the two link functions
by plotting their fitted germination probabilities
across the range of the storage covariate,
alongside the observed proportions:

:::{#fig-anthers-fitted-scales}

```{r}
#| code-fold: true
library(tidyr)

anthers_fitted_grid <-
  data.frame(
    storage = seq(
      min(anthers_sum$storage),
      max(anthers_sum$storage),
      length.out = 100
    )
  ) |>
  mutate(
    `Log link` = predict(anthers_glm_log, pick(storage), type = "response"),
    `Logit link` = predict(anthers_glm_logit, pick(storage), type = "response")
  ) |>
  pivot_longer(
    cols = c(`Log link`, `Logit link`),
    names_to = "Link function",
    values_to = "fitted"
  )

anthers_observed <-
  anthers_sum |>
  mutate(proportion = y / n)

ggplot() +
  geom_line(
    data = anthers_fitted_grid,
    mapping = aes(x = storage, y = fitted, color = `Link function`)
  ) +
  geom_point(
    data = anthers_observed,
    mapping = aes(x = storage, y = proportion),
    size = 3
  ) +
  labs(
    x = "Storage condition",
    y = "Fitted probability of germination"
  ) +
  theme_bw()
```

Fitted germination probabilities from the log-link and logit-link models for the `anthers` data,
with the observed proportions as points.
The two link functions nearly coincide here because the model is saturated
(two parameters fit to two storage conditions),
so both reproduce the observed proportions at the data points
and differ only slightly in between.
:::

{{< slidebreak >}}

In practice, fitting a binomial model with `link = "log"` often fails to converge,
producing errors about not finding good starting values for the estimation procedure.
This convergence failure is likely because the model is producing
fitted probabilities greater than 1.

When this happens,
you can instead switch the outcome distribution from binomial to Poisson
and fit a Poisson regression model
(we will see those soon!).
Note that this is a change in model *structure* —
a different outcome distribution —
not just a different way of fitting the same model.
With a Poisson outcome distribution,
you won't get warnings about fitted probabilities greater than 1,
but that distribution isn't quite right for binary outcomes.
Despite this mismatch,
Poisson regression is the usual fallback
when the log-binomial model fails to converge.

### WCGS: link functions

```{r}
wcgs_glm_logit_link <- chd_grouped_data |>
  mutate(type = relevel(dibpat, ref = "Type B")) |>
  glm(
    "formula" = cbind(x, `n - x`) ~ dibpat * age,
    "data" = _,
    "family" = binomial(link = "logit")
  )

wcgs_glm_identity_link <-
  chd_grouped_data |>
  mutate(type = relevel(dibpat, ref = "Type B")) |>
  glm(
    "formula" = cbind(x, `n - x`) ~ dibpat * age,
    "data" = _,
    "family" = binomial(link = "identity")
  )
wcgs_glm_identity_link |>
  coef() |>
  pander()
```

:::{#fig-diagnostics-glm1 layout-ncol=2}

:::{#fig-diagnostics-glm1-logit}

```{r}
#| code-fold: true

library(ggfortify)
wcgs_glm_logit_link |> autoplot(which = c(1), ncol = 1) + facet_wrap(~dibpat)
```

Logistic link

:::

:::{#fig-diagnostics-glm1-identity}

```{r}
#| code-fold: true

wcgs_glm_identity_link |> autoplot(which = c(1), ncol = 1) + facet_wrap(~dibpat)
```

Identity link

:::

Residuals vs Fitted plot for `wcgs` models

:::

:::{#fig-diagnostics-beetles layout-ncol=2}

:::{#fig-diagnostics-beetles-logit}

```{r}
#| code-fold: true

beetles_lm <-
  beetles_long |>
  lm(formula = died ~ dose)

beetles_glm_grouped <- beetles |>
  glm(formula = cbind(died, survived) ~ dose, family = "binomial")


beetles <-
  beetles |> mutate(
    resid_logit = beetles_glm_grouped |> resid(type = "response")
  )
beetles_glm_grouped |> autoplot(which = c(1), ncol = 1)
```

Logistic link

:::

:::{#fig-diagnostics-beetles-identity}

```{r}
#| code-fold: true

beetles_lm |> autoplot(which = c(1), ncol = 1)
```

Identity link

:::

Residuals vs Fitted plot for `BeetleMortality` models

:::
