
### Deviance test

<!-- slidebreak omitted: the section heading immediately precedes the div; do not re-flag -->
:::{#def-deviance-test}

#### Deviance goodness-of-fit test

We can compare the maximized log-likelihood of our model,
$\ell(\hat\beta; \mathbf x)$,
versus the log-likelihood of the full
model (aka saturated model aka maximal model), $\ell_{\text{full}}$,
which has one parameter per covariate pattern.
The **deviance** statistic is:

$$D \eqdef 2(\ell_{\text{full}} - \ell(\hat\beta; \mathbf x))$$

With enough data,
$D \dot \sim \chi^2(q - p)$,
where $q$ is the number of distinct covariate patterns
and $p$ is the number of $\beta$ parameters in our model.
The **deviance test** compares $D$ to that $\chi^2(q-p)$ distribution;
a significant p-value for this **deviance** statistic
indicates that there's some detectable pattern in the data
that our model isn't flexible enough to catch.

:::

::: callout-caution
The deviance statistic needs to have a large amount of data **for each
covariate pattern** for the $\chi^2$ approximation to hold. A guideline
from Dobson is that if there are $q$ distinct covariate patterns
$x_1...,x_q$, with $n_1,...,n_q$ observations per pattern, then the
expected frequencies $n_k \cdot \pi(x_k)$ should be at least 1 for every
pattern $k\in 1:q$.
:::

If you have covariates measured on a continuous scale, you may not be
able to use the deviance tests to assess goodness of fit.

{{< slidebreak >}}

:::{#exm-deviance-test-wcgs}

#### Deviance test for the WCGS CHD model

Our CHD model `chd_glm_contrasts` was fit to individual-level observations,
but its covariate patterns are just the combinations of
personality type (`dibpat`) and whole-year `age`,
so we can refit the same model to the grouped data
and perform the deviance test:

```{r}
#| label: deviance-test-wcgs

wcgs_gof_grouped <-
  wcgs |>
  summarize(
    .by = c(dibpat, age),
    n = n(),
    chd = sum(chd69 == "Yes"),
    no_chd = sum(chd69 == "No")
  )

chd_glm_gof_grouped <- glm(
  "formula" = cbind(chd, no_chd) ~ dibpat * age,
  "data" = wcgs_gof_grouped,
  "family" = binomial(link = "logit")
)

deviance_gof <- deviance(chd_glm_gof_grouped)
q_gof <- nrow(wcgs_gof_grouped)
p_gof <- length(coef(chd_glm_gof_grouped))
pval_deviance_gof <- pchisq(
  deviance_gof,
  df = q_gof - p_gof,
  lower.tail = FALSE
)
min_exp_freq <- min(wcgs_gof_grouped$n * fitted(chd_glm_gof_grouped))
```

The smallest expected event frequency across the covariate patterns is
$\min_k \left\{ n_k \cdot \hat\pi(x_k) \right\} = `r round(min_exp_freq, 2)`$,
which is at least 1,
so these grouped data meet
the minimum-expected-frequency guideline from @dobson4e
for using the $\chi^2$ approximation.

The grouped data have $q = `r q_gof`$ distinct covariate patterns,
and the model has $p = `r p_gof`$ parameters,
so the deviance statistic
$D = `r round(deviance_gof, 2)`$
has $q - p = `r q_gof - p_gof`$ degrees of freedom.
The corresponding p-value is
$p(\chi^2(`r q_gof - p_gof`) > `r round(deviance_gof, 2)`) = `r round(pval_deviance_gof, 3)`$,
so the deviance test finds no significant evidence of lack of fit
for this model.

:::

### Hosmer-Lemeshow test

If our covariate patterns produce groups that are too small, a
reasonable solution is to make bigger groups by merging some of the
covariate-pattern groups together.

{{< slidebreak >}}

:::{#def-hosmer-lemeshow}

#### Hosmer-Lemeshow test

Hosmer and Lemeshow (1980) proposed that we group the patterns by their
predicted probabilities according to the model of interest. For example,
you could group all of the observations with predicted probabilities of
10% or less together, then group the observations with 11%-20%
probability together, and so on; $g=10$ categories in all.

Then we can construct a statistic
$$X^2 \eqdef \sum_{c=1}^g \frac{(o_c - e_c)^2}{e_c}$$ where $o_c$ is the
number of events *observed* in group $c$, and $e_c$ is the number of
events expected in group $c$ (based on the sum of the fitted values
$\hat\pi_i$ for observations in group $c$).

If each group has enough observations in it, you can compare $X^2$ to a
$\chi^2$ distribution; by simulation, the degrees of freedom has been
found to be approximately $g-2$.

:::

{{< slidebreak >}}

::::{#exm-hosmer-lemeshow-wcgs}

#### Hosmer-Lemeshow test for the WCGS CHD model

For our CHD model, this procedure would be:

```{r}
wcgs <-
  wcgs |>
  mutate(
    pred_probs_glm1 = chd_glm_contrasts |> fitted(),
    pred_prob_cats1 = pred_probs_glm1 |>
      cut(
        breaks = seq(0, 1, by = .1),
        include.lowest = TRUE
      )
  )

HL_table <- # nolint: object_name_linter
  wcgs |>
  summarize(
    .by = pred_prob_cats1,
    n = n(),
    o = sum(chd69 == "Yes"),
    e = sum(pred_probs_glm1)
  )

library(pander)
HL_table |> pander()

X2 <- HL_table |> # nolint: object_name_linter
  summarize(
    `X^2` = sum((o - e)^2 / e)
  ) |>
  pull(`X^2`)
print(X2)

pval1 <- pchisq(X2, lower = FALSE, df = nrow(HL_table) - 2)
```

Our statistic is $X^2 = `r X2`$; $p(\chi^2(1) > `r X2`) = `r pval1`$,
which is our p-value for detecting a lack of goodness of fit.

Unfortunately that grouping plan left us with
just three categories with any observations,
so instead of grouping by 10% increments of predicted probability,
typically analysts use deciles of the predicted probabilities:

```{r}
wcgs <-
  wcgs |>
  mutate(
    pred_probs_glm1 = chd_glm_contrasts |> fitted(),
    pred_prob_cats1 = pred_probs_glm1 |>
      cut(
        breaks = quantile(pred_probs_glm1, seq(0, 1, by = .1)),
        include.lowest = TRUE
      )
  )

HL_table <- # nolint: object_name_linter
  wcgs |>
  summarize(
    .by = pred_prob_cats1,
    n = n(),
    o = sum(chd69 == "Yes"),
    e = sum(pred_probs_glm1)
  )

HL_table |> pander()

X2 <- HL_table |> # nolint: object_name_linter
  summarize(
    `X^2` = sum((o - e)^2 / e)
  ) |>
  pull(`X^2`)

print(X2)

pval1 <- pchisq(X2, lower = FALSE, df = nrow(HL_table) - 2)
```

Now we have more evenly split categories. The p-value is $`r pval1`$,
still not significant.

Graphically, we have compared:

```{r}
HL_plot <- # nolint: object_name_linter
  HL_table |>
  ggplot(aes(x = pred_prob_cats1)) +
  geom_line(
    aes(y = e, x = pred_prob_cats1, group = "Expected", col = "Expected")
  ) +
  geom_point(aes(y = e, size = n, col = "Expected")) +
  geom_point(aes(y = o, size = n, col = "Observed")) +
  geom_line(aes(y = o, col = "Observed", group = "Observed")) +
  scale_size(range = c(1, 4)) +
  theme_bw() +
  ylab("number of CHD events") +
  theme(axis.text.x = element_text(angle = 45))
```

::: {.content-visible when-format="html"}
```{r}
ggplotly(HL_plot)
```
:::

::: {.content-visible when-format="pdf"}
```{r}
print(HL_plot)
```
:::

::::

### Comparing models

The deviance test (@def-deviance-test)
and the Hosmer-Lemeshow test (@def-hosmer-lemeshow)
assess the fit of a single model in absolute terms.
To choose between two competing models,
we can instead compare their fits against each other,
using criteria that trade off fit against complexity:

-   AIC = $-2 * \ell(\hat\theta) + 2 * p$ \[lower is better\]
    (see [the AIC definition](Linear-models-overview.qmd#def-aic))
-   BIC = $-2 * \ell(\hat\theta) + p * \text{log}(n)$ \[lower is
better\]
    (see [the BIC definition](Linear-models-overview.qmd#def-bic))

A larger maximized log-likelihood is not, by itself,
evidence that one model is better than another:
adding parameters to a model can never decrease
its maximized log-likelihood,
so the more complex of two nested models always fits at least as well.
For **nested** models,
the likelihood ratio test accounts for this complexity difference
by testing whether the improvement in fit
is larger than chance alone would produce:

{{< slidebreak >}}

:::{#def-lrt}

#### Likelihood ratio test (LRT)

Suppose a **reduced** model with $p_0$ parameters
is nested in a **full** model with $p_1 > p_0$ parameters;
that is, the reduced model is a special case of the full model.
The **likelihood ratio test statistic** is:

$$\Lambda \eqdef 2(\ell_{\text{full}}(\hat\theta_1) - \ell_{\text{red}}(\hat\theta_0))$$

where $\hat\theta_1$ and $\hat\theta_0$ are
the maximum likelihood estimates
of the full and reduced models, respectively.
Under the null hypothesis that the reduced model is adequate,
$\Lambda \dot \sim \chi^2(p_1 - p_0)$
by [Wilks' theorem](intro-MLEs.qmd#thm-wilks).
The **likelihood ratio test (LRT)** rejects the reduced model
when $\Lambda$ is large relative to that $\chi^2(p_1 - p_0)$ distribution.

:::

::: notes
See also [the likelihood ratio test for linear models](Linear-models-overview.qmd#sec-linreg-LRT).
:::

{{< slidebreak >}}

:::{#exm-lrt-wcgs}

#### Testing the WCGS interaction term with a likelihood ratio test

Our CHD model `chd_glm_contrasts` includes a `dibpat:age` interaction term.
Does that interaction term improve the fit
enough to justify its extra parameter?
Let's fit the nested model without the interaction term
and compare the two models with a likelihood ratio test:

```{r}
#| label: lrt-wcgs

chd_glm_additive <-
  wcgs |>
  glm(
    "data" = _,
    "formula" = chd69 == "Yes" ~ dibpat + age,
    "family" = binomial(link = "logit")
  )

lrt_chd <- anova(
  chd_glm_additive,
  chd_glm_contrasts,
  test = "LRT"
)
lrt_chd

llik_full_chd <- logLik(chd_glm_contrasts) |> as.numeric()
llik_red_chd <- logLik(chd_glm_additive) |> as.numeric()
lambda_chd <- lrt_chd$Deviance[2]
df_chd <- lrt_chd$Df[2]
pval_lrt_chd <- lrt_chd$`Pr(>Chi)`[2]
```

Here the test statistic is:

$$
\ba
\Lambda
&\eqdef 2(\ell_{\text{full}}(\hat\theta_1) - \ell_{\text{red}}(\hat\theta_0))
&& \text{(definition of the LRT statistic)}
\\ &= 2 \times (`r round(llik_full_chd, 2)` - (`r round(llik_red_chd, 2)`))
&& \text{(substituting the two maximized log-likelihoods)}
\\ &= 2 \times `r round(llik_full_chd - llik_red_chd, 2)`
&& \text{(subtraction)}
\\ &= `r round(lambda_chd, 2)`
&& \text{(multiplication)}
\ea
$$

The full model has $p_1 = `r length(coef(chd_glm_contrasts))`$ parameters
and the reduced model has $p_0 = `r length(coef(chd_glm_additive))`$,
so the null distribution is
$\chi^2(p_1 - p_0) = \chi^2(`r df_chd`)$,
and the p-value is
$p(\chi^2(`r df_chd`) > `r round(lambda_chd, 2)`) = `r round(pval_lrt_chd, 3)`$.
The LRT finds no significant evidence
that the interaction term improves the fit.

The `lrtest()` function from the
[{lmtest}](https://cran.r-project.org/package=lmtest) package
performs the same test:

```{r}
#| label: lrt-wcgs-lmtest

lmtest::lrtest(chd_glm_contrasts, chd_glm_additive)
```

AIC and BIC can also compare these two models:

```{r}
#| label: aic-bic-wcgs

AIC(chd_glm_additive, chd_glm_contrasts)
BIC(chd_glm_additive, chd_glm_contrasts)
```

Both criteria are lower for the additive model
(AIC: $`r round(AIC(chd_glm_additive), 2)`$ vs $`r round(AIC(chd_glm_contrasts), 2)`$;
BIC: $`r round(BIC(chd_glm_additive), 2)`$ vs $`r round(BIC(chd_glm_contrasts), 2)`$),
agreeing with the LRT
that the interaction term is not worth its extra parameter.

:::
