Contrasting marginal risks across exposure levels
gives the corresponding marginal risk differences:

{{< slidebreak >}}

:::{#def-observed-marginal-rd}

#### Observed marginal risk difference

The **observed marginal risk difference** (**OMRD**) is:

$$
\text{OMRD} \eqdef \pi(1) - \pi(0)
$$

:::

{{< slidebreak >}}

:::{#exm-observed-marginal-rd}

#### Estimated observed marginal risk difference

Continuing from @exm-wcgs-observed-marginal-risks,
the observed marginal risk difference estimate is:

```{r}
#| label: wcgs-observed-marginal-rd
rd_obs <- pi1_obs - pi0_obs
```

$$
\ba
\widehat{\text{OMRD}}
&= \hat\pi(1) - \hat\pi(0)
&& \text{[definition of OMRD]}
\\
&= `r round(pi1_obs, 3)` - `r round(pi0_obs, 3)`
&& \text{[substitute estimates]}
\\
&= `r round(rd_obs, 3)`
&& \text{[arithmetic, computed from full-precision } \hat\pi(1) - \hat\pi(0) \text{]}
\ea
$$

:::

{{< slidebreak >}}

:::{#def-causal-marginal-rd}

#### Causal marginal risk difference

The **causal marginal risk difference** (**CMRD**) is:

$$
\text{CMRD} \eqdef \pi_1 - \pi_0
$$

:::

{{< slidebreak >}}

:::{#exm-causal-marginal-rd}

#### Estimated causal marginal risk difference

Continuing from @exm-wcgs-causal-marginal-risks,
the causal marginal risk difference estimate is:

```{r}
#| label: wcgs-causal-marginal-rd
rd_hat <- pi1_hat - pi0_hat
```

$$
\ba
\widehat{\text{CMRD}}
&= \hat\pi_1 - \hat\pi_0
&& \text{[definition of CMRD]}
\\
&= `r round(pi1_hat, 3)` - `r round(pi0_hat, 3)`
&& \text{[substitute estimates]}
\\
&= `r round(rd_hat, 3)`
&& \text{[arithmetic, computed from full-precision } \hat\pi_1 - \hat\pi_0 \text{]}
\ea
$$

:::
