
::: notes
When the outcome is rare (i.e., its probability is small) for both groups being compared in an odds ratio, the odds of the outcome will be similar to the probability of the outcome, and thus the risk ratio will be similar to the odds ratio.
:::

##### Case 1: rare events

For rare events, odds ratios and probability (a.k.a. risk, a.k.a.
prevalence) ratios will be close:

$$\pi_1 = .01$$
$$\pi_2 = .02$$

```{r "rare events or-rr"}
#| code-fold: show
pi1 <- .01
pi2 <- .02
pi2 / pi1
odds(pi2) / odds(pi1)
```

{{< slidebreak >}}

:::{#exm-or-rr-OC-MI}
In @exm-oc-mi, the outcome is rare for both OC and non-OC participants, so the odds for both groups are similar to the corresponding probabilities, and the odds ratio is similar the risk ratio.
:::

{{< slidebreak >}}

##### Case 2: frequent events

$$\pi_1 = .4$$

$$\pi_2 = .5$$

For more frequently-occurring outcomes, this won't be the case:

```{r}
#| code-fold: show
pi1 <- .4
pi2 <- .5
pi2 / pi1
odds(pi2) / odds(pi1)
```

{{< slidebreak >}}

{{< include _subfiles/logistic-regression/_fig-rd-rr-or.qmd >}}

{{< slidebreak >}}

Testing whether an odds ratio = 1 is equivalent to
testing whether the corresponding risk ratio = 1,
and also equivalent to
testing whether the risk difference = 0.
Therefore,
in **hypothesis testing**,
if the null hypothesis is no effect,
then we can shift between RD, RR, and OR.
But when we're talking about **point estimates** and **CIs**,
we need to limit our conclusions to the effect measure(s) that we actually estimated,
because the *sizes* of RDs, RRs, and ORs don't have a simple relationship to each other,
except when $\pi_1=\pi_2$ (as shown by @fig-rd-rr-or).
