{{< include _subfiles/math-prereqs/_calc_regularity_defs.qmd >}}

{{< slidebreak >}}

Before stating the Fundamental Theorem of Calculus, we record two
prerequisite results.
The FTC requires the integrand $f$ to be continuous,
and the two theorems below establish where continuity comes from
(differentiability $\Rightarrow$ continuity)
and what it buys us (continuity $\Rightarrow$ integrability).

:::{#thm-diff-implies-cont}
#### Differentiability implies continuity

If $f$ is differentiable at $x = c$, then $f$ is continuous at $x = c$.

[@larsonCalc11e, Theorem 2.1, p. 106]

:::

:::{#exm-diff-implies-cont}
#### Differentiable, hence continuous: $x^3 - x$

$f(x) = x^3 - x$ is differentiable everywhere
(with derivative $f'(x) = 3x^2 - 1$),
so by @thm-diff-implies-cont it is continuous everywhere.

:::

{{< slidebreak >}}

::::{#exm-cont-not-diff}
#### Continuous but not differentiable: $\abs{x}$

The absolute-value function $f(x) = \abs{x}$ is continuous at $x = 0$
($\lim_{x \to 0}\abs{x} = 0 = \abs{0}$),
but it is not differentiable at $x = 0$:
the left-derivative is $-1$ and the right-derivative is $+1$.

This shows that the converse of @thm-diff-implies-cont fails:
continuity does not imply differentiability.
See @fig-abs-value.

{{< slidebreak >}}

:::{#fig-abs-value}

```{r}
#| label: fig-abs-value-code
#| code-fold: true
ggplot() +
  geom_function(fun = abs, xlim = c(-2, 2), linewidth = 1) +
  geom_point(aes(x = 0, y = 0), size = 3) +
  labs(x = "x", y = expression(f(x) == abs(x))) +
  theme_minimal()
```

$f(x) = \abs{x}$ has a sharp corner at $x = 0$ (not differentiable there)
but is continuous everywhere: no gaps or jumps.

:::

::::

{{< slidebreak >}}

:::{#thm-cont-implies-int}
#### Continuity implies integrability

If $f$ is continuous on the closed interval $[a, b]$,
then $f$ is integrable on $[a, b]$
(i.e., the Riemann integral $\int_a^b f(x)\,dx$ exists and is finite).

[@larsonCalc11e, Theorem 4.4, p. 272]

:::

:::{#exm-cont-implies-int}
#### Continuous, hence integrable: polynomials

Every polynomial is continuous on $\mathbb{R}$,
so by @thm-cont-implies-int every polynomial is integrable on every
closed interval $[a, b]$.

:::

{{< slidebreak >}}

::::{#exm-int-not-cont}
#### Integrable but not continuous: a step function

Let $f(x) = 0$ for $x < \tfrac{1}{2}$ and $f(x) = 1$ for $x \ge \tfrac{1}{2}$.
Then $f$ is discontinuous at $x = \tfrac{1}{2}$,
but it is integrable on $[0, 1]$:

$$
\int_0^1 f(x)\,dx
= \int_0^{1/2} 0\,dx + \int_{1/2}^1 1\,dx
= 0 + \tfrac{1}{2}
= \tfrac{1}{2}.
$$

This shows that the converse of @thm-cont-implies-int fails:
integrability does not imply continuity.
See @fig-step.

{{< slidebreak >}}

:::{#fig-step}

```{r}
#| label: fig-step-code
#| code-fold: true
step_df <- data.frame(
  x = c(0, 0.5, 0.5, 1),
  y = c(0, 0, 1, 1),
  segment = c("left", "left", "right", "right")
)

ggplot() +
  geom_rect(
    aes(xmin = 0.5, xmax = 1, ymin = 0, ymax = 1),
    fill = "steelblue", alpha = 0.3
  ) +
  geom_line(
    data = step_df,
    aes(x = x, y = y, group = segment),
    linewidth = 1
  ) +
  geom_point(aes(x = 0.5, y = 0), shape = 1, size = 3) +
  geom_point(aes(x = 0.5, y = 1), shape = 16, size = 3) +
  scale_x_continuous(breaks = c(0, 0.5, 1), labels = c("0", "1/2", "1")) +
  scale_y_continuous(limits = c(-0.1, 1.2)) +
  labs(x = "x", y = "f(x)") +
  theme_minimal()
```

Step function: $f(x) = 0$ on $[0, \tfrac{1}{2})$ (open circle at the jump)
and $f(x) = 1$ on $[\tfrac{1}{2}, 1]$ (filled circle).
The shaded rectangle has area $\tfrac{1}{2}$, matching the integral
computed above.

:::

::::

{{< slidebreak >}}

Together, @thm-diff-implies-cont and @thm-cont-implies-int establish the chain:

$$\text{differentiable} \;\Rightarrow\; \text{continuous} \;\Rightarrow\; \text{integrable}$$

@exm-cont-not-diff and @exm-int-not-cont
show that neither implication reverses in general.
