:::{#thm-ftc}
#### Fundamental Theorem of Calculus

Let $f$ be a continuous function on a closed interval $[a, b]$.

**Part 1 (Derivative of an integral).**
Define $F(x) = \int_a^x f(t)\,dt$ for $x \in [a, b]$.
Then $F$ is differentiable and:

$$\deriv{x}\int_a^x f(t)\,dt = f(x)$$ {#eq-ftc-deriv-of-integral}

::: {.callout-note}
Continuity on all of $[a, b]$ is a sufficient condition.
More generally, Part 1 holds at any individual point $x$ where $f$ is
integrable on $[a, b]$ (see @def-integrable) and continuous at $x$
(see @def-continuous), even if $f$ has jump discontinuities elsewhere
[@rudin1976principles, Theorem 6.20, p. 133].
:::

[@larsonCalc11e, Theorem 4.11, p. 288]

**Part 2 (Evaluation theorem).**
The $F$ here may be *any* antiderivative of $f$ — not just the
accumulation function from Part 1.
If $F$ is an antiderivative of $f$ on $[a, b]$
(i.e., $\deriv{x} F(x) = f(x)$ for all $x \in [a, b]$), then:

$$\int_a^b f(x)\,dx = F(b) - F(a)$$ {#eq-ftc-part2}

Equivalently, with $b$ replaced by a variable upper limit $x$,
integrating the derivative of $F$ recovers the net change in $F$:

$$\int_a^x F'(t)\,dt = F(x) - F(a)$$ {#eq-ftc-integral-of-deriv}

or equivalently in Leibniz notation:

$$\int_a^x \dif{F}{t}\,dt = F(x) - F(a)$$

[@calclifesaver, Chapter 18; @larsonCalc11e, Theorem 4.9, p. 282]

:::

{{< slidebreak >}}

The two parts of the FTC together express that **differentiation and
integration are inverse operations**:

- Part 1: differentiating the integral of $f$ recovers $f$
  (@eq-ftc-deriv-of-integral).
- Part 2: the integral of $f$ over $[a, b]$ equals the difference of
  any antiderivative's values at the endpoints
  (@eq-ftc-part2),
  which rearranges to "integrating the derivative of $F$ recovers the
  net change in $F$" (@eq-ftc-integral-of-deriv).

The standard form of the FTC assumes $f$ is continuous on $[a, b]$;
continuity is *sufficient* but not strictly necessary (see the
preceding callout note for the more general statement).
Since differentiability implies continuity (@thm-diff-implies-cont),
the FTC applies in particular whenever $f$ is differentiable —
a common situation in Epi 204.

{{< slidebreak >}}

::::{#exm-ftc-part1-accumulation}
#### FTC Part 1 visualized: accumulation function for $f(t) = 2t$

Take $f(t) = 2t$ on $[0, 2]$. The accumulation function from $0$ is

$$F(x) \;\eqdef\; \int_0^x 2t\,dt \;=\; \sb{t^2}_{t=0}^{t=x} \;=\; x^2 - 0^2 \;=\; x^2,$$

so $F(x) = x^2$, and indeed $F'(x) = 2x = f(x)$, as @thm-ftc Part 1
predicts.
@fig-ftc-part1 shows the integrand on the left
(shaded area equals $F(x)$ at each $x$)
and the accumulation function $F(x) = x^2$ on the right
(its slope at $x$ equals $f(x) = 2x$).

{{< slidebreak >}}

```{r}
#| label: fig-ftc-part1-setup
#| include: false
x_marks <- c(1, 1.5, 2)
x_focus <- 1.5
```

:::{#fig-ftc-part1 layout-ncol="2"}

:::{#fig-ftc-part1-left}

```{r}
#| label: fig-ftc-part1-left-code
#| code-fold: true
ggplot() +
  geom_area(
    data = data.frame(t = seq(0, x_focus, length.out = 200)),
    aes(x = t, y = 2 * t),
    fill = "steelblue", alpha = 0.4
  ) +
  geom_function(fun = \(t) 2 * t, xlim = c(0, 2.2), linewidth = 1) +
  geom_vline(
    data = data.frame(x = x_marks),
    aes(xintercept = x, color = factor(x)),
    linetype = "dashed", linewidth = 0.6
  ) +
  labs(x = "t", y = "f(t) = 2t", color = "x") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

$f(t) = 2t$; shaded area equals $F(`r x_focus`) = `r x_focus^2`$; vertical lines mark $x \in \{`r paste(x_marks, collapse = ", ")`\}$.
:::

:::{#fig-ftc-part1-right}

```{r}
#| label: fig-ftc-part1-right-code
#| code-fold: true
slope_df <- data.frame(
  x = x_marks,
  Fx = x_marks^2,
  slope = 2 * x_marks
)

ggplot() +
  geom_function(fun = \(x) x^2, xlim = c(0, 2.2), linewidth = 1) +
  geom_point(
    data = slope_df,
    aes(x = x, y = Fx, color = factor(x)),
    size = 3
  ) +
  geom_segment(
    data = slope_df,
    aes(
      x = x - 0.3, xend = x + 0.3,
      y = Fx - 0.3 * slope, yend = Fx + 0.3 * slope,
      color = factor(x)
    ),
    linewidth = 0.8
  ) +
  labs(x = "x", y = expression(F(x) == x^2), color = "x") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

$F(x) = x^2$; tangent slope at each marked $x$ equals $f(x) = 2x$.
:::

Left: $f(t) = 2t$; the shaded area $\int_0^{`r x_focus`} 2t\,dt = F(`r x_focus`) = `r x_focus^2`$; vertical lines mark $x \in \{`r paste(x_marks, collapse = ", ")`\}$.
Right: $F(x) = x^2$; for each marked $x$, the tangent slope equals $f(x) = 2x$.

:::

::::

{{< slidebreak >}}

::::{#exm-ftc-cdf-pdf}
#### CDF and PDF of the exponential distribution

In what follows, $f$ denotes the PDF and $F$ the CDF — the same letters
as the antiderivative pair in @def-antiderivative, because the FTC will
show $F$ is exactly an antiderivative of $f$.

For the exponential distribution with rate parameter $\lambda > 0$,
the probability density function (PDF) is
[@kleinbaum2012survival, Section II, p. 295, "Survival and Hazard
Functions for Selected Distributions"]:

$$f(t) = \lambda \ef{-\lambda t}, \quad t \ge 0$$

**FTC Part 2** gives the cumulative distribution function (CDF) from the PDF.
Apply the $\ef{cx}$ rule from @def-integral-rules with $c = -\lambda$ to
antidifferentiate the integrand:

$$
\ba
F(t)
&= \int_0^t \lambda \ef{-\lambda u}\,du \\
&= \sb{\lambda \cd \frac{1}{-\lambda}\ef{-\lambda u}}_{u=0}^{u=t} \\
&= \sb{(-1)\ef{-\lambda u}}_{u=0}^{u=t} \\
&= \sb{-\ef{-\lambda u}}_{u=0}^{u=t} \\
&= -\ef{-\lambda t} - \paren{-\ef{0}} \\
&= -\ef{-\lambda t} - (-1) \\
&= 1 - \ef{-\lambda t}
\ea
$$

**FTC Part 1** recovers the PDF from the CDF:

$$
\ba
\deriv{t} F(t)
&= \deriv{t}\paren{1 - \ef{-\lambda t}} \\
&= 0 - (-\lambda)\ef{-\lambda t} \\
&= \lambda\ef{-\lambda t} \\
&= f(t)
\ea
$$

For a concrete instance: with $\lambda = 1$ (standard exponential),
the probability that $T \le 2$ is:

$$
F(2) = 1 - \ef{-1 \cd 2} = 1 - \ef{-2} \approx 1 - 0.135 = 0.865
$$

See @fig-exp-pdf-cdf.

```{r}
#| label: fig-exp-pdf-cdf-setup
#| include: false
lambda <- 1
t_focus <- 2
t_max <- 5
F_at_focus <- 1 - exp(-lambda * t_focus)
```

{{< slidebreak >}}

:::{#fig-exp-pdf-cdf layout-ncol="2"}

:::{#fig-exp-pdf-cdf-pdf}

```{r}
#| label: fig-exp-pdf-cdf-pdf-code
#| code-fold: true
ggplot() +
  geom_area(
    data = data.frame(t = seq(0, t_focus, length.out = 300)),
    aes(x = t, y = lambda * exp(-lambda * t)),
    fill = "steelblue", alpha = 0.4
  ) +
  geom_function(
    fun = \(t) lambda * exp(-lambda * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  labs(x = "t", y = "f(t)") +
  theme_minimal()
```

PDF with $\lambda = `r lambda`$; shaded area equals $F(`r t_focus`) \approx `r round(F_at_focus, 3)`$.
:::

:::{#fig-exp-pdf-cdf-cdf}

```{r}
#| label: fig-exp-pdf-cdf-cdf-code
#| code-fold: true
ggplot() +
  geom_function(
    fun = \(t) 1 - exp(-lambda * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  geom_point(
    aes(x = t_focus, y = F_at_focus),
    size = 3, color = "steelblue"
  ) +
  geom_segment(
    aes(x = t_focus, xend = t_focus, y = 0, yend = F_at_focus),
    linetype = "dashed", color = "steelblue"
  ) +
  geom_segment(
    aes(x = 0, xend = t_focus, y = F_at_focus, yend = F_at_focus),
    linetype = "dashed", color = "steelblue"
  ) +
  labs(x = "t", y = "F(t)") +
  theme_minimal()
```

CDF with $\lambda = `r lambda`$; point marks $F(`r t_focus`) \approx `r round(F_at_focus, 3)`$.
:::

Exponential distribution with $\lambda = `r lambda`$.
Left: the PDF $f(t) = \lambda \ef{-\lambda t}$; the shaded area under the curve from $0$ to $`r t_focus`$ equals $F(`r t_focus`) \approx `r round(F_at_focus, 3)`$.
Right: the CDF $F(t) = 1 - \ef{-\lambda t}$; the dashed lines mark the value $F(`r t_focus`)$ computed via FTC Part 2.

:::

::::
