:::{#def-antiderivative}
#### Antiderivative

A function $F$ is an **antiderivative** of $f$ on an interval $I$ if:

$$\deriv{x} F(x) = f(x), \quad \forall x \in I$$

The family of all antiderivatives of $f$ is written as the **indefinite integral**:

$$\int f(x)\,dx = F(x) + C$$

where $C$ is an arbitrary constant of integration.

[@larsonCalc11e, Section 4.1, pp. 248--249]

:::

{{< slidebreak >}}

::::{#exm-antiderivative-power}
#### Antiderivative of a power function

For $f(x) = x^2$, an antiderivative is $F(x) = \frac{x^3}{3}$,
since $\deriv{x}\frac{x^3}{3} = x^2 = f(x)$.

Adding any constant $C$ gives another antiderivative;
for example, with $C = 7$, $F(x) = \frac{x^3}{3} + 7$ also satisfies
$F'(x) = x^2$, since adding a constant does not change the derivative.
See @fig-antiderivatives.

{{< slidebreak >}}

```{r}
#| label: fig-antiderivatives-setup
#| include: false
C_vals <- c(-2, -1, 0, 1, 2)
x_lim <- c(-2, 2)
```

:::{#fig-antiderivatives layout-ncol="2"}

:::{#fig-antiderivatives-f}

```{r}
#| label: fig-antiderivatives-f-code
#| code-fold: true
ggplot() +
  geom_function(fun = \(x) x^2, xlim = x_lim, linewidth = 1) +
  labs(x = "x", y = expression(f(x))) +
  theme_minimal()
```

The function $f(x) = x^2$.
:::

:::{#fig-antiderivatives-F}

```{r}
#| label: fig-antiderivatives-F-code
#| code-fold: true
x_seq <- seq(x_lim[1], x_lim[2], length.out = 200)
df <- do.call(rbind, lapply(C_vals, \(C) {
  data.frame(
    x = x_seq,
    y = x_seq^3 / 3 + C,
    C = factor(C)
  )
}))

ggplot(df, aes(x = x, y = y, color = C)) +
  geom_line(linewidth = 0.8) +
  labs(x = "x", y = expression(F(x)), color = "C") +
  theme_minimal()
```

Family of antiderivatives $F(x) = x^3/3 + C$.
:::

The function $f(x) = x^2$ and five antiderivatives $F(x) = x^3/3 + C$ for $C \in \{`r paste(C_vals, collapse = ", ")`\}$.
Each antiderivative has the same derivative $f$; they differ only by a vertical shift.

:::

::::

{{< slidebreak >}}

:::{#def-integral-rules}
#### Basic integration rules

Each antiderivative below is defined only up to an arbitrary constant
$C$ (see @def-antiderivative);
the table omits $+ C$ from every row for brevity.

| Function $f(x)$ | Antiderivative $F(x)$ | Condition |
|:---:|:---:|:---|
| $c$ | $cx$ | — |
| $x^n$ | $\dfrac{x^{n+1}}{n+1}$ | $n \ne -1$ |
| $\dfrac{1}{x}$ | $\ln\abs{x}$ | $x \ne 0$ |
| $\ef{x}$ | $\ef{x}$ | (self-antiderivative) |
| $\ef{cx}$ | $\dfrac{1}{c}\ef{cx}$ | $c \ne 0$ |
| $c \cd f(x)$ | $c \cd F(x)$ | — |
| $f(x) + g(x)$ | $F(x) + G(x)$ | — |

The first two rows and the bottom two rows (linearity) are from
[@larsonCalc11e, Section 4.1, p. 250 "Basic Integration Rules"];
$1/x$ is from [@larsonCalc11e, Section 5.2, Theorem 5.5, p. 324];
$\ef{x}$ and $\ef{cx}$ are from
[@larsonCalc11e, Section 5.4, Theorem 5.12, p. 346].

:::

{{< slidebreak >}}

:::{#exm-integral-rules-quadratic}
#### Antiderivative of $3x^2 - 1$

By the power rule ($n = 2$) and linearity from @def-integral-rules:

$$
\int \paren{3x^2 - 1}\,dx
= 3 \cd \frac{x^3}{3} - x + C
= x^3 - x + C.
$$

Verify by differentiating:
$\deriv{x}\paren{x^3 - x + C} = 3x^2 - 1 = f(x)$, as required.

:::
