::::{#exm-fubini-nonrect}

#### Changing the order of integration for a non-rectangular region

Adapted from [@larsonCalc11e, Section 14.2, Example 4, pp. 984--985].

Let $X$ and $Y$ be independent $\operatorname{Uniform}(0, 1)$ random variables,
with joint density $f(x, y) = 1$ on the unit square $[0, 1]^2$.
Define the function $g(x, y) = \ef{-x^2}\,\indicp{y \le x}$,
and compute its expectation $\E{g(X, Y)}$.

Because the joint density equals $1$ on $[0, 1]^2$,
this expectation is the double integral of $g$ over the unit square:

$$\E{g(X, Y)} = \iint_{[0, 1]^2} g(x, y)\,dA.$$

The indicator factor $\indicp{y \le x}$ equals $1$ on the triangular
region where $y \le x$ and $0$ elsewhere,
so only that region,
namely $D = \{(x, y) : x \in [0, 1],\; y \in [0, x]\}$
(@fig-fubini-nonrect-region),
contributes, and there $g(x, y) = \ef{-x^2}$:

$$\E{g(X, Y)} = \iint_D \ef{-x^2}\,dA.$$

:::{#fig-fubini-nonrect-region}

```{r}
#| code-fold: true
#| message: false

region <- data.frame(x = c(0, 1, 1), y = c(0, 0, 1))
ggplot2::ggplot(region, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_polygon(
    fill = "steelblue", alpha = 0.4, color = "black", linewidth = 0.7
  ) +
  ggplot2::annotate(
    "text", x = 0.65, y = 0.28, label = "D", size = 8, fontface = "italic"
  ) +
  ggplot2::annotate(
    "text", x = 0.36, y = 0.52, label = "y == x",
    parse = TRUE, angle = 45
  ) +
  ggplot2::annotate(
    "text", x = 1.08, y = 0.5, label = "x == 1",
    parse = TRUE, angle = 90, hjust = 0.5
  ) +
  ggplot2::annotate(
    "text", x = 0.5, y = -0.1, label = "y == 0", parse = TRUE
  ) +
  ggplot2::scale_x_continuous(breaks = c(0, 0.5, 1), limits = c(-0.1, 1.3)) +
  ggplot2::scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(-0.2, 1.2)) +
  ggplot2::coord_equal() +
  ggplot2::labs(x = "x", y = "y") +
  ggplot2::theme_minimal()
```

Triangular integration region $D = \{(x, y) : x \in [0, 1],\; y \in [0, x]\}$,
bounded below by $y = 0$, above-left by $y = x$, and on the right by $x = 1$.

:::

**Order $dx\,dy$ is intractable.** Re-describing $D$ as
$D = \{(x, y) : y \in [0, 1],\; x \in [y, 1]\}$, the inner integral is

$$\int_y^1 \ef{-x^2}\,dx,$$

which has no elementary antiderivative.

**Order $dy\,dx$ works.** Applying @thm-fubini Part 1
($\ef{-x^2}$ is continuous and $D$ is the vertically simple region
$x \in [0, 1]$, $y \in [0, x]$):

$$
\ba
\iint_D \ef{-x^2}\,dA
&= \int_0^1\!\int_0^x \ef{-x^2}\,dy\,dx
\\&= \int_0^1 \ef{-x^2}\paren{\int_0^x dy}\,dx
\\&= \int_0^1 x\,\ef{-x^2}\,dx
\\&= \sb{-\frac{1}{2}\,\ef{-x^2}}_0^1
\\&= -\frac{1}{2}\paren{\ef{-1} - 1}
\\&= \frac{e - 1}{2e}
\\&\approx 0.316
\ea
$$

The solid whose volume equals this integral is shown in @fig-fubini-nonrect.

:::{#fig-fubini-nonrect}

```{r}
#| code-fold: true
#| message: false

n_grid <- 51
x_seq <- seq(0, 1, length.out = n_grid)
y_seq <- seq(0, 1, length.out = n_grid)

z_mat <- outer(x_seq, y_seq, function(x, y) {
  z <- exp(-x^2)
  z[y > x] <- NA
  z
})

plotly::plot_ly(x = ~x_seq, y = ~y_seq, z = ~t(z_mat)) |>
  plotly::add_surface(showscale = FALSE) |>
  plotly::layout(scene = list(
    xaxis = list(title = "x"),
    yaxis = list(title = "y"),
    zaxis = list(title = "z = exp(-x^2)"),
    camera = list(eye = list(x = 1.6, y = -1.6, z = 0.8))
  ))
```

Surface $z = e^{-x^2}$ over the region $D = \{(x, y) : x \in [0, 1],\; y \in [0, x]\}$.
The surface depends only on $x$ (constant in $y$), so for each $x$ the
inner integral over $y \in [0, x]$ contributes $x \cdot e^{-x^2}$.

:::

::::
