::::{#exm-fubini-rect}

#### Evaluating a double integral on a rectangle

Structure adapted from [@larsonCalc11e, Section 14.2, Example 2, pp. 982--983];
the integrand $x^2 + y^2$ is original, chosen so the integral equals
$\E{g(X, Y)}$ for $g(x, y) = x^2 + y^2$.

Let $X$ and $Y$ be independent $\operatorname{Uniform}(0, 1)$ random variables,
with joint density $f(x, y) = 1$ on the unit square
$R = \{(x, y) : x \in [0, 1],\; y \in [0, 1]\}$
(@fig-fubini-rect-region).
Define the function $g(x, y) = x^2 + y^2$,
and compute its expectation $\E{g(X, Y)}$.

Because the joint density equals $1$ on $R$,
this expectation is the double integral of $g$ over $R$:

$$\E{g(X, Y)} = \iint_R \paren{x^2 + y^2}\,dA.$$

:::{#fig-fubini-rect-region}

```{r}
#| code-fold: true
#| message: false

ggplot2::ggplot() +
  ggplot2::annotate(
    "rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1,
    fill = "steelblue", alpha = 0.4, color = "black", linewidth = 0.7
  ) +
  ggplot2::annotate(
    "text", x = 0.5, y = 0.5, label = "R", size = 8, fontface = "italic"
  ) +
  ggplot2::scale_x_continuous(breaks = c(0, 1), limits = c(-0.15, 1.25)) +
  ggplot2::scale_y_continuous(breaks = c(0, 1), limits = c(-0.15, 1.25)) +
  ggplot2::coord_equal() +
  ggplot2::labs(x = "x", y = "y") +
  ggplot2::theme_minimal()
```

Integration region $R = [0, 1]^2$, the unit square.

:::

The integrand is continuous on $R$, so @cor-fubini-rect applies and
either order of integration yields the same value.

**Integrating $y$ first, then $x$:**

$$
\ba
\E{g(X, Y)}
&= \int_0^1\!\int_0^1 \paren{x^2 + y^2}\,dy\,dx
\\&= \int_0^1 \sb{x^2 y + \frac{y^3}{3}}_0^1\,dx
\\&= \int_0^1 \paren{x^2 + \frac{1}{3}}\,dx
\\&= \sb{\frac{x^3}{3} + \frac{x}{3}}_0^1
\\&= \frac{2}{3}
\ea
$$

**Integrating $x$ first, then $y$** (verifying the order can be swapped):

$$
\ba
\int_0^1\!\int_0^1 \paren{x^2 + y^2}\,dx\,dy
&= \int_0^1 \sb{\frac{x^3}{3} + y^2 x}_0^1\,dy
\\&= \int_0^1 \paren{\frac{1}{3} + y^2}\,dy
\\&= \sb{\frac{y}{3} + \frac{y^3}{3}}_0^1
\\&= \frac{2}{3}
\ea
$$

Both orders give $\frac{2}{3}$, as @cor-fubini-rect guarantees.

As a cross-check, linearity of expectation gives the same value:
since $\E{X^2} = \int_0^1 x^2\,dx = \frac{1}{3}$ for $X \sim \operatorname{Uniform}(0, 1)$
(and likewise for $Y$),

$$\E{g(X, Y)} = \E{X^2 + Y^2} = \E{X^2} + \E{Y^2} = \frac{1}{3} + \frac{1}{3} = \frac{2}{3}.$$

The solid whose volume equals this integral is shown in @fig-fubini-rect.

:::{#fig-fubini-rect}

```{r}
#| code-fold: true
#| message: false

n_grid <- 41
x_seq <- seq(0, 1, length.out = n_grid)
y_seq <- seq(0, 1, length.out = n_grid)
z_mat <- outer(x_seq, y_seq, function(x, y) x^2 + y^2)

plotly::plot_ly(x = ~x_seq, y = ~y_seq, z = ~t(z_mat)) |>
  plotly::add_surface(showscale = FALSE) |>
  plotly::layout(scene = list(
    xaxis = list(title = "x"),
    yaxis = list(title = "y"),
    zaxis = list(title = "z", range = c(0, 2))
  ))
```

Surface $z = x^2 + y^2$ over the unit square $[0, 1]^2$. The double
integral $\tfrac{2}{3}$ is the volume between this surface and the
$xy$-plane, and equals $\E{X^2 + Y^2}$.

:::

::::
