
The **Fubini–Tonelli theorem** states conditions under which the order
of integration in a double integral can be exchanged.
We state two versions:
the Riemann version (@thm-fubini) is what Epi 204 directly uses
for double integrals of continuous functions on simple regions;
the σ-finite measure-theoretic version (@thm-fubini-tonelli) is
included to make the probability-measure corollaries
([probability-measure form](probability.qmd#cor-fubini-prob),
[joint-density form](probability.qmd#cor-fubini-joint))
in the probability chapter
follow from a stated theorem rather than from an aside.

{{< slidebreak >}}

:::{#thm-fubini}

### Fubini's theorem (Riemann version)

Let $f$ be **continuous** on a plane region $R \subseteq \mathbb{R}^2$.

1. **Vertically simple region.**
   If $R$ is defined by $a \le x \le b$ and $g_1(x) \le y \le g_2(x)$,
   where $g_1$ and $g_2$ are continuous on $[a, b]$, then

   $$\iint_R f(x, y)\,dA
   \;=\; \int_a^b \int_{g_1(x)}^{g_2(x)} f(x, y)\,dy\,dx.$$

2. **Horizontally simple region.**
   If $R$ is defined by $c \le y \le d$ and $h_1(y) \le x \le h_2(y)$,
   where $h_1$ and $h_2$ are continuous on $[c, d]$, then

   $$\iint_R f(x, y)\,dA
   \;=\; \int_c^d \int_{h_1(y)}^{h_2(y)} f(x, y)\,dx\,dy.$$

When $R$ can be described both ways, the two iterated integrals are
equal — so the order of integration can be exchanged.

[@larsonCalc11e, Theorem 14.2, p. 982]

:::

{{< slidebreak >}}

:::{#thm-fubini-tonelli}

### Fubini–Tonelli theorem (measure-theoretic form)

Let $(\Omega_1, \mathcal F_1, \mu_1)$ and $(\Omega_2, \mathcal F_2, \mu_2)$
be **σ-finite** measure spaces,
and let $f : \Omega_1 \times \Omega_2 \to \mathbb{R}$ be measurable
with respect to the product σ-algebra
$\mathcal F_1 \otimes \mathcal F_2$.
If either

(a) $f \ge 0$ almost everywhere with respect to $\mu_1 \otimes \mu_2$
(**Tonelli's theorem**), or

(b) $\int_{\Omega_1 \times \Omega_2} \abs{f}\,d(\mu_1 \otimes \mu_2) < \infty$
(**Fubini's theorem**),

then both iterated integrals exist, agree with the double integral,
and equal each other:

$$
\int_{\Omega_1 \times \Omega_2} f\,d(\mu_1 \otimes \mu_2)
= \int_{\Omega_1} \paren{\int_{\Omega_2} f(\omega_1, \omega_2)\,d\mu_2(\omega_2)}\,d\mu_1(\omega_1)
= \int_{\Omega_2} \paren{\int_{\Omega_1} f(\omega_1, \omega_2)\,d\mu_1(\omega_1)}\,d\mu_2(\omega_2).
$$

[@billingsley1995probability, Theorem 18.3; @gut2013, Theorem 9.1, p. 65;
@fubini1907; @wp:fubini]

:::

::: notes
This generalization is not required for Epi 204 itself, but it is
what justifies the probability-measure corollaries
([probability-measure form](probability.qmd#cor-fubini-prob),
[joint-density form](probability.qmd#cor-fubini-joint))
used later in the probability
chapter: probability measures are finite (hence σ-finite), so the
σ-finiteness hypothesis is automatic. The integrability conditions
(nonnegativity or absolute integrability) still need to be verified
in each application.
:::

{{< slidebreak >}}

:::{#cor-fubini-rect}

### Continuous functions on a rectangle (corollary of @thm-fubini)

If $f : [a, b] \times [c, d] \to \mathbb{R}$ is **continuous** on the
closed bounded rectangle $[a, b] \times [c, d]$, then:

$$\int_a^b \paren{\int_c^d f(x, y)\,dy}\,dx
= \int_c^d \paren{\int_a^b f(x, y)\,dx}\,dy
= \iint_{[a,b]\times[c,d]} f(x, y)\,dx\,dy.$$

[@larsonCalc11e, Theorem 14.2, p. 982]

:::

::: proof

A closed bounded rectangle $[a, b] \times [c, d]$ is both vertically
simple (with $g_1 \equiv c$, $g_2 \equiv d$) and horizontally simple
(with $h_1 \equiv a$, $h_2 \equiv b$).
Applying both parts of @thm-fubini to $f$ on this rectangle gives the
two iterated forms shown.

:::

{{< slidebreak >}}

::::{#exm-fubini-rect}

#### Evaluating a double integral on a rectangle

Adapted from [@larsonCalc11e, Section 14.2, Example 2, pp. 982--983].

Evaluate $\displaystyle\iint_R \paren{1 - \frac{1}{2}x^2 - \frac{1}{2}y^2}\,dA$
on the unit square $R = \{(x, y) : 0 \le x \le 1,\; 0 \le y \le 1\}$.

The integrand is continuous on $R$, so @cor-fubini-rect applies and
either order of integration yields the same value.

**Integrating $y$ first, then $x$** (the order Larson chooses):

$$
\ba
\iint_R \paren{1 - \frac{1}{2}x^2 - \frac{1}{2}y^2}\,dA
&= \int_0^1\!\int_0^1 \paren{1 - \frac{1}{2}x^2 - \frac{1}{2}y^2}\,dy\,dx
\\&= \int_0^1 \sb{\paren{1 - \frac{1}{2}x^2}y - \frac{y^3}{6}}_0^1\,dx
\\&= \int_0^1 \paren{\frac{5}{6} - \frac{1}{2}x^2}\,dx
\\&= \sb{\frac{5}{6}x - \frac{x^3}{6}}_0^1
\\&= \frac{2}{3}
\ea
$$

**Integrating $x$ first, then $y$** (verifying the order can be swapped):

The integrand is symmetric in $x$ and $y$, so the same arithmetic with
the roles swapped gives:

$$
\int_0^1\!\int_0^1 \paren{1 - \frac{1}{2}x^2 - \frac{1}{2}y^2}\,dx\,dy
= \frac{2}{3}
$$

Both orders give $\frac{2}{3}$, as @cor-fubini-rect guarantees.

:::{#fig-fubini-rect}

```{r}
#| code-fold: true
#| message: false

n_grid <- 41
x_seq <- seq(0, 1, length.out = n_grid)
y_seq <- seq(0, 1, length.out = n_grid)
z_mat <- outer(x_seq, y_seq, function(x, y) 1 - x^2 / 2 - y^2 / 2)

plotly::plot_ly(x = ~x_seq, y = ~y_seq, z = ~t(z_mat)) |>
  plotly::add_surface(showscale = FALSE) |>
  plotly::layout(scene = list(
    xaxis = list(title = "x"),
    yaxis = list(title = "y"),
    zaxis = list(title = "z", range = c(0, 1))
  ))
```

Surface $z = 1 - \tfrac{1}{2}x^2 - \tfrac{1}{2}y^2$ over the unit square
$[0, 1]^2$. The double integral $\tfrac{2}{3}$ is the volume between
this surface and the $xy$-plane.

:::

::::

{{< slidebreak >}}

::::{#exm-fubini-nonrect}

#### Changing the order of integration for a non-rectangular region

Adapted from [@larsonCalc11e, Section 14.2, Example 4, pp. 984--985].

Find the volume of the solid bounded by the surface $z = \ef{-x^2}$
and the planes $z = 0$, $y = 0$, $y = x$, and $x = 1$.

The base of the solid in the $xy$-plane is the triangular region
$D = \{(x, y) : 0 \le x \le 1,\; 0 \le y \le x\}$,
so the volume is

$$\iint_D \ef{-x^2}\,dA.$$

**Order $dx\,dy$ is intractable.** Re-describing $D$ as
$D = \{(x, y) : 0 \le y \le 1,\; y \le x \le 1\}$, the inner integral is

$$\int_y^1 \ef{-x^2}\,dx,$$

which has no elementary antiderivative.

**Order $dy\,dx$ works.** Applying @thm-fubini Part 1
($\ef{-x^2}$ is continuous and $D$ is the vertically simple region
$0 \le x \le 1$, $0 \le y \le x$):

$$
\ba
\iint_D \ef{-x^2}\,dA
&= \int_0^1\!\int_0^x \ef{-x^2}\,dy\,dx
\\&= \int_0^1 \ef{-x^2}\paren{\int_0^x dy}\,dx
\\&= \int_0^1 x\,\ef{-x^2}\,dx
\\&= \sb{-\frac{1}{2}\,\ef{-x^2}}_0^1
\\&= -\frac{1}{2}\paren{\ef{-1} - 1}
\\&= \frac{e - 1}{2e}
\\&\approx 0.316
\ea
$$

:::{#fig-fubini-nonrect}

```{r}
#| code-fold: true
#| message: false

n_grid <- 51
x_seq <- seq(0, 1, length.out = n_grid)
y_seq <- seq(0, 1, length.out = n_grid)

z_mat <- outer(x_seq, y_seq, function(x, y) {
  z <- exp(-x^2)
  z[y > x] <- NA
  z
})

plotly::plot_ly(x = ~x_seq, y = ~y_seq, z = ~t(z_mat)) |>
  plotly::add_surface(showscale = FALSE) |>
  plotly::layout(scene = list(
    xaxis = list(title = "x"),
    yaxis = list(title = "y"),
    zaxis = list(title = "z = exp(-x^2)"),
    camera = list(eye = list(x = 1.6, y = -1.6, z = 0.8))
  ))
```

Surface $z = e^{-x^2}$ over the triangular region $0 \le y \le x \le 1$.
The surface depends only on $x$ (constant in $y$), so for each $x$ the
inner integral over $y \in [0, x]$ contributes $x \cdot e^{-x^2}$.

:::

::::

{{< slidebreak >}}

::::{#exm-fubini-fail}

#### When conditions fail: a counterexample

The conditions in @thm-fubini are not merely technical —
when they fail, iterated integrals can exist yet disagree.

Let
$$f(x, y) = \frac{x^2 - y^2}{(x^2 + y^2)^2}$$
on $R = [0, 1] \times [0, 1]$.

**Integrating $y$ first, then $x$:**

Using $\displaystyle\frac{\partial}{\partial y}\frac{y}{x^2 + y^2}
= \frac{x^2 - y^2}{(x^2 + y^2)^2}$:

$$
\ba
\int_0^1\!\int_0^1 f(x, y)\,dy\,dx
&= \int_0^1 \sb{\frac{y}{x^2 + y^2}}_{y=0}^{y=1}\,dx
\\&= \int_0^1 \frac{1}{x^2 + 1}\,dx
\\&= \sb{\arctan(x)}_0^1
\\&= \frac{\pi}{4}
\ea
$$

**Integrating $x$ first, then $y$:**

Using $\displaystyle\frac{\partial}{\partial x}\paren{-\frac{x}{x^2 + y^2}}
= \frac{x^2 - y^2}{(x^2 + y^2)^2}$:

$$
\ba
\int_0^1\!\int_0^1 f(x, y)\,dx\,dy
&= \int_0^1 \sb{-\frac{x}{x^2 + y^2}}_{x=0}^{x=1}\,dy
\\&= \int_0^1 \paren{-\frac{1}{1 + y^2}}\,dy
\\&= -\sb{\arctan(y)}_0^1
\\&= -\frac{\pi}{4}
\ea
$$

**Conclusion:** $\dfrac{\pi}{4} \neq -\dfrac{\pi}{4}$,
so the two iterated integrals are unequal.
Neither @thm-fubini nor @cor-fubini-rect applies here.

**Why the hypotheses fail:**
@thm-fubini requires $f$ to be **continuous** on $R$.
The denominator $(x^2 + y^2)^2$ vanishes at the origin $(0, 0) \in R$,
so $f$ is *not even defined* there — let alone continuous — and the
theorem does not apply.

For @thm-fubini-tonelli, the failure shows up as
$\iint_R |f|\,dA = \infty$, which violates the absolute-integrability
hypothesis (b).
Switching to polar coordinates $(r, \theta)$ near the origin, the
integrand satisfies
$|f(x, y)| = \abs{x^2 - y^2}/(x^2 + y^2)^2 = \abs{\cos 2\theta}/r^2$,
so

$$
\iint_R |f|\,dA
\;\ge\; \int_0^{\pi/2}\!\int_0^{\epsilon}
   \frac{\abs{\cos 2\theta}}{r^2}\, r\,dr\,d\theta
\;=\; \paren{\int_0^{\pi/2}\abs{\cos 2\theta}\,d\theta}
       \int_0^{\epsilon} \frac{dr}{r}
\;=\; +\infty,
$$

since $\int_0^{\epsilon} dr/r$ diverges.

[@wp:fubini]

:::{#fig-fubini-fail}

```{r}
#| code-fold: true
#| message: false

n_grid <- 81
eps <- 0.04
x_seq <- seq(eps, 1, length.out = n_grid)
y_seq <- seq(eps, 1, length.out = n_grid)

z_mat <- outer(x_seq, y_seq, function(x, y) (x^2 - y^2) / (x^2 + y^2)^2)
z_clip <- 50
z_mat[z_mat >  z_clip] <-  z_clip
z_mat[z_mat < -z_clip] <- -z_clip

plotly::plot_ly(x = ~x_seq, y = ~y_seq, z = ~t(z_mat)) |>
  plotly::add_surface(
    showscale = FALSE,
    colorscale = list(
      list(0,   "#3b4cc0"),
      list(0.5, "#dddddd"),
      list(1,   "#b40426")
    )
  ) |>
  plotly::layout(scene = list(
    xaxis = list(title = "x"),
    yaxis = list(title = "y"),
    zaxis = list(title = "f(x, y)", range = c(-z_clip, z_clip)),
    camera = list(eye = list(x = 1.6, y = 1.6, z = 0.6))
  ))
```

Surface $f(x, y) = (x^2 - y^2)/(x^2 + y^2)^2$ on $[0, 1]^2$,
sampled away from the origin and clipped to $[-50, 50]$ for display.
The function diverges to $+\infty$ along the $x$-axis (red ridge,
$f > 0$ when $|x| > |y|$) and to $-\infty$ along the $y$-axis (blue
ridge, $f < 0$ when $|y| > |x|$). The singularity at $(0, 0)$ is why
$\iint_R |f|\,dA = \infty$ and Fubini–Tonelli's absolute-integrability
hypothesis fails.

:::

::::
