{{< include latex-macros/macros.qmd >}}

Under the causal assumptions
(consistency, conditional exchangeability, and positivity),
the causal effect can be estimated
by adjusting for the confounders $\vec{L}$
in a regression model.

## Direct standardization (G-computation) {#sec-causal-g-computation}

The **G-computation** estimator
(also called **direct standardization** or **standardization**)
estimates the ATE by:

1. Fitting a regression model:
   $\hat{\mu}(a, \vec{l}) = \widehat{\E{Y \mid A = a, \vec{L} = \vec{l}}}$

2. Predicting the potential outcome mean for each individual
   under both treatment levels $a = 1$ and $a = 0$:
   $$\hat{Y}_i(a) = \hat{\mu}(a, \vec{L}_i)$$

3. Averaging over the study population:
   $$\widehat{\E{Y(a)}} = \frac{1}{n} \sum_{i=1}^n \hat{Y}_i(a)$$

4. Estimating the ATE as the contrast:
   $$\widehat{\text{ATE}} = \widehat{\E{Y(1)}} - \widehat{\E{Y(0)}}$$

::: notes

G-computation is model-based:
it relies on the regression model being correctly specified.
A misspecified model will produce biased causal estimates
even if confounders are included.

G-computation is easily generalized to:

- Binary outcomes (using logistic regression)
- Survival outcomes (using Cox or Weibull models)
- Multiple treatments or treatment levels

:::

## Regression adjustment in linear models {#sec-causal-linear-adjustment}

When the outcome is continuous
and the regression model is linear,
G-computation simplifies to the
**adjusted treatment coefficient** from a linear regression:

$$Y = \beta_0 + \beta_A A + \beta_L \vec{L} + \varepsilon$$

Under consistency, conditional exchangeability,
and positivity,
the coefficient $\beta_A$ estimates
the average causal effect of $A$ on $Y$
(the ATE, if the model is correctly specified
and effect modification by $\vec{L}$ is absent).

::: notes

The regression adjustment estimator for the ATE
is equivalent to G-computation
when the outcome model is linear and has no treatment-covariate interactions.
When there are interactions,
G-computation and regression adjustment can give different estimates.
G-computation is preferred in that case.

:::

## Example: regression adjustment for LDL in HERS {#sec-causal-hers-ldl-adj}

The HERS study DAG (see @sec-pred-sel-hers-dag)
shows that `LDL` and `HDL` are mediators of the HT → CHD pathway.
For estimating the total causal effect of HT on CHD,
we should not adjust for `LDL` or `HDL`.

However, to estimate the causal effect of **HT on LDL** itself,
we use a separate regression model
with `LDL` as the outcome.
In this model,
`LDL` is no longer a mediator but the outcome of interest.

Since HT was randomized in HERS,
no adjustment is needed for causal identification of HT → LDL.
However, adjusting for strong predictors of LDL
improves precision:

```{r}
#| label: causal-hers-ldl-setup
#| code-fold: true

library(haven)
library(dplyr)

hers <- read_stata(fs::path_package("rme", "extdata/hersdata.dta"))

hers_ldl <-
  hers |>
  filter(!is.na(LDL)) |>
  mutate(
    HT = factor(HT, levels = c(0, 1), labels = c("Placebo", "HT")),
    statins = factor(statins, levels = c(0, 1), labels = c("No", "Yes")),
    diabetes = factor(diabetes, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

::: {#tbl-causal-hers-ldl-adj}

```{r}
#| code-fold: true

library(gtsummary)

model_ldl_adj <-
  lm(LDL ~ HT + age + BMI + diabetes + statins, data = hers_ldl)

model_ldl_adj |>
  tbl_regression() |>
  bold_labels()
```

Adjusted regression for the causal effect of HT on LDL in HERS.
Since HT was randomized, adjustment is for precision only.

:::

The estimated causal effect of HT on LDL
is approximately −11 mg/dL (negative = lower LDL on HT),
with narrower confidence intervals than the unadjusted estimate.
