{{< include latex-macros/macros.qmd >}}

**Propensity score methods** are an alternative approach
to confounding adjustment in observational studies.
They replace adjustment for a high-dimensional confounder set $\vec{L}$
with adjustment for a single scalar summary:
the propensity score.

:::{#def-propensity-score}
#### Propensity score

The **propensity score** is the conditional probability of treatment
given the observed covariates:
$$e(\vec{L}) = \P\paren{A = 1 \mid \vec{L}}$$

@rosenbaum1983central showed that
if $A \perp\!\!\!\perp (Y(0), Y(1)) \mid \vec{L}$
(conditional exchangeability given $\vec{L}$),
then also:
$$A \perp\!\!\!\perp (Y(0), Y(1)) \mid e(\vec{L})$$

That is, the propensity score is a **balancing score**:
conditioning on the propensity score is sufficient
to remove confounding.

:::

## Propensity score estimation {#sec-ps-estimation}

The propensity score is estimated by fitting a model
for the probability of treatment given covariates,
typically using logistic regression:

$$\text{logit}(\hat{e}(\vec{L}_i)) =
\hat{\beta}_0 + \hat{\beta}_1 L_{i1} + \cdots + \hat{\beta}_p L_{ip}$$

The estimated propensity score $\hat{e}(\vec{L}_i)$
is the predicted probability of treatment for each individual.

## Propensity score methods {#sec-ps-methods}

There are four main ways to use the propensity score:

1. **Matching**: Match each treated individual to one or more
   untreated individuals with similar propensity scores.
   The matched dataset is then analyzed as a simple comparison.

2. **Stratification (subclassification)**: Divide individuals
   into strata based on propensity score quantiles.
   Estimate the causal effect within each stratum,
   then combine across strata.

3. **Inverse probability weighting (IPW)**:
   Weight each individual by the inverse of their probability
   of receiving the treatment they actually received.
   See @def-ipw below.

4. **Covariate adjustment using the propensity score**:
   Include the estimated propensity score as a covariate
   in a regression model for the outcome.

:::{#def-ipw}
#### Inverse probability weighting (IPW)

The **inverse probability weighted (IPW)** estimator
of the ATE uses weights:
$$w_i = \frac{A_i}{\hat{e}(\vec{L}_i)} +
\frac{1 - A_i}{1 - \hat{e}(\vec{L}_i)}$$

The IPW estimator of $\E{Y(a)}$ is:
$$\widehat{\E{Y(a)}}_{\text{IPW}} =
\frac{\sum_{i: A_i = a} w_i Y_i}{\sum_{i: A_i = a} w_i}$$

Reweighting by $w_i$ creates a **pseudo-population**
in which treatment is independent of covariates,
effectively removing confounding.

:::

## Propensity score example: WCGS {#sec-ps-wcgs}

In the WCGS study,
we can use propensity score methods
to estimate the causal effect of Type A behavior on CHD.
We first fit a logistic regression model
predicting Type A behavior from the confounders
identified by the DAG (@sec-pred-sel-wcgs-dag):

```{r}
#| label: ps-wcgs-setup
#| code-fold: true

library(dplyr)

here::here() |>
  fs::path("Data/wcgs.rda") |>
  load()

wcgs_ps <-
  wcgs |>
  mutate(
    dibpat_A = as.integer(dibpat == "Type A"),
    chd69_yn = as.integer(chd69 == "Yes")
  ) |>
  filter(
    !is.na(age), !is.na(dibpat), !is.na(chd69),
    !is.na(bmi), !is.na(sbp), !is.na(chol), !is.na(smoke)
  )
```

```{r}
#| label: ps-wcgs-model
#| code-fold: true

ps_model <- glm(
  dibpat_A ~ age + bmi + sbp + chol + smoke,
  data = wcgs_ps,
  family = binomial()
)

wcgs_ps <- wcgs_ps |>
  mutate(
    ps = predict(ps_model, type = "response"),
    ipw = dibpat_A / ps + (1 - dibpat_A) / (1 - ps)
  )
```

::: {#fig-ps-wcgs}

```{r}
#| code-fold: true

library(ggplot2)

ggplot(wcgs_ps, aes(x = ps, fill = factor(dibpat_A))) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  scale_fill_manual(
    values = c("steelblue", "tomato"),
    labels = c("Type B", "Type A"),
    name = "Personality type"
  ) +
  labs(
    x = "Estimated propensity score",
    y = "Count",
    title = "Propensity score distributions by personality type (WCGS)"
  ) +
  theme_bw()
```

Distribution of estimated propensity scores
by personality type in the WCGS study.
Good overlap between the two groups
(the **overlap** or **common support** assumption)
supports positivity.

:::

```{r}
#| label: ps-wcgs-ipw
#| code-fold: true

library(gtsummary)

outcome_ipw <- glm(
  chd69_yn ~ dibpat_A,
  data = wcgs_ps,
  weights = ipw,
  family = quasibinomial()
)

outcome_ipw |>
  tbl_regression(exponentiate = TRUE) |>
  bold_labels()
```

::: notes

The IPW-weighted odds ratio for Type A vs. Type B behavior
estimates the causal odds ratio for CHD,
adjusting for confounding by age, BMI, SBP, cholesterol, and smoking.

Compare this to the unadjusted estimate:
the difference reflects the degree of confounding
in the WCGS data.

:::
