{{< include latex-macros/macros.qmd >}}

This section is adapted from [@vittinghoff2e, Chapter 6].

### What Are Competing Risks?

::: {#def-competing-risks}

#### Competing Risks Data

**Competing risks data** arise when multiple events can occur,
and follow-up can end due to occurrence of one or more of those types of events,
precluding observation of at least one of the other event types.

:::

Competing risks are common in medical research.
For example,
in a study of bone fracture risk in elderly men,
participants may:

- Experience the event of interest (fracture)
- Die before experiencing a fracture (competing event)
- Be lost to follow-up or reach the end of the study period (censored)

#### Why Competing Risks Matter

When analyzing time-to-event data,
standard survival analysis methods make assumptions about censoring.
The **independent censoring assumption** assumes that
the future risk of censored observations
can be represented by those who remain under follow-up.

This assumption is reasonable for:

- Administrative censoring (end of study)
- Loss to follow-up (if unrelated to outcome risk)

However,
this assumption is questionable for competing events like death,
because:

1. We cannot extrapolate beyond participant lifetimes
2. Death fundamentally alters the risk set for the primary outcome
3. Projecting to a setting "without death" creates a hypothetical population

#### Approaches to Analyzing Competing Risks Data

Two main approaches exist for analyzing competing risks:

1. **Elimination approach**:
   Extrapolates to a scenario where a competing event is not possible
   (appropriate for losses to follow-up)

2. **Accommodation approach**:
   Acknowledges and allows for the competing risks in the analysis
   (appropriate for death and other true competing events)

### Notation for Competing Risks

We denote competing risks outcome data using two variables:

- $Y$: time of the first observed event of any type
- $\delta = k$: if the $k$-th event type occurs first

where each of the $K$ possible types of failure
are denoted by a numerical code.

**Example coding scheme:**

- 0: loss to follow-up (censored)
- 1: event of interest (e.g., fracture)
- 2: competing event (e.g., death)

A participant followed for 18 months who dies
before experiencing a fracture
would have $Y = 18$ months and $\delta = 2$.

### Summaries for Competing Risks Data

#### Cause-Specific Hazard Functions

::: {#def-cause-specific-hazard}

#### Cause-Specific Hazard Function

The **cause-specific hazard function** for event type $k$,
denoted $h_k(t)$,
is the short-term rate at which participants experience
the onset of the $k$-th event
among those who have not yet experienced
the event of interest or a competing event prior to time $t$.

:::

**Key properties:**

- The numerator counts only events of type $k$
- The denominator includes all participants
  who could have developed the event by time $t$
- Reduces to the ordinary hazard function when there is only one failure type

**Estimation:**

To estimate and model cause-specific hazard functions:

1. Set up data as ordinary survival data
2. Define the $k$-th failure type as the only "event"
3. Treat all competing causes (including death) as "censored"

You can then examine predictor effects on the cause-specific hazard
using standard Cox proportional hazards models.

#### Cumulative Incidence Functions

::: {#def-cumulative-incidence}

#### Cumulative Incidence Function

The **cumulative incidence function** (CIF) for cause type $k$ at time $t$,
denoted $F_k(t)$,
is the proportion of the population
who have experienced the $k$-th event prior to time $t$.

:::

The cumulative incidence function:

- Measures the prevalence of a particular event at each time $t$
- Accounts for all competing risks
- Differs from cause-specific hazards in how it treats competing events

```{r}
#| label: mros-cif-setup
#| include: false
library(rmb)
library(cmprsk)
library(survival)
library(ggplot2)
library(dplyr)

mros_data <- mros

# Compute cumulative incidence functions (CIF) for each event type
# (no group argument, so there is exactly one group)
mros_cif <- cuminc(
  ftime   = mros_data$years,
  fstatus = mros_data$status
)

# Extract CIF by cause code using endsWith() for robustness
cif_names <- names(mros_cif)[names(mros_cif) != "Tests"]
fracture_name <- cif_names[endsWith(cif_names, " 1")]  # cause 1 = fracture
death_name    <- cif_names[endsWith(cif_names, " 2")]  # cause 2 = death

fracture_cif <- tibble(
  time  = mros_cif[[fracture_name]]$time,
  cif   = mros_cif[[fracture_name]]$est,
  event = "Hip fracture"
)

death_cif <- tibble(
  time  = mros_cif[[death_name]]$time,
  cif   = mros_cif[[death_name]]$est,
  event = "Death without fracture"
)

cif_df <- bind_rows(fracture_cif, death_cif)

# 5-year CIF: last time point <= 5 (right-continuous step function)
fracture_cif_five_yr <-
  tail(fracture_cif$cif[fracture_cif$time <= 5], 1)
death_cif_five_yr <-
  tail(death_cif$cif[death_cif$time <= 5], 1)
```

**Interpretation (MrOS study):**

In the MrOS study,
at 5 years of follow-up:

- `r round(fracture_cif_five_yr * 100, 1)`% of men had experienced a hip fracture
- `r round(death_cif_five_yr * 100, 1)`% had died without a fracture
- `r round((1 - fracture_cif_five_yr - death_cif_five_yr) * 100, 1)`%
  remained alive without fracture

### Estimation of Cumulative Incidence

The cumulative incidence at time $t$ is calculated as:

$$
\hat{F}_k(t) = \sum_{t_i \le t} \hsurvkmf{t_{i-1}} \times \frac{d_{ki}}{n_i}
$$

where:

- $t_i$ are the ordered event times
- $d_{ki}$ is the number of type-$k$ events at time $t_i$
- $n_i$ is the number at risk at time $t_i$
- $\hsurvkmf{t_{i-1}}$ is the Kaplan-Meier estimate of event-free survival
  just before time $t_i$

The estimation proceeds in steps:

1. Calculate the overall event-free probability at each time point
   using the Kaplan-Meier method
   (combining all event types)
2. For each time interval,
   calculate the probability of a new type-$k$ event as:
   - Probability of being event-free at the start of the interval
   - Times the rate of type-$k$ events during the interval
3. The cumulative incidence is the cumulative sum
   of these time-specific probabilities

#### Numerical Example: MrOS Study

The `mros` dataset from the `rmb` package
contains data from the
Osteoporotic Fractures in Men (MrOS) study [@orwoll2005design],
a prospective cohort study of older men.
The outcome variable is coded as:

- 0: censored (no fracture or death during follow-up)
- 1: hip fracture (event of interest)
- 2: death without fracture (competing event)

```{r}
#| label: fig-mros-cif
#| code-fold: true
ggplot(cif_df) +
  aes(x = time, y = cif, color = event) +
  geom_step() +
  labs(
    x       = "Years of follow-up",
    y       = "Cumulative incidence",
    color   = "Event type",
    caption = "Data: rmb::mros (MrOS study, Orwoll et al. 2005)"
  ) +
  scale_y_continuous(limits = c(0, 0.20)) +
  theme_bw() +
  theme(legend.position = "bottom")
```

At approximately 5 years of follow-up:

- About `r round(fracture_cif_five_yr * 100, 1)`% of men had experienced a hip fracture
- About `r round(death_cif_five_yr * 100, 1)`% had died without a fracture

These two cumulative incidences sum to less than the overall
event probability because many men are still event-free at 5 years.

#### Naive vs. Proper Cumulative Incidence

A naive approach treats competing events as censored
and uses the standard Kaplan-Meier method:
$1 - \hsurvkmf{t}$.
This overestimates the true cumulative incidence
because it assumes censored individuals
(including those who died)
have the same risk as those still event-free.

```{r}
#| label: fig-mros-cif-vs-naive
#| code-fold: true
# Naive KM estimate for fracture (treating death as censored)
km_naive <- survfit(
  Surv(years, status == 1) ~ 1,
  data = mros_data
)

naive_df <- tibble(
  time  = km_naive$time,
  cif   = 1 - km_naive$surv,
  method = "Naive KM (1 - S(t))"
)

proper_df <- fracture_cif |>
  mutate(method = "Proper CIF")

comparison_df <- bind_rows(naive_df, proper_df)

ggplot(comparison_df) +
  aes(x = time, y = cif, color = method) +
  geom_step() +
  labs(
    x     = "Years of follow-up",
    y     = "Estimated probability of fracture",
    color = "Method",
    caption = "Data: rmb::mros (MrOS study)"
  ) +
  scale_y_continuous(limits = c(0, 0.15)) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The naive KM curve overestimates the probability of fracture
because it imagines a hypothetical world without competing mortality.
The proper CIF curve represents the actual clinical probability
of experiencing a fracture.

### Modeling Competing Risks

When analyzing competing risks data with predictors,
two main approaches are available:

#### Cause-Specific Hazards Models

Fit separate Cox models for each event type,
treating other event types as censored:

**Advantages:**

- Standard software can be used
- Straightforward interpretation of hazard ratios
- Can assess effects on each failure type separately

**Limitations:**

- Results do not directly estimate cumulative incidence
- Interpretation can be challenging when effects differ across event types

#### Subdistribution Hazards Models (Fine-Gray Model)

An alternative approach directly models the cumulative incidence function
using subdistribution hazards [@fine1999proportional].

The Fine-Gray model:

- Directly estimates the effect on cumulative incidence
- Keeps participants who experience competing events in the risk set
- Requires specialized software (e.g., `cmprsk` package in R)
