The prior distribution $p(\theta)$ encodes what is known about
$\theta$ before the current data are seen.
Choosing it is the step that most distinguishes Bayesian practice from
frequentist practice, and it is where most of the controversy and most
of the craft live [@dobson4e, Chapter 12, p. 281].

## Conjugate priors {#sec-bayes-conjugate}

A prior is **conjugate** to a likelihood when the resulting posterior
belongs to the same parametric family as the prior.
Conjugate priors make the posterior available in closed form, as in
@exm-normal-normal.
The Beta family is conjugate to the Bernoulli/binomial likelihood:

{{< slidebreak >}}

:::{#exm-beta-bernoulli}
#### Beta-Bernoulli updating

Let $Y_1, \ldots, Y_n \siid \Bernoulli(\pi)$ with $r = \sum_i y_i$
successes, and place a $\operatorname{Beta}(a, b)$ prior on $\pi$.
By @eq-bayes-proportional,

$$
p(\pi \mid \vy)
\propto \underbrace{\pi^{r}(1-\pi)^{n-r}}_{\text{likelihood}}
\cdot \underbrace{\pi^{a-1}(1-\pi)^{b-1}}_{\text{prior}}
= \pi^{a + r - 1}(1-\pi)^{b + n - r - 1},
$$

which is the kernel of a $\operatorname{Beta}(a + r,\; b + n - r)$ density.
The prior parameters $a$ and $b$ act as **pseudo-counts** of prior
successes and failures: the posterior simply adds the observed $r$
successes and $n - r$ failures to them.
With a uniform $\operatorname{Beta}(1, 1)$ prior, $n = 91$, and $r = 55$,
the posterior is $\operatorname{Beta}(56, 37)$:

```{r}
a_prior <- 1
b_prior <- 1
n_obs <- 91
r_obs <- 55
a_post <- a_prior + r_obs
b_post <- b_prior + n_obs - r_obs
round(c(
  post_mean = a_post / (a_post + b_post),
  post_lower = qbeta(0.025, a_post, b_post),
  post_upper = qbeta(0.975, a_post, b_post)
), 4)
```
:::

## Informative, weakly informative, and flat priors {#sec-bayes-informativeness}

Priors range along a spectrum of how much they constrain $\theta$
[@dobson4e, Chapter 12, p. 281]:

- An **informative** prior concentrates mass in a region supported by
  external knowledge (previous studies, biological limits, expert
  judgment). It pulls the posterior toward that region, most strongly
  when the data are sparse.
- A **weakly informative** prior rules out implausible values
  (for example, an odds ratio of $10^6$) while remaining flat across the
  range of scientifically plausible values.
- A **flat** (or **noninformative**) prior aims to "let the data speak,"
  approximating constant density over $\Theta$. Flat priors can be
  *improper* (not integrating to one) and are not invariant to
  reparameterization, so they require care.

## A skeptical prior {#sec-bayes-skeptical}

A useful special case is a **skeptical prior**, centered on "no effect"
and deliberately tight, used to ask how strong the data must be to
overturn a default of no effect [@dobson4e, Chapter 12, p. 281].

{{< slidebreak >}}

:::{#exm-skeptical-prior}
#### A skeptical prior for a normal mean

Reusing the normal-mean model of @exm-normal-normal but with a skeptical
$\ndist{0, \tau^2}$ prior of standard deviation $\tau$,
we derive the exact posterior density by combining the normal likelihood and the $\ndist{0, \tau^2}$ prior:

$$
\ba
p(M=\mu \mid X=x)
&\propto p(X=x \mid M=\mu)\, p(M=\mu) \\
&\quad \text{(by Bayes' theorem, dropping marginal likelihood factor } p(X=x)\text{)} \\
&\propto \expf{-\frac{1}{2}\sum_{i=1}^n (x_i - \mu)^2} \cdot \expf{-\frac{1}{2\tau^2}\mu^2} \\
&\quad \text{(substituting normal likelihood and prior kernels)} \\
&= \expf{-\frac{1}{2}\sb{\sum_{i=1}^n (x_i^2 - 2x_i \mu + \mu^2) + \frac{\mu^2}{\tau^2}}} \\
&\quad \text{(expanding quadratic term inside exponent)} \\
&= \expf{-\frac{1}{2}\sb{\sum_{i=1}^n x_i^2 - 2\mu \sum_{i=1}^n x_i + n\mu^2 + \frac{\mu^2}{\tau^2}}} \\
&\quad \text{(distributing summation across terms)} \\
&\propto \expf{-\frac{1}{2}\sb{\paren{n + \frac{1}{\tau^2}}\mu^2 - 2n\bar{x}\mu}} \\
&\quad \text{(substituting } \sum x_i = n\bar{x} \text{ and dropping factor independent of } \mu\text{)} \\
&= \expf{-\frac{1}{2}\paren{n + \frac{1}{\tau^2}}\sb{\mu^2 - 2 \frac{n\bar{x}}{n + 1/\tau^2}\mu}} \\
&\quad \text{(factoring out precision } n + 1/\tau^2\text{)} \\
&\propto \expf{-\frac{1}{2}\paren{n + \frac{1}{\tau^2}}\paren{\mu - \frac{n\bar{x}}{n + 1/\tau^2}}^2} \\
&\quad \text{(completing square in } \mu \text{ and dropping factor independent of } \mu\text{)}
\ea
$$

This calculation yields the normal posterior density kernel:

$$
M \mid X = x \sim \ndist{\frac{n \bar{x}}{n + 1/\tau^2}, \frac{1}{n + 1/\tau^2}}.
$$

The posterior mean is therefore the precision-weighted average:

$$
\E{M \mid X = x}
= \frac{n\,\bar{x}}{n + 1/\tau^2},
$$

which shrinks toward $0$ more aggressively as the prior becomes more
skeptical (smaller $\tau$). We tabulate the posterior mean across a range
of prior skepticism, holding the data fixed:

```{r}
xbar_obs <- 2
n_obs2 <- 20
tau_grid <- c(0.1, 0.25, 0.5, 1, 2, 10)
post_mean <- (n_obs2 * xbar_obs) / (n_obs2 + 1 / tau_grid^2)
round(
  data.frame(prior_sd = tau_grid, posterior_mean = post_mean),
  3
)
```

A very skeptical prior ($\tau = 0.1$) drags the posterior mean almost to
zero despite $\bar{x} = 2$; a diffuse prior ($\tau = 10$) leaves it
essentially at the sample mean.
:::
