When several candidate models are plausible, **Bayesian model averaging
(BMA)** forms predictions by averaging over models, weighting each by its
posterior probability rather than committing to a single "best" model
[@dobson4e, Chapter 14, p. 338].
This propagates *model* uncertainty — not just parameter uncertainty —
into the final inference.
Unlike the single-model fits above, BMA requires fitting and comparing a
*collection* of models and computing their posterior weights, which
connects directly to the model-comparison criteria of
@sec-dic and to our
[predictor-selection chapter](predictor-selection.qmd).

:::{#exm-bayes-bma}
#### A BIC approximation to Bayesian model averaging

Marginal likelihoods are hard to compute, but the Bayesian information
criterion provides a convenient approximation to the posterior model
probability [@schwarz1978estimating]: for model $m$,
$p(m \mid \vy) \propto \expf{-\tfrac{1}{2}\operatorname{BIC}_m}$.
We simulate data in which only $x_1$ and $x_2$ affect the outcome
(with $x_3$ irrelevant), enumerate every subset of the three predictors,
and form BIC-based model weights.

```{r}
#| code-fold: show
set.seed(2027)
N <- 120L
x1 <- rnorm(N)
x2 <- rnorm(N)
x3 <- rnorm(N) # irrelevant
y <- 1 + 0.8 * x1 + 0.5 * x2 + rnorm(N)
dat <- data.frame(y, x1, x2, x3)

predictors <- c("x1", "x2", "x3")
subsets <- unlist(
  lapply(0:3, function(k) combn(predictors, k, simplify = FALSE)),
  recursive = FALSE
)
models <- lapply(subsets, function(vars) {
  rhs <- if (length(vars)) paste(vars, collapse = " + ") else "1"
  lm(as.formula(paste("y ~", rhs)), data = dat)
})

# BIC approximation to posterior model probabilities:
bic <- vapply(models, BIC, numeric(1))
weight <- exp(-0.5 * (bic - min(bic)))
weight <- weight / sum(weight)
```

The **posterior inclusion probability** of each predictor is the total
weight of the models that contain it:

```{r}
#| code-fold: show
pip <- vapply(
  predictors,
  function(v) sum(weight[vapply(subsets, function(s) v %in% s, logical(1))]),
  numeric(1)
)
round(pip, 3)
```

and the **model-averaged** slope for $x_1$ averages each model's estimate
(taken as $0$ when $x_1$ is absent from that model):

```{r}
#| code-fold: show
beta_x1 <- vapply(models, function(m) {
  cf <- coef(m)
  if ("x1" %in% names(cf)) cf[["x1"]] else 0
}, numeric(1))
round(c(bma_slope_x1 = sum(weight * beta_x1)), 3)
```

The real predictors $x_1$ and $x_2$ receive inclusion probabilities near
$1$, the irrelevant $x_3$ a small one, and the model-averaged slope for
$x_1$ sits near its true value $0.8$ — with the averaging across models,
rather than a single selected model, accounting for which predictors
belong.
:::
