We begin with the simplest possible model — a single Bernoulli
probability — fit with **JAGS** ("Just Another Gibbs Sampler"),
which we drive from R.
This example introduces the mechanics (specifying a model, supplying
data, monitoring parameters, and inspecting draws) that the later
analyses reuse.

:::{#exm-bayes-proportion}
#### Bernoulli model via JAGS

First, load required packages:

```{r}
#| code-fold: show
library(rjags)
library(runjags)
runjags::findJAGS()
```

We'll use a simple Bernoulli model to estimate a probability parameter using Bayesian inference.

```{r}
#| label: run-dobson-model
#| warning: false
#| message: false
#| code-fold: show

# JAGS chain initialization function
initsfunction <- function(chain) {
  stopifnot(chain %in% (1:4)) # max 4 chains allowed
  rng_seed <- (1:4)[chain]
  rng_name <- c(
    "base::Wichmann-Hill", "base::Marsaglia-Multicarry",
    "base::Super-Duper", "base::Mersenne-Twister"
  )[chain]
  return(list(".RNG.seed" = rng_seed, ".RNG.name" = rng_name))
}

# Generate sample data
set.seed(1)
data1 <- rbinom(n = 91, size = 1, prob = .6)

# Run JAGS model
jags_post0 <- run.jags(
  n.chains = 2,
  inits = initsfunction,
  model = system.file("extdata/model.dobson.jags", package = "rme"),
  data = list(r = data1, N = length(data1)),
  monitor = "p"
)
```

The results are examined by displaying the first few posterior draws:

```{r}
#| code-fold: show
jags_post0$mcmc |> as.array() |> head()
```
:::
