The engine of Bayesian inference is **Bayes' theorem**,
applied to the parameter and the data.
Writing $p(\theta)$ for the prior and $\Lik(\theta) = p(\vy \mid \theta)$
for the likelihood,
the posterior is

:::{#def-posterior}
#### Posterior distribution

$$
p(\theta \mid \vy)
\eqdef
\frac{p(\vy \mid \theta)\,p(\theta)}{p(\vy)},
\qquad
p(\vy) = \int p(\vy \mid \theta)\,p(\theta)\,d\theta.
$$ {#eq-bayes-posterior}
:::

The denominator $p(\vy)$ — the **marginal likelihood** or
*evidence* — does not depend on $\theta$;
it is the constant that makes the posterior integrate to one.
Consequently the posterior is determined up to proportionality by the
numerator alone [@dobson4e, Chapter 12, p. 272]:

$$
\underbrace{p(\theta \mid \vy)}_{\text{posterior}}
\;\propto\;
\underbrace{p(\vy \mid \theta)}_{\text{likelihood}}
\cdot
\underbrace{p(\theta)}_{\text{prior}}.
$$ {#eq-bayes-proportional}

This proportionality is the workhorse of applied Bayesian analysis:
it lets us recognize the posterior from the shape of
$\text{likelihood} \times \text{prior}$ without ever computing the
integral $p(\vy)$, and it is what makes the simulation methods of
@sec-foundations possible.

## A Gaussian mean with a Gaussian prior {#sec-bayes-normal-normal}

When prior and likelihood combine to give a posterior in the same family
as the prior, the prior is called **conjugate**.
The normal-mean model is the cleanest example.

:::{#exm-normal-normal}
#### Posterior for a normal mean

Suppose $X_1, \ldots, X_n \siid \ndist{\mu, 1}$ with the variance known,
and place a $\ndist{0, 1}$ prior on the mean: $M \sim \ndist{0, 1}$.
Collecting the terms of $p(M = \mu \mid X = x) \propto p(X = x \mid M = \mu)\,p(M = \mu)$
that involve $\mu$ and completing the square:

$$
\ba
p(M=\mu \mid X=x)
&\propto p(X=x \mid M=\mu)\, p(M=\mu)\\
&\propto \expf{-\frac{1}{2}\paren{n\mu^2 - 2\mu n\bar{x}}}
         \cdot \expf{-\frac{1}{2}\mu^2}\\
&= \expf{-\frac{1}{2}\paren{(n+1)\mu^2 - 2\mu n\bar{x}}}\\
&\propto \expf{-\frac{1}{2}(n+1)\paren{\mu - \frac{n}{n+1}\bar{x}}^2}.
\ea
$$

The last line is the kernel of a normal density, so the posterior is

$$
M \mid X = x \;\sim\; \ndist{\frac{n}{n+1}\bar{x},\; \frac{1}{n+1}}.
$$

The posterior mean $\frac{n}{n+1}\bar{x}$ is the sample mean shrunk
toward the prior mean $0$, and the shrinkage vanishes as $n \to \infty$.
:::

## The parameter space {#sec-bayes-parameter-space}

Because the Bayesian treats $\theta$ as random, the prior and posterior
are distributions over the **parameter space** $\Theta$ —
the set of values $\theta$ can take
[@dobson4e, Chapter 12, p. 274].
The parameter space must be respected when choosing a prior:
a probability $\pi \in (0,1)$ calls for a prior supported on the unit
interval (such as a Beta distribution),
a variance $\sigma^2 > 0$ calls for a prior on the positive half-line,
and a vector of regression coefficients calls for a multivariate prior
on $\reals^p$.
A prior that places no mass on part of $\Theta$ forces the posterior to
do the same, no matter what the data say —
so the support of the prior is itself a modeling assumption.

:::{#exm-normal-credible-vs-confidence}
#### Credible vs. confidence interval for a normal mean

We compare the two interval estimates from @sec-bayes-interval-interp on
simulated data from the model of @exm-normal-normal.
First, the frequentist interval:

```{r}
set.seed(1)
mu <- 2
sigma <- 1
n <- 20
x <- rnorm(n = n, mean = mu, sd = sigma)
xbar <- mean(x)
se <- sigma / sqrt(n)
CI_freq <- xbar + se * qnorm(c(.025, .975))
print(CI_freq)
```

The likelihood as a function of $\mu$:

```{r}
lik <- function(mu) {
  (2 * pi * sigma^2)^(-n / 2) *
    exp(
      -1 / (2 * sigma^2) *
        (sum(x^2) - 2 * mu * sum(x) + n * (mu^2))
    )
}
ngraph <- 1001
plot1 <- ggplot() +
  geom_function(fun = lik, aes(col = "likelihood"), n = ngraph) +
  xlim(c(-5, 10)) +
  theme_bw() +
  labs(col = "") +
  theme(legend.position = "bottom")
print(plot1)
```

Next, the Bayesian credible interval, using the closed-form posterior
$\ndist{\frac{n}{n+1}\bar{x},\, (n+1)^{-1}}$ from @exm-normal-normal:

```{r}
mu_prior_mean <- 0
mu_prior_sd <- 1
mu_post_mean <- n / (n + 1) * xbar
mu_post_var <- 1 / (n + 1)
mu_post_sd <- sqrt(mu_post_var)
CI_bayes <- qnorm(
  p = c(.025, .975),
  mean = mu_post_mean,
  sd = mu_post_sd
)
print(CI_bayes)
prior <- function(mu) dnorm(mu, mean = mu_prior_mean, sd = mu_prior_sd)
posterior <- function(mu) dnorm(mu, mean = mu_post_mean, sd = mu_post_sd)
plot2 <- plot1 +
  geom_function(fun = prior, aes(col = "prior"), n = ngraph) +
  geom_function(fun = posterior, aes(col = "posterior"), n = ngraph)
print(plot2 + scale_y_log10())
```

Finally, the quantity a credible interval reports —
the posterior probability that $M$ lies in the *frequentist* interval —
is well defined for the Bayesian but not for the frequentist:

```{r}
pr_in_CI <- pnorm(
  CI_freq,
  mean = mu_post_mean,
  sd = mu_post_sd
) |> diff()
print(pr_in_CI)
```
:::
