
```{r, printr.help.sections = c("description", "usage")}
?survdiff
```

:::{#exm-survdiff-drug6mp}
#### `survdiff()` with `drug6mp` data

Now we are going to compare the placebo and 6-MP data. We need to
reshape the data to make it usable with the standard `survival`
workflow:

```{r}
library(survival)
library(tidyr)
data_v4 <-
  data_v3 |>
  select(pair, remstat, t1, t2, outcome) |>
  # here we are going to change the data from a wide format to long:
  pivot_longer(
    cols = c(t1, t2),
    names_to = "treatment",
    values_to = "exit_time"
  ) |>
  mutate(
    treatment = treatment |>
      case_match(
        "t1" ~ "placebo",
        "t2" ~ "6-MP"
      ),
    outcome = if_else(
      treatment == "placebo",
      "relapsed",
      outcome
    ),
    surv = Surv(
      time = exit_time,
      event = (outcome == "relapsed")
    )
  )
```

{{< slidebreak >}}

Using this long data format, we can fit a Kaplan-Meier curve for each
treatment group simultaneously:

```{r}
km_model2 <-
  survfit(
    formula = surv ~ treatment,
    data = data_v4
  )
```

{{< slidebreak >}}

We can plot the curves in the same graph:

```{r}
km_model2 |> autoplot()
```

{{< slidebreak >}}

We can also perform something like a t-test, where the null hypothesis
is that the curves are the same:

:::{#tbl-logrank-6mp}
```{r}

o_e <- data_v4 |>
  arrange(exit_time) |>
  mutate(
    .by = treatment,
    n_exited = row_number(),
    n_at_risk = n() - n_exited + 1
  ) |>
  dplyr::summarize(
    .by = all_of(c("exit_time", "treatment")),
    n_at_risk = max(n_at_risk),
    n_events = sum(outcome == "relapsed")
  ) |>
  tidyr::pivot_wider(
    names_from = "treatment",
    values_from = c(n_at_risk, n_events)
  ) |>
  tidyr::fill(
    starts_with("n_at_risk"),
    .direction = "up"
  ) |>
  replace_na(list("n_events_placebo" = 0,
                  "n_events_6-MP" = 0)) |>
  mutate(
    n_at_risk = rowSums(across(starts_with("n_at_risk"))),
    n_events = rowSums(across(starts_with("n_events"))),
    marginal_hazard = n_events / n_at_risk,
    expected_6mp = marginal_hazard * `n_at_risk_6-MP`,
    expected_plc = marginal_hazard * n_at_risk_placebo,
    diff_6mp = `n_events_6-MP` - expected_6mp,
    diff_plc = n_events_placebo - expected_plc
  ) |>
  filter(n_events > 0)

o_e

```
Observed and expected event counts for the 6-MP data, for log-rank test
:::

:::{#tbl-sum-obs-exp}
```{r}
o_e_summ <- o_e |>
  summarize(
    across(starts_with("expected"), sum),
    across(starts_with("n_events_"), sum)
  )
pander::pander(o_e_summ)
```
Observed and expected sums for the 6-MP data, for log-rank test

:::

{{< slidebreak >}}

::: notes
The exact variance formula for each of two groups is:
:::

$$
\Varf{O_i - E_i} = \sum_j \frac
{n_{1j} n_{2j} (m_j) (n_j - m_j)}
{(n_j)^2 (n_j - 1)}
$$

::: notes
See @kleinbaum2012survival, Chapter 2 Appendix for the exact variance formula
for more than two groups.
:::

{{< slidebreak >}}

::: notes
Or we can use an approximate statistic:
:::
$$X^2 \approx \sum_{i = 1}^p \frac{(O_i - E_i)^2}{E_i}$$

```{r}
with(
  o_e_summ,
  tibble(
    "6mp" = (`n_events_6-MP` - expected_6mp)^2 / expected_6mp,
    "placebo" = (n_events_placebo - expected_plc)^2 / expected_plc,
    sum = `6mp` + placebo
  )
) |>
  pander::pander()
```

{{< slidebreak >}}

R gives us both the exact and approximate results:

```{r}
survdiff(
  formula = surv ~ treatment,
  data = data_v4
)
```

{{< slidebreak >}}

::: notes
By default, `survdiff()` ignores any pairing,
but we can use `strata()` to perform something similar to a paired t-test:
:::

```{r}
lrank_test <- survdiff(
  formula = surv ~ treatment + strata(pair),
  data = data_v4
)
lrank_test
```

::: notes
Interestingly, accounting for pairing reduces the significance of the
difference.
:::

:::
