
We can also perform something like a t-test, where the null hypothesis
is that the curves are the same. The table below shows the per-time
$(d_{gi}, r_{gi})$ counts and the expected counts $e_{gi}$ from
@eq-logrank-expected:

:::{#tbl-logrank-6mp}
```{r}
#| code-fold: true

o_e <- data_v4 |>
  arrange(exit_time) |>
  mutate(
    .by = treatment,
    n_exited = row_number(),
    n_at_risk = n() - n_exited + 1
  ) |>
  dplyr::summarize(
    .by = all_of(c("exit_time", "treatment")),
    n_at_risk = max(n_at_risk),
    n_events = sum(outcome == "relapsed")
  ) |>
  tidyr::pivot_wider(
    names_from = "treatment",
    values_from = c(n_at_risk, n_events)
  ) |>
  tidyr::fill(
    starts_with("n_at_risk"),
    .direction = "up"
  ) |>
  replace_na(list("n_events_placebo" = 0,
                  "n_events_6-MP" = 0)) |>
  mutate(
    n_at_risk = rowSums(across(starts_with("n_at_risk"))),
    n_events = rowSums(across(starts_with("n_events"))),
    marginal_hazard = n_events / n_at_risk,
    expected_6mp = marginal_hazard * `n_at_risk_6-MP`,
    expected_plc = marginal_hazard * n_at_risk_placebo,
    diff_6mp = `n_events_6-MP` - expected_6mp,
    diff_plc = n_events_placebo - expected_plc
  ) |>
  filter(n_events > 0)

o_e

```
Observed and expected event counts for the 6-MP data, for the log-rank
test. Columns: `exit_time` is $t_i$;
`n_at_risk_placebo`/`n_at_risk_6-MP` are $r_{gi}$;
`n_events_placebo`/`n_events_6-MP` are $d_{gi}$;
`n_at_risk`/`n_events` are pooled $r_i$, $d_i$;
`marginal_hazard` is $\hat{\haz}_i = d_i / r_i$;
`expected_*` are $e_{gi}$ from @eq-logrank-expected;
`diff_*` are the per-time contributions $d_{gi} - e_{gi}$ to $O_g - E_g$.
:::

The summed observed $O_g$ and expected $E_g$ totals (one row, one column per group):

:::{#tbl-sum-obs-exp}
```{r}
#| code-fold: true
o_e_summ <- o_e |>
  summarize(
    across(starts_with("expected"), sum),
    across(starts_with("n_events_"), sum)
  )
pander::pander(o_e_summ)
```
Sums of observed and expected events for the 6-MP data.
:::

{{< slidebreak >}}

The Pearson-style approximation to @eq-logrank-stat-2grp is

$$X^2 \;\approx\; \sum_{g} \frac{(O_g - E_g)^2}{E_g}$$

which we compute directly from @tbl-sum-obs-exp:

```{r}
#| code-fold: show
with(
  o_e_summ,
  tibble(
    "6mp" = (`n_events_6-MP` - expected_6mp)^2 / expected_6mp,
    "placebo" = (n_events_placebo - expected_plc)^2 / expected_plc,
    sum = `6mp` + placebo
  )
) |>
  pander::pander()
```

{{< slidebreak >}}

R gives us both the exact and approximate results:

```{r}
#| code-fold: show
survdiff(
  formula = surv ~ treatment,
  data = data_v4
)
```

{{< slidebreak >}}

::: notes
By default, `survdiff()` ignores any pairing,
but we can use `strata()` to perform something similar to a paired t-test:
:::

```{r}
#| code-fold: show
lrank_test <- survdiff(
  formula = surv ~ treatment + strata(pair),
  data = data_v4
)
lrank_test
```

::: notes
Interestingly, accounting for pairing reduces the significance of the
difference.
:::
