## Comparing nonparametric survival function estimates

@tbl-surv1 lists some simulated survival data, where:

- $Y$ is study exit time
- $D$ is reason for study exit:
  - $D=1$: event
  - $D=0$: censored
- $Y^*$ is the combination of $Y$ and $D$ into a single `Surv()` variable.

```{r}
#| label: tbl-surv1
#| tbl-cap: "Simulated survival data"
#| tbl-pos: H
#| echo: true
#| code-summary: "simulated survival data"
library(tibble)
data1 <- tibble(
  Y = c(1, 3, 7, 10, 13, 15, 16, 16, 16),
  D = c(1, 1, 1,  1,  0,  1,  0,  0,  0),
  S = survival::Surv(Y, D)
)
print(data1)
```

:::{#exr-km-curve-by-hand}
Compute the Kaplan-Meier (KM) estimated survival curve, by hand.
:::

::::{#sol-km-curve-by-hand .solution}

Let:

- $n$: total sample size
- $y$: study exit time;
- $d(y)$: number of events at $y$;
- $c(y)$: number exiting without events ("censored") at $y$
- $q(y)$: number censored or continuing in the study at $y$;
- $e(y) = d(y) + q(y)$: total exiting at $y$;
- $E(y)$: cumulative exits prior to $y$;
- $r(y) = n - E(y)$: number at risk at $y$;
- $r^*(y) = r(y) - e(y)$: number at risk after $y$;
- $\hat{\kappa}(y) = \hat{\Pr}(Y > y \mid Y \geq y) = [r(y) - d(y)]/r(y)$: conditional survival factor;
- $\hat{\lambda}(y) = \hat{\Pr}(Y = y \mid Y \geq y) = \frac{d(y)}{r(y)} = 1 - \hat{\kappa}(y)$: estimated discrete hazard;
- $\hat{S}_{KM}(y) = \prod_{k:\, y_k \leq y} \hat{\kappa}(y_k)$: Kaplan-Meier survival estimate.

:::{#tbl-km-curve}
```{r}
library(pander)
library(tidyverse)

start_data <- tibble(
  Y = 0,
  events = 0,
  censored = 0
)
km_curve <-
  data1 |>
  summarise(.by  = Y,
            events = sum(D == 1),
            censored = sum(D == 0)) |>
  # bind_rows(start_data) |>
  arrange(Y) |>
  mutate(
    study_size = nrow(data1),
    exiting = events + censored,
    exited = cumsum(exiting) |> dplyr::lag(default = 0),
    at_risk = study_size - exited,
    r_star = at_risk - exiting,
    hazard = events / at_risk,
    km_factor = 1 - hazard,
    km_surv_curve = cumprod(km_factor)
  ) |>
  select(-study_size) |>
  relocate(km_factor, .before = hazard)
km_curve |>
  mutate(
    hazard_math = paste0("\\frac{", events, "}{", at_risk, "}"),
    km_factor_math = paste0("\\frac{", at_risk - events, "}{", at_risk, "}"),
    km_surv_curve = paste0(
      "$",
      purrr::map_chr(
        row_number(),
        \(i) paste(km_factor_math[seq_len(i)], collapse = " \\times ")
      ),
      " = ",
      round(km_surv_curve, 4),
      "$"
    ),
    hazard = paste0(
      "$", hazard_math, " = ",
      round(hazard, 4), "$"
    ),
    km_factor = paste0(
      "$", km_factor_math, " = ",
      round(km_factor, 4), "$"
    )
  ) |>
  select(-hazard_math, -km_factor_math) |>
  rename(
    `$y$`                        = Y,
    `$d(y)$`                     = events,
    `$c(y)$`                     = censored,
    `$e(y)$`                     = exiting,
    `$E(y)$`                     = exited,
    `$r(y)$`                     = at_risk,
    `$r^*(y)$`                   = r_star,
    `$\\hat{\\kappa}(y)$`         = km_factor,
    `$\\hat{\\lambda}(y)$`       = hazard,
    `$\\hat{S}_{KM}(y)$`         = km_surv_curve
  ) |>
  pander()
```

Kaplan-Meier survival curve calculations

:::

::::

:::{#exr-km-survfit}
Implement the KM survival curve estimator in R.
Check the output of your implementation against the version in the `survival` package.
:::

:::{#sol-km-survfit .solution}

See above for implementation.
Here's the version from the `survival` package:

```{r}
#| label: survfit
library(survival)
data1 |>
  survfit(Surv(time = Y, event = D) ~ 1, data = _) |>
  summary()
```

:::

:::{#exr-km-graph-1}
Graph the KM estimated survival curve for the data in @tbl-surv1.
:::

:::{#sol-km-graph-1 .solution}

```{r}
#| label: km-fig
library(ggplot2)
km_curve |>
  ggplot(aes(x = Y, y = km_surv_curve)) +
  geom_step() +
  geom_point(data = km_curve |> filter(censored > 0), shape = 3) +
  ylab("KM Survival Curve") +
  expand_limits(y = c(0, 1), x = 0) +
  scale_y_continuous(labels = scales::percent)
```

or

```{r}
#| label: autoplot-fig
library(ggfortify)
data1 <- data1 |>
  mutate(
    surv = Surv(time = Y, event = D)
  )
KM_model <- data1 |> survfit(formula = surv ~ 1)
KM_model |> autoplot()

```

:::

:::{#exr-km-median}
Find the KM estimate of median survival time.
:::

:::{#sol-km-median .solution}

```{r}
#| label: km-median
KM_model |> quantile(prob = 0.5) |> as_tibble()
```

:::

:::{#exr-na-curve-by-hand}
Compute the Nelson-Aalen (NA) estimated survival curve, by hand.
:::

::::{#sol-na-curve-by-hand .solution}

Let:

- $n$: total sample size
- $y$: study exit time;
- $d(y)$: number of events at $y$;
- $c(y)$: number exiting without events ("censored") at $y$
- $e(y) = d(y) + c(y)$: total exiting at $y$;
- $E(y)$: cumulative exits prior to $y$;
- $r(y) = n - E(y)$: number at risk at $y$;
- $r^*(y) = r(y) - e(y)$: number at risk after $y$;
- $\hat{\lambda}(y) = \frac{d(y)}{r(y)}$: Nelson-Aalen hazard increment;
- $\hat{\Lambda}_{NA}(y) = \sum_{k:\, y_k \leq y} \hat{\lambda}(y_k)$: Nelson-Aalen cumulative hazard estimate;
- $\hat{\eta}(y) = \exp[-\hat{\lambda}(y)]$: Nelson-Aalen survival factor;
- $\hat{S}_{NA}(y) = \exp[-\hat{\Lambda}_{NA}(y)] = \prod_{k:\, y_k \leq y} \hat{\eta}(y_k)$: Nelson-Aalen survival estimate.

:::{#tbl-na-curve}
```{r}
#| label: NA-fit-implement
library(dplyr)
na_curve <-
  data1 |>
  summarise(.by  = Y,
            events = sum(D == 1),
            censored = sum(D == 0)) |>
  arrange(Y) |>
  mutate(
    study_size = nrow(data1),
    exiting = events + censored,
    exited = cumsum(exiting) |> dplyr::lag(default = 0),
    at_risk = study_size - exited,
    r_star = at_risk - exiting,
    hazard = events / at_risk,
    na_factor = exp(-hazard),
    cusum_hazard = cumsum(hazard),
    na_surv_curve = cumprod(na_factor)
  ) |>
  select(-study_size) |>
  relocate(na_factor, .before = hazard)
na_curve |>
  mutate(
    hazard_math = paste0("\\frac{", events, "}{", at_risk, "}"),
    na_factor_math = paste0("e^{-", hazard_math, "}"),
    na_surv_curve = paste0(
      "$",
      purrr::map_chr(
        row_number(),
        \(i) paste(na_factor_math[seq_len(i)], collapse = " \\times ")
      ),
      " = ",
      round(na_surv_curve, 4),
      "$"
    ),
    cusum_hazard = paste0(
      "$",
      purrr::map_chr(
        row_number(),
        \(i) paste(hazard_math[seq_len(i)], collapse = " + ")
      ),
      " = ",
      round(cusum_hazard, 4),
      "$"
    ),
    hazard = paste0(
      "$", hazard_math, " = ",
      round(hazard, 4), "$"
    ),
    na_factor = paste0(
      "$", na_factor_math, " = ",
      round(na_factor, 4), "$"
    )
  ) |>
  select(-hazard_math, -na_factor_math) |>
  rename(
    `$y$`                        = Y,
    `$d(y)$`                     = events,
    `$c(y)$`                     = censored,
    `$e(y)$`                     = exiting,
    `$E(y)$`                     = exited,
    `$r(y)$`                     = at_risk,
    `$r^*(y)$`                   = r_star,
    `$\\hat{\\eta}(y)$`           = na_factor,
    `$\\hat{\\lambda}(y)$`       = hazard,
    `$\\hat{\\Lambda}_{NA}(y)$`  = cusum_hazard,
    `$\\hat{S}_{NA}(y)$`         = na_surv_curve
  ) |>
  pander()
```

Nelson-Aalen survival curve calculations

:::

::::

:::{#exr-na-survfit}
Implement the NA survival curve estimator in R.
Check the output of your implementation against the version in the `survival` package.
:::

:::{#sol-na-survfit .solution}

```{r}
#| label: NA-survfit
NA_model <-
  data1 |>
  survfit(formula = surv ~ 1,
          type = "fleming-harrington")

NA_model |> summary()
```

:::

:::{#exr-km-na-graph}
Add the NA estimated survival curve to the graph from @exr-km-graph-1.
:::

:::{#sol-km-na-graph .solution}

```{r}
#| label: km-na-fig
km_and_na <-
  bind_rows(
    .id = "model",
    "Kaplan-Meier" = KM_model |> fortify(surv.connect = FALSE),
    "Nelson-Aalen" = NA_model |> fortify(surv.connect = FALSE)
  ) |>
  as_tibble()

km_and_na |>
  ggplot(aes(x = time, y = surv, col = model)) +
  geom_step() +
  theme_bw() +
  ylab("S(t) = P(T>=t)") +
  xlab("Survival time (t, days)") +
  theme(legend.position = "bottom")
```

:::

:::{#exr-na-median}
Find the NA estimate of median survival time.
:::

:::{#sol-na-median .solution}

```{r}
#| label: na-median
NA_model |> quantile(prob = 0.5) |> as_tibble()
```

:::

:::{#exr-compare-km-na}
Describe the similarities and differences between the graphs of the KM and NA survival function estimates.
:::

:::{#sol-compare-km-na .solution}

The curves are very similar for this data set, with the KM curve returning slightly lower estimates of survival.

:::
