:::{#exr-km-curve-by-hand}

@tbl-surv1 lists some simulated survival data, where:

- $Y$ is study exit time
- $D$ is reason for study exit:
  - $D=1$: event
  - $D=0$: censored
- `S` is the combination of $Y$ and $D$ into a single `Surv()` variable.

:::{#tbl-surv1}

```{r}
#| tbl-pos: H
library(tibble)
data1 <- tibble(
  Y = c(1, 3, 7, 10, 13, 15, 16, 16, 16),
  D = c(1, 1, 1,  1,  0,  1,  0,  0,  0),
  S = survival::Surv(Y, D)
)
knitr::kable(data1)
```
Simulated survival data for @exr-km-curve-by-hand.
:::

Compute the Kaplan-Meier (KM) estimated survival curve by hand.

:::

::::{#sol-km-curve-by-hand .solution}

Let:

- $n$: total sample size
- $y$: study exit time;
- $d(y)$: number of events at $y$;
- $c(y)$: number exiting without events ("censored") at $y$;
- $e(y) = d(y) + c(y)$: total exiting at $y$;
- $E(y)$: cumulative exits prior to $y$;
- $r(y) = n - E(y)$: number at risk at $y$;
- $r^*(y) = r(y) - e(y)$: number at risk after $y$;
- $\hat{\cs}(y) = \hat{\Pr}(T > y \mid T \geq y) = [r(y) - d(y)]/r(y)$: conditional survival factor;
- $\hat{\haz}(y) = \hat{\Pr}(T = y \mid T \geq y) = \frac{d(y)}{r(y)} = 1 - \hat{\cs}(y)$: estimated discrete hazard;
- $\hskmf{y} = \prod_{k:\, y_k \leq y} \hat{\cs}(y_k)$: Kaplan-Meier survival estimate.

:::{#tbl-km-curve}
```{r}
library(pander)
library(tidyverse)

km_curve <-
  data1 |>
  summarise(.by  = Y,
            events = sum(D == 1),
            censored = sum(D == 0)) |>
  arrange(Y) |>
  mutate(
    study_size = nrow(data1),
    exiting = events + censored,
    exited = cumsum(exiting) |> dplyr::lag(default = 0),
    at_risk = study_size - exited,
    r_star = at_risk - exiting,
    hazard = events / at_risk,
    km_factor = 1 - hazard,
    km_surv_curve = cumprod(km_factor)
  ) |>
  select(-study_size) |>
  relocate(km_factor, .before = hazard)
km_curve |>
  mutate(
    hazard_math = paste0("\\frac{", events, "}{", at_risk, "}"),
    km_factor_math = paste0("\\frac{", at_risk - events, "}{", at_risk, "}"),
    km_surv_curve = paste0(
      "$",
      purrr::map_chr(
        row_number(),
        \(i) paste(km_factor_math[seq_len(i)], collapse = " \\times ")
      ),
      " = ",
      round(km_surv_curve, 4),
      "$"
    ),
    hazard = paste0(
      "$", hazard_math, " = ",
      round(hazard, 4), "$"
    ),
    km_factor = paste0(
      "$", km_factor_math, " = ",
      round(km_factor, 4), "$"
    )
  ) |>
  select(-hazard_math, -km_factor_math) |>
  rename(
    `$y$`                        = Y,
    `$d(y)$`                     = events,
    `$c(y)$`                     = censored,
    `$e(y)$`                     = exiting,
    `$E(y)$`                     = exited,
    `$r(y)$`                     = at_risk,
    `$r^*(y)$`                   = r_star,
    `$\\hat{\\cs}(y)$`         = km_factor,
    `$\\hat{\\haz}(y)$`          = hazard,
    `$\\hskmf{y}$`            = km_surv_curve
  ) |>
  pander()
```

Kaplan-Meier survival curve calculations

:::

::::

:::{#exr-km-survfit}
Implement the KM survival curve estimator in R.
Check the output of your implementation against the version in the `survival` package.
:::

:::{#sol-km-survfit .solution}

See above for implementation.
Here's the version from the `survival` package:

```{r}
#| label: survfit
library(survival)
data1 |>
  survfit(Surv(time = Y, event = D) ~ 1, data = _) |>
  summary()
```

:::

:::{#exr-km-graph-1}
Graph the KM estimated survival curve for the data in @tbl-surv1.
:::

:::{#sol-km-graph-1 .solution}

```{r}
#| label: km-fig
library(ggplot2)
km_curve |>
  ggplot(aes(x = Y, y = km_surv_curve)) +
  geom_step() +
  geom_point(data = km_curve |> filter(censored > 0), shape = 3) +
  ylab("KM Survival Curve") +
  expand_limits(y = c(0, 1), x = 0) +
  scale_y_continuous(labels = scales::percent)
```

or

```{r}
#| label: autoplot-fig
library(ggfortify)
data1 <- data1 |>
  mutate(
    surv = Surv(time = Y, event = D)
  )
KM_model <- data1 |> survfit(formula = surv ~ 1)
KM_model |> autoplot()

```

:::

:::{#exr-km-median}
Find the KM estimate of median survival time.
:::

:::{#sol-km-median .solution}

```{r}
#| label: km-median
KM_model |> quantile(prob = 0.5) |> as_tibble()
```

:::

