:::{#exr-na-curve-by-hand}
Compute the Nelson-Aalen (NA) estimated survival curve for the data in @tbl-surv1, by hand.
:::

::::{#sol-na-curve-by-hand .solution}

Let:

- $n$: total sample size
- $y$: study exit time;
- $d(y)$: number of events at $y$;
- $c(y)$: number exiting without events ("censored") at $y$
- $e(y) = d(y) + c(y)$: total exiting at $y$;
- $E(y)$: cumulative exits prior to $y$;
- $r(y) = n - E(y)$: number at risk at $y$;
- $r^*(y) = r(y) - e(y)$: number at risk after $y$;
- $\hat{\haz}(y) = \frac{d(y)}{r(y)}$: Nelson-Aalen hazard increment;
- $\hat{\cs}(y) = \expf{-\hat{\haz}(y)}$: Nelson-Aalen conditional survival factor;
- $\hat{\cuhaz}_{NA}(y) = \sum_{k:\, y_k \leq y} \hat{\haz}(y_k)$: Nelson-Aalen cumulative hazard estimate;
- $\hsurvnaaf{y} = \expf{-\hat{\cuhaz}_{NA}(y)} = \prod_{k:\, y_k \leq y} \hat{\cs}(y_k)$: Nelson-Aalen survival estimate.

:::{#tbl-na-curve}
```{r}
#| label: NA-fit-implement
library(dplyr)
na_curve <-
  data1 |>
  summarise(.by  = Y,
            events = sum(D == 1),
            censored = sum(D == 0)) |>
  arrange(Y) |>
  mutate(
    study_size = nrow(data1),
    exiting = events + censored,
    exited = cumsum(exiting) |> dplyr::lag(default = 0),
    at_risk = study_size - exited,
    r_star = at_risk - exiting,
    hazard = events / at_risk,
    kappa = exp(-hazard),
    cusum_hazard = cumsum(hazard),
    na_surv_curve = exp(-cusum_hazard)
  ) |>
  select(-study_size)
na_curve |>
  mutate(
    hazard_math = paste0("\\frac{", events, "}{", at_risk, "}"),
    cusum_hazard_round = round(cusum_hazard, 4),
    na_surv_curve = paste0(
      "$\\expf{-", cusum_hazard_round, "}",
      " = ",
      round(na_surv_curve, 4),
      "$"
    ),
    cusum_hazard = paste0(
      "$",
      purrr::map_chr(
        row_number(),
        \(i) paste(hazard_math[seq_len(i)], collapse = " + ")
      ),
      " = ",
      cusum_hazard_round,
      "$"
    ),
    kappa = paste0(
      "$\\expf{-", hazard_math, "} = ",
      round(kappa, 4), "$"
    ),
    hazard = paste0(
      "$", hazard_math, " = ",
      round(hazard, 4), "$"
    )
  ) |>
  select(-hazard_math, -cusum_hazard_round) |>
  rename(
    `$y$`                                  = Y,
    `$d(y)$`                               = events,
    `$c(y)$`                               = censored,
    `$e(y)$`                               = exiting,
    `$E(y)$`                               = exited,
    `$r(y)$`                               = at_risk,
    `$r^*(y)$`                             = r_star,
    `$\\hat{\\haz}(y)$`                    = hazard,
    `$\\hat{\\cs}(y)$`                     = kappa,
    `$\\hat{\\cuhaz}_{NA}(y)$`             = cusum_hazard,
    `$\\hsurvnaaf{y}$`                     = na_surv_curve
  ) |>
  pander()
```

Nelson-Aalen survival curve calculations

:::

::::

:::{#exr-na-survfit}
Implement the NA survival curve estimator in R.
Check the output of your implementation against the version in the `survival` package.
:::

:::{#sol-na-survfit .solution}

```{r}
#| label: NA-survfit
NA_model <-
  data1 |>
  survfit(formula = surv ~ 1,
          type = "fleming-harrington")

NA_model |> summary()
```

:::

:::{#exr-km-na-graph}
Add the NA estimated survival curve to the graph from @exr-km-graph-1.
:::

:::{#sol-km-na-graph .solution}

```{r}
#| label: km-na-fig
km_and_na <-
  bind_rows(
    .id = "model",
    "Kaplan-Meier" = KM_model |> fortify(surv.connect = FALSE),
    "Nelson-Aalen" = NA_model |> fortify(surv.connect = FALSE)
  ) |>
  as_tibble()

km_and_na |>
  ggplot(aes(x = time, y = surv, col = model)) +
  geom_step() +
  theme_bw() +
  ylab("S(t) = P(T>t)") +
  xlab("Survival time (t, days)") +
  theme(legend.position = "bottom")
```

:::

:::{#exr-na-median}
Find the NA estimate of median survival time.
:::

:::{#sol-na-median .solution}

```{r}
#| label: na-median
NA_model |> quantile(prob = 0.5) |> as_tibble()
```

:::

:::{#exr-compare-km-na}
Describe the similarities and differences between the graphs of the KM and NA survival function estimates.
:::

:::{#sol-compare-km-na .solution}

The curves are very similar for this data set, with the KM curve returning slightly lower estimates of survival.

:::
