
### Estimating survival in datasets without censoring

::: notes
In the `drug6mp` dataset, the estimated survival function for the placebo patients is easy to compute. 
For any time $t$ in months, $\surv(t)$ is the fraction of patients with times greater than $t$:
:::

```{r}
#| include: false
#| code-summary: example-km-curve

y <- c(1, 2, 3, 3, 3, 5, 9, 13, 23, 100)
d <- c(1, 1, 1, 0, 1, 0, 1, 1, 1, 1)
library(survival)
temp1 <- survfit(Surv(y, d) ~ 1)
```


### Estimating survival in datasets with censoring

-   For the 6-MP patients, we cannot ignore the censored data because we
know that the time to relapse is greater than the censoring time.

-   For any time $t$ in months, we know that 6-MP patients with times
greater than $t$ have not relapsed, and those with relapse time less
than $t$ have relapsed, but we don't know if patients with censored
time less than $t$ have relapsed or not.

-   The procedure we usually use is the Kaplan-Meier product-limit
estimator of the survival function.

-   The Kaplan-Meier estimator is a step function (like the empirical
cdf), which changes value only at the event times, not at the
censoring times.

-   At each event time $t_i$, we compute the at-risk set, which
includes all observations whose event time or censoring time is at
least $t_i$.
    Its size is $r_i$.

-   If $d_i$ of the observations have an event time (not a censoring time)
of $t_i$, then the group of survivors immediately following time $t_i$
is reduced by the fraction $$\frac{r_i-d_i}{r_i}=1-\frac{d_i}{r_i}$$

:::{#fig-risk-set-dropout-cartoon}

```{r}
#| echo: false

risk_cartoon <- tibble::tibble(
  person = paste("Person", 1:8),
  exit_time = c(2, 3, 4, 5, 6, 6, 7, 8),
  status = c(
    "Event",
    "Censored",
    "Event",
    "Censored",
    "Event",
    "Censored",
    "Event",
    "Censored"
  )
) |>
  dplyr::mutate(person_id = dplyr::row_number())

risk_counts <- tibble::tibble(time = c(0, 2, 4, 6, 8)) |>
  dplyr::mutate(
    at_risk = purrr::map_int(time, \(t) sum(risk_cartoon$exit_time > t)),
    label = paste0("r(", time, "^+) = ", at_risk)
  )

ggplot2::ggplot(risk_cartoon) +
  ggplot2::aes(y = person_id) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = 0,
      xend = exit_time,
      yend = person_id
    ),
    linewidth = 1,
    color = "grey50"
  ) +
  ggplot2::geom_point(
    ggplot2::aes(
      x = exit_time,
      shape = status,
      fill = status
    ),
    size = 3,
    color = "black"
  ) +
  ggplot2::geom_vline(
    data = dplyr::filter(risk_counts, time > 0),
    ggplot2::aes(xintercept = time),
    linetype = "dashed",
    color = "grey85"
  ) +
  ggplot2::geom_text(
    data = risk_counts,
    ggplot2::aes(
      x = time,
      y = max(risk_cartoon$person_id) + 0.8,
      label = label
    ),
    size = 3.4,
    vjust = 0
  ) +
  ggplot2::scale_shape_manual(values = c(Event = 21, Censored = 22)) +
  ggplot2::scale_fill_manual(
    values = c(Event = "#F8766D", Censored = "#00BFC4")
  ) +
  ggplot2::scale_y_continuous(
    breaks = risk_cartoon$person_id,
    labels = risk_cartoon$person
  ) +
  ggplot2::coord_cartesian(ylim = c(0.5, max(risk_cartoon$person_id) + 1.6)) +
  ggplot2::labs(
    x = "Time",
    y = NULL,
    shape = NULL,
    fill = NULL
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank()
  )
```

Cartoon risk-set view.
Each line is one individual.
Circles are events.
Squares are censoring times.
Dashed lines mark selected times.
Labels above show the number at risk immediately after each marked time.

:::

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_def-km-estimator.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_sec-km-product-limit-math.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_exr-nonparametric-survival.qmd >}}

:::{#thm-KM-est-no-cens}
#### Kaplan-Meier Estimate with No Censored Observations

If there are no censored data, and there are $n$ data points, then just
after (say) the third event time 

$$
\begin{aligned}
\hsurv(t) 
&= \prod_{t_i \le t}\sb{1-\frac{d_i}{r_i}}
\\ &= \sb{\frac{n-d_1}{n}} \sb{\frac{n-d_1-d_2}{n-d_1}} \sb{\frac{n-d_1-d_2-d_3}{n-d_1-d_2}}
\\ &= \frac{n-d_1-d_2-d_3}{n}
\\ &=1-\frac{d_1+d_2+d_3}{n}
\\ &=1-\hat F(t)
\end{aligned}
$$

where $\hat F(t)$ is the usual empirical CDF estimate.

:::

### Kaplan-Meier curve for `drug6mp` data

Here is the Kaplan-Meier estimated survival curve for the patients who
received 6-MP in the `drug6mp` dataset (we will see code to produce
figures like this one shortly):

```{r}
#| fig-cap: "Kaplan-Meier Survival Curve for 6-MP Patients"
#| label: fig-KM-mp6
#| echo: false

require(KMsurv)
data(drug6mp)
library(dplyr)
library(survival)

km_model1 <-
  drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, data = _)

library(ggfortify)
km_model1 |>
  autoplot(
    mark.time = TRUE,
    conf.int = FALSE
  ) +
  expand_limits(y = 0) +
  xlab("Time since diagnosis (months)") +
  ylab("KM Survival Curve")
```

### Kaplan-Meier calculations {.smaller}

Let's compute these estimates and build the chart by hand:

```{r}
library(KMsurv)
library(dplyr)
data(drug6mp)

data_v2 <-
  drug6mp |>
  as_tibble() |>
  mutate(
    remstat = remstat |>
      case_match(
        1 ~ "partial",
        2 ~ "complete"
      ),
    # renaming to "outcome" while relabeling is just a style choice:
    outcome = relapse |>
      case_match(
        0 ~ "censored",
        1 ~ "relapsed"
      )
  )

km_table <-
  data_v2 |>
  summarize(
    .by = t2,
    Relapses = sum(outcome == "relapsed"),
    Censored = sum(outcome == "censored")
  ) |>
  # here we add a start time row, so the graph starts at time 0:
  bind_rows(
    tibble(
      t2 = 0,
      Relapses = 0,
      Censored = 0
    )
  ) |>
  # sort in time order:
  arrange(t2) |>
  mutate(
    Exiting = Relapses + Censored,
    `Study Size` = sum(Exiting),
    Exited = cumsum(Exiting) |> dplyr::lag(default = 0),
    `At Risk` = `Study Size` - Exited,
    Hazard = Relapses / `At Risk`,
    `KM Factor` = 1 - Hazard,
    `Cumulative Hazard` = cumsum(`Hazard`),
    `KM Survival Curve` = cumprod(`KM Factor`)
  )

library(pander)
pander(km_table)
```

---

#### Summary

For the 6-MP patients at time 6 months, there are 21 patients at risk.
At $t=6$ there are 3 relapses and 1 censored observations.

The Kaplan-Meier factor is $(21-3)/21 = 0.857$. The number at risk for
the next time ($t=7$) is $21-3-1=17$.

At time 7 months, there are 17 patients at risk. At $t=7$ there is 1
relapse and 0 censored observations. The Kaplan-Meier factor is
$(17-1)/17 = 0.941$. The Kaplan Meier estimate is
$0.857\times0.941=0.807$. The number at risk for the next time ($t=9$)
is $17-1=16$.

---

Now, let's graph this estimated survival curve using `ggplot()`:

```{r}
#| label: fig-km-by-hand
#| fig-cap: "KM curve for 6MP patients, calculated by hand"
library(ggplot2)
conflicts_prefer(dplyr::filter)
km_table |>
  ggplot(aes(x = t2, y = `KM Survival Curve`)) +
  geom_step() +
  geom_point(data = km_table |> filter(Censored > 0), shape = 3) +
  expand_limits(y = c(0, 1), x = 0) +
  xlab("Time since diagnosis (months)") +
  ylab("KM Survival Curve") +
  scale_y_continuous(labels = scales::percent)
```

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_exr_survival_prediction_intervals.qmd >}}
