::: notes
For survival data, a more important quantity is the **survival
function**:
:::

:::{#def-surv-fn}

#### Survival function

:::: notes
Given a random time-to-event variable $T$,
the **survival function** or **survivor function**,
denoted $\surv(t)$,
is the probability
that the event time is later than $t$.
If the event in a clinical trial is death,
then $\surv(t)$ is the expected fraction
of the original population at time 0
who have survived up to time $t$
and are still alive at time $t$; that is:
::::

$$\surv(t) \eqdef \Pr(T > t)$$

:::

---

:::{#thm-survival-expressions-1}

$$
\begin{aligned}
\surv(t) &\eqdef \Pr(T > t)\\
&=\int_{u=t}^\infty p(u) du\\
&=1-F(t)\\
\end{aligned}
$$
:::

---

:::{#exm-exp-survfn}
##### exponential distribution

Since $\surv(t) = 1 - F(t)$, the survival function of the exponential
distribution family of models is:

$$
P(T> t) =
\begin{cases}
\ef{-\haz t}, & t \ge 0 \\
1, & t \le 0
\end{cases}
$$ where $\haz > 0$.

@fig-exp-survfuns shows some examples of exponential survival functions.

:::

---

```{r}
#| fig-cap: "Exponential Survival Functions"
#| label: fig-exp-survfuns
library(ggplot2)
ggplot() +
  geom_function(
    aes(col = "0.5"),
    fun = pexp,
    args = list(lower.tail = FALSE, rate = 0.5)
  ) +
  geom_function(
    aes(col = "1"),
    fun = pexp,
    args = list(lower.tail = FALSE, rate = 1)
  ) +
  geom_function(
    aes(col = "1.5"),
    fun = pexp,
    args = list(lower.tail = FALSE, rate = 1.5)
  ) +
  geom_function(
    aes(col = "5"),
    fun = pexp,
    args = list(lower.tail = FALSE, rate = 5)
  ) +
  theme_bw() +
  ylab("S(t)") +
  guides(col = guide_legend(title = expr(lambda))) +
  xlab("Time (t)") +
  xlim(0, 2.5) +
  theme(
    legend.position = "bottom",
    axis.title.x =
      element_text(
        angle = 0,
        vjust = 1,
        hjust = 1
      ),
    axis.title.y =
      element_text(
        angle = 0,
        vjust = 1,
        hjust = 1
      )
  )
```

---

:::{#thm-surv-fn-as-mean-status}

If $A_t$ represents survival status at time $t$, with $A_t = 1$ denoting alive at time $t$ and $A_t = 0$ denoting deceased at time $t$, then:

$$\surv(t) = \P(A_t=1) = \E{A_t}$$

:::

---

:::{#thm-surv-and-mean}

If $T$ is a nonnegative random variable, then:

$$\E{T} = \int_{t=0}^{\infty} \surv(t)dt$$

:::

---

:::{#fig-surv-and-mean-integration-bounds}

```{r}
#| code-fold: true
u_max <- 6

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
on.exit(par(old_par), add = TRUE)

plot(
  NA,
  xlim = c(0, u_max),
  ylim = c(0, u_max),
  xlab = "t",
  ylab = "u",
  main = "Before reversing order"
)
polygon(
  x = c(0, u_max, 0),
  y = c(0, u_max, u_max),
  col = grDevices::adjustcolor("steelblue", alpha.f = 0.3),
  border = NA
)
abline(a = 0, b = 1, lty = 2)
text(x = 0.6 * u_max, y = 0.2 * u_max, labels = "0 ≤ t ≤ u")
text(x = 0.2 * u_max, y = 0.93 * u_max, labels = "0 ≤ u < ∞")

plot(
  NA,
  xlim = c(0, u_max),
  ylim = c(0, u_max),
  xlab = "t",
  ylab = "u",
  main = "After reversing order"
)
polygon(
  x = c(0, u_max, 0),
  y = c(0, u_max, u_max),
  col = grDevices::adjustcolor("tomato", alpha.f = 0.3),
  border = NA
)
abline(a = 0, b = 1, lty = 2)
text(x = 0.2 * u_max, y = 0.93 * u_max, labels = "0 ≤ t < ∞")
text(x = 0.6 * u_max, y = 0.2 * u_max, labels = "t ≤ u < ∞")
```

The same triangular region
in the $(t, u)$-plane
can be written in two equivalent ways:
$0 \le t \le u < \infty$
or
$0 \le t < \infty,\ t \le u < \infty$.
These correspond
to the two integration orders
used in the proof.

:::

---

:::{.proof}

Adapted from the StatProofBook proof
"Mean of non-negative random variable"
[@statproofbook]
at
<https://statproofbook.github.io/P/mean-nnrvar.html>.

Figure @fig-surv-and-mean-integration-bounds
shows the same triangular integration region
written in two equivalent ways
before and after
reversing the order of integration.

$$
\ba
\E{T}
&= \int_{u=0}^{\infty} u\pdf(u)\,du\\
&= \int_{u=0}^{\infty}\paren{\int_{t=0}^{u} 1\,dt}\pdf(u)\,du\\
&= \int_{u=0}^{\infty}\int_{t=0}^{u} \pdf(u)\,dt\,du\\
&= \int_{t=0}^{\infty}\int_{u=t}^{\infty} \pdf(u)\,du\,dt\\
&= \int_{t=0}^{\infty}\P(T>t)\,dt\\
&= \int_{t=0}^{\infty}\surv(t)\,dt.
\ea
$$

:::
