
## Estimating survival in datasets without censoring

::: notes
When the data contain no censored observations (@def-right-censor),
we can easily estimate the survival function (@def-surv-fn).
We observe every patient's event time,
so for any time $t$ we simply count the fraction of patients whose event time is greater than $t$.
:::

:::{#def-emp-surv-no-cens}

#### Empirical survival function without censoring

:::: notes
Given event times $Y_1, \dots, Y_n$ with no censoring,
the **empirical survival function** $\hsurv(t)$
is the fraction of observed event times greater than $t$:
::::

$$\hsurv(t) \eqdef \frac{1}{n}\sumin \indicp{Y_i > t}$$

:::

---

:::{#thm-emp-surv-cdf}

#### Empirical survival function and empirical CDF

Without censoring,
the empirical survival function equals one minus the [empirical CDF](nonparametric-models.qmd#def-empirical-cdf) $\hcdf(t)$ of the event times:

$$\hsurv(t) = 1 - \hcdf(t)$$ {#eq-emp-surv-no-cens}

:::

::: proof
$$
\begin{aligned}
\hsurv(t)
&\eqdef \frac{1}{n}\sumin \indicp{Y_i > t}
\\ &= \frac{1}{n}\sumin \sb{1 - \indicp{Y_i \le t}}
\\ &= \frac{1}{n}\sumin 1 - \frac{1}{n}\sumin \indicp{Y_i \le t}
\\ &= 1 - \frac{1}{n}\sumin \indicp{Y_i \le t}
\\ &= 1 - \hcdf(t)
\end{aligned}
$$
:::

::: notes
$\hsurv(t)$ is a right-continuous step function
that starts at $\hsurv(0) = 1$
and drops by $d_j/n$ at each distinct event time $t_j$,
where $d_j$ is the number of events at $t_j$.
Without censoring,
@eq-emp-surv-no-cens matches the Kaplan-Meier estimator,
as we prove later in @thm-KM-est-no-cens.
:::

---

:::{#exm-surv-6mp-placebo}

#### Estimated survival function for placebo arm of 6-MP study

::: notes
In the `drug6mp` dataset (@exm-6mp-leukemia),
all of the placebo patients ($n = 21$) relapsed (exit times `t1`, no censoring),
so we can estimate their survival function using @def-emp-surv-no-cens:
:::

::: {#tbl-surv-placebo-no-cens}

```{r}
#| code-fold: true

library(KMsurv)
library(dplyr)
data(drug6mp)

placebo_times <- drug6mp$t1
emp_cdf <- ecdf(placebo_times)

placebo_surv <-
  tibble(t = sort(unique(placebo_times))) |>
  mutate(
    `n with time > t` = purrr::map_int(t, \(time) sum(placebo_times > time)),
    `S-hat(t)` = 1 - emp_cdf(t)
  )

placebo_surv
```

Empirical survival function $\hsurv(t)$ for the placebo arm of `drug6mp`
:::

::: {#fig-surv-placebo-no-cens}

```{r}
#| code-fold: true

library(ggplot2)

placebo_surv_plot <-
  placebo_surv |>
  select(t, `S-hat(t)`) |>
  add_row(t = 0, `S-hat(t)` = 1, .before = 1)

ggplot(placebo_surv_plot) +
  aes(x = t, y = `S-hat(t)`) +
  geom_step(direction = "hv") +
  expand_limits(y = c(0, 1)) +
  xlab("Time since diagnosis (months)") +
  ylab(expression(hat(S)(t))) +
  theme_bw()
```

Empirical survival curve $\hsurv(t)$ for the placebo arm of `drug6mp`
:::

:::

## Estimating survival in datasets with censoring

{{< include _subfiles/intro-to-survival-analysis/_sec-surv-KM-censoring.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_def-risk-set.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_def-km-estimator.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_sec-km-product-limit-math.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_exr-nonparametric-survival.qmd >}}

:::{#thm-KM-est-no-cens}
#### Kaplan-Meier Estimate with No Censored Observations

If there are no censored data, and there are $n$ data points, then just
after (say) the third event time

$$
\begin{aligned}
\hsurv(t)
&= \prod_{t_i \le t}\sb{1-\frac{d_i}{r_i}}
\\ &= \sb{\frac{n-d_1}{n}} \sb{\frac{n-d_1-d_2}{n-d_1}} \sb{\frac{n-d_1-d_2-d_3}{n-d_1-d_2}}
\\ &= \frac{n-d_1-d_2-d_3}{n}
\\ &=1-\frac{d_1+d_2+d_3}{n}
\\ &=1-\hcdf(t)
\end{aligned}
$$

where $\hcdf(t)$ is the usual empirical CDF estimate.

:::

{{< slidebreak >}}

## Variance of the Kaplan-Meier estimator

{{< include _subfiles/intro-to-survival-analysis/_sec-surv-KM-greenwood.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_exm-KM-drug6mp.qmd >}}

## Kaplan-Meier calculations {.smaller}

{{< include _subfiles/intro-to-survival-analysis/_sec-surv-KM-calculations.qmd >}}
