
---

:::{#def-na-cuhaz-est}
#### Nelson-Aalen Cumulative Hazard Estimator
:::: notes
The point hazard at time $t_i$ can be estimated by $d_i/Y_i$, which
leads to the **Nelson-Aalen estimator of the cumulative hazard**:
::::

{{< include _subfiles/intro-to-survival-analysis/_eq-NA-cuhaz-est.qmd >}}

:::

---

:::{#thm-var-NA-est}

#### Variance of Nelson-Aalen estimator

:::: notes
The variance of this estimator is approximately:
::::

$$
\begin{aligned}
\hat{\text{Var}}\left(\hat H_{NA} (t)\right)
&= \sum_{t_i <t}\frac{(d_i/Y_i)(1-d_i/Y_i)}{Y_i}\\
&\approx \sum_{t_i <t}\frac{d_i}{Y_i^2}
\end{aligned}
$${#eq-var-NA-cuhaz-est}

:::

---

Since $\surv(t)=\expf{-\cuhaz(t)}$, the Nelson-Aalen cumulative
hazard estimate can be converted into an alternate estimate of the
survival function:

$$
\ba
\hat S_{NA}(t)
&= \expf{-\hat H_{NA}(t)}\\
&= \expf{-\sum_{t_i < t}\frac{d_i}{Y_i}}\\
&= \prod_{t_i < t}\expf{-\frac{d_i}{Y_i}}\\
\ea
$$

---

Compare these with the corresponding Kaplan-Meier estimates:

$$
\ba
\hat H_{KM}(t) &= -\sum_{t_i < t}\logf{1-\frac{d_i}{Y_i}}\\
\hat S_{KM}(t) &= \prod_{t_i < t}\left[1-\frac{d_i}{Y_i}\right]
\ea
$$

::: notes
The product limit estimate and the Nelson-Aalen estimate often do not
differ by much. The latter is considered more accurate in small samples
and also directly estimates the cumulative hazard.
The `"fleming-harrington"` method for `survfit()` reduces to Nelson-Aalen
when the data are unweighted.
We can also estimate the cumulative hazard
as the negative log of the KM survival function estimate.
:::

### Application to `bmt` dataset

```{r}
na_fit <- survfit(
  formula = surv ~ group,
  type = "fleming-harrington",
  data = bmt
)

km_fit <- survfit(
  formula = surv ~ group,
  type = "kaplan-meier",
  data = bmt
)

km_and_na <-
  bind_rows(
    .id = "model",
    "Kaplan-Meier" = km_fit |> fortify(surv.connect = TRUE),
    "Nelson-Aalen" = na_fit |> fortify(surv.connect = TRUE)
  ) |>
  as_tibble()
```

```{r}
#| fig-cap: "Kaplan-Meier and Nelson-Aalen Survival Function Estimates, stratified by disease group"

km_and_na |>
  ggplot(aes(x = time, y = surv, col = model)) +
  geom_step() +
  facet_grid(. ~ strata) +
  theme_bw() +
  ylab("S(t) = P(T>=t)") +
  xlab("Survival time (t, days)") +
  theme(legend.position = "bottom")
```

The Kaplan-Meier and Nelson-Aalen survival estimates are very similar
for this dataset.
