{{< include latex-macros/macros.qmd >}}

This section is adapted from [@vittinghoff2e, Chapter 6].

### Choosing the Time Origin

Survival times are measured from some initial time,
with more than one possible choice of origin.

::: {#exm-pbc-time-origin}

#### Time origin choice in the PBC study

In the @dickson1989prognosis Primary Biliary Cholangitis (PBC) study,
survival time could be measured from:

- **Cohort enrollment**:
  time from when the patient entered the study
- **Diagnosis**:
  time from PBC diagnosis (more biologically meaningful)

When using time-since-diagnosis as the timescale,
patients who had already survived some time since diagnosis
before enrolling introduce **left-truncation**.

:::

### What is Left-Truncation?

**Left-truncation** occurs when some survival times are not observed
because the sampling scheme tends to miss short survival times.

::: {#def-left-truncation}

#### Left-Truncation

**Left-truncated** data arise when
individuals can only enter the study
if they survive beyond a certain time point.
Those who experience the event
before that time point are never observed.

:::

**Key feature:**
There must be a time delay between:

1. The event that defines the time origin (e.g., diagnosis)
2. Entry into the study cohort (e.g., enrollment)

### Left-Truncation vs. Staggered Entry

**Important distinction:**

- **Staggered entry**: Participants enroll at different calendar times
  - Does NOT necessarily imply left-truncation
  - Example: enrolling patients at the time of diagnosis
    (truncation times = 0)

- **Left-truncation**: Participants must survive some period
  before they can be enrolled
  - Example: recruiting from a referral center
    where months/years elapse between diagnosis and enrollment

### Why Left-Truncation Matters

Patients with rapid disease progression
are less likely to be enrolled because they may:

- Die before referral to the study center
- Die before recruitment into the study

This leads to:

- **Undercounting** of short survival times
- **Selection bias** toward longer survivors
- **Overestimation** of survival probabilities if ignored

### Truncated vs. Censored Data

**Censored data:**

- Event time falls outside the follow-up period
- Data are **incomplete** but the individual is observed

**Truncated data:**

- Individual is not observed at all
  if they would have had an event outside the observation window
- Truncated individuals leave no trace
  - Called "ghosts" in the survival analysis literature

### Independent Truncation Assumption

Analysis under left-truncation
requires the **independent truncation assumption**:

The time of delayed entry and subsequent survival are independent.

This is satisfied when:

- Disease incidence and post-diagnosis survival are independent

### Implementation

Denote:

- $V$: truncation time (delay between origin and study entry)
- $X$: follow-up time relative to the time origin
- $\delta$: censoring indicator

**In R:**

Use the `Surv()` function with entry time:

```r
Surv(time  = entry_time,
     time2 = exit_time,
     event = status)
```

**In Stata:**

```stata
stset years_since_diag, failure(status) entry(disease_dur)
```

### Numerical Example: PBC Study

The `pbc` dataset from the `rmb` package
contains data from a clinical trial
of D-penicillamine treatment in primary biliary cholangitis (PBC)
[@dickson1989prognosis].

Here we demonstrate the effect of left-truncation
by comparing analyses using two different time scales:

1. Time from enrollment (standard approach, no truncation)
2. Age as the time scale (introduces left-truncation,
   since patients are observed from their current age, not from birth)

::: {#fig-pbc-km-truncation}

```{r}
#| code-fold: true
library(rmb)
library(survival)
library(ggplot2)
library(dplyr)

pbc_data <- rmb::pbc

# Analysis 1: time from enrollment (no truncation)
km_enroll <- survfit(
  Surv(years, as.numeric(status) == 1) ~ 1,
  data = pbc_data
)

# Analysis 2: age as time scale (left-truncated)
km_age <- survfit(
  Surv(time = age, time2 = age + years, event = as.numeric(status) == 1) ~ 1,
  data = pbc_data
)

# Extract KM estimates at specific times for each analysis
km1_summary <- summary(km_enroll, times = seq(0, 12, by = 0.5))
km2_summary <- summary(km_age, times = seq(26, 78, by = 0.5))

km1_df <- tibble(
  time = km1_summary$time,
  surv = km1_summary$surv,
  analysis = "Time from enrollment"
)

km2_df <- tibble(
  time = km2_summary$time - median(pbc_data$age),
  surv = km2_summary$surv,
  analysis = paste0(
    "Age as time scale\n",
    "(left-truncated; centered at median enrollment age)"
  )
)

km_combined <- bind_rows(km1_df, km2_df)

ggplot(km_combined) +
  aes(x = time, y = surv, color = analysis) +
  geom_step() +
  labs(
    x = "Time (years)",
    y = "Estimated survival probability",
    color = "Analysis",
    caption = "Data: rmb::pbc (PBC clinical trial, Dickson et al. 1989)"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Kaplan-Meier survival curves for the PBC study
using two different time scales:
time from enrollment (no truncation)
vs. age centered at the median enrollment age (left-truncated).

:::

::: {.notes}
The curve using age as the time scale
accounts for the fact that participants
were only observed from their age at enrollment.
When centered at the median enrollment age,
the two analyses can be compared on a common scale.
Note that ignoring this truncation
would lead to overestimated survival:
participants who died at younger ages ---
before reaching their enrollment age ---
were never observed and are excluded from the data,
leading to a risk set that over-represents longer survivors.
:::

### Effect of Ignoring Left-Truncation

When left-truncation is ignored:

- Survival probabilities are **overestimated**
  (especially at early time points)
- The analysis fails to account
  for undersampling of short survival times
- Hazard ratios in Cox models may be attenuated
  (effect is less predictable)

The following example illustrates the bias using the PBC data.
The naive analysis treats all participants as entering the risk set
at the earliest observed age,
whereas the proper analysis uses counting-process notation
to enter each participant at their actual enrollment age.

```{r}
#| label: pbc-truncation-bias-setup
#| code-fold: true
# Naive analysis: ignoring delayed entry
# (treats all subjects as if they entered the risk set at age 0)
km_naive_age <- survfit(
  Surv(age + years, as.numeric(status) == 1) ~ 1,
  data = pbc_data
)

# Extract estimates at age 60 for comparison
surv_naive_at_sixty  <- summary(km_naive_age, times = 60)$surv
surv_proper_at_sixty <- summary(km_age, times = 60)$surv
```

At age 60,
the naive analysis (ignoring delayed entry)
estimates survival at
`r scales::percent(surv_naive_at_sixty, accuracy = 0.1)`,
compared to
`r scales::percent(surv_proper_at_sixty, accuracy = 0.1)`
with the proper left-truncation correction.

The naive estimate is inflated
because the risk set at each age incorrectly includes
participants who had not yet been enrolled
(and thus had not actually been observed to be at risk).

### Right-Truncation

**Right-truncation** can also occur if:

- Study recruits based on an endpoint
- People with large event times (or who never had the event)
  are not recruited

**Example:**
A fecundability study that excludes couples
who never conceive
exhibits right-truncation.
