
(a.k.a. the Mantel-Cox test)

{{< slidebreak >}}

:::{#exr-compare-surv-ests}
How do we test the null hypothesis that two or more groups
have the same time-to-event distribution?
:::

{{< slidebreak >}}

:::{#sol-compare-surv-ests}
One option is the **log-rank test**,
which compares the Kaplan-Meier estimates of the survival functions of
those groups.
:::

{{< slidebreak >}}

::: notes
Adapted from [@kleinbaum2012survival, Section IV, pp. 67--68]:

- The log-rank test is a large-sample chi-square test.

- The log-rank test uses a test statistic that compares KM curves
between groups across all survival times.

- Like many other statistics used in chi-square tests,
the log-rank statistic compares observed versus expected cell counts
over categories of outcomes.

- The categories for the log-rank statistic
are defined by each of the ordered failure times for the entire set of
data being analyzed.

:::

## Notation {#sec-logrank-notation}

Suppose we are comparing $G \ge 2$ groups, indexed by $g = 1, \ldots, G$.
Pool the failure times across all groups and let

- $t_1 < t_2 < \cdots < t_k$ be the **distinct ordered failure times**
in the pooled data ($i = 1, \ldots, k$ indexes these times).

For each ordered failure time $t_i$ and each group $g$, define:

| Symbol | Meaning |
|---|---|
| $d_{gi}$ | number of failures in group $g$ at time $t_i$ |
| $r_{gi}$ | number at risk in group $g$ just before time $t_i$ |
| $d_i = \sum_{g=1}^{G} d_{gi}$ | total failures at $t_i$ (pooled across groups) |
| $r_i = \sum_{g=1}^{G} r_{gi}$ | total at risk at $t_i$ (pooled across groups) |

This matches the $(d_i, r_i)$ notation used for the Kaplan-Meier estimator
elsewhere in this chapter; here we additionally split each count by group.

[@kleinbaum2012survival, pp. 67--69] uses $m_{if}$ and $n_{if}$ in place of
$d_{gi}$ and $r_{gi}$; we have changed letters to stay consistent with the
rest of the chapter.

## Expected counts under the null

The null hypothesis is

$$H_0:\; \text{all $G$ groups have the same survival function.}$$

Under $H_0$, the pooled marginal hazard estimate at time $t_i$ is

$$\hat{\haz}_i = \frac{d_i}{r_i}.$$

The **expected** number of failures in group $g$ at time $t_i$ under $H_0$
is the pooled hazard applied to that group's risk set:

$$
e_{gi}
\;=\; r_{gi}\,\hat{\haz}_i
\;=\; \underbrace{\frac{r_{gi}}{r_i}}_{\text{fraction at risk}}\;\cdot\;
      \underbrace{d_i}_{\text{total failures}}
$$ {#eq-logrank-expected}

[@kleinbaum2012survival, p. 68, expected-cell-counts box].

The **summed observed minus expected score** for group $g$ over all
failure times is:

$$O_g - E_g \;\eqdef\; \sum_{i=1}^{k} \paren{d_{gi} - e_{gi}}.$$

## Log-rank statistic (two groups)

For two groups ($G = 2$), the observed-minus-expected scores satisfy
$O_1 - E_1 = -(O_2 - E_2)$ (the total failures at each $t_i$ are by
definition partitioned across the two groups), so it does not matter
which group we use to form the test statistic.

:::{#thm-logrank-two-group}
#### Log-rank statistic, two groups

$$
\chi^2_{\text{LR}}
\;=\; \frac{(O_g - E_g)^2}{\widehat{\Varf{O_g - E_g}}}
\;\stackrel{H_0}{\sim}\; \chi^2_1
$$ {#eq-logrank-stat-2grp}

where the variance estimate is

$$
\widehat{\Varf{O_g - E_g}}
\;=\; \sum_{i=1}^{k}
   \frac{r_{1i}\, r_{2i}\, d_i\, (r_i - d_i)}{r_i^2\,(r_i - 1)}.
$$ {#eq-logrank-var-2grp}

For two groups, the variance is the same whether computed for group 1 or
group 2.

[@kleinbaum2012survival, pp. 69--70, two-group variance formula]

:::

::: notes
@eq-logrank-var-2grp is the hypergeometric variance for the count of
failures from group 1 at $t_i$ conditional on $(d_i, r_{1i}, r_{2i})$,
summed over failure times. Each summand has the form
$r_{1i} r_{2i} d_i (r_i - d_i) / [r_i^2 (r_i - 1)]$.
:::

## Log-rank statistic (more than two groups)

For $G \ge 2$ groups, the log-rank statistic still uses the
observed-minus-expected scores $O_g - E_g$, but the variance--covariance
matrix of the vector
$\mathbf{O} - \mathbf{E} \eqdef (O_1 - E_1, \ldots, O_G - E_G)'$
is nontrivial because the scores are linearly dependent
($\sum_g (O_g - E_g) = 0$). The statistic takes the matrix-quadratic
form

$$
\chi^2_{\text{LR}}
\;=\; (\mathbf{O} - \mathbf{E})'\,\widehat{\Sigma}^{-}\,(\mathbf{O} - \mathbf{E})
\;\stackrel{H_0}{\sim}\; \chi^2_{G - 1}
$$

where $\widehat\Sigma$ is the estimated variance--covariance matrix of
$\mathbf{O} - \mathbf{E}$ and $\widehat\Sigma^{-}$ is a generalized
inverse (any $G - 1$ of the rows/columns suffice, since the scores sum
to zero).

[@kleinbaum2012survival, pp. 71--72 and Chapter 2 Appendix for the matrix
formula]

::: notes
In practice we don't compute this by hand — every standard survival
package (`survival::survdiff` in R, `sts test` in Stata, `LIFETEST` in
SAS) returns the exact log-rank statistic.
:::

{{< slidebreak >}}

## Approximate (Pearson-style) statistic

When only $(O_g, E_g)$ are available — e.g., from a printed summary —
the variance--covariance matrix can be replaced by the diagonal $E_g$
entries to yield the classic Pearson-style approximation:

$$X^2 \;\approx\; \sum_{g=1}^{G} \frac{(O_g - E_g)^2}{E_g}.$$

This is what some textbooks call "the log-rank test"; it is slightly
smaller than the exact statistic but typically close.

[@kleinbaum2012survival, p. 70 (approximate formula)]

## The `survdiff` function

```{r, printr.help.sections = c("description", "usage")}
?survdiff
```

:::::{#exm-survdiff-drug6mp}
#### `survdiff()` with `drug6mp` data

Now we are going to compare the placebo and 6-MP data. We need to
reshape the data to make it usable with the standard `survival`
workflow:

```{r}
library(survival)
library(tidyr)
data_v4 <-
  data_v3 |>
  select(pair, remstat, t1, t2, outcome) |>
  # here we are going to change the data from a wide format to long:
  pivot_longer(
    cols = c(t1, t2),
    names_to = "treatment",
    values_to = "exit_time"
  ) |>
  mutate(
    treatment = treatment |>
      case_match(
        "t1" ~ "placebo",
        "t2" ~ "6-MP"
      ),
    outcome = if_else(
      treatment == "placebo",
      "relapsed",
      outcome
    ),
    surv = Surv(
      time = exit_time,
      event = (outcome == "relapsed")
    )
  )
```

{{< slidebreak >}}

Using this long data format, we can fit a Kaplan-Meier curve for each
treatment group simultaneously:

```{r}
km_model2 <-
  survfit(
    formula = surv ~ treatment,
    data = data_v4
  )
```

{{< slidebreak >}}

We can plot the curves in the same graph:

```{r}
km_model2 |> autoplot()
```

{{< slidebreak >}}

We can also perform something like a t-test, where the null hypothesis
is that the curves are the same. The table below shows the per-time
$(d_{gi}, r_{gi})$ counts and the expected counts $e_{gi}$ from
@eq-logrank-expected:

:::{#tbl-logrank-6mp}
```{r}

o_e <- data_v4 |>
  arrange(exit_time) |>
  mutate(
    .by = treatment,
    n_exited = row_number(),
    n_at_risk = n() - n_exited + 1
  ) |>
  dplyr::summarize(
    .by = all_of(c("exit_time", "treatment")),
    n_at_risk = max(n_at_risk),
    n_events = sum(outcome == "relapsed")
  ) |>
  tidyr::pivot_wider(
    names_from = "treatment",
    values_from = c(n_at_risk, n_events)
  ) |>
  tidyr::fill(
    starts_with("n_at_risk"),
    .direction = "up"
  ) |>
  replace_na(list("n_events_placebo" = 0,
                  "n_events_6-MP" = 0)) |>
  mutate(
    n_at_risk = rowSums(across(starts_with("n_at_risk"))),
    n_events = rowSums(across(starts_with("n_events"))),
    marginal_hazard = n_events / n_at_risk,
    expected_6mp = marginal_hazard * `n_at_risk_6-MP`,
    expected_plc = marginal_hazard * n_at_risk_placebo,
    diff_6mp = `n_events_6-MP` - expected_6mp,
    diff_plc = n_events_placebo - expected_plc
  ) |>
  filter(n_events > 0)

o_e

```
Observed and expected event counts for the 6-MP data, for the log-rank
test. Columns: `exit_time` is $t_i$;
`n_at_risk_placebo`/`n_at_risk_6-MP` are $r_{gi}$;
`n_events_placebo`/`n_events_6-MP` are $d_{gi}$;
`n_at_risk`/`n_events` are pooled $r_i$, $d_i$;
`marginal_hazard` is $\hat{\haz}_i = d_i / r_i$;
`expected_*` are $e_{gi}$ from @eq-logrank-expected;
`diff_*` are the per-time contributions $d_{gi} - e_{gi}$ to $O_g - E_g$.
:::

The summed observed $O_g$ and expected $E_g$ totals (one row per group):

:::{#tbl-sum-obs-exp}
```{r}
o_e_summ <- o_e |>
  summarize(
    across(starts_with("expected"), sum),
    across(starts_with("n_events_"), sum)
  )
pander::pander(o_e_summ)
```
Sums of observed and expected events for the 6-MP data.
:::

{{< slidebreak >}}

Plugging $(O_g - E_g)^2$ and the variance into @eq-logrank-stat-2grp
gives the exact log-rank statistic; the Pearson-style approximation is

$$X^2 \;\approx\; \sum_{g} \frac{(O_g - E_g)^2}{E_g}$$

which we compute directly from @tbl-sum-obs-exp:

```{r}
with(
  o_e_summ,
  tibble(
    "6mp" = (`n_events_6-MP` - expected_6mp)^2 / expected_6mp,
    "placebo" = (n_events_placebo - expected_plc)^2 / expected_plc,
    sum = `6mp` + placebo
  )
) |>
  pander::pander()
```

{{< slidebreak >}}

R gives us both the exact and approximate results:

```{r}
survdiff(
  formula = surv ~ treatment,
  data = data_v4
)
```

{{< slidebreak >}}

::: notes
By default, `survdiff()` ignores any pairing,
but we can use `strata()` to perform something similar to a paired t-test:
:::

```{r}
lrank_test <- survdiff(
  formula = surv ~ treatment + strata(pair),
  data = data_v4
)
lrank_test
```

::: notes
Interestingly, accounting for pairing reduces the significance of the
difference.
:::

:::::
