:::{#def-emp-inv-surv-fn}

#### Empirical inverse survival function

Given a time-to-event dataset,
the **empirical inverse survival function** (empirical ISF),
also called the **empirical survival quantile function** (empirical SQF),
$\hsurv^{-1}(p)$
replaces the population survival function $\surv(t)$
with an estimated survival function $\hsurv(t)$:

$$\hsurv^{-1}(p) \eqdef \inf\cb{t \ge 0 : \hsurv(t) \le p}, \quad 0 < p < 1.$$

:::

---

Since $\hsurv(t)$ is a non-increasing step function,
$\hsurv^{-1}(p)$ equals the smallest observed event time
at which the estimated survival probability has dropped to or below $p$.

**From the Kaplan-Meier estimator.**
Substituting $\hsurv_{KM}(t)$ from @def-KM-estimator:

$$
\hsurv_{KM}^{-1}(p)
= \inf\cb{t \ge 0 : \prod_{t_i \le t}\sb{1 - \frac{d_i}{r_i}} \le p}.
$$

**From the Nelson-Aalen estimator.**
Substituting $\hsurv_{NA}(t) = \expf{-\hat{\cuhaz}_{NA}(t)}$
from @def-na-cuhaz-est:

$$
\hsurv_{NA}^{-1}(p)
= \inf\cb{t \ge 0 : \expf{-\sum_{t_i \le t}\frac{d_i}{r_i}} \le p}.
$$

In R, both estimators are computed from a `survfit()` object using `quantile()`.

---

:::{#exm-emp-isf-drug6mp}
##### Empirical ISF for 6-MP patients

Using the `drug6mp` dataset,
we compute the empirical ISF for the 6-MP patients.
The KM and Nelson-Aalen survival estimates at each event time are:

```{r}
#| code-fold: true
library(KMsurv)
library(dplyr)
library(survival)
data(drug6mp)

km_fit <- drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, data = _)

na_fit <- drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, type = "fleming-harrington", data = _)

tbl <- tibble(
  time = km_fit$time,
  n.risk = km_fit$n.risk,
  n.event = km_fit$n.event,
  surv_km = km_fit$surv,
  surv_na = na_fit$surv
) |>
  filter(n.event > 0)

library(knitr)
kable(
  tbl,
  digits = 3,
  col.names = c(
    "Time (months)",
    "At risk $r_i$",
    "Events $d_i$",
    "$\\hat{S}_{KM}(t)$",
    "$\\hat{S}_{NA}(t)$"
  )
)
```

:::{#fig-emp-surv-drug6mp}

```{r}
#| code-fold: true

t_all_km <- c(0, km_fit$time)
s_all_km <- c(1, km_fit$surv)
t_all_na <- c(0, na_fit$time)
s_all_na <- c(1, na_fit$surv)
t_max_plot <- max(km_fit$time, na_fit$time) * 1.05

draw_surv_step <- function(t_all, s_all, t_end, col) {
  n <- length(t_all)
  segments(t_all, s_all, c(t_all[-1], t_end), s_all, col = col, lwd = 2)
  points(t_all[-1], s_all[-n], pch = 1, col = col)
  points(t_all, s_all, pch = 19, col = col)
}

plot(
  0, 0, type = "n",
  xlim = c(0, t_max_plot), ylim = c(0, 1.05),
  xlab = "Time (months)",
  ylab = expression(hat(S)(t))
)
draw_surv_step(t_all_km, s_all_km, t_max_plot, "steelblue")
draw_surv_step(t_all_na, s_all_na, t_max_plot, "darkorange")
legend(
  "topright",
  legend = c("Kaplan-Meier", "Nelson-Aalen"),
  col = c("steelblue", "darkorange"),
  lwd = 2, pch = 19, bty = "n"
)
```

Estimated survival functions for 6-MP patients from the KM and
Nelson-Aalen estimators.

:::

:::{#fig-emp-isf-drug6mp}

```{r}
#| code-fold: true

draw_isf_step <- function(s_ev, t_ev, col) {
  p_right <- c(1, s_ev[-length(s_ev)])
  segments(s_ev, t_ev, p_right, t_ev, col = col, lwd = 2)
  points(s_ev, t_ev, pch = 19, col = col)
  points(p_right, t_ev, pch = 1, col = col)
}

ev_km <- km_fit$n.event > 0
t_ev_km <- km_fit$time[ev_km]
s_ev_km <- km_fit$surv[ev_km]

ev_na <- na_fit$n.event > 0
t_ev_na <- na_fit$time[ev_na]
s_ev_na <- na_fit$surv[ev_na]

t_max_isf <- max(t_ev_km, t_ev_na) * 1.05

plot(
  0, 0, type = "n",
  xlim = c(0, 1), ylim = c(0, t_max_isf),
  xlab = "p",
  ylab = expression(hat(S)^{-1}*(p))
)
draw_isf_step(s_ev_km, t_ev_km, "steelblue")
draw_isf_step(s_ev_na, t_ev_na, "darkorange")
legend(
  "topright",
  legend = c("Kaplan-Meier", "Nelson-Aalen"),
  col = c("steelblue", "darkorange"),
  lwd = 2, pch = 19, bty = "n"
)
```

Empirical inverse survival function $\hsurv^{-1}(p)$ for 6-MP patients
from the KM and Nelson-Aalen estimators.

:::

The empirical median survival time is the smallest event time $t$
at which the estimated survival probability falls to or below 0.5.

From the table:

- $\hsurv_{KM}(22) \approx 0.538 > 0.5$
  and $\hsurv_{KM}(23) \approx 0.448 \le 0.5$,
  so $\hsurv_{KM}^{-1}(0.5) = 23$ months.

- $\hsurv_{NA}(22) \approx 0.557 > 0.5$
  and $\hsurv_{NA}(23) \approx 0.471 \le 0.5$,
  so $\hsurv_{NA}^{-1}(0.5) = 23$ months.

In R:

```{r}
#| code-fold: show
quantile(km_fit, probs = 0.5)
```

```{r}
#| code-fold: show
quantile(na_fit, probs = 0.5)
```

Both the KM and Nelson-Aalen estimators give an estimated median
relapse-free survival time of **23 months** for the 6-MP patients.

:::

