:::{#def-emp-inv-surv-fn}

#### Empirical inverse survival function

Given a time-to-event dataset,
the **empirical inverse survival function**
$\hsurv^{-1}(p)$
replaces the population survival function $\surv(t)$
with an estimated survival function $\hsurv(t)$:

$$\hsurv^{-1}(p) \eqdef \inf\cb{t \ge 0 : \hsurv(t) \le p}, \quad 0 < p < 1.$$

:::

---

Since $\hsurv(t)$ is a non-increasing step function,
$\hsurv^{-1}(p)$ equals the smallest observed event time
at which the estimated survival probability has dropped to or below $p$.

**From the Kaplan-Meier estimator.**
Substituting $\hsurv_{KM}(t)$ from @def-KM-estimator:

$$
\hsurv_{KM}^{-1}(p)
= \inf\cb{t \ge 0 : \prod_{t_i \le t}\sb{1 - \frac{d_i}{r_i}} \le p}.
$$

**From the Nelson-Aalen estimator.**
Substituting $\hsurv_{NA}(t) = \expf{-\hat{\cuhaz}_{NA}(t)}$
from @def-na-cuhaz-est:

$$
\hsurv_{NA}^{-1}(p)
= \inf\cb{t \ge 0 : \expf{-\sum_{t_i \le t}\frac{d_i}{r_i}} \le p}.
$$

In R, both estimators are computed from a `survfit()` object using `quantile()`.

---

:::{#exm-emp-isf-drug6mp}
##### Empirical ISF for 6-MP patients

Using the `drug6mp` dataset,
we compute the empirical ISF for the 6-MP patients.
The KM and Nelson-Aalen survival estimates at each event time are:

```{r}
#| code-fold: true
library(KMsurv)
library(dplyr)
library(survival)
data(drug6mp)

km_fit <- drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, data = _)

na_fit <- drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, type = "fleming-harrington", data = _)

tbl <- tibble(
  time = km_fit$time,
  n.risk = km_fit$n.risk,
  n.event = km_fit$n.event,
  surv_km = km_fit$surv,
  surv_na = na_fit$surv
) |>
  filter(n.event > 0)

library(knitr)
kable(
  tbl,
  digits = 3,
  col.names = c(
    "Time (months)",
    "At risk $r_i$",
    "Events $d_i$",
    "$\\hat{S}_{KM}(t)$",
    "$\\hat{S}_{NA}(t)$"
  )
)
```

@fig-drug6mp-surv-est shows the estimated survival curves from both
estimators, and @fig-drug6mp-isf-est shows the corresponding empirical
inverse survival functions.

:::{#fig-drug6mp-surv-est}
```{r}
#| code-fold: true
library(ggplot2)

surv_plot_data <- bind_rows(
  tibble(
    time = c(0, km_fit$time),
    surv = c(1, km_fit$surv),
    estimator = "KM"
  ),
  tibble(
    time = c(0, na_fit$time),
    surv = c(1, na_fit$surv),
    estimator = "Nelson-Aalen"
  )
)

ggplot(surv_plot_data, aes(x = time, y = surv, col = estimator)) +
  geom_step() +
  theme_bw() +
  xlab("Time (months)") +
  ylab(expression(hat(S)(t))) +
  expand_limits(y = 0) +
  labs(color = "Estimator") +
  theme(
    axis.title.x = element_text(angle = 0, vjust = 1, hjust = 1),
    axis.title.y = element_text(angle = 0, vjust = 1, hjust = 1)
  )
```
Estimated survival functions for 6-MP patients using the Kaplan-Meier
and Nelson-Aalen estimators.
:::

:::{#fig-drug6mp-isf-est}
```{r}
#| code-fold: true
library(ggplot2)
library(dplyr)

km_event_mask <- km_fit$n.event > 0
na_event_mask <- na_fit$n.event > 0

isf_plot_data <- bind_rows(
  tibble(
    p = c(km_fit$surv[km_event_mask], 1),
    t = c(
      km_fit$time[km_event_mask],
      km_fit$time[km_event_mask][1]
    ),
    estimator = "KM"
  ),
  tibble(
    p = c(na_fit$surv[na_event_mask], 1),
    t = c(
      na_fit$time[na_event_mask],
      na_fit$time[na_event_mask][1]
    ),
    estimator = "Nelson-Aalen"
  )
) |>
  arrange(estimator, p)

ggplot(isf_plot_data, aes(x = p, y = t, col = estimator)) +
  geom_step(direction = "hv") +
  theme_bw() +
  xlab("p") +
  ylab(expression(hat(S)^{-1}(p))) +
  labs(color = "Estimator") +
  theme(
    axis.title.x = element_text(angle = 0, vjust = 1, hjust = 1),
    axis.title.y = element_text(angle = 0, vjust = 1, hjust = 1)
  )
```
Empirical inverse survival functions for 6-MP patients from the
Kaplan-Meier and Nelson-Aalen estimators.
Each step indicates the smallest event time at which the estimated
survival probability drops to or below $p$.
:::

The empirical median survival time is the smallest event time $t$
at which the estimated survival probability falls to or below 0.5.

From the table:

- $\hsurv_{KM}(22) \approx 0.538 > 0.5$
  and $\hsurv_{KM}(23) \approx 0.448 \le 0.5$,
  so $\hsurv_{KM}^{-1}(0.5) = 23$ months.

- $\hsurv_{NA}(22) \approx 0.557 > 0.5$
  and $\hsurv_{NA}(23) \approx 0.471 \le 0.5$,
  so $\hsurv_{NA}^{-1}(0.5) = 23$ months.

In R:

```{r}
#| code-fold: show
quantile(km_fit, probs = 0.5)
```

```{r}
#| code-fold: show
quantile(na_fit, probs = 0.5)
```

Both the KM and Nelson-Aalen estimators give an estimated median
relapse-free survival time of **23 months** for the 6-MP patients.

:::
