:::{#def-emp-inv-surv-fn}

#### Empirical inverse survival function

Given a time-to-event dataset,
the **empirical inverse survival function** (empirical ISF),
also called the **empirical survival quantile function** (empirical SQF),
$\hsurv^{-1}(p)$
replaces the population survival function $\surv(t)$
with an estimated survival function $\hsurv(t)$:

$$\hsurv^{-1}(p) \eqdef \inf\cb{t \ge 0 : \hsurv(t) \le p}, \quad 0 < p < 1.$$

:::

{{< slidebreak >}}

Since $\hsurv(t)$ is a non-increasing step function,
$\hsurv^{-1}(p)$ equals the smallest observed event time
at which the estimated survival probability has dropped to or below $p$.

**From the Kaplan-Meier estimator.**
Substituting $\hsurvkmf{t}$ from @def-KM-estimator:

$$
\hsurvkm^{-1}(p)
= \inf\cb{t \ge 0 : \prod_{t_i \le t}\sb{1 - \frac{d_i}{r_i}} \le p}.
$$

**From the Nelson-Aalen estimator.**
Substituting $\hsurvnaaf{t} = \expf{-\hat{\cuhaz}_{NA}(t)}$
from @def-na-cuhaz-est:

$$
\hsurvnaa^{-1}(p)
= \inf\cb{t \ge 0 : \expf{-\sum_{t_i \le t}\frac{d_i}{r_i}} \le p}.
$$

In R, both estimators are computed from a `survfit()` object using `quantile()`.

{{< slidebreak >}}

:::{#exm-emp-isf-drug6mp}
##### Empirical ISF for 6-MP patients

Using the `drug6mp` dataset,
we compute the empirical ISF for the 6-MP patients.
The KM and Nelson-Aalen survival estimates at each event time are:

```{r}
#| code-fold: true
library(KMsurv)
library(dplyr)
library(survival)
data(drug6mp)

km_fit <- drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, data = _)

na_fit <- drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, type = "fleming-harrington", data = _)

tbl <- tibble(
  time = km_fit$time,
  n.risk = km_fit$n.risk,
  n.event = km_fit$n.event,
  surv_km = km_fit$surv,
  surv_na = na_fit$surv
) |>
  filter(n.event > 0)

library(knitr)
kable(
  tbl,
  digits = 3,
  col.names = c(
    "Time (months)",
    "At risk $r_i$",
    "Events $d_i$",
    "$\\hsurvkmf{t}$",
    "$\\hsurvnaaf{t}$"
  )
)
```

:::{#fig-emp-surv-drug6mp}

```{r}
#| code-fold: true
library(ggplot2)

surv_steps <- rbind(
  data.frame(
    time = c(0, km_fit$time), surv = c(1, km_fit$surv),
    estimator = "Kaplan-Meier"
  ),
  data.frame(
    time = c(0, na_fit$time), surv = c(1, na_fit$surv),
    estimator = "Nelson-Aalen"
  )
)

cens_marks <- rbind(
  data.frame(
    time = km_fit$time[km_fit$n.censor > 0],
    surv = km_fit$surv[km_fit$n.censor > 0],
    estimator = "Kaplan-Meier"
  ),
  data.frame(
    time = na_fit$time[na_fit$n.censor > 0],
    surv = na_fit$surv[na_fit$n.censor > 0],
    estimator = "Nelson-Aalen"
  )
)

ev_idx_km <- km_fit$n.event > 0
s_after_km <- km_fit$surv[ev_idx_km]
s_before_km <- c(1, head(s_after_km, -1))
ev_idx_na <- na_fit$n.event > 0
s_after_na <- na_fit$surv[ev_idx_na]
s_before_na <- c(1, head(s_after_na, -1))

open_surv <- rbind(
  data.frame(time = km_fit$time[ev_idx_km], surv = s_before_km, estimator = "Kaplan-Meier"),
  data.frame(time = na_fit$time[ev_idx_na], surv = s_before_na, estimator = "Nelson-Aalen")
)
closed_surv <- rbind(
  data.frame(time = km_fit$time[ev_idx_km], surv = s_after_km, estimator = "Kaplan-Meier"),
  data.frame(time = na_fit$time[ev_idx_na], surv = s_after_na, estimator = "Nelson-Aalen")
)

ggplot(surv_steps, aes(x = time, y = surv, color = estimator)) +
  geom_step() +
  geom_point(data = cens_marks, shape = 3, size = 2, show.legend = FALSE) +
  geom_point(data = open_surv, shape = 1, size = 2) +
  geom_point(data = closed_surv, shape = 19, size = 2) +
  scale_color_manual(
    values = c("Kaplan-Meier" = "steelblue", "Nelson-Aalen" = "darkorange")
  ) +
  coord_cartesian(ylim = c(0, 1.05)) +
  labs(x = "Time (months)", y = expression(hat(S)(t)), color = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

Estimated survival functions for 6-MP patients from the KM and
Nelson-Aalen estimators.

:::

:::{#fig-emp-isf-drug6mp}

```{r}
#| code-fold: true

ev_km <- km_fit$n.event > 0
t_ev_km <- km_fit$time[ev_km]
s_ev_km <- km_fit$surv[ev_km]

ev_na <- na_fit$n.event > 0
t_ev_na <- na_fit$time[ev_na]
s_ev_na <- na_fit$surv[ev_na]

isf_steps <- rbind(
  data.frame(
    p = c(1, s_ev_km), t = c(t_ev_km[1], t_ev_km),
    estimator = "Kaplan-Meier"
  ),
  data.frame(
    p = c(1, s_ev_na), t = c(t_ev_na[1], t_ev_na),
    estimator = "Nelson-Aalen"
  )
)
isf_steps <- isf_steps[order(isf_steps$estimator, isf_steps$p), ]

open_isf <- rbind(
  data.frame(p = head(s_ev_km, -1), t = tail(t_ev_km, -1), estimator = "Kaplan-Meier"),
  data.frame(p = head(s_ev_na, -1), t = tail(t_ev_na, -1), estimator = "Nelson-Aalen")
)
closed_isf <- rbind(
  data.frame(p = s_ev_km, t = t_ev_km, estimator = "Kaplan-Meier"),
  data.frame(p = s_ev_na, t = t_ev_na, estimator = "Nelson-Aalen")
)

ggplot(isf_steps, aes(x = p, y = t, color = estimator)) +
  geom_step() +
  geom_point(data = open_isf, shape = 1, size = 2) +
  geom_point(data = closed_isf, shape = 19, size = 2) +
  scale_color_manual(
    values = c("Kaplan-Meier" = "steelblue", "Nelson-Aalen" = "darkorange")
  ) +
  coord_cartesian(xlim = c(0, 1)) +
  labs(x = "p", y = expression(hat(S)^{-1}*(p)), color = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

Empirical inverse survival function $\hsurv^{-1}(p)$ for 6-MP patients
from the KM and Nelson-Aalen estimators.

:::

The rightmost step is drawn out to $p = 1$ for visual continuity only.
The inverse survival function is defined for $0 < p < 1$;
at the boundary, $\hsurv^{-1}(1) = \inf\cb{t \ge 0 : \hsurv(t) \le 1} = 0$,
since $\hsurv(0) = 1$.

The empirical median survival time is the smallest event time $t$
at which the estimated survival probability falls to or below 0.5.

From the table:

- $\hsurvkmf{22} \approx 0.538 > 0.5$
  and $\hsurvkmf{23} \approx 0.448 \le 0.5$,
  so $\hsurvkm^{-1}(0.5) = 23$ months.

- $\hsurvnaaf{22} \approx 0.557 > 0.5$
  and $\hsurvnaaf{23} \approx 0.471 \le 0.5$,
  so $\hsurvnaa^{-1}(0.5) = 23$ months.

In R:

```{r}
#| code-fold: show
quantile(km_fit, probs = 0.5)
```

```{r}
#| code-fold: show
quantile(na_fit, probs = 0.5)
```

Both the KM and Nelson-Aalen estimators give an estimated median
relapse-free survival time of **23 months** for the 6-MP patients.

:::
