
The `pbc` dataset from the `rmb` package
contains data from a clinical trial
of D-penicillamine treatment in primary biliary cholangitis (PBC)
[@dickson1989prognosis].

Here we demonstrate the effect of left-truncation
by comparing analyses using two different time scales:

1. Time from enrollment (standard approach, no truncation)
2. Age as the time scale (introduces left-truncation,
   since patients are observed from their current age, not from birth)

::: {#fig-pbc-km-truncation}

```{r}
#| code-fold: true
library(rmb)
library(survival)
library(ggplot2)
library(dplyr)

pbc_data <- rmb::pbc

# Analysis 1: time from enrollment (no truncation)
km_enroll <- survfit(
  Surv(years, as.numeric(status) == 1) ~ 1,
  data = pbc_data
)

# Analysis 2: age as time scale (left-truncated)
km_age <- survfit(
  Surv(time = age, time2 = age + years, event = as.numeric(status) == 1) ~ 1,
  data = pbc_data
)

# Extract KM estimates at specific times for each analysis
km1_summary <- summary(km_enroll, times = seq(0, 12, by = 0.5))
km2_summary <- summary(km_age, times = seq(26, 78, by = 0.5))

km1_df <- tibble(
  time = km1_summary$time,
  surv = km1_summary$surv,
  analysis = "Time from enrollment"
)

km2_df <- tibble(
  time = km2_summary$time - median(pbc_data$age),
  surv = km2_summary$surv,
  analysis = paste0(
    "Age as time scale\n",
    "(left-truncated; centered at median enrollment age)"
  )
)

km_combined <- bind_rows(km1_df, km2_df)

ggplot(km_combined) +
  aes(x = time, y = surv, color = analysis) +
  geom_step() +
  labs(
    x = "Time (years)",
    y = "Estimated survival probability",
    color = "Analysis",
    caption = "Data: rmb::pbc (PBC clinical trial, Dickson et al. 1989)"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Kaplan-Meier survival curves for the PBC study
using two different time scales:
time from enrollment (no truncation)
vs. age centered at the median enrollment age (left-truncated).

:::

::: {.notes}
The curve using age as the time scale
accounts for the fact that participants
were only observed from their age at enrollment.
When centered at the median enrollment age,
the two analyses can be compared on a common scale.
Note that ignoring this truncation
would lead to overestimated survival:
participants who died at younger ages ---
before reaching their enrollment age ---
were never observed and are excluded from the data,
leading to a risk set that over-represents longer survivors.
:::

## Effect of Ignoring Left-Truncation

When left-truncation is ignored:

- Survival probabilities are **overestimated**
  (especially at early time points)
- The analysis fails to account
  for undersampling of short survival times
- Hazard ratios in Cox models may be attenuated
  (effect is less predictable)

The following example illustrates the bias using the PBC data.
The naive analysis treats all participants as entering the risk set
at the earliest observed age,
whereas the proper analysis uses counting-process notation
to enter each participant at their actual enrollment age.

```{r}
#| label: pbc-truncation-bias-setup
#| code-fold: true
km_naive_age <- survfit(
  Surv(age + years, as.numeric(status) == 1) ~ 1,
  data = pbc_data
)

# Extract estimates at age 60 for comparison
surv_naive_at_sixty  <- summary(km_naive_age, times = 60)$surv
surv_proper_at_sixty <- summary(km_age, times = 60)$surv
```

At age 60,
the naive analysis (ignoring delayed entry)
estimates survival at
`r scales::percent(surv_naive_at_sixty, accuracy = 0.1)`,
compared to
`r scales::percent(surv_proper_at_sixty, accuracy = 0.1)`
with the proper left-truncation correction.

The naive estimate is inflated
because the risk set at each age incorrectly includes
participants who had not yet been enrolled
(and thus had not actually been observed to be at risk).
