
## Estimating survival in datasets without censoring

::: notes
In the `drug6mp` dataset, the estimated survival function for the placebo patients is easy to compute. 
For any time $t$ in months, $\surv(t)$ is the fraction of patients with times greater than $t$:
:::

```{r}
#| include: false
#| code-summary: example-km-curve

y <- c(1, 2, 3, 3, 3, 5, 9, 13, 23, 100)
d <- c(1, 1, 1, 0, 1, 0, 1, 1, 1, 1)
library(survival)
temp1 <- survfit(Surv(y, d) ~ 1)
```


## Estimating survival in datasets with censoring

-   For the 6-MP patients, we cannot ignore the censored data because we
know that the time to relapse is greater than the censoring time.

-   For any time $t$ in months, we know that 6-MP patients with times
greater than $t$ have not relapsed, and those with relapse time less
than $t$ have relapsed, but we don't know if patients with censored
time less than $t$ have relapsed or not.

-   The procedure we usually use is the Kaplan-Meier product-limit
estimator of the survival function.

-   The Kaplan-Meier estimator is a step function (like the empirical
cdf), which changes value only at the event times, not at the
censoring times.

-   At each event time $t_i$, we compute the at-risk set, which
includes all observations whose event time or censoring time is at
least $t_i$.
    Its size is $r_i$.

-   If $d_i$ of the observations have an event time (not a censoring time)
of $t_i$, then the group of survivors immediately following time $t_i$
is reduced by the fraction $$\frac{r_i-d_i}{r_i}=1-\frac{d_i}{r_i}$$

:::{#fig-risk-set-dropout-cartoon}

```{r}
#| code-fold: true
#| echo: false

risk_cartoon <- tibble::tibble(
  person = paste("Person", 1:8),
  exit_time = c(2, 3, 4, 5, 6, 6, 7, 8),
  status = c(
    "Event",
    "Censored",
    "Event",
    "Censored",
    "Event",
    "Censored",
    "Event",
    "Censored"
  )
) |>
  dplyr::mutate(person_id = dplyr::row_number())

risk_counts <- tibble::tibble(time = c(0, 2, 4, 6, 8)) |>
  dplyr::mutate(
    at_risk = purrr::map_int(
      time,
      \(current_time) sum(risk_cartoon$exit_time >= current_time)
    ),
    label = paste0("At risk at ", time, ": ", at_risk)
  )

ggplot2::ggplot(risk_cartoon) +
  ggplot2::aes(y = person_id) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = 0,
      xend = exit_time,
      yend = person_id
    ),
    linewidth = 1,
    color = "grey50"
  ) +
  ggplot2::geom_point(
    ggplot2::aes(
      x = exit_time,
      shape = status,
      fill = status
    ),
    size = 3,
    color = "black"
  ) +
  ggplot2::geom_vline(
    data = dplyr::filter(risk_counts, time > 0),
    ggplot2::aes(xintercept = time),
    linetype = "dashed",
    color = "grey85"
  ) +
  ggplot2::geom_text(
    data = risk_counts,
    ggplot2::aes(
      x = time,
      y = max(risk_cartoon$person_id) + 0.8,
      label = label
    ),
    size = 3.4,
    vjust = 0
  ) +
  ggplot2::scale_shape_manual(values = c(Event = 21, Censored = 22)) +
  ggplot2::scale_fill_manual(
    values = c(Event = "#F8766D", Censored = "#00BFC4")
  ) +
  ggplot2::scale_y_continuous(
    breaks = risk_cartoon$person_id,
    labels = risk_cartoon$person
  ) +
  ggplot2::coord_cartesian(ylim = c(0.5, max(risk_cartoon$person_id) + 1.6)) +
  ggplot2::labs(
    x = "Time",
    y = NULL,
    shape = NULL,
    fill = NULL
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank()
  )
```

Cartoon risk-set view showing individual follow-up lines.
Filled circles indicate events,
filled squares indicate censoring times,
and dashed lines mark selected times.
Labels above show the number at risk at each marked time.

:::

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_def-km-estimator.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_sec-km-product-limit-math.qmd >}}

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_exr-nonparametric-survival.qmd >}}

:::{#thm-KM-est-no-cens}
#### Kaplan-Meier Estimate with No Censored Observations

If there are no censored data, and there are $n$ data points, then just
after (say) the third event time 

$$
\begin{aligned}
\hsurv(t) 
&= \prod_{t_i \le t}\sb{1-\frac{d_i}{r_i}}
\\ &= \sb{\frac{n-d_1}{n}} \sb{\frac{n-d_1-d_2}{n-d_1}} \sb{\frac{n-d_1-d_2-d_3}{n-d_1-d_2}}
\\ &= \frac{n-d_1-d_2-d_3}{n}
\\ &=1-\frac{d_1+d_2+d_3}{n}
\\ &=1-\hat F(t)
\end{aligned}
$$

where $\hat F(t)$ is the usual empirical CDF estimate.

:::

{{< slidebreak >}}

## Variance of the Kaplan-Meier estimator

{{< include _subfiles/intro-to-survival-analysis/_sec-surv-KM-greenwood.qmd >}}

{{< slidebreak >}}

## Kaplan-Meier curve for `drug6mp` data

Here is the Kaplan-Meier estimated survival curve for the patients who
received 6-MP in the `drug6mp` dataset (we will see code to produce
figures like this one shortly):

```{r}
#| fig-cap: "Kaplan-Meier Survival Curve for 6-MP Patients"
#| label: fig-KM-mp6
#| echo: false

require(KMsurv)
data(drug6mp)
library(dplyr)
library(survival)

km_model1 <-
  drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, data = _)

library(ggfortify)
km_model1 |>
  autoplot(
    mark.time = TRUE,
    conf.int = FALSE
  ) +
  expand_limits(y = 0) +
  xlab("Time since diagnosis (months)") +
  ylab("KM Survival Curve")
```

## Kaplan-Meier calculations {.smaller}

{{< include _subfiles/intro-to-survival-analysis/_sec-surv-KM-calculations.qmd >}}
