:::{#def-inv-surv-fn}

#### Inverse survival function

Given a random time-to-event variable $T$ with
[survival function](#def-surv-fn) $\surv(t)$,
the **inverse survival function**
$\surv^{-1}(p)$
is the earliest time $t$ at which the survival probability
has fallen to or below $p$:

$$\surv^{-1}(p) \eqdef \inf\{t \ge 0 : \surv(t) \le p\}, \quad 0 < p < 1.$$

:::

---

:::{#exm-inv-surv-exp}
##### Exponential distribution

For an exponential distribution with rate $\haz > 0$,
the survival function is $\surv(t) = \ef{-\haz t}$.

Setting $\surv(t) = p$ and solving for $t$:

$$
\ba
\ef{-\haz t} &= p\\
-\haz t &= \logf{p}\\
t &= \frac{-\logf{p}}{\haz},
\ea
$$

so the inverse survival function of the exponential distribution is:

$$\surv^{-1}(p) = \frac{-\logf{p}}{\haz}.$$

@fig-exp-isf-rates shows the ISF for several values of $\haz$.

:::{#fig-exp-isf-rates}
```{r}
#| code-fold: true
library(ggplot2)
ggplot() +
  geom_function(
    aes(col = "0.5"),
    fun = qexp,
    args = list(lower.tail = FALSE, rate = 0.5),
    xlim = c(0.01, 0.99)
  ) +
  geom_function(
    aes(col = "1"),
    fun = qexp,
    args = list(lower.tail = FALSE, rate = 1),
    xlim = c(0.01, 0.99)
  ) +
  geom_function(
    aes(col = "1.5"),
    fun = qexp,
    args = list(lower.tail = FALSE, rate = 1.5),
    xlim = c(0.01, 0.99)
  ) +
  geom_function(
    aes(col = "5"),
    fun = qexp,
    args = list(lower.tail = FALSE, rate = 5),
    xlim = c(0.01, 0.99)
  ) +
  theme_bw() +
  ylab(expression(S^{-1}(p))) +
  xlab("p") +
  guides(col = guide_legend(title = expr(lambda))) +
  theme(
    axis.title.x = element_text(angle = 0, vjust = 1, hjust = 1),
    axis.title.y = element_text(angle = 0, vjust = 1, hjust = 1)
  )
```
Exponential inverse survival functions for various rate parameters $\haz$.
Higher $\haz$ compresses survival toward earlier times, so the ISF is
lower for any given $p$.
:::

**Numerical example.**
Suppose patients in a clinical trial have exponentially distributed
survival times with rate $\haz = 2$ events per year.
What is the median survival time?

The median survival time is the time $t$ at which
half the patients are still alive, i.e. $\surv(t) = 0.5$:

$$
\surv^{-1}(0.5)
= \frac{-\logf{0.5}}{2}
= \frac{\logf{2}}{2}
\approx \frac{0.693}{2}
\approx 0.347 \text{ years.}
$$

In R:

```{r}
#| code-fold: show
qexp(p = 0.5, rate = 2)
```

@fig-exp-isf-median shows this point in both the survival function and
the inverse survival function.

:::{#fig-exp-isf-median layout-ncol=2}

```{r}
#| label: fig-exp-survfn-with-median
#| fig-cap: "Survival function $S(t) = e^{-2t}$."
#| code-fold: true
library(ggplot2)
lambda_val <- 2
t_med <- log(2) / lambda_val

ggplot() +
  geom_function(
    fun = pexp,
    args = list(lower.tail = FALSE, rate = lambda_val),
    xlim = c(0, 1.5)
  ) +
  geom_segment(
    aes(x = 0, xend = t_med, y = 0.5, yend = 0.5),
    linetype = "dashed",
    color = "red"
  ) +
  geom_segment(
    aes(x = t_med, xend = t_med, y = 0, yend = 0.5),
    linetype = "dashed",
    color = "red"
  ) +
  geom_point(aes(x = t_med, y = 0.5), size = 3, color = "red") +
  theme_bw() +
  xlab("t (years)") +
  ylab("S(t)") +
  theme(
    axis.title.x = element_text(angle = 0, vjust = 1, hjust = 1),
    axis.title.y = element_text(angle = 0, vjust = 1, hjust = 1)
  )
```

```{r}
#| label: fig-exp-isf-with-median
#| fig-cap: "Inverse survival function $S^{-1}(p) = -\\log(p)/2$."
#| code-fold: true
library(ggplot2)
lambda_val <- 2
t_med <- log(2) / lambda_val

ggplot() +
  geom_function(
    fun = qexp,
    args = list(lower.tail = FALSE, rate = lambda_val),
    xlim = c(0.01, 0.99)
  ) +
  geom_segment(
    aes(x = 0, xend = 0.5, y = t_med, yend = t_med),
    linetype = "dashed",
    color = "red"
  ) +
  geom_segment(
    aes(x = 0.5, xend = 0.5, y = 0, yend = t_med),
    linetype = "dashed",
    color = "red"
  ) +
  geom_point(aes(x = 0.5, y = t_med), size = 3, color = "red") +
  theme_bw() +
  xlab("p") +
  ylab(expression(S^{-1}(p))) +
  theme(
    axis.title.x = element_text(angle = 0, vjust = 1, hjust = 1),
    axis.title.y = element_text(angle = 0, vjust = 1, hjust = 1)
  )
```

The median survival time ($p = 0.5$, $t \approx 0.347$ years) shown in
the survival function (left) and inverse survival function (right).
The dashed red lines mark the coordinates of this point.

:::

:::

---

:::{#thm-inv-surv-is-quantile}

The inverse survival function equals the $(1-p)$th
[population quantile](probability.qmd#def-quantile-function)
of $T$:

$$\surv^{-1}(p) = Q(1 - p),$$

where $Q(u) = \inf\{t : \cdf(t) \ge u\}$ is the quantile function of $T$.

:::

::: proof

Since $\surv(t) = 1 - \cdf(t)$,

$$
\ba
\surv(t) \le p
&\iff 1 - \cdf(t) \le p\\
&\iff \cdf(t) \ge 1 - p.
\ea
$$

Therefore,

$$
\surv^{-1}(p)
= \inf\cb{t \ge 0 : \surv(t) \le p}
= \inf\cb{t \ge 0 : \cdf(t) \ge 1-p}
= Q(1-p).
$$

:::
