
### Overview of study {.smaller}

This is from a clinical trial in 1963 for 6-MP treatment vs. placebo for
Acute Leukemia in 42 children.

-   Pairs of children:

-   matched by remission status at the time of treatment (`remstat`:
`1` = partial, `2` = complete)
-   randomized to 6-MP (exit times in `t2`) or placebo (exit times
in `t1`)

-   Followed until relapse or end of study.

-   All of the placebo group relapsed, but some of the 6-MP group were
censored (which means they were still in remission); indicated by
`relapse` variable (`0` = censored, `1` = relapse).

-   6-MP = 6-Mercaptopurine (Purinethol) is an anti-cancer
("antineoplastic" or "cytotoxic") chemotherapy drug used currently
for Acute lymphoblastic leukemia (ALL). It is classified as an
antimetabolite.

### Study design {.smaller}

- Clinical trial in 1963 for 6-MP treatment vs. placebo for Acute Leukemia
in 42 children.
- Pairs of children:
- matched by remission status at the time of treatment (`remstat`)
- `remstat` = 1: partial
- `remstat` = 2: complete
- randomized to 6-MP (exit time: `t2`) or placebo (`t1`).
- Followed until relapse or end of study.
- All of the placebo group relapsed,
- Some of the 6-MP group were censored.

---

```{r}
#| tbl-cap: "`drug6mp` pediatric acute leukemia data"
#| label: tbl-drug6mp
library(KMsurv)
data(drug6mp)
drug6mp <- drug6mp |>
  tibble::as_tibble() |>
  print()
```

### Data documentation for `drug6mp`

```{r, printr.help.sections = c("description", "format")}
#| fig-cap: Data documentation for `drug6mp`
#| label: fig-drug6mp-helpdoc
# library(printr) # inserts help-file output into markdown output
library(KMsurv)
?drug6mp
```

### Descriptive Statistics {.smaller}

```{r}
#| tbl-cap: "Summary statistics for `drug6mp` data"
#| label: tbl-drug6mp-summary
#| code-fold: show
summary(drug6mp)
```

::: notes

-   The average time in each group is not useful. Some of the 6-MP
patients have not relapsed at the time recorded, while all of the
placebo patients have relapsed.
-   The median time is not really useful either because so many of the
6-MP patients have not relapsed (12/21).
-   Both are biased down in the 6-MP group. Remember that lower times
are worse since they indicate sooner recurrence.
:::

### Exponential model

::: notes
-   We *can* compute the hazard rate, assuming an exponential model:
number of relapses divided by the sum of the exit times (@eq-mle-exp).
:::

$$\hat\lambda = \frac{\sumin D_i}{\sumin Y_i}$$

::: notes
-   For the placebo, that is just the reciprocal of the mean time:
:::
$$
\ba
\hat \lambda_{\text{placebo}}
&= \frac{\sumin D_i}{\sumin Y_i}
\\ &= \frac{\sumin 1}{\sumin Y_i}
\\ &= \frac{n}{\sumin Y_i}
\\ &= \frac{1}{\bar{Y}}
\\ &= \frac{1}{`r drug6mp |> pull(t1) |> mean()`}
\\ &= `r 1/(drug6mp |> pull(t1) |> mean())`
\ea
$$

---

-   For the 6-MP group, $\hat\lambda = 9/359 = 0.025$

$$
\ba
\hat \lambda_{\text{6-MP}}
&= \frac{\sumin D_i}{\sumin Y_i}
\\ &= \frac{9}{359}
\\ &= `r 9/359`
\ea
$$

-   The estimated hazard in the placebo group is 4.6 times as
large as in the 6-MP group (assuming the hazard is constant over time).
