:::{#def-inv-surv-fn}

#### Inverse survival function

Given a random time-to-event variable $T$ with
survival function $\surv(t)$ (@def-surv-fn),
the **inverse survival function** (ISF),
also called the **survival quantile function** (SQF),
$\surv^{-1}(p)$
is the earliest time $t$ at which the survival probability
has fallen to or below $p$:

$$\surv^{-1}(p) \eqdef \inf\{t \ge 0 : \surv(t) \le p\}, \quad 0 < p < 1.$$

:::

---

:::{#exm-inv-surv-exp}
##### Exponential distribution

For an exponential distribution with rate $\haz > 0$,
the survival function is $\surv(t) = \ef{-\haz t}$.

Setting $\surv(t) = p$ and solving for $t$:

$$
\ba
\ef{-\haz t} &= p\\
-\haz t &= \logf{p}\\
t &= \frac{-\logf{p}}{\haz},
\ea
$$

so the inverse survival function of the exponential distribution is:

$$\surv^{-1}(p) = \frac{-\logf{p}}{\haz}.$$

:::{#fig-isf-exp-rates}

```{r}
#| code-fold: true
library(ggplot2)

p_seq <- seq(0.001, 0.999, length.out = 500)
rates <- c(0.5, 1, 2, 4)

df_isf <- do.call(rbind, lapply(rates, function(lam) {
  data.frame(
    p = p_seq,
    isf = -log(p_seq) / lam,
    lambda = factor(
      paste0("lambda = ", lam),
      levels = paste0("lambda = ", rates)
    )
  )
}))

ggplot(df_isf, aes(x = p, y = isf, color = lambda)) +
  geom_line() +
  labs(
    x = expression(p),
    y = expression(S^{-1}*(p)),
    color = NULL
  ) +
  theme_bw()
```

Exponential inverse survival function $\surv^{-1}(p) = -\log(p)/\haz$
for various rate parameters $\haz$.

:::

**Numerical example.**
Suppose patients in a clinical trial have exponentially distributed
survival times with rate $\haz = 2$ events per year.
What is the median survival time?

The median survival time is the time $t$ at which
half the patients are still alive, i.e. $\surv(t) = 0.5$:

$$
\surv^{-1}(0.5)
= \frac{-\logf{0.5}}{2}
= \frac{\logf{2}}{2}
\approx \frac{0.693}{2}
\approx 0.347 \text{ years.}
$$

@fig-isf-surv-exp-median and @fig-isf-isf-exp-median show the survival
function $\surv(t) = \ef{-2t}$ and the inverse survival function
$\surv^{-1}(p) = -\log(p)/2$, each with the median survival point
highlighted.

::::{#fig-isf-exp-median layout-ncol=2}

:::{#fig-isf-surv-exp-median}
```{r}
#| code-fold: true
library(ggplot2)

lam <- 2
t_med <- log(2) / lam
t_seq <- seq(0, 2, length.out = 500)
p_seq <- seq(0.001, 0.999, length.out = 500)

ggplot(
  data.frame(t = t_seq, s = exp(-lam * t_seq)),
  aes(x = t, y = s)
) +
  geom_line() +
  geom_segment(
    aes(x = 0, xend = t_med, y = 0.5, yend = 0.5),
    linetype = "dashed", color = "red"
  ) +
  geom_segment(
    aes(x = t_med, xend = t_med, y = 0, yend = 0.5),
    linetype = "dashed", color = "red"
  ) +
  geom_point(aes(x = t_med, y = 0.5), color = "red", size = 3) +
  labs(x = "t (years)", y = "S(t)") +
  theme_bw()
```

Survival function $\surv(t)$.
:::

:::{#fig-isf-isf-exp-median}
```{r}
#| code-fold: true
ggplot(
  data.frame(p = p_seq, isf = -log(p_seq) / lam),
  aes(x = p, y = isf)
) +
  geom_line() +
  geom_segment(
    aes(x = 0.5, xend = 0.5, y = 0, yend = t_med),
    linetype = "dashed", color = "red"
  ) +
  geom_segment(
    aes(x = 0, xend = 0.5, y = t_med, yend = t_med),
    linetype = "dashed", color = "red"
  ) +
  geom_point(aes(x = 0.5, y = t_med), color = "red", size = 3) +
  labs(x = "p", y = expression(S^{-1}*(p))) +
  theme_bw()
```

Inverse survival function $\surv^{-1}(p)$.
:::

For $\haz = 2$, the survival function $\surv(t)$ (left) and
inverse survival function $\surv^{-1}(p)$ (right),
with the median survival time $\surv^{-1}(0.5) \approx 0.347$ years
shown as a red point.
::::

In R:

```{r}
#| code-fold: show
qexp(p = 0.5, rate = 2)
```

:::

---

:::{#thm-inv-surv-is-quantile}

The inverse survival function equals the $(1-p)$th
[population quantile](probability.qmd#def-quantile-function)
of $T$:

$$\surv^{-1}(p) = Q(1 - p),$$

where $Q(u) = \inf\{t : \cdf(t) \ge u\}$ is the quantile function of $T$.

:::

::: proof

Since $\surv(t) = 1 - \cdf(t)$,

$$
\ba
\surv(t) \le p
&\iff 1 - \cdf(t) \le p\\
&\iff \cdf(t) \ge 1 - p.
\ea
$$

Therefore,

$$
\surv^{-1}(p)
= \inf\cb{t \ge 0 : \surv(t) \le p}
= \inf\cb{t \ge 0 : \cdf(t) \ge 1-p}
= Q(1-p).
$$

:::

---

