
Let's compute these estimates and build the chart by hand:

```{r}
#| code-fold: true
library(KMsurv)
library(dplyr)
data(drug6mp)

data_v2 <-
  drug6mp |>
  as_tibble() |>
  mutate(
    remstat = remstat |>
      case_match(
        1 ~ "partial",
        2 ~ "complete"
      ),
    # renaming to "outcome" while relabeling is just a style choice:
    outcome = relapse |>
      case_match(
        0 ~ "censored",
        1 ~ "relapsed"
      )
  )

km_table <-
  data_v2 |>
  summarize(
    .by = t2,
    Relapses = sum(outcome == "relapsed"),
    Censored = sum(outcome == "censored")
  ) |>
  # here we add a start time row, so the graph starts at time 0:
  bind_rows(
    tibble(
      t2 = 0,
      Relapses = 0,
      Censored = 0
    )
  ) |>
  # sort in time order:
  arrange(t2) |>
  mutate(
    Exiting = Relapses + Censored,
    `Study Size` = sum(Exiting),
    Exited = cumsum(Exiting) |> dplyr::lag(default = 0),
    `At Risk` = `Study Size` - Exited,
    Hazard = Relapses / `At Risk`,
    `KM Factor` = 1 - Hazard,
    `Cumulative Hazard` = cumsum(`Hazard`),
    `KM Survival Curve` = cumprod(`KM Factor`)
  )

library(pander)
pander(km_table)
```

{{< slidebreak >}}

### Summary

For the 6-MP patients at time 6 months, there are 21 patients at risk.
At $t=6$ there are 3 relapses and 1 censored observations.

The Kaplan-Meier factor is $(21-3)/21 = 0.857$. The number at risk for
the next time ($t=7$) is $21-3-1=17$.

At time 7 months, there are 17 patients at risk. At $t=7$ there is 1
relapse and 0 censored observations. The Kaplan-Meier factor is
$(17-1)/17 = 0.941$. The Kaplan Meier estimate is
$0.857\times0.941=0.807$. The number at risk for the next time ($t=9$)
is $17-1=16$.

{{< slidebreak >}}

Now, let's graph this estimated survival curve using `ggplot()`:

```{r}
#| label: fig-km-by-hand
#| fig-cap: "KM curve for 6MP patients, calculated by hand"
library(ggplot2)
conflicts_prefer(dplyr::filter)
km_table |>
  ggplot(aes(x = t2, y = `KM Survival Curve`)) +
  geom_step() +
  geom_point(data = km_table |> filter(Censored > 0), shape = 3) +
  expand_limits(y = c(0, 1), x = 0) +
  xlab("Time since diagnosis (months)") +
  ylab("KM Survival Curve") +
  scale_y_continuous(labels = scales::percent)
```

{{< slidebreak >}}

{{< include _subfiles/intro-to-survival-analysis/_exr_survival_prediction_intervals.qmd >}}
