:::{#def-emp-inv-surv-fn}

#### Empirical inverse survival function

Given a time-to-event dataset,
the **empirical inverse survival function**
$\hsurv^{-1}(p)$
replaces the population survival function $\surv(t)$
with an estimated survival function $\hsurv(t)$:

$$\hsurv^{-1}(p) \eqdef \inf\cb{t \ge 0 : \hsurv(t) \le p}, \quad 0 < p < 1.$$

:::

---

Since $\hsurv(t)$ is a non-increasing step function,
$\hsurv^{-1}(p)$ equals the smallest observed event time
at which the estimated survival probability has dropped to or below $p$.

**From the Kaplan-Meier estimator.**
Substituting $\hsurv_{KM}(t)$ from @def-KM-estimator:

$$
\hsurv_{KM}^{-1}(p)
= \inf\cb{t \ge 0 : \prod_{t_i \le t}\sb{1 - \frac{d_i}{r_i}} \le p}.
$$

**From the Nelson-Aalen estimator.**
Substituting $\hsurv_{NA}(t) = \expf{-\hat{\cuhaz}_{NA}(t)}$
from @def-na-cuhaz-est:

$$
\hsurv_{NA}^{-1}(p)
= \inf\cb{t \ge 0 : \expf{-\sum_{t_i \le t}\frac{d_i}{r_i}} \le p}.
$$

In R, both estimators are computed from a `survfit()` object using `quantile()`.

---

:::{#exm-emp-isf-drug6mp}
##### Empirical ISF for 6-MP patients

Using the `drug6mp` dataset,
we compute the empirical ISF for the 6-MP patients.
The KM and Nelson-Aalen survival estimates at each event time are:

```{r}
#| code-fold: true
library(KMsurv)
library(dplyr)
library(survival)
data(drug6mp)

km_fit <- drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, data = _)

na_fit <- drug6mp |>
  mutate(surv = Surv(t2, relapse)) |>
  survfit(formula = surv ~ 1, type = "fleming-harrington", data = _)

tbl <- tibble(
  time = km_fit$time,
  n.risk = km_fit$n.risk,
  n.event = km_fit$n.event,
  surv_km = km_fit$surv,
  surv_na = na_fit$surv
) |>
  filter(n.event > 0)

library(knitr)
kable(
  tbl,
  digits = 3,
  col.names = c(
    "Time (months)",
    "At risk $r_i$",
    "Events $d_i$",
    "$\\hat{S}_{KM}(t)$",
    "$\\hat{S}_{NA}(t)$"
  )
)
```

The empirical median survival time is the smallest event time $t$
at which the estimated survival probability falls to or below 0.5.

From the table:

- $\hsurv_{KM}(22) \approx 0.538 > 0.5$
  and $\hsurv_{KM}(23) \approx 0.448 \le 0.5$,
  so $\hsurv_{KM}^{-1}(0.5) = 23$ months.

- $\hsurv_{NA}(22) \approx 0.557 > 0.5$
  and $\hsurv_{NA}(23) \approx 0.471 \le 0.5$,
  so $\hsurv_{NA}^{-1}(0.5) = 23$ months.

In R:

```{r}
#| code-fold: show
quantile(km_fit, probs = 0.5)
```

```{r}
#| code-fold: show
quantile(na_fit, probs = 0.5)
```

Both the KM and Nelson-Aalen estimators give an estimated median
relapse-free survival time of **23 months** for the 6-MP patients.

:::
