Integration is the inverse operation of differentiation: it recovers a
function from its derivative and accumulates quantities such as areas,
totals, and probabilities. We begin with antiderivatives, then state
basic integration rules, and conclude with the Fundamental Theorem of
Calculus and a worked example from probability.

:::{#def-antiderivative}
#### Antiderivative

A function $F$ is an **antiderivative** of $f$ on an interval $I$ if:

$$\deriv{x} F(x) = f(x), \quad \forall x \in I$$

The family of all antiderivatives of $f$ is written as the **indefinite integral**:

$$\int f(x)\,dx = F(x) + C$$

where $C$ is an arbitrary constant of integration.

[@larsonCalc11e, Section 4.1, pp. 248--249]

:::

{{< slidebreak >}}

::::{#exm-antiderivative-power}
#### Antiderivative of a power function

For $f(x) = x^2$, an antiderivative is $F(x) = \frac{x^3}{3}$,
since $\deriv{x}\frac{x^3}{3} = x^2 = f(x)$.

Adding any constant $C$ gives another antiderivative;
for example, with $C = 7$, $F(x) = \frac{x^3}{3} + 7$ also satisfies
$F'(x) = x^2$, since adding a constant does not change the derivative.
See @fig-antiderivatives.

{{< slidebreak >}}

:::{#fig-antiderivatives}

```{r}
#| code-fold: true
#| layout-ncol: 2
#| fig-subcap:
#|   - "The function $f(x) = x^2$"
#|   - "Family of antiderivatives $F(x) = x^3/3 + C$"
library(ggplot2)

ggplot() +
  geom_function(fun = \(x) x^2, xlim = c(-2, 2), linewidth = 1) +
  labs(x = "x", y = expression(f(x))) +
  theme_minimal()

C_vals <- c(-2, -1, 0, 1, 2)
x_seq <- seq(-2, 2, length.out = 200)
df <- do.call(rbind, lapply(C_vals, \(C) {
  data.frame(
    x = x_seq,
    y = x_seq^3 / 3 + C,
    C = factor(C)
  )
}))

ggplot(df, aes(x = x, y = y, color = C)) +
  geom_line(linewidth = 0.8) +
  labs(x = "x", y = expression(F(x)), color = "C") +
  theme_minimal()
```

The function $f(x) = x^2$ and five antiderivatives $F(x) = x^3/3 + C$ for $C \in \{-2,-1,0,1,2\}$.
Each antiderivative has the same derivative $f$; they differ only by a vertical shift.

:::

::::

{{< slidebreak >}}

:::{#thm-integral-rules}
#### Basic integration rules

| Function $f(x)$ | Antiderivative $F(x)$ | Condition |
|:---:|:---:|:---:|
| $c$ | $cx$ | — |
| $x^n$ | $\dfrac{x^{n+1}}{n+1}$ | $n \ne -1$ |
| $\dfrac{1}{x}$ | $\ln\abs{x}$ | $x \ne 0$ |
| $\ef{x}$ | $\ef{x}$ | — |
| $\ef{cx}$ | $\dfrac{1}{c}\ef{cx}$ | $c \ne 0$ |
| $c \cdot f(x)$ | $c \cdot F(x)$ | — |
| $f(x) + g(x)$ | $F(x) + G(x)$ | — |

Each antiderivative in this table is defined only up to an arbitrary
constant $C$ (see @def-antiderivative);
the table omits $+ C$ from every row for brevity.

The first two rows and the bottom two rows (linearity) are from
[@larsonCalc11e, Section 4.1, p. 250 "Basic Integration Rules"];
$1/x$ is from [@larsonCalc11e, Section 5.2, Theorem 5.5, p. 324];
$\ef{x}$ and $\ef{cx}$ are from
[@larsonCalc11e, Section 5.4, Theorem 5.12, p. 346].

:::

{{< slidebreak >}}

:::{#exm-integral-rules-quadratic}
#### Antiderivative of $3x^2 - 1$

By the power rule ($n = 2$) and linearity from @thm-integral-rules:

$$
\int \paren{3x^2 - 1}\,dx
= 3 \cd \frac{x^3}{3} - x + C
= x^3 - x + C.
$$

Verify by differentiating:
$\deriv{x}\paren{x^3 - x + C} = 3x^2 - 1 = f(x)$, as required.

:::

{{< slidebreak >}}

Before stating the Fundamental Theorem of Calculus, we record two
prerequisite results.
The FTC requires the integrand $f$ to be continuous,
and the two theorems below establish where continuity comes from
(differentiability ⇒ continuity)
and what it buys us (continuity ⇒ integrability).

:::{#thm-diff-implies-cont}
#### Differentiability implies continuity

If $f$ is differentiable at $x = c$, then $f$ is continuous at $x = c$.

[@larsonCalc11e, Theorem 2.1, p. 106]

:::

:::{#exm-diff-implies-cont}
#### Differentiable, hence continuous: $x^3 - x$

$f(x) = x^3 - x$ is differentiable everywhere
(with derivative $f'(x) = 3x^2 - 1$),
so by @thm-diff-implies-cont it is continuous everywhere.

:::

{{< slidebreak >}}

:::{#exm-cont-not-diff}
#### Continuous but not differentiable: $\abs{x}$

The absolute-value function $f(x) = \abs{x}$ is continuous at $x = 0$
($\lim_{x \to 0}\abs{x} = 0 = \abs{0}$),
but it is not differentiable at $x = 0$:
the left-derivative is $-1$ and the right-derivative is $+1$.

This shows that the converse of @thm-diff-implies-cont fails:
continuity does not imply differentiability.

:::

{{< slidebreak >}}

:::{#thm-cont-implies-int}
#### Continuity implies integrability

If $f$ is continuous on the closed interval $[a, b]$,
then $f$ is integrable on $[a, b]$;
that is, $\int_a^b f(x)\,dx$ exists.

[@larsonCalc11e, Theorem 4.4, p. 272]

:::

:::{#exm-cont-implies-int}
#### Continuous, hence integrable: polynomials

Every polynomial is continuous on $\mathbb{R}$,
so by @thm-cont-implies-int every polynomial is integrable on every
closed interval $[a, b]$.

:::

{{< slidebreak >}}

:::{#exm-int-not-cont}
#### Integrable but not continuous: a step function

Let $f(x) = 0$ for $x < \tfrac{1}{2}$ and $f(x) = 1$ for $x \ge \tfrac{1}{2}$.
Then $f$ is discontinuous at $x = \tfrac{1}{2}$,
but it is integrable on $[0, 1]$:

$$
\int_0^1 f(x)\,dx
= \int_0^{1/2} 0\,dx + \int_{1/2}^1 1\,dx
= 0 + \tfrac{1}{2}
= \tfrac{1}{2}.
$$

This shows that the converse of @thm-cont-implies-int fails:
integrability does not imply continuity.

:::

{{< slidebreak >}}

Together, @thm-diff-implies-cont and @thm-cont-implies-int establish the chain:

$$\text{differentiable} \;\Rightarrow\; \text{continuous} \;\Rightarrow\; \text{integrable}$$

@exm-cont-not-diff and @exm-int-not-cont show that neither implication
reverses in general.

{{< slidebreak >}}

:::{#thm-ftc}
#### Fundamental Theorem of Calculus

Let $f$ be a continuous function on a closed interval $[a, b]$.

**Part 1 (Derivative of an integral).**
Define $F(x) = \int_a^x f(t)\,dt$ for $x \in [a, b]$.
Then $F$ is differentiable and:

$$\deriv{x}\int_a^x f(t)\,dt = f(x)$$ {#eq-ftc-deriv-of-integral}

[@larsonCalc11e, Theorem 4.11, p. 288]

**Part 2 (Evaluation theorem).**
If $F$ is an antiderivative of $f$ on $[a, b]$
(i.e., $\deriv{x} F(x) = f(x)$ for all $x \in [a, b]$), then:

$$\int_a^b f(x)\,dx = F(b) - F(a)$$ {#eq-ftc-part2}

Equivalently, with $b$ replaced by a variable upper limit $x$,
integrating the derivative of $F$ recovers the net change in $F$:

$$\int_a^x F'(t)\,dt = F(x) - F(a)$$ {#eq-ftc-integral-of-deriv}

[@calclifesaver, Chapter 18; @larsonCalc11e, Theorem 4.9, p. 282]

:::

{{< slidebreak >}}

The two parts of the FTC together express that **differentiation and
integration are inverse operations**:

- Part 1: differentiating the integral of $f$ recovers $f$
  (@eq-ftc-deriv-of-integral).
- Part 2: integrating the derivative of $F$ recovers the net change in
  $F$ (@eq-ftc-integral-of-deriv).

The FTC requires $f$ to be continuous, so by the chain established by
@thm-diff-implies-cont and @thm-cont-implies-int, the FTC applies
whenever $f$ is differentiable.

{{< slidebreak >}}

::::{#exm-ftc-cdf-pdf}
#### CDF and PDF of the exponential distribution

For the exponential distribution with rate parameter $\haz > 0$,
the probability density function (PDF) is
[@kleinbaum2012survival, Section II, p. 295, "Survival and Hazard
Functions for Selected Distributions"]:

$$f(t) = \haz \ef{-\haz t}, \quad t \ge 0$$

**FTC Part 2** gives the cumulative distribution function (CDF) from the PDF.
Apply the $\ef{cx}$ rule from @thm-integral-rules with $c = -\haz$ to
antidifferentiate the integrand:

$$
\ba
F(t)
&= \int_0^t \haz \ef{-\haz u}\,du \\
&= \haz \cd \frac{1}{-\haz}\ef{-\haz u} \bigg|_{u=0}^{u=t} \\
&= \left(-\ef{-\haz u}\right)\bigg|_{u=0}^{u=t} \\
&= -\ef{-\haz t} - \left(-\ef{0}\right) \\
&= -\ef{-\haz t} - (-1) \\
&= 1 - \ef{-\haz t}
\ea
$$

**FTC Part 1** recovers the PDF from the CDF:

$$
\ba
\deriv{t} F(t)
&= \deriv{t}\left(1 - \ef{-\haz t}\right) \\
&= 0 - (-\haz)\ef{-\haz t} \\
&= \haz\ef{-\haz t} \\
&= f(t)
\ea
$$

For a concrete instance: with $\haz = 1$ (standard exponential),
the probability that $T \le 2$ is:

$$
F(2) = 1 - \ef{-1 \cd 2} = 1 - \ef{-2} \approx 1 - 0.135 = 0.865
$$

See @fig-exp-pdf-cdf.

{{< slidebreak >}}

:::{#fig-exp-pdf-cdf}

```{r}
#| code-fold: true
#| layout-ncol: 2
#| fig-subcap:
#|   - "PDF $f(t) = e^{-t}$; shaded area = $F(2) \\approx 0.865$"
#|   - "CDF $F(t) = 1 - e^{-t}$; point marks $F(2)$"
library(ggplot2)
lam <- 1
t_max <- 5

ggplot() +
  geom_area(
    data = data.frame(t = seq(0, 2, length.out = 300)),
    aes(x = t, y = lam * exp(-lam * t)),
    fill = "steelblue", alpha = 0.4
  ) +
  geom_function(
    fun = \(t) lam * exp(-lam * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  labs(x = "t", y = "f(t)") +
  theme_minimal()

ggplot() +
  geom_function(
    fun = \(t) 1 - exp(-lam * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  geom_point(aes(x = 2, y = 1 - exp(-lam * 2)), size = 3, color = "steelblue") +
  geom_segment(
    aes(x = 2, xend = 2, y = 0, yend = 1 - exp(-lam * 2)),
    linetype = "dashed", color = "steelblue"
  ) +
  geom_segment(
    aes(x = 0, xend = 2, y = 1 - exp(-lam * 2), yend = 1 - exp(-lam * 2)),
    linetype = "dashed", color = "steelblue"
  ) +
  labs(x = "t", y = "F(t)") +
  theme_minimal()
```

Exponential distribution with $\haz = 1$.
Left: the PDF $f(t) = \ef{-t}$; the shaded area under the curve from $0$ to $2$ equals $F(2) \approx 0.865$.
Right: the CDF $F(t) = 1 - \ef{-t}$; the dashed lines mark the value $F(2)$ computed via FTC Part 2.

:::

::::
