:::{#def-antiderivative}
#### Antiderivative

A function $F$ is an **antiderivative** of $f$ on an interval $I$ if:

$$\deriv{x} F(x) = f(x), \quad \forall x \in I$$

The family of all antiderivatives of $f$ is written as the **indefinite integral**:

$$\int f(x)\,dx = F(x) + C$$

where $C$ is an arbitrary constant of integration.

:::

{{< slidebreak >}}

:::{#exm-antiderivative-power}
#### Antiderivative of a power function

For $f(x) = x^2$, an antiderivative is $F(x) = \frac{x^3}{3}$,
since $\deriv{x}\frac{x^3}{3} = x^2 = f(x)$.

Adding any constant $C$ gives another antiderivative: $F(x) = \frac{x^3}{3} + 7$,
since adding a constant does not change the derivative.
See @fig-antiderivatives.

:::

{{< slidebreak >}}

:::{#fig-antiderivatives}

```{r}
#| code-fold: true
#| layout-ncol: 2
#| fig-subcap:
#|   - "The derivative $f(x) = x^2$"
#|   - "Family of antiderivatives $F(x) = x^3/3 + C$"
library(ggplot2)

ggplot() +
  geom_function(fun = \(x) x^2, xlim = c(-2, 2), linewidth = 1) +
  labs(x = "x", y = expression(f(x))) +
  theme_minimal()

C_vals <- c(-2, -1, 0, 1, 2)
x_seq <- seq(-2, 2, length.out = 200)
df <- do.call(rbind, lapply(C_vals, \(C) data.frame(
  x = x_seq,
  y = x_seq^3 / 3 + C,
  C = factor(C)
)))

ggplot(df, aes(x = x, y = y, color = C)) +
  geom_line(linewidth = 0.8) +
  labs(x = "x", y = expression(F(x)), color = "C") +
  theme_minimal()
```

The function $f(x) = x^2$ and five antiderivatives $F(x) = x^3/3 + C$ for $C \in \{-2,-1,0,1,2\}$.
Each antiderivative has the same derivative $f$; they differ only by a vertical shift.

:::

{{< slidebreak >}}

:::{#thm-integral-rules}
#### Basic integration rules

| Function $f(x)$ | Antiderivative $F(x)$ | Condition |
|---|---|---|
| $c$ | $cx$ | — |
| $x^n$ | $\dfrac{x^{n+1}}{n+1}$ | $n \ne -1$ |
| $\dfrac{1}{x}$ | $\ln\abs{x}$ | $x \ne 0$ |
| $\ef{x}$ | $\ef{x}$ | — |
| $\ef{cx}$ | $\dfrac{1}{c}\ef{cx}$ | $c \ne 0$ |
| $c \cdot f(x)$ | $c \cdot F(x)$ | — |
| $f(x) + g(x)$ | $F(x) + G(x)$ | — |

:::

{{< slidebreak >}}

:::{#thm-ftc}
#### Fundamental Theorem of Calculus

Let $f$ be a continuous function on a closed interval $[a, b]$.

**Part 1 (Derivative of an integral).**
Define $F(x) = \int_a^x f(t)\,dt$ for $x \in [a, b]$.
Then $F$ is differentiable and:

$$\deriv{x}\int_a^x f(t)\,dt = f(x)$$

**Part 2 (Evaluation theorem).**
If $F$ is an antiderivative of $f$ on $[a, b]$
(i.e., $\deriv{x} F(x) = f(x)$ for all $x \in [a, b]$), then:

$$\int_a^b f(x)\,dx = F(b) - F(a)$$

[@calclifesaver, Chapter 18]

:::

{{< slidebreak >}}

:::{.notes}

The two parts of the FTC together express that **differentiation and integration are inverse operations**:

- Part 1: differentiating the integral of $f$ recovers $f$.
- Part 2: integrating the derivative of $F$ recovers the net change in $F$.

:::

{{< slidebreak >}}

:::{#exm-ftc-cdf-pdf}
#### CDF and PDF of the exponential distribution

For the exponential distribution with rate parameter $\haz > 0$,
the probability density function (PDF) is:

$$f(t) = \haz \ef{-\haz t}, \quad t \ge 0$$

**FTC Part 2** gives the cumulative distribution function (CDF) from the PDF:

$$
\ba
F(t)
&= \int_{u=0}^t \haz \ef{-\haz u}\,du \\
&= \haz \cd \frac{1}{-\haz}\ef{-\haz u} \bigg|_{u=0}^{u=t} \\
&= \left(-\ef{-\haz u}\right)\bigg|_{u=0}^{u=t} \\
&= -\ef{-\haz t} - \left(-\ef{0}\right) \\
&= 1 - \ef{-\haz t}
\ea
$$

**FTC Part 1** recovers the PDF from the CDF:

$$
\ba
\deriv{t} F(t)
&= \deriv{t}\left(1 - \ef{-\haz t}\right) \\
&= 0 - (-\haz)\ef{-\haz t} \\
&= \haz\ef{-\haz t} \\
&= f(t)
\ea
$$

For a concrete instance: with $\haz = 1$ (standard exponential),
the probability that $T \le 2$ is:

$$
F(2) = 1 - \ef{-1 \cd 2} = 1 - \ef{-2} \approx 1 - 0.1353 = 0.8647
$$

See @fig-exp-pdf-cdf.

:::

{{< slidebreak >}}

:::{#fig-exp-pdf-cdf}

```{r}
#| code-fold: true
#| layout-ncol: 2
#| fig-subcap:
#|   - "PDF $f(t) = e^{-t}$; shaded area = $F(2) \\approx 0.865$"
#|   - "CDF $F(t) = 1 - e^{-t}$; point marks $F(2)$"
library(ggplot2)
lam <- 1
t_max <- 5

ggplot() +
  geom_area(
    data = data.frame(t = seq(0, 2, length.out = 300)),
    aes(x = t, y = lam * exp(-lam * t)),
    fill = "steelblue", alpha = 0.4
  ) +
  geom_function(
    fun = \(t) lam * exp(-lam * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  labs(x = "t", y = "f(t)") +
  theme_minimal()

ggplot() +
  geom_function(
    fun = \(t) 1 - exp(-lam * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  geom_point(aes(x = 2, y = 1 - exp(-2)), size = 3, color = "steelblue") +
  geom_segment(
    aes(x = 2, xend = 2, y = 0, yend = 1 - exp(-2)),
    linetype = "dashed", color = "steelblue"
  ) +
  geom_segment(
    aes(x = 0, xend = 2, y = 1 - exp(-2), yend = 1 - exp(-2)),
    linetype = "dashed", color = "steelblue"
  ) +
  labs(x = "t", y = "F(t)") +
  theme_minimal()
```

Exponential distribution with $\haz = 1$.
Left: the PDF $f(t) = \ef{-t}$; the shaded area under the curve from $0$ to $2$ equals $F(2) \approx 0.865$.
Right: the CDF $F(t) = 1 - \ef{-t}$; the dashed lines mark the value $F(2)$ computed via FTC Part 2.

:::

{{< slidebreak >}}
