---
title: "Concept Map: Definitions and Results"
---

{{< include shared-config.qmd >}}

This appendix shows how the definitions and results
(`def`, `thm`, `lem`, `cor`, `prp` callouts)
in the notes depend on one another.

We say result $B$ is a **descendant** of result $A$
if $A$ is referenced inside the statement or proof of $B$
(directly, or transitively through a chain of intermediate results).
The more descendants a result has,
the more of the rest of the notes rests on it ---
so descendant count is a rough measure of how foundational a result is.

The dependency graph is built by `data-raw/callout-graph.R`,
which scans the source `.qmd` files and saves the result to
`inst/extdata/callout-graph.rds`.
This appendix reads that saved file rather than re-scanning the notes on
every render, so **re-run that script whenever divs are added, removed, or
re-titled** to refresh the diagram and table below.

```{r}
#| label: load-concept-graph
#| code-fold: true
#| message: false
#| warning: false

library(igraph)
library(ggraph)

cg <- readRDS(here::here("inst/extdata/callout-graph.rds"))

type_levels <- c("def", "thm", "lem", "cor", "prp")
type_labels <- c(
  def = "Definition", thm = "Theorem", lem = "Lemma",
  cor = "Corollary", prp = "Proposition"
)
type_palette <- c(
  def = "#1b9e77", thm = "#d95f02", lem = "#7570b3",
  cor = "#e7298a", prp = "#66a61e"
)
```

There are `r nrow(cg$nodes)` labeled definitions and results in the notes,
connected by `r nrow(cg$edges)` direct dependency links.

## Dependency diagram

@fig-concept-map shows the results that participate in at least one
dependency link, laid out so that connected results sit near each other.
Color encodes the type of result,
and **node size grows with the number of descendants**,
so the most foundational results stand out
(the largest, most-connected nodes are also labeled).
Results with no detected dependency links
(foundational definitions used only in prose, or standalone results)
are omitted from the diagram but appear in @sec-descendants-table
if they have descendants.

:::{#fig-concept-map}

```{r}
#| label: fig-concept-map-plot
#| code-fold: true
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 9

connected <- cg$nodes$id[cg$nodes$id %in% c(cg$edges$from, cg$edges$to)]
core <- graph_from_data_frame(
  cg$edges,
  vertices = cg$nodes[cg$nodes$id %in% connected, ],
  directed = TRUE
)
V(core)$type <- factor(V(core)$type, levels = type_levels)

set.seed(204)
ggraph(core, layout = "stress") +
  geom_edge_link(
    arrow = arrow(length = unit(1.8, "mm"), type = "closed"),
    end_cap = circle(1.6, "mm"),
    edge_alpha = 0.25, edge_width = 0.3
  ) +
  geom_node_point(aes(size = n_desc + 1, color = type)) +
  geom_node_text(
    aes(label = ifelse(n_desc >= 3, name, "")),
    repel = TRUE, size = 2.7, max.overlaps = 30, segment.alpha = 0.3
  ) +
  scale_color_manual(
    values = type_palette, labels = type_labels, name = "Type", drop = FALSE
  ) +
  scale_size_continuous(range = c(1.5, 9), guide = "none") +
  theme_void() +
  theme(
    legend.position = "bottom",
    text = element_text(family = "serif")
  )
```

Dependency structure of the definitions and results in the notes.
An arrow points from a result to each result that uses it.
Node and label size are proportional to the total number of descendants.

:::

## Descendants of each result {#sec-descendants-table}

@tbl-descendants lists every result that has at least one descendant,
sorted by the number of **direct** descendants.
For each result, the first column links to its direct descendants
(results that reference it directly)
and the second column links to its indirect descendants
(results that reach it only through a chain of intermediate results).
All entries are live cross-reference links.

::: {#tbl-descendants}

```{r}
#| label: tbl-descendants-build
#| code-fold: true
#| message: false
#| warning: false

ref_list <- function(ids) {
  if (!length(ids)) return("—")
  ids <- ids[order(match(sub("-.*", "", ids), type_levels), ids)]
  paste0("@", ids, collapse = ", ")
}

ranked <- cg$nodes[cg$nodes$n_direct >= 1, ]
ranked <- ranked[order(-ranked$n_direct, -ranked$n_desc, ranked$id), ]

direct_col <- vapply(
  ranked$id, function(v) ref_list(cg$descendants[[v]]$direct), character(1)
)
indirect_col <- vapply(
  ranked$id, function(v) ref_list(cg$descendants[[v]]$indirect), character(1)
)
descendant_table <- data.frame(
  Result = paste0("@", ranked$id),
  `Direct descendants` = direct_col,
  `Indirect descendants` = indirect_col,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)

knitr::kable(descendant_table, format = "pipe", align = "lll")
```

Direct and indirect descendants of every result that has at least one
descendant, sorted by number of direct descendants.

:::
