---
title: "Concept Map: Definitions and Results"
---

{{< include shared-config.qmd >}}

This appendix shows how the definitions and results
(`def`, `thm`, `lem`, `cor`, `prp` callouts)
in the notes depend on one another.

We say result $B$ is a **descendant** of result $A$
if $A$ is referenced inside the statement or proof of $B$
(directly, or transitively through a chain of intermediate results).
The more descendants a result has,
the more of the rest of the notes rests on it ---
so descendant count is a rough measure of how foundational a result is.

The dependency graph is built by `data-raw/callout-graph.R`,
which scans the source `.qmd` files and saves the result to
`inst/extdata/callout-graph.rds`.
This appendix reads that saved file rather than re-scanning the notes on
every render, so **re-run that script whenever divs are added, removed, or
re-titled** to refresh the diagram and table below.

```{r}
#| label: load-concept-graph
#| code-fold: true
#| message: false
#| warning: false

library(igraph)
library(ggraph)
library(ggrepel)

cg <- readRDS(here::here("inst/extdata/callout-graph.rds"))

type_levels <- c("def", "thm", "lem", "cor", "prp")
type_labels <- c(
  def = "Definition", thm = "Theorem", lem = "Lemma",
  cor = "Corollary", prp = "Proposition"
)
type_palette <- c(
  def = "#1b9e77", thm = "#d95f02", lem = "#7570b3",
  cor = "#e7298a", prp = "#66a61e"
)
```

There are `r nrow(cg$nodes)` labeled definitions and results in the notes,
connected by `r nrow(cg$edges)` direct dependency links.

## Dependency diagram

::: {.content-visible unless-format="pdf"}
::: {.content-visible unless-format="docx"}

@fig-concept-map shows the results that participate in at least one
dependency link, laid out so that connected results sit near each other.
Color encodes the type of result.
Results with no detected dependency links are omitted from the diagram.
Node labels are larger for results with more total descendants,
so the most foundational results are visually prominent.
The most foundational results (most descendants) appear near the center;
peripheral results have no or few descendants.
Long names are abbreviated in the diagram;
the table below gives each result's full name.
Labels are shifted to avoid overlap;
a grey line connects a shifted label to its node position.
Zoom in with Ctrl+scroll (or pinch on mobile) to read individual labels.

:::{#fig-concept-map}

```{r}
#| label: concept-map-ggraph
#| code-fold: true
#| message: false
#| warning: false
#| fig-width: 50
#| fig-height: 50
#| out-width: "100%"

connected <- cg$nodes$id[cg$nodes$id %in% c(cg$edges$from, cg$edges$to)]
core <- graph_from_data_frame(
  cg$edges,
  vertices = cg$nodes[cg$nodes$id %in% connected, ],
  directed = TRUE
)
V(core)$type <- factor(V(core)$type, levels = type_levels)

set.seed(204)
# Use Fruchterman-Reingold for angular arrangement (connected nodes cluster
# together), then override each node's radius so the most foundational nodes
# land near the center.
fr_layout <- create_layout(core, layout = "fr", niter = 2000)
n_desc_vals <- V(core)$n_desc
fr_angle <- atan2(fr_layout$y, fr_layout$x)
# log1p radius: nodes with more descendants land closer to center (r = 0);
# nodes with 0 descendants land at the periphery (r = 1). The log scale
# compresses the high end so the few top hubs don't crowd the center, and
# pulls nodes with even a handful of descendants well off the outer rim.
max_desc <- max(n_desc_vals)
target_r <- 1 - log1p(n_desc_vals) / log1p(max_desc)
layout <- fr_layout
layout$x <- target_r * cos(fr_angle)
layout$y <- target_r * sin(fr_angle)

# Abbreviate long node labels for the diagram only (the descendants table below
# keeps the full titles). First drop synonym lists -- the text after the first
# comma -- so e.g. "Expectation, expected value, population mean" becomes
# "Expectation". Then apply curated short forms for the few names that are
# still long; extend this map as needed.
layout$label <- sub(",.*$", "", layout$title)
short_forms <- c("def-hazard" = "Hazard")
hit <- match(layout$name, names(short_forms))
layout$label <- ifelse(is.na(hit), layout$label, short_forms[hit])
layout$label <- stringr::str_wrap(layout$label, width = 15)

ggraph(layout) +
  geom_edge_link(
    arrow = arrow(length = unit(1.8, "mm"), type = "closed"),
    end_cap = circle(2, "mm"),
    edge_alpha = 0.25, edge_width = 0.3
  ) +
  geom_label_repel(
    data = as.data.frame(layout),
    aes(x = x, y = y, label = label, fill = type, size = n_desc),
    inherit.aes = FALSE,
    # point.size = NA: suppress the dot so the label itself is the node;
    # ggrepel still repels labels away from each other to avoid overlap.
    point.size = NA,
    color = "white",
    fontface = "bold",
    max.overlaps = Inf,
    force = 3,
    force_pull = 1,
    box.padding = unit(0.15, "lines"),
    label.padding = unit(0.1, "lines"),
    label.r = unit(0.05, "lines"),
    label.size = 0.1,
    alpha = 0.88,
    segment.color = "grey60",
    segment.alpha = 0.5,
    segment.size = 0.3,
    min.segment.length = unit(0.5, "lines"),
    seed = 204
  ) +
  scale_fill_manual(
    values = type_palette, labels = type_labels, name = "Type", drop = FALSE
  ) +
  scale_size_continuous(range = c(8.0, 16.0), guide = "none") +
  theme_void() +
  theme(legend.position = "bottom")
```

Dependency structure of the definitions and results in the notes.
An arrow points from a result to each result that uses it.
Color indicates result type (see legend).
Label font size is proportional to the number of total descendants;
results with more descendants appear closer to the center.

:::

:::
:::

::: {.content-visible when-format="pdf"}

The dependency diagram is only available in the
[HTML version of the notes](https://d-morrison.github.io/rme/chapters/concept-map.html).

:::

::: {.content-visible when-format="docx"}

The dependency diagram is only available in the
[HTML version of the notes](https://d-morrison.github.io/rme/chapters/concept-map.html).

:::

## Descendants of each result {#sec-descendants-table}

@tbl-descendants lists every result that has at least one descendant,
sorted by the number of total descendants (direct or transitive).

::: {#tbl-descendants}

```{r}
#| label: descendants-table-build
#| code-fold: true
#| message: false
#| warning: false

ranked <- cg$nodes[cg$nodes$n_desc >= 1, ]
ranked <- ranked[order(-ranked$n_desc, ranked$id), ]

descendant_table <- data.frame(
  Result               = ranked$title,
  Type                 = type_labels[ranked$type],
  `Direct descendants`  = ranked$n_direct,
  `Total descendants`   = ranked$n_desc,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)

knitr::kable(descendant_table, format = "pipe", align = "llrr")
```

All results with at least one descendant,
sorted by number of total descendants (most-foundational first).

:::
