---
title: "Concept Map: Definitions and Results"
---

{{< include shared-config.qmd >}}

This appendix shows how the definitions and results
(`def`, `thm`, `lem`, `cor`, `prp` callouts)
in the notes depend on one another.

We say result $B$ is a **descendant** of result $A$
if $A$ is referenced inside the statement or proof of $B$
(directly, or transitively through a chain of intermediate results).
The more descendants a result has,
the more of the rest of the notes rests on it ---
so descendant count is a rough measure of how foundational a result is.

The dependency graph is built by `data-raw/callout-graph.R`,
which scans the source `.qmd` files and saves the result to
`inst/extdata/callout-graph.rds`.
This appendix reads that saved file rather than re-scanning the notes on
every render, so **re-run that script whenever divs are added, removed, or
re-titled** to refresh the diagram and table below.

```{r}
#| label: load-concept-graph
#| code-fold: true
#| message: false
#| warning: false

library(igraph)
library(ggraph)

cg <- readRDS(here::here("inst/extdata/callout-graph.rds"))

type_levels <- c("def", "thm", "lem", "cor", "prp")
type_labels <- c(
  def = "Definition", thm = "Theorem", lem = "Lemma",
  cor = "Corollary", prp = "Proposition"
)
type_palette <- c(
  def = "#1b9e77", thm = "#d95f02", lem = "#7570b3",
  cor = "#e7298a", prp = "#66a61e"
)
```

There are `r nrow(cg$nodes)` labeled definitions and results in the notes,
connected by `r nrow(cg$edges)` direct dependency links.

## Dependency diagram

::: {.content-visible when-format="html"}

@fig-concept-map shows the results that participate in at least one
dependency link, laid out so that connected results sit near each other.
Color encodes the type of result.
Results with no detected dependency links are omitted from the diagram.
Zoom in with Ctrl+scroll (or pinch on mobile) to read individual labels.

:::{#fig-concept-map}

```{r}
#| label: fig-concept-map-plot
#| code-fold: true
#| message: false
#| warning: false
#| fig-width: 50
#| fig-height: 40
#| out-width: "100%"
#| fig-format: "svg"

connected <- cg$nodes$id[cg$nodes$id %in% c(cg$edges$from, cg$edges$to)]
core <- graph_from_data_frame(
  cg$edges,
  vertices = cg$nodes[cg$nodes$id %in% connected, ],
  directed = TRUE
)
V(core)$type <- factor(V(core)$type, levels = type_levels)

set.seed(204)
ggraph(core, layout = "stress") +
  geom_edge_link(
    arrow = arrow(length = unit(1.8, "mm"), type = "closed"),
    end_cap = circle(3, "mm"),
    edge_alpha = 0.25, edge_width = 0.3
  ) +
  geom_node_label(
    aes(label = title, fill = type),
    color = "white", alpha = 0.85,
    label.padding = unit(0.12, "lines"),
    label.r = unit(0.08, "lines"),
    fontface = "bold",
    size = 2.5
  ) +
  scale_fill_manual(
    values = type_palette, labels = type_labels, name = "Type", drop = FALSE
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    text = element_text(family = "serif")
  )
```

Dependency structure of the definitions and results in the notes.
An arrow points from a result to each result that uses it.
Color indicates result type (see legend).

:::

:::

::: {.content-hidden when-format="html"}

The dependency diagram is only available in the
[HTML version of the notes](https://d-morrison.github.io/rme/chapters/concept-map.html).

:::

## Descendants of each result {#sec-descendants-table}

@tbl-descendants lists every result that has at least one descendant,
sorted by the number of **direct** descendants (results that reference it directly).
The total column counts all transitive descendants.

::: {#tbl-descendants}

```{r}
#| label: tbl-descendants-build
#| code-fold: true
#| message: false
#| warning: false

ranked <- cg$nodes[cg$nodes$n_direct >= 1, ]
ranked <- ranked[order(-ranked$n_direct, -ranked$n_desc, ranked$id), ]

descendant_table <- data.frame(
  Result               = ranked$title,
  Type                 = type_labels[ranked$type],
  `Direct descendants` = ranked$n_direct,
  `Total descendants`  = ranked$n_desc,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)

knitr::kable(descendant_table, format = "pipe", align = "llrr")
```

All results with at least one descendant,
sorted by number of direct descendants (most-foundational first).

:::
