---
title: "Concept Map: Definitions and Results"
---

{{< include shared-config.qmd >}}

This appendix shows how the definitions and results
(`def`, `thm`, `lem`, `cor`, `prp` callouts)
in the notes depend on one another.

We say result $B$ is a **descendant** of result $A$
if $A$ is referenced inside the statement or proof of $B$
(directly, or transitively through a chain of intermediate results).
The more descendants a result has,
the more of the rest of the notes rests on it ---
so descendant count is a rough measure of how foundational a result is.

The dependency graph is built by `data-raw/callout-graph.R`,
which scans the source `.qmd` files and saves the result to
`inst/extdata/callout-graph.rds`.
This appendix reads that saved file rather than re-scanning the notes on
every render, so **re-run that script whenever divs are added, removed, or
re-titled** to refresh the diagram and table below.

```{r}
#| label: load-concept-graph
#| code-fold: true
#| message: false
#| warning: false

library(igraph)
library(ggraph)
library(ggrepel)

cg <- readRDS(here::here("inst/extdata/callout-graph.rds"))
cg$nodes$title <- trimws(gsub("\\\\index\\{[^}]*\\}", "", cg$nodes$title))

type_levels <- c("def", "thm", "lem", "cor", "prp")
type_labels <- c(
  def = "Definition", thm = "Theorem", lem = "Lemma",
  cor = "Corollary", prp = "Proposition"
)
type_palette <- c(
  def = "#1b9e77", thm = "#d95f02", lem = "#7570b3",
  cor = "#e7298a", prp = "#66a61e"
)
```

There are `r nrow(cg$nodes)` labeled definitions and results in the notes,
connected by `r nrow(cg$edges)` direct dependency links.

## Dependency diagram

::: {.content-visible unless-format="pdf"}
::: {.content-visible unless-format="docx"}

@fig-concept-map shows the results that participate in at least one
dependency link, laid out so that connected results sit near each other.
Color encodes the type of result.
Results with no detected dependency links are omitted from the diagram.
Node labels are larger for results with more total descendants,
so the most foundational results are visually prominent.
Zoom in with Ctrl+scroll (or pinch on mobile) to read individual labels.
Lines from labels to dots indicate where ggrepel moved a label to avoid overlap.

:::{#fig-concept-map}

```{r}
#| label: concept-map-ggraph
#| code-fold: true
#| message: false
#| warning: false
#| fig-width: 50
#| fig-height: 50
#| out-width: "100%"
#| fig-format: "svg"

connected <- cg$nodes$id[cg$nodes$id %in% c(cg$edges$from, cg$edges$to)]
core <- graph_from_data_frame(
  cg$edges,
  vertices = cg$nodes[cg$nodes$id %in% connected, ],
  directed = TRUE
)
V(core)$type <- factor(V(core)$type, levels = type_levels)

set.seed(204)
# Stress layout (from graphlayouts, via ggraph) minimises a stress function so
# connected nodes sit close together and nodes are distributed evenly across
# the canvas — much less dead whitespace than the Sugiyama hierarchical layout.
layout <- create_layout(core, layout = "stress")

ggraph(layout) +
  geom_edge_link(
    arrow = arrow(length = unit(1.8, "mm"), type = "closed"),
    end_cap = circle(2, "mm"),
    edge_alpha = 0.25, edge_width = 0.3
  ) +
  geom_node_point(color = "grey60", size = 0.8, alpha = 0.7) +
  geom_label_repel(
    data = as.data.frame(layout),
    aes(x = x, y = y, label = title, fill = type, size = n_desc),
    inherit.aes = FALSE,
    color = "white",
    fontface = "bold",
    max.overlaps = Inf,
    force = 3,
    force_pull = 0.5,
    box.padding = unit(0.25, "lines"),
    label.padding = unit(0.1, "lines"),
    label.r = unit(0.05, "lines"),
    label.size = 0.1,
    alpha = 0.88,
    segment.color = "grey60",
    segment.alpha = 0.6,
    segment.size = 0.3,
    min.segment.length = 0,
    seed = 204
  ) +
  scale_fill_manual(
    values = type_palette, labels = type_labels, name = "Type", drop = FALSE
  ) +
  scale_size_continuous(range = c(4.0, 8.0), guide = "none") +
  theme_void() +
  theme(legend.position = "bottom")
```

Dependency structure of the definitions and results in the notes.
An arrow points from a result to each result that uses it.
Color indicates result type (see legend).
Label font size is proportional to the number of total descendants.

:::

:::
:::

::: {.content-visible when-format="pdf"}

The dependency diagram is only available in the
[HTML version of the notes](https://d-morrison.github.io/rme/chapters/concept-map.html).

:::

::: {.content-visible when-format="docx"}

The dependency diagram is only available in the
[HTML version of the notes](https://d-morrison.github.io/rme/chapters/concept-map.html).

:::

## Descendants of each result {#sec-descendants-table}

@tbl-descendants lists every result that has at least one descendant,
sorted by the number of total descendants (direct or transitive).

::: {#tbl-descendants}

```{r}
#| label: descendants-table-build
#| code-fold: true
#| message: false
#| warning: false

ranked <- cg$nodes[cg$nodes$n_desc >= 1, ]
ranked <- ranked[order(-ranked$n_desc, ranked$id), ]

descendant_table <- data.frame(
  Result               = ranked$title,
  Type                 = type_labels[ranked$type],
  `Direct descendants`  = ranked$n_direct,
  `Total descendants`   = ranked$n_desc,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)

knitr::kable(descendant_table, format = "pipe", align = "llrr")
```

All results with at least one descendant,
sorted by number of total descendants (most-foundational first).

:::
