```{r}
library(MASS) # for glm.nb()
glm1_nb <- glm.nb(
  formula = shared_syr ~ homeless +
    age + sex + ethn + polydrug + hivstat + sexabuse + hplsns,
  data = needles
)

equatiomatic::extract_eq(glm1_nb)
```

```{r}
#| tbl-cap: "Negative binomial model for needle-sharing data"
#| label: tbl-needles-nb
summary(glm1_nb)
```

---

```{r}
#| tbl-cap: "Poisson versus Negative Binomial Regression coefficient estimates"
#| label: tbl-compare-poisson-nb
tibble(name = names(coef(glm1)), poisson = coef(glm1), nb = coef(glm1_nb))
```

#### zero-inflation

```{r}
#| tbl-cap: "Zero-inflated Poisson model"
#| label: tbl-zeroinf-poisson
library(glmmTMB)
zinf_pois <- glmmTMB(
  family = "poisson",
  data = needles,
  formula = shared_syr ~ homeless +
    age + sex + ethn + polydrug + hivstat + sexabuse + hplsns,
  # the zero-inflation submodel uses the exposure, matching Vittinghoff's
  # `inflate(i.homeless)` (@vittinghoff2e, §8.3.1)
  ziformula = ~ homeless
)

zinf_pois |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

::: notes

Another R package for zero-inflated models is [`pscl`](https://cran.r-project.org/web/packages/pscl/index.html) (@pscl08).

:::

#### zero-inflated negative binomial model

```{r}
#| tbl-cap: "Zero-inflated negative binomial model"
#| label: tbl-zeroinf-nb
zinf_nb <- glmmTMB(
  family = nbinom2,
  data = needles,
  formula = shared_syr ~ homeless +
    age + sex + ethn + polydrug + hivstat + sexabuse + hplsns,
  ziformula = ~ homeless
)

zinf_nb |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

::: notes
Both zero-inflated models keep the zero-inflation (structural-zero) submodel
parsimonious — just the exposure, `ziformula = ~ homeless` —
matching @vittinghoff2e's `inflate(i.homeless)` [§8.3.1].

A richer zero-inflation submodel is poorly identified here.
With the negative binomial in particular,
the overdispersion parameter and the zero-inflation component
both compete to explain the excess zeros,
so adding several covariates to the submodel makes the logistic fit
*separate*: its coefficients diverge and the exponentiated estimates
(odds ratios) run off to $\pm\infty$.
Keeping the submodel small avoids that.
:::
