```{r}
library(tidyverse)
library(haven)
needles <-
  here::here("inst/extdata/needle_sharing.dta") |>
  read_dta() |>
  as_tibble() |>
  mutate(
    hivstat = hivstat |>
      case_match(1 ~ "HIV+", 0 ~ "HIV-") |>
      factor() |>
      relevel(ref = "HIV-"),
    polydrug = polydrug |>
      case_match(1 ~ "multiple drugs used", 0 ~ "one drug used") |>
      factor() |>
      relevel(ref = "one drug used"),
    homeless = homeless |>
      case_match(1 ~ "homeless", 0 ~ "not homeless") |>
      factor() |>
      relevel(ref = "not homeless"),
    ethn = ethn |>
      factor() |>
      forcats::fct_lump_min(min = 5, other_level = "Other") |>
      relevel(ref = "White"),
    sex = sex |> factor() |> relevel(ref = "M")
  ) |>
  labelled::set_variable_labels(
    "id" = "Participant identifier",
    "sex" = "Biological sex (reference = Male)",
    "ethn" = "Self-reported ethnicity (reference = White; n<5 lumped to Other)",
    "age" = "Age at first interview (years)",
    "dprsn_dx" = "Depression diagnosis indicator",
    "sexabuse" = "History of sexual abuse (1 = yes, 0 = no)",
    "shared_syr" = "Number of shared syringes (in 30 days)",
    "hivstat" = "HIV serostatus (reference = HIV-)",
    "hplsns" = "Hopelessness score",
    "nivdu" = "Number of injections (in 30 days)",
    "shsyryn" = "Shared syringe? (1 = yes, 0 = no)",
    "sqrtnivd" = "sqrt(No. of injections in 30 days)",
    "logshsyr" = "log(No. of shared needles)",
    "polydrug" = "Multiple non-injection drugs used?",
    "sqrtninj" = "sqrt(No. of injections in 30 days); duplicate of sqrtnivd",
    "homeless" = "Homeless? (reference = not homeless)",
    "shsyr" = "Shared needles (30 days); sparse duplicate of shared_syr"
  )
```

Dataset source: Chapter 8 of *Regression Methods in Biostatistics*, 2e
([UCSF companion website](https://regression.ucsf.edu/second-edition/data-examples-and-problems)) —
a longitudinal panel study of injection drug users measuring
repeated needle-sharing behaviour.
$n = 128$ participants, 17 variables.

:::{#tbl-needles-data-dict}

```{r}
dict <- tibble(
  variable = names(needles),
  description = labelled::get_variable_labels(needles) |> 
    sapply(function(x) ifelse(is.null(x), "", x)),
)
dict |> pander::pander()
```

Data dictionary for `needles` data

:::

---

@tbl-needles-demographics summarizes the demographic and behavioural
variables for the participants.

:::{#tbl-needles-demographics}

```{r}
#| code-fold: true
needles |>
  dplyr::select(-id) |>
  gtsummary::tbl_summary()
```

Demographics and behavioural variables for the needle-sharing dataset
(all participants).

:::

{{< slidebreak >}}

Before modeling, we sketch a hypothesized causal structure for these
variables.

::: {#fig-needles-dag}

```{r}
#| fig-width: 7
#| fig-height: 5
#| code-fold: true

library(dagitty)

needles_dag <- dagitty(
  'dag {
    age      [pos="0,-2"]
    sex      [pos="0,-1"]
    ethn     [pos="0,0"]
    homeless [pos="0,1"]
    polydrug [pos="0,2"]
    nivdu    [pos="2,0"]
    shared_syr [outcome, pos="4,0"]

    age -> nivdu
    sex -> nivdu
    ethn -> nivdu
    homeless -> nivdu
    polydrug -> nivdu

    age -> shared_syr
    sex -> shared_syr
    ethn -> shared_syr
    homeless -> shared_syr
    polydrug -> shared_syr

    nivdu -> shared_syr
  }'
)

plot(needles_dag)
```

Hypothesized causal structure for the needle-sharing example.
Demographic factors (`age`, `sex`, `ethn`) and social factors
(`homeless`, `polydrug`) influence both how often a participant injects
(`nivdu`) and how many of those injections involve a shared syringe
(`shared_syr`, the outcome).

:::

::: notes
In @fig-needles-dag, injection frequency `nivdu` is a **mediator**:
each covariate can affect the number of shared syringes both
**directly** (its effect on the propensity to share per injection)
and **indirectly** (through how often the participant injects).

Because `nivdu` lies *on* these causal paths, we do **not** adjust for it
(neither as a covariate nor as an offset): conditioning on a mediator
would block the indirect path and leave only a partial effect.
Modeling the count of shared syringes on the covariates alone therefore
estimates each covariate's **total** association with needle-sharing —
the approach taken by @vittinghoff2e [§8.3.1].

This DAG is simplified; a fuller model might include
HIV status, depression, or hopelessness.
:::

```{r}
#| tbl-cap: "Needle-sharing data"
#| label: tbl-needle-data
needles
```

---

```{r}
#| fig-cap: "Rates of needle sharing"
#| label: fig-needles
library(ggplot2)

needles |>
  dplyr::filter(nivdu > 0) |>
  ggplot(
    aes(
      x = age,
      y = shared_syr / nivdu,
      shape = sex,
      col = ethn
    )
  ) +
  geom_point(aes(size = nivdu), alpha = .5) +
  scale_size_area(max_size = 4) +
  facet_grid(cols = vars(polydrug), rows = vars(homeless)) +
  labs(y = "Shared needles per injection") +
  theme(legend.position = "bottom")
```


{{< slidebreak >}}

@fig-needles-raw shows the underlying counts:
the number of shared syringes against the number of injections,
with a dashed $y = x$ line marking the ceiling
where every injection involves a shared syringe.

::: {#fig-needles-raw}

```{r}
#| code-fold: true
library(ggplot2)

needles |>
  ggplot(
    aes(
      x = nivdu,
      y = shared_syr,
      shape = sex,
      col = ethn
    )
  ) +
  geom_point(alpha = .5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", alpha = .5) +
  facet_grid(cols = vars(polydrug), rows = vars(homeless)) +
  labs(
    x = "Number of injections (in 30 days)",
    y = "Number of shared syringes (in 30 days)"
  ) +
  guides(
    col = guide_legend(nrow = 2),
    shape = guide_legend(nrow = 2)
  ) +
  theme(legend.position = "bottom", legend.box = "vertical")
```

Number of shared syringes versus number of injections
(raw counts, 30-day window).
The dashed line marks $y = x$ (every injection involves a shared syringe).

:::

---

## Covariate counts

```{r}
#| tbl-cap: "Counts in `needles` by sex, housing status, and multiple drug use"
#| label: tbl-count-needles
#| code-fold: show
needles |>
  dplyr::select(sex, homeless, polydrug) |>
  summary()
```

---

There's only one individual with `sex = Trans`, 
which unfortunately isn't enough data to analyze.
We will remove that individual:

```{r}
#| label: remove-trans-obs
#| code-summary: 'remove singleton observation with sex == Trans'
needles <- needles |> filter(sex != "Trans")

```

---

## Model {.smaller}

Following @vittinghoff2e [§8.3.1],
we model the **count** of shared syringes directly
with a Poisson regression (log link),
rather than modeling a rate.

**Why not use `nivdu` as an offset?**
A Poisson *rate* model would add `offset(log(nivdu))`,
forcing the expected count to be proportional to injection frequency
(the coefficient of `log(nivdu)` fixed at exactly 1).
We avoid that here for two reasons:

- **`nivdu` is a mediator, not an exposure window.**
An offset is appropriate when its variable is a known denominator —
person-time or population at risk — that is external to the causal
question. Injection frequency is instead *caused* by the covariates
(@fig-needles-dag), so conditioning on it (whether as an offset or as a
covariate) would block the indirect path and estimate only a partial
effect. We want each covariate's *total* association with needle-sharing.
- **It imposes an untested proportionality assumption.**
Fixing the `log(nivdu)` coefficient at 1 assumes sharing scales
one-for-one with injecting, which the data need not support.

Vittinghoff's analysis uses neither an offset nor `nivdu` as a covariate.

We also use **main effects only**:
a `homeless × polydrug` interaction is not identifiable here,
because no non-homeless participant who used multiple drugs shared any
needles, leaving that cell empty for the outcome.

```{r}
glm1 <- glm(
  formula = shared_syr ~ age + sex + homeless + polydrug,
  family = stats::poisson,
  data = needles
)
library(equatiomatic)
equatiomatic::extract_eq(glm1)
```

---

```{r}
#| tbl-cap: "Poisson count model for needle-sharing data"
#| label: tbl-needles-pois


library(parameters)
glm1 |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

---

```{r}
#| fig-cap: "Fitted Poisson count model and observed shared-syringe counts"
#| label: fig-needles-pois
#| code-fold: true
library(ggplot2)

# Predicted expected counts across age and the categorical covariates.
# The model has no offset, so predict(type = "response") returns the
# expected number of shared syringes directly.
pred_grid <- with(glm1$model, expand.grid(
  age = seq(min(age), max(age), length.out = 100),
  sex = levels(droplevels(sex)),
  homeless = levels(homeless),
  polydrug = levels(polydrug)
))
pred_grid$shared_syr <- predict(glm1, newdata = pred_grid, type = "response")

ggplot(glm1$model, aes(x = age, y = shared_syr, color = sex)) +
  geom_point(alpha = 0.4) +
  geom_line(data = pred_grid, linewidth = 0.8) +
  facet_grid(
    rows = vars(homeless),
    cols = vars(polydrug),
    labeller = label_both
  ) +
  labs(y = "Number of shared syringes (30 days)") +
  guides(color = guide_legend(nrow = 1)) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Interpreting the coefficients {.smaller}

For a Poisson model with a log link,
exponentiating a coefficient turns it into a *ratio of expected counts*
(an **incidence rate ratio**, IRR, in @vittinghoff2e's terminology):
$\exp{\b_j}$ is the multiplicative change in the expected number of
shared syringes per one-unit increase in $x_j$,
holding the other covariates fixed.
Because we do not condition on the mediator `nivdu`,
each ratio is a *total* association.
The fitted ratios are reported in @tbl-needles-pois.

```{r}
#| code-fold: true
irr <- exp(coef(glm1))
```

**Baseline (intercept).**
$\exp{\hat\b_0} = `r round(irr[["(Intercept)"]], 3)`$
is the expected count of shared syringes for the reference participant
(male, not homeless, one drug used), extrapolated to age 0.
Age 0 is outside the data, so the intercept is a mathematical anchor
rather than a directly interpretable quantity.

**Age.**
$\exp{\hat\b_{\text{age}}} = `r round(irr[["age"]], 3)`$ per year:
essentially no association between age and the expected count.

**Sex.**
$\exp{\hat\b_{\text{sexF}}} = `r round(irr[["sexF"]], 2)`$:
women's expected count of shared syringes is about
`r round(irr[["sexF"]], 2)` times men's,
adjusting for age, housing status, and drug-use pattern.

**Homelessness.**
$\exp{\hat\b_{\text{homeless}}} = `r round(irr[["homelesshomeless"]], 2)`$:
homeless participants share syringes about
`r round(irr[["homelesshomeless"]], 2)` times as often as the non-homeless,
adjusting for the other covariates.
@vittinghoff2e [§8.3.1] reports a comparable unadjusted IRR of about 3.3
for homelessness in a single-predictor model.

**Multiple-drug use.**
$\exp{\hat\b_{\text{polydrug}}} = `r round(irr[["polydrugmultiple drugs used"]], 2)`$:
multiple-drug users have a *lower* expected count than single-drug users,
though only `r sum(glm1$model[["polydrug"]] == "multiple drugs used")`
participants used multiple drugs, so this estimate is imprecise.

**Overdispersion caveat.**
These data are highly overdispersed — the variance far exceeds the mean —
so the model-based Poisson standard errors are too small.
@vittinghoff2e [§8.3.1] recommends robust standard errors or a
negative-binomial model for this example.

