```{r}
library(tidyverse)
library(haven)
needles = 
  here::here("inst/extdata/needle_sharing.dta") |> 
  read_dta() |>
  as_tibble() |>
  mutate(
    hivstat =
      hivstat |>
      case_match(
        1 ~ "HIV+",
        0 ~ "HIV-") |> 
      factor() |> 
      relevel(ref = "HIV-"),
    polydrug =
      polydrug |>
      case_match(
        1 ~ "multiple drugs used",
        0 ~ "one drug used") |>
      factor() |>
      relevel(ref = "one drug used"),
    homeless = 
      homeless |> 
      case_match(
        1 ~ "homeless", 
        0 ~ "not homeless") |>
      factor() |>
      relevel(ref = "not homeless"),
    ethn = ethn |> factor() |> relevel(ref = "White"),
    sex = sex |> factor() |> relevel(ref = "M")
  ) |> 
  labelled::set_variable_labels(
    "id" = "Participant identifier",
    "sex" = "Biological sex (reference = Male)",
    "ethn" = "Self-reported ethnicity (reference = White)",
    "age" = "Age at first interview (years)",
    "dprsn_dx" = "Depression diagnosis indicator",
    "sexabuse" = "History of sexual abuse (1 = yes, 0 = no)",
    "shared_syr" = "Number of shared syringes (in 30 days)",
    "hivstat" = "HIV serostatus (reference = HIV-)",
    "hplsns" = "Hopelessness score",
    "nivdu" = "Number of injections (in 30 days)",
    "shsyryn" = "Shared syringe? (1 = yes, 0 = no)",
    "sqrtnivd" = "sqrt(No. of injections in 30 days)",
    "logshsyr" = "log(No. of shared needles)",
    "polydrug" = "Multiple non-injection drugs used?",
    "sqrtninj" = "sqrt(No. of injections in 30 days)",
    "homeless" = "Homeless? (reference = not homeless)",
    "shsyr" = "Number of shared needles (in 30 days)"
  )
```

Dataset source: Chapter 8 of *Regression Methods in Biostatistics*, 2e
([UCSF companion website](https://regression.ucsf.edu/second-edition/data-examples-and-problems)) —
a longitudinal panel study of injection drug users measuring
repeated needle-sharing behaviour.
$n = 128$ participants, 17 variables.

:::{#tbl-needles-data-dict}

```{r}
dict <- tibble(
  variable = names(needles),
  description = labelled::get_variable_labels(needles) |> 
    sapply(function(x) ifelse(is.null(x), "", x)),
)
dict |> pander::pander()
```

Data dictionary for `needles` data

:::

---

:::{#tbl-needles-demographics}

```{r}
#| code-fold: true

demog_data <- needles |>
  dplyr::filter(sex != "Trans")

med_iqr <- function(x) {
  q <- stats::quantile(x, c(0.25, 0.5, 0.75), na.rm = TRUE)
  sprintf("%.0f (%.0f, %.0f)", q[2], q[1], q[3])
}

n_pct <- function(x, level) {
  sprintf(
    "%d (%.0f%%)",
    sum(x == level, na.rm = TRUE),
    100 * mean(x == level, na.rm = TRUE)
  )
}

ethn_levels <- levels(demog_data$ethn)
ethn_rows <- vapply(
  ethn_levels,
  function(L) n_pct(demog_data$ethn, L),
  character(1)
)

demog_tbl <- tibble::tibble(
  Characteristic = c(
    "N",
    "Age (years), median (Q1, Q3)",
    "Female, n (%)",
    paste0("Ethnicity — ", ethn_levels, ", n (%)"),
    "HIV+, n (%)",
    "Homeless, n (%)",
    "Multiple drugs used, n (%)",
    "Injections in 30 days, median (Q1, Q3)",
    "Shared syringes in 30 days, median (Q1, Q3)"
  ),
  Value = c(
    as.character(nrow(demog_data)),
    med_iqr(demog_data$age),
    n_pct(demog_data$sex, "F"),
    ethn_rows,
    n_pct(demog_data$hivstat, "HIV+"),
    n_pct(demog_data$homeless, "homeless"),
    n_pct(demog_data$polydrug, "multiple drugs used"),
    med_iqr(demog_data$nivdu),
    med_iqr(demog_data$shared_syr)
  )
)

demog_tbl |> pander::pander()
```

Demographics and key behavioural variables for the needle-sharing dataset
(excluding one individual with `sex == "Trans"`).

:::

---

```{r}
#| tbl-cap: "Needle-sharing data"
#| label: tbl-needle-data
needles
```

---

```{r}
#| fig-cap: "Rates of needle sharing"
#| label: fig-needles
library(ggplot2)

needles |>
  dplyr::filter(nivdu > 0) |>
  ggplot(
    aes(
      x = age,
      y = shared_syr / nivdu,
      shape = sex,
      col = ethn
    )
  ) +
  geom_point(aes(size = nivdu), alpha = .5) +
  scale_size_area(max_size = 4) +
  facet_grid(cols = vars(polydrug), rows = vars(homeless)) +
  labs(y = "Shared needles per injection") +
  theme(legend.position = "bottom")
```

---

#### Covariate counts

```{r}
#| tbl-cap: "Counts of observations in `needles` dataset by sex, unhoused status, and multiple drug use"
#| label: tbl-count-needles
#| code-fold: show
needles |> 
  dplyr::select(sex, homeless, polydrug) |> 
  summary()
```

---

There's only one individual with `sex = Trans`, 
which unfortunately isn't enough data to analyze.
We will remove that individual:

```{r}
#| label: remove-trans-obs
#| code-summary: 'remove singleton observation with sex == Trans'
needles = needles |> filter(sex != "Trans")

```

---

### Model {.smaller}

```{r}
glm1 = 
  needles |> 
  dplyr::filter(nivdu > 0) |> 
  glm(
    offset = log(nivdu),
    family = stats::poisson,
    formula = shared_syr ~ age + sex + homeless*polydrug
  )
library(equatiomatic)
equatiomatic::extract_eq(glm1)
```

---

```{r}
#| tbl-cap: "Poisson model for needle-sharing data"
#| label: tbl-needles-pois


library(parameters)
glm1 |> parameters(exponentiate = TRUE) |> 
  print_md()
```

---

```{r}
#| fig-cap: "fitted Poisson model for needle-sharing data"
#| label: fig-needles-pois
library(sjPlot)
glm1 |> 
  sjPlot::plot_model(
    type = "pred",
    terms = c("age", "sex", "homeless", "polydrug"),
    show.data = TRUE
  ) +
  theme(legend.position = "bottom")
```

