
```{r}
library(MASS) #need this for glm.nb()
glm1.nb = glm.nb(
  data = needles,
  shared_syr ~ age + sex + homeless + polydrug + hivstat
)

equatiomatic::extract_eq(glm1.nb)
```

```{r}
#| tbl-cap: "Negative binomial model for needle-sharing data"
#| label: tbl-needles-nb
summary(glm1.nb)
```

---

```{r}
#| tbl-cap: "Poisson versus Negative Binomial Regression coefficient estimates"
#| label: tbl-compare-poisson-nb
tibble(name = names(coef(glm1)), poisson = coef(glm1), nb = coef(glm1.nb))
```

#### zero-inflation

```{r}
#| tbl-cap: "Zero-inflated poisson model"
#| label: tbl-zeroinf-poisson
library(glmmTMB)
zinf_fit1 = glmmTMB(
  family = "poisson",
  data  = needles,
  formula = shared_syr ~ age + sex + homeless + polydrug + hivstat,
  # hivstat omitted from ziformula because only ~8 of 128 participants are
  # HIV+, which risks convergence failure in the zero-inflation submodel
  # (mirroring why the homeless*polydrug interaction was avoided above).
  ziformula = ~ age + sex + homeless + polydrug
)

zinf_fit1 |>
  parameters(exponentiate = TRUE) |>
  print_md()

```

::: notes

Another R package for zero-inflated models is [`pscl`](https://cran.r-project.org/web/packages/pscl/index.html) (@pscl08).

:::

#### zero-inflated negative binomial model

```{r}
#| tbl-cap: "Zero-inflated negative binomial model"
#| label: tbl-zeroinf-nb
library(glmmTMB)
zinf_fit_nb = glmmTMB(
  family = nbinom2,
  data  = needles,
  formula = shared_syr ~ age + sex + homeless + polydrug + hivstat,
  # Intercept-only zero-inflation submodel (see note below).
  ziformula = ~ 1
)

zinf_fit_nb |>
  parameters(exponentiate = TRUE) |>
  print_md()

```

::: notes
Unlike the zero-inflated Poisson model above,
here the zero-inflation submodel uses only an intercept (`ziformula = ~ 1`).
Two reasons to drop the covariates from this submodel:

- **The DAG gives it no covariates.**
@fig-needles-dag posits a *single* process
in which the covariates act on `shared_syr`
(directly and through the mediator `nivdu`);
it does not posit a separate latent class of "structural never-sharers"
for those covariates to predict.
So a covariate-laden zero-inflation submodel
is not identified by the assumed causal structure.

- **The data cannot identify it either.**
The negative binomial's overdispersion parameter and the zero-inflation component
both compete to explain the excess zeros,
so with all four covariates in the zero-inflation submodel
the logistic fit *separates*:
its coefficients diverge and the exponentiated estimates (odds ratios)
run off to $\pm\infty$.
Reducing the zero-inflation submodel to a single constant
removes the separation
while still allowing a fixed fraction of structural zeros.
:::

