```{r}
library(tidyverse)
library(haven)
needles <-
  here::here("inst/extdata/needle_sharing.dta") |>
  read_dta() |>
  as_tibble() |>
  mutate(
    hivstat = hivstat |>
      case_match(1 ~ "HIV+", 0 ~ "HIV-") |>
      factor() |>
      relevel(ref = "HIV-"),
    polydrug = polydrug |>
      case_match(1 ~ "multiple drugs used", 0 ~ "one drug used") |>
      factor() |>
      relevel(ref = "one drug used"),
    homeless = homeless |>
      case_match(1 ~ "homeless", 0 ~ "not homeless") |>
      factor() |>
      relevel(ref = "not homeless"),
    ethn = ethn |>
      factor() |>
      forcats::fct_lump_min(min = 5, other_level = "Other") |>
      relevel(ref = "White"),
    sex = sex |> factor() |> relevel(ref = "M")
  ) |>
  labelled::set_variable_labels(
    "id" = "Participant identifier",
    "sex" = "Biological sex (reference = Male)",
    "ethn" = "Self-reported ethnicity (reference = White; n<5 lumped to Other)",
    "age" = "Age at first interview (years)",
    "dprsn_dx" = "Depression diagnosis indicator",
    "sexabuse" = "History of sexual abuse (1 = yes, 0 = no)",
    "shared_syr" = "Number of shared syringes (in 30 days)",
    "hivstat" = "HIV serostatus (reference = HIV-)",
    "hplsns" = "Hopelessness score",
    "nivdu" = "Number of injections (in 30 days)",
    "shsyryn" = "Shared syringe? (1 = yes, 0 = no)",
    "sqrtnivd" = "sqrt(No. of injections in 30 days)",
    "logshsyr" = "log(No. of shared needles)",
    "polydrug" = "Multiple non-injection drugs used?",
    "sqrtninj" = "sqrt(No. of injections in 30 days); duplicate of sqrtnivd",
    "homeless" = "Homeless? (reference = not homeless)",
    "shsyr" = "Number of shared needles (in 30 days)"
  )
```

Dataset source: Chapter 8 of *Regression Methods in Biostatistics*, 2e
([UCSF companion website](https://regression.ucsf.edu/second-edition/data-examples-and-problems)) —
a longitudinal panel study of injection drug users measuring
repeated needle-sharing behaviour.
$n = 128$ participants, 17 variables.

:::{#tbl-needles-data-dict}

```{r}
dict <- tibble(
  variable = names(needles),
  description = labelled::get_variable_labels(needles) |> 
    sapply(function(x) ifelse(is.null(x), "", x)),
)
dict |> pander::pander()
```

Data dictionary for `needles` data

:::

---

:::{#tbl-needles-demographics}

```{r}
#| code-fold: true
needles |>
  dplyr::select(-id) |>
  gtsummary::tbl_summary()
```

Demographics and behavioural variables for the needle-sharing dataset
(all participants; the `id` identifier column is omitted).

:::

{{< slidebreak >}}

Before modeling, we sketch a hypothesized causal structure for these
variables.

::: {#fig-needles-dag}

```{r}
#| fig-width: 7
#| fig-height: 5
#| code-fold: true

library(dagitty)

needles_dag <- dagitty(
  'dag {
    age      [pos="0,-2"]
    sex      [pos="0,-1"]
    ethn     [pos="0,0"]
    homeless [pos="0,1"]
    polydrug [pos="0,2"]
    nivdu    [pos="2,0"]
    shared_syr [outcome, pos="4,0"]

    age -> nivdu
    sex -> nivdu
    ethn -> nivdu
    homeless -> nivdu
    polydrug -> nivdu

    age -> shared_syr
    sex -> shared_syr
    ethn -> shared_syr
    homeless -> shared_syr
    polydrug -> shared_syr

    nivdu -> shared_syr
  }'
)

plot(needles_dag)
```

Hypothesized causal structure for the needle-sharing example.
Demographic factors (`age`, `sex`, `ethn`) and social factors
(`homeless`, `polydrug`) influence both how often a participant injects
(`nivdu`) and how many of those injections involve a shared syringe
(`shared_syr`, the outcome).

:::

::: notes
In @fig-needles-dag, injection frequency `nivdu` is a **mediator**:
each covariate can affect the number of shared syringes both
**directly** (its effect on the propensity to share *per injection*)
and **indirectly** (through how often the participant injects).

This distinction drives the two models we fit below:

- Putting `nivdu` in the model as an **offset** (`offset(log(nivdu))`)
conditions on the mediator. The covariate coefficients then describe the
*per-injection sharing rate* — the **direct** path, holding injection
frequency fixed.
- Leaving `nivdu` out of the model lets the covariate coefficients
describe the *total* number of shared syringes — the **total** effect,
including the indirect path through `nivdu`.

This DAG is simplified; a fuller model might include
HIV status, depression, or hopelessness.
:::

```{r}
#| tbl-cap: "Needle-sharing data"
#| label: tbl-needle-data
needles
```

---

```{r}
#| fig-cap: "Rates of needle sharing"
#| label: fig-needles
library(ggplot2)

needles |>
  dplyr::filter(nivdu > 0) |>
  ggplot(
    aes(
      x = age,
      y = shared_syr / nivdu,
      shape = sex,
      col = ethn
    )
  ) +
  geom_point(aes(size = nivdu), alpha = .5) +
  scale_size_area(max_size = 4) +
  facet_grid(cols = vars(polydrug), rows = vars(homeless)) +
  labs(y = "Shared needles per injection") +
  theme(legend.position = "bottom")
```


{{< slidebreak >}}

@fig-needles-raw shows the underlying counts:
the number of shared syringes against the number of injections,
with a dashed $y = x$ line marking the ceiling
where every injection involves a shared syringe.

::: {#fig-needles-raw}

```{r}
#| code-fold: true
library(ggplot2)

needles |>
  ggplot(
    aes(
      x = nivdu,
      y = shared_syr,
      shape = sex,
      col = ethn
    )
  ) +
  geom_point(alpha = .5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", alpha = .5) +
  facet_grid(cols = vars(polydrug), rows = vars(homeless)) +
  labs(
    x = "Number of injections (in 30 days)",
    y = "Number of shared syringes (in 30 days)"
  ) +
  guides(
    col = guide_legend(nrow = 2),
    shape = guide_legend(nrow = 2)
  ) +
  theme(legend.position = "bottom", legend.box = "vertical")
```

Number of shared syringes versus number of injections
(raw counts, 30-day window).
The dashed line marks $y = x$ (every injection involves a shared syringe).

:::

---

## Covariate counts

```{r}
#| tbl-cap: "Counts in `needles` by sex, housing status, and multiple drug use"
#| label: tbl-count-needles
#| code-fold: show
needles |>
  dplyr::select(sex, homeless, polydrug) |>
  summary()
```

---

There's only one individual with `sex = Trans`, 
which unfortunately isn't enough data to analyze.
We will remove that individual:

```{r}
#| label: remove-trans-obs
#| code-summary: 'remove singleton observation with sex == Trans'
needles <- needles |> filter(sex != "Trans")

```

---

## Model {.smaller}

As @fig-needles-dag suggests, injection frequency `nivdu` is a mediator,
so we fit two complementary Poisson models:

- a **rate model** that includes `nivdu` as an offset, estimating the
effect of each covariate on the *per-injection* sharing rate
(the direct path, holding injection frequency fixed); and
- a **count model** that omits `nivdu`, estimating the *total* effect of
each covariate on the number of shared syringes
(direct plus the indirect path through `nivdu`).

### Rate model {.smaller}

The offset `offset(log(nivdu))` fixes the coefficient of `log(nivdu)`
at 1, which is equivalent to modeling the *rate* of shared syringes
per injection (`shared_syr / nivdu`):

```{r}
glm1 <-
  needles |>
  dplyr::filter(nivdu > 0) |>
  glm(
    family = stats::poisson,
    formula = shared_syr ~ age + sex + homeless * polydrug +
      offset(log(nivdu))
  )
library(equatiomatic)
equatiomatic::extract_eq(glm1)
```

---

```{r}
#| tbl-cap: "Poisson model for needle-sharing data"
#| label: tbl-needles-pois


library(parameters)
glm1 |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

---

```{r}
#| fig-cap: "Fitted Poisson rate model and observed needle-sharing rates"
#| label: fig-needles-pois
#| code-fold: true
library(ggplot2)

# Build a prediction grid that varies age and the categorical covariates,
# with nivdu = 1 so the offset log(nivdu) = 0 and predict(type = "response")
# returns the Poisson rate per injection (matching the y-axis scale of
# @fig-needles).
pred_grid <- with(glm1$model, expand.grid(
  age = seq(min(age), max(age), length.out = 100),
  sex = levels(droplevels(sex)),
  homeless = levels(homeless),
  polydrug = levels(polydrug),
  nivdu = 1
))
pred_grid$rate <- predict(glm1, newdata = pred_grid, type = "response")

obs_data <- needles |>
  dplyr::filter(
    nivdu > 0,
    !is.na(homeless), !is.na(polydrug), !is.na(sex), !is.na(age)
  ) |>
  dplyr::mutate(rate = shared_syr / nivdu)

ggplot(obs_data, aes(x = age, y = rate, color = sex)) +
  geom_point(aes(size = nivdu), alpha = 0.4) +
  geom_line(data = pred_grid, linewidth = 0.8) +
  scale_size_area(max_size = 4) +
  facet_grid(
    rows = vars(homeless),
    cols = vars(polydrug),
    labeller = label_both
  ) +
  labs(y = "Shared needles per injection") +
  guides(
    color = guide_legend(nrow = 1),
    size = guide_legend(nrow = 1)
  ) +
  theme_bw() +
  theme(legend.position = "bottom", legend.box = "vertical")
```

---

### Count model {.smaller}

Dropping the offset models the *count* of shared syringes directly.
We keep the same sample of injectors (`nivdu > 0`) so the two models are
comparable; the only difference is whether `nivdu` enters as an offset:

```{r}
glm2 <-
  needles |>
  dplyr::filter(nivdu > 0) |>
  glm(
    family = stats::poisson,
    formula = shared_syr ~ age + sex + homeless * polydrug
  )
equatiomatic::extract_eq(glm2)
```

---

```{r}
#| tbl-cap: "Count model (no offset) for needle-sharing data"
#| label: tbl-needles-count
glm2 |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

## Interpreting the coefficients {.smaller}

For a Poisson model with a log link,
exponentiating a coefficient turns it into a *ratio*:
$\exp(\beta_j)$ is the multiplicative change in the expected outcome
per one-unit increase in $x_j$,
holding the other covariates fixed
(this is the rate-ratio interpretation introduced earlier in this chapter).
In the **rate model** this ratio is a *rate ratio* (shared syringes per
injection); in the **count model** it is a *count ratio*
(total shared syringes).

```{r}
#| code-fold: true
rr_rate <- exp(coef(glm1))
rr_count <- exp(coef(glm2))
```

**Baseline (intercept).**
In the rate model, $\exp(\hat\beta_0) = `r round(rr_rate[["(Intercept)"]], 3)`$
is the estimated sharing rate per injection for the reference participant
(male, not homeless, one drug used),
extrapolated to age 0.
Age 0 is outside the data, so the intercept is a mathematical anchor
rather than a directly interpretable quantity.

**Age.**
The rate ratio per additional year of age is
$\exp(\hat\beta_{\text{age}}) = `r round(rr_rate[["age"]], 3)`$,
i.e., essentially no change in the per-injection sharing rate with age.

**Sex.**
The rate ratio for female versus male is
$\exp(\hat\beta_{\text{sexF}}) = `r round(rr_rate[["sexF"]], 2)`$:
women share needles at about `r round(rr_rate[["sexF"]], 2)` times
the per-injection rate of men with the same age, housing status, and
drug-use pattern.
In the count model the corresponding count ratio is larger,
about `r round(rr_count[["sexF"]], 2)`.
Because the total-count ratio exceeds the per-injection rate ratio,
part of women's higher total sharing operates **indirectly** —
through injecting more often
(the `sex → nivdu → shared_syr` path in @fig-needles-dag) —
on top of their higher per-injection rate.

**Homelessness.**
Among participants using one drug,
the rate ratio for homeless versus not-homeless is
$\exp(\hat\beta_{\text{homeless}}) = `r round(rr_rate[["homelesshomeless"]], 2)`$,
versus a count ratio of about `r round(rr_count[["homelesshomeless"]], 2)`
in the count model —
again, a larger *total* association than *direct* (per-injection) one.

**Interaction (caution).**
The `homeless × polydrug` cell is sparse,
which produces an extreme, unstable interaction estimate
(a rate ratio in the hundreds of thousands).
We would not interpret that coefficient without more data;
it is a reminder to check cell counts (see @tbl-count-needles)
before trusting an interaction term.

