To evaluate whether overdispersion and excess zeros distort inference
in the needle-sharing dataset,
we fit a negative binomial model and compare its coefficient estimates
and standard errors against the baseline Poisson model.

```{r}
#| label: fit-nb-model
library(MASS) # for glm.nb()
glm1_nb <- glm.nb(
  formula = shared_syr ~ homeless +
    age + sex + ethn + polydrug + hivstat + sexabuse + hplsns,
  data = needles
)

equatiomatic::extract_eq(glm1_nb)
```

```{r}
#| tbl-cap: "Negative binomial model for needle-sharing data"
#| label: tbl-needles-nb
summary(glm1_nb)
```

---

Evaluating the estimated parameters across models reveals
how accounting for overdispersion affects inference.
While the estimated log-rate coefficients remain comparable
between the Poisson and negative binomial specifications,
the standard errors under the negative binomial model appropriately widen
to reflect the excess outcome variability.

```{r}
#| tbl-cap: "Poisson vs. Negative Binomial regression estimates"
#| label: tbl-compare-poisson-nb
tibble(
  Parameter = names(coef(glm1)),
  `Poisson Coef` = coef(glm1),
  `Poisson SE` = summary(glm1)$coefficients[, "Std. Error"],
  `NB Coef` = coef(glm1_nb),
  `NB SE` = summary(glm1_nb)$coefficients[, "Std. Error"]
) |>
  pander::pander()
```

{{< slidebreak >}}

### Zero-inflated models for needle-sharing

Because many participants report zero shared syringes in the past 30 days,
we fit a zero-inflated Poisson (ZIP) model
to distinguish structural non-sharers from count variability.

```{r}
#| tbl-cap: "Zero-inflated Poisson model"
#| label: tbl-zeroinf-poisson
library(glmmTMB)
zinf_pois <- glmmTMB(
  family = "poisson",
  data = needles,
  formula = shared_syr ~ homeless +
    age + sex + ethn + polydrug + hivstat + sexabuse + hplsns,
  # the zero-inflation submodel uses the exposure, matching Vittinghoff's
  # `inflate(i.homeless)` (@vittinghoff2e, §8.3.1)
  ziformula = ~ homeless
)

zinf_pois |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

::: notes

Another R package for zero-inflated models is
[`pscl`](https://cran.r-project.org/web/packages/pscl/index.html) (@pscl08).

:::

{{< slidebreak >}}

To combine flexible dispersion modeling with zero-inflation,
we also fit a zero-inflated negative binomial (ZINB) model.

```{r}
#| tbl-cap: "Zero-inflated negative binomial model"
#| label: tbl-zeroinf-nb
zinf_nb <- glmmTMB(
  family = nbinom2,
  data = needles,
  formula = shared_syr ~ homeless +
    age + sex + ethn + polydrug + hivstat + sexabuse + hplsns,
  ziformula = ~ homeless
)

zinf_nb |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

::: notes
Both zero-inflated models keep the zero-inflation (structural-zero) submodel
parsimonious — just the exposure, `ziformula = ~ homeless` —
matching @vittinghoff2e's `inflate(i.homeless)` [§8.3.1].

A richer zero-inflation submodel is poorly identified here.
With the negative binomial in particular,
the overdispersion parameter and the zero-inflation component
both compete to explain the excess zeros,
so adding several covariates to the submodel makes the logistic fit
*separate*: its coefficients diverge and the exponentiated estimates
(odds ratios) run off to $\pm\infty$.
Keeping the submodel small avoids that.
:::
