```{r}
library(tidyverse)
library(haven)
needles <-
  here::here("inst/extdata/needle_sharing.dta") |>
  read_dta() |>
  as_tibble() |>
  mutate(
    hivstat = hivstat |>
      case_match(1 ~ "HIV+", 0 ~ "HIV-") |>
      factor() |>
      relevel(ref = "HIV-"),
    polydrug = polydrug |>
      case_match(1 ~ "multiple drugs used", 0 ~ "one drug used") |>
      factor() |>
      relevel(ref = "one drug used"),
    homeless = homeless |>
      case_match(1 ~ "homeless", 0 ~ "not homeless") |>
      factor() |>
      relevel(ref = "not homeless"),
    sexabuse = sexabuse |>
      case_match(1 ~ "yes", 0 ~ "no") |>
      factor() |>
      relevel(ref = "no"),
    ethn = ethn |>
      factor() |>
      forcats::fct_lump_min(min = 5, other_level = "Other") |>
      relevel(ref = "White"),
    sex = sex |> factor() |> relevel(ref = "M")
  ) |>
  labelled::set_variable_labels(
    "id" = "Participant identifier",
    "sex" = "Biological sex (reference = Male)",
    "ethn" = "Self-reported ethnicity (reference = White; n<5 lumped to Other)",
    "age" = "Age at first interview (years)",
    "dprsn_dx" = "Depression diagnosis indicator",
    "sexabuse" = "History of sexual abuse (1 = yes, 0 = no)",
    "shared_syr" = "Number of shared syringes (in 30 days)",
    "hivstat" = "HIV serostatus (reference = HIV-)",
    "hplsns" = "Hopelessness score",
    "nivdu" = "Number of injections (in 30 days)",
    "shsyryn" = "Shared syringe? (1 = yes, 0 = no)",
    "sqrtnivd" = "sqrt(No. of injections in 30 days)",
    "logshsyr" = "log(No. of shared needles)",
    "polydrug" = "Multiple non-injection drugs used?",
    "sqrtninj" = "sqrt(No. of injections in 30 days); duplicate of sqrtnivd",
    "homeless" = "Homeless? (reference = not homeless)",
    "shsyr" = "Shared needles (30 days); sparse duplicate of shared_syr"
  )
```

## Introduction

This example, from @vittinghoff2e [§8.3.1],
models *risky drug-use behavior* — needle sharing — among
injection drug users (IDUs),
a behavior of public-health interest because shared syringes can transmit
blood-borne infections.
We examine how needle-sharing frequency relates to participants' age, sex,
ethnicity, housing status (homelessness), drug-use patterns, HIV serostatus,
sexual-abuse history, and hopelessness score;
injection frequency reflects how often each participant has the
opportunity to share a syringe.

These data are a sample of injection drug users
distributed with @vittinghoff2e [§8.3.1]
via the
[UCSF companion website](https://regression.ucsf.edu/second-edition/data-examples-and-problems);
the original study is not documented in the textbook or the companion materials.
One possible (though unconfirmed) source is @tsui2009hcv,
a study of injection-drug-using participants in San Francisco
co-authored by one of the textbook's authors (E. Vittinghoff).
There are $n = 128$ participants and 17 variables.
Our outcome is `shared_syr`,
the number of times a participant shared a syringe in the past 30 days.

:::{#tbl-needles-data-dict}

```{r}
dict <- tibble(
  variable = names(needles),
  description = labelled::get_variable_labels(needles) |> 
    sapply(function(x) ifelse(is.null(x), "", x)),
)
dict |> pander::pander()
```

Data dictionary for `needles` data

:::

---

@tbl-needles-demographics summarizes the demographic and behavioural
variables for the participants.

:::{#tbl-needles-demographics}

```{r}
#| code-fold: true
needles |>
  dplyr::select(-id) |>
  gtsummary::tbl_summary()
```

Demographics and behavioural variables for the needle-sharing dataset
(all participants).

:::

{{< slidebreak >}}

Before modeling, we sketch a hypothesized causal structure for these
variables.

::: {#fig-needles-dag}

```{r}
#| fig-width: 8.5
#| fig-height: 6.5
#| code-fold: true

library(dagitty)
library(ggdag)
library(dplyr)

# This is a cross-sectional study: every covariate is measured at a single
# interview, so we cannot causally order the time-varying covariates among
# themselves. We draw their mutual associations -- from shared, unobserved
# earlier circumstances -- as non-directional (bidirected) edges, instead of
# giving each covariate its own separate latent "past state" node.
demographics <- c("age", "sex", "ethn") # fixed at (or determined by) birth
covariates <- c(
  "homeless", "polydrug", "hivstat", "sexabuse", "dprsn_dx", "hplsns"
)

# every unordered pair of time-varying covariates gets a bidirected edge
covariate_pairs <- utils::combn(covariates, 2)
bidirected_edges <- paste(covariate_pairs[1, ], "<->", covariate_pairs[2, ])
demographic_edges <- as.vector(
  outer(demographics, covariates, function(a, b) paste(a, "->", b))
)

needles_dag <- dagitty(paste0("dag {", paste(c(
  # birth-fixed demographics cause the current states, injection frequency,
  # and the outcome
  demographic_edges,
  paste(demographics, "-> nivdu"),
  paste(demographics, "-> shared_syr"),
  # time-varying covariates are mutually associated, with no causal ordering
  bidirected_edges,
  # each time-varying covariate can affect injection frequency and the outcome
  paste(covariates, "-> nivdu"),
  paste(covariates, "-> shared_syr"),
  # injecting is the only mediator: sharing a syringe requires injecting
  "nivdu -> shared_syr",
  # study inclusion is conditioned on; several covariates affect being sampled,
  # so conditioning on `selected` induces associations (selection bias)
  paste(c(demographics, "homeless", "polydrug", "hivstat"), "-> selected")
), collapse = "; "), "}"))

# layout: time flows left -> right (birth-fixed demographics -> current
# covariates -> injection frequency -> shared syringes)
y_cov <- seq(4, -4, length.out = length(covariates))
coordinates(needles_dag) <- list(
  x = c(
    age = -1, sex = -1, ethn = -1,
    setNames(rep(2, length(covariates)), covariates),
    nivdu = 5, shared_syr = 7, selected = 5
  ),
  y = c(
    age = 2, sex = 0, ethn = -2,
    setNames(y_cov, covariates),
    nivdu = 3.2, shared_syr = 0, selected = -5.5
  )
)

needles_tidy <- needles_dag |>
  tidy_dagitty() |>
  mutate(role = case_when(
    name == "homeless" ~ "exposure",
    name == "shared_syr" ~ "outcome",
    name == "nivdu" ~ "mediator",
    name == "selected" ~ "study inclusion (conditioned on)",
    name %in% demographics ~ "demographic (birth-fixed)",
    TRUE ~ "time-varying covariate (current/recent status)"
  ))

# split edges so associations and causal arrows can be styled separately
directed_edges <- function(d) dplyr::filter(d, .data$direction == "->")
association_edges <- function(d) dplyr::filter(d, .data$direction == "<->")

needles_tidy |>
  ggplot() +
  aes(x = x, y = y, xend = xend, yend = yend) +
  # non-directional associations among time-varying covariates: dashed grey arcs
  geom_dag_edges_arc(
    data = association_edges, curvature = 0.4, arrow = NULL,
    edge_colour = "grey65", edge_linetype = "dashed", edge_width = 0.4,
    start_cap = ggraph::circle(5, "mm"), end_cap = ggraph::circle(5, "mm")
  ) +
  # causal (directed) edges: solid arrows
  geom_dag_edges_link(
    data = directed_edges, edge_width = 0.35, edge_colour = "grey20",
    start_cap = ggraph::circle(5, "mm"), end_cap = ggraph::circle(5, "mm")
  ) +
  geom_dag_point(aes(colour = role, shape = role), size = 7) +
  geom_dag_text(colour = "black", size = 1.7) +
  scale_shape_manual(values = c(
    "exposure" = 16, "outcome" = 16, "mediator" = 16,
    "study inclusion (conditioned on)" = 15,
    "demographic (birth-fixed)" = 16,
    "time-varying covariate (current/recent status)" = 16
  )) +
  scale_colour_manual(values = c(
    "exposure" = "#d73027", "outcome" = "#4575b4", "mediator" = "#1a9850",
    "study inclusion (conditioned on)" = "grey30",
    "demographic (birth-fixed)" = "grey55",
    "time-varying covariate (current/recent status)" = "grey80"
  )) +
  guides(colour = guide_legend(nrow = 2), shape = guide_legend(nrow = 2)) +
  theme_dag() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

Hypothesized causal structure for the needle-sharing example,
drawn for a **cross-sectional** study.
The exposure of interest is homelessness (`homeless`, red)
and the outcome is the count of shared syringes (`shared_syr`, blue).
Injection frequency (`nivdu`, green) is the one genuine mediator:
sharing a syringe requires injecting.
Demographic variables (`age`, `sex`, `ethn`, darker grey) are fixed at
(or determined by) birth, so they sit upstream of every other variable,
with arrows into the time-varying covariates, injection frequency, and the outcome.
Because every other covariate is measured at the same single interview,
we cannot order the time-varying covariates (lighter grey) causally among themselves;
their mutual associations — from shared, unobserved earlier circumstances —
are drawn as dashed, non-directional edges.
Each time-varying covariate may still affect injection frequency and the 30-day
outcome, shown as solid arrows into `nivdu` and `shared_syr`.
The study-inclusion indicator (`selected`, square) is conditioned on by design;
because several covariates influence who is sampled,
conditioning on it can induce associations among them (selection bias).

:::

::: notes
We treat **homelessness as the exposure of interest**
and ask how it affects the number of shared syringes —
following @vittinghoff2e [§8.3.1], who model this outcome on `homeless`.

Injection frequency `nivdu` is a **mediator**:
a shared syringe is necessarily one of the injections,
so homelessness can raise sharing partly by raising how often a person injects.
Because `nivdu` lies *on* the causal path from `homeless` to `shared_syr`,
we do **not** adjust for it
(neither as a covariate nor as an offset);
doing so would block that path and estimate only a partial effect.
We want the **total** effect of homelessness on needle-sharing.

Because the study is **cross-sectional**,
the remaining covariates are measured at the same interview as the exposure,
so we cannot order them causally relative to `homeless` or to one another.
We draw their mutual associations as non-directional (bidirected) edges
and adjust for the measured covariates to control the confounding they carry.
A single fitted coefficient is therefore an **adjusted association**,
not an isolated causal effect.

Conditioning on study inclusion (`selected`) is unavoidable —
we only observe people who were sampled —
but because several covariates influence who is sampled,
it can induce **selection bias** that no covariate adjustment removes.
:::

```{r}
#| tbl-cap: "Needle-sharing data"
#| label: tbl-needle-data
needles
```

---

@fig-needles plots the shared-syringe counts against age,
split by the covariate subgroups (point size shows injection frequency).

```{r}
#| fig-cap: "Needle-sharing counts by age"
#| label: fig-needles
#| fig-width: 8.5
#| fig-height: 5
library(ggplot2)

needles |>
  ggplot() +
  aes(
    x = age,
    y = shared_syr,
    shape = sex,
    col = hivstat
  ) +
  geom_point(aes(size = nivdu), alpha = .5) +
  scale_size_area(max_size = 4) +
  scale_y_continuous(
    transform = "pseudo_log",
    breaks = c(0, 1, 3, 10, 30, 60)
  ) +
  facet_grid(cols = vars(polydrug), rows = vars(homeless)) +
  labs(y = "Number of shared syringes (30 days)") +
  theme(legend.position = "bottom")
```


{{< slidebreak >}}

@fig-needles-raw shows the underlying counts:
the number of shared syringes against the number of injections,
with a dashed $y = x$ line marking the ceiling
where every injection involves a shared syringe.

::: {#fig-needles-raw}

```{r}
#| code-fold: true
#| fig-width: 8.5
#| fig-height: 5
library(ggplot2)

needles |>
  ggplot() +
  aes(
    x = nivdu,
    y = shared_syr,
    shape = sex,
    col = hivstat
  ) +
  geom_point(alpha = .5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", alpha = .5) +
  # both axes are counts, so pseudo-log both: a shared transform keeps the
  # y = x reference line straight (data diagonal -> panel diagonal)
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 3, 10, 30, 100, 300, 900)
  ) +
  scale_y_continuous(
    transform = "pseudo_log",
    breaks = c(0, 1, 3, 10, 30, 60)
  ) +
  facet_grid(cols = vars(polydrug), rows = vars(homeless)) +
  labs(
    x = "Number of injections (in 30 days)",
    y = "Number of shared syringes (in 30 days)"
  ) +
  theme(legend.position = "bottom")
```

Number of shared syringes versus number of injections
(raw counts, 30-day window).
The dashed line marks $y = x$ (every injection involves a shared syringe).

:::

---

## Covariate counts

```{r}
#| tbl-cap: "Counts in `needles` by sex, housing status, and multiple drug use"
#| label: tbl-count-needles
#| code-fold: show
needles |>
  dplyr::select(sex, homeless, polydrug) |>
  summary()
```

---

There's only one individual with `sex = Trans`, 
which unfortunately isn't enough data to analyze.
We will remove that individual:

```{r}
#| label: remove-trans-obs
#| code-summary: 'remove singleton observation with sex == Trans'
needles <- needles |> filter(sex != "Trans")

```

---

## Model {.smaller}

Following @vittinghoff2e [§8.3.1],
we take **homelessness as the exposure of interest**
and ask how it affects the **count** of shared syringes,
using a Poisson regression (log link) rather than modeling a rate.
The remaining covariates enter as adjustments for confounding
(@fig-needles-dag): because the study is cross-sectional,
they are mutually associated with homelessness through shared, unobserved
earlier circumstances, which we cannot resolve into a causal order.

:::{#exr-nivdu-offset}
A Poisson *rate* model would add `offset(log(nivdu))`,
forcing the expected count to be proportional to injection frequency
(the coefficient of `log(nivdu)` fixed at exactly 1).
Why not use `nivdu` as an offset here?
:::

{{< slidebreak >}}

:::{#sol-nivdu-offset}
We avoid an offset here for two reasons:

- **`nivdu` is a mediator, not an exposure window.**
An offset is appropriate when its variable is a known denominator —
person-time or population at risk — that is external to the causal
question. Injection frequency is instead on the causal path from
homelessness to sharing (@fig-needles-dag): being homeless can raise
sharing partly by raising how often a person injects. Conditioning on it
(whether as an offset or as a covariate) would block that path and estimate
only a partial effect. We want the *total* effect of homelessness on
needle-sharing.
- **It imposes an untested proportionality assumption.**
Fixing the `log(nivdu)` coefficient at 1 assumes sharing scales
one-for-one with injecting, which the data need not support.
:::

Vittinghoff's analysis uses neither an offset nor `nivdu` as a covariate.

::: {.callout-tip title="Rule of thumb"}
An offset's exposure is usually something fixed by the *study design* —
person-time of follow-up, population at risk, area surveyed, number of
trials — a known denominator that is *exogenous* to the question
(not caused by the predictors of interest).
When the candidate denominator is instead something the participants
*did*, and the predictors also influence it (here, how often they inject),
it is a mediator rather than an exposure window, and we model the count
directly instead of dividing by it.
:::

We adjust for the measured covariates to control the confounding they carry
in @fig-needles-dag — demographic (`age`, `sex`, `ethn`),
social (`polydrug`), serostatus (`hivstat`), and trauma/mental-health
(`sexabuse`, hopelessness score `hplsns`) — using **main effects only**.
We exclude the mediator `nivdu` (above) and the depression indicator
`dprsn_dx`, whose `1`/`5` coding is undocumented in the source data and
so cannot be interpreted.

```{r}
glm1 <- glm(
  formula = shared_syr ~ homeless +
    age + sex + ethn + polydrug + hivstat + sexabuse + hplsns,
  family = stats::poisson,
  data = needles
)
library(equatiomatic)
equatiomatic::extract_eq(glm1)
```

---

```{r}
#| tbl-cap: "Poisson count model for needle-sharing data"
#| label: tbl-needles-pois


library(parameters)
glm1 |>
  parameters(exponentiate = TRUE) |>
  print_md()
```

---

```{r}
#| fig-cap: "Fitted Poisson model: expected sharing by age and housing status"
#| label: fig-needles-pois
#| code-fold: true
library(ggplot2)

# Predicted expected counts across age for the two homelessness groups,
# holding the other covariates at their reference levels (hplsns at its
# median). The model has no offset, so predict(type = "response") returns
# the expected number of shared syringes directly.
ref_lvl <- function(v) levels(glm1$model[[v]])[1]
pred_grid <- expand.grid(
  age = seq(min(glm1$model$age), max(glm1$model$age), length.out = 100),
  homeless = levels(glm1$model$homeless),
  sex = ref_lvl("sex"),
  ethn = ref_lvl("ethn"),
  polydrug = ref_lvl("polydrug"),
  hivstat = ref_lvl("hivstat"),
  sexabuse = ref_lvl("sexabuse"),
  hplsns = median(glm1$model$hplsns)
)
pred_grid$shared_syr <- predict(glm1, newdata = pred_grid, type = "response")

ggplot(glm1$model) +
  aes(x = age, y = shared_syr, colour = homeless) +
  geom_point(alpha = 0.4) +
  geom_line(data = pred_grid, linewidth = 1) +
  scale_y_continuous(
    transform = "pseudo_log",
    breaks = c(0, 1, 3, 10, 30, 60)
  ) +
  labs(
    y = "Number of shared syringes (30 days)",
    colour = "Housing status"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Interpreting the coefficients {.smaller}

For a Poisson model with a log link,
exponentiating a coefficient turns it into a *ratio of expected counts*
(an **incidence rate ratio**, IRR, in @vittinghoff2e's terminology):
$\exp{\b_j}$ is the multiplicative change in the expected number of
shared syringes per one-unit increase in $x_j$,
holding the other covariates fixed.
Our **focal estimate is the IRR for homelessness**;
the remaining coefficients are adjustments for confounding,
not effects of separate interest.
Because we do not condition on the mediator `nivdu`,
the homelessness IRR is a *total* effect.
The fitted ratios are reported in @tbl-needles-pois.

```{r}
#| code-fold: true
irr <- exp(coef(glm1))
```

**Baseline (intercept).**
$\exp{\hat\b_0} = `r round(irr[["(Intercept)"]], 3)`$
is the expected count of shared syringes for the reference participant
(male, not homeless, White, one drug used, HIV-negative,
no sexual-abuse history, hopelessness score 0), extrapolated to age 0.
Age 0 is outside the data, so the intercept is a mathematical anchor
rather than a directly interpretable quantity.

**Age.**
$\exp{\hat\b_{\text{age}}} = `r round(irr[["age"]], 3)`$ per year:
essentially no association between age and the expected count.

**Sex.**
$\exp{\hat\b_{\text{sexF}}} = `r round(irr[["sexF"]], 2)`$:
women's expected count of shared syringes is about
`r round(irr[["sexF"]], 2)` times men's,
adjusting for the other covariates.

**Homelessness (the focal exposure).**
$\exp{\hat\b_{\text{homeless}}} = `r round(irr[["homelesshomeless"]], 2)`$:
homeless participants share syringes about
`r round(irr[["homelesshomeless"]], 2)` times as often as the non-homeless,
adjusting for the other covariates.
@vittinghoff2e [§8.3.1] reports an *unadjusted* IRR of about 3.3
for homelessness in a single-predictor model;
adjusting for the confounder-proxies here moves the estimate
to about `r round(irr[["homelesshomeless"]], 1)`.

**Multiple-drug use.**
$\exp{\hat\b_{\text{polydrug}}} = `r round(irr[["polydrugmultiple drugs used"]], 2)`$:
multiple-drug users have a *lower* expected count than single-drug users,
though only `r sum(glm1$model[["polydrug"]] == "multiple drugs used")`
participants used multiple drugs, so this estimate is imprecise.

**HIV serostatus.**
$\exp{\hat\b_{\text{hivstatHIV+}}} = `r round(irr[["hivstatHIV+"]], 2)`$:
HIV-positive participants have a lower expected count of shared syringes
than HIV-negative participants, adjusting for the other covariates.
Only `r sum(glm1$model[["hivstat"]] == "HIV+")` participants are HIV+
in this sample, so this estimate is imprecise.

**Ethnicity.**
The `ethn` coefficients compare each retained group to the White reference
(small groups were lumped into "Other"); they adjust for any confounding
of the homelessness effect by ethnicity but are not of separate interest.

**Sexual-abuse history.**
$\exp{\hat\b_{\text{sexabuse}}} = `r round(irr[["sexabuseyes"]], 2)`$:
participants with a history of sexual abuse share syringes about
`r round(irr[["sexabuseyes"]], 2)` times as often as those without,
adjusting for the other covariates.

**Hopelessness.**
$\exp{\hat\b_{\text{hplsns}}} = `r round(irr[["hplsns"]], 2)`$ per point:
each one-point increase in the hopelessness score multiplies the expected
count by `r round(irr[["hplsns"]], 2)`,
adjusting for the other covariates.

::: {.callout-warning title="Overdispersion caveat"}
These data are highly overdispersed — the variance far exceeds the mean —
so the model-based Poisson standard errors are too small.
@vittinghoff2e [§8.3.1] recommends robust standard errors or a
negative-binomial model for this example.
:::

