---
title: "Schedule"
format:
  html: default
  pdf:
    output-file: schedule-handout.pdf
  docx:
    output-file: schedule-handout.docx

---

::: callout-note
This is the **tentative** schedule for the Spring 2026 offering of Epi 204.
It is generated directly from the course schedule spreadsheet
(`inst/extdata/epi-204-schedule-2026.xlsx`) and will be updated as the
quarter progresses. The authoritative, continuously-updated version is on
[Canvas](https://canvas.ucdavis.edu/courses/1077308/files/folder/Logistics).
:::

## Class sessions

::: {#tbl-schedule}

```{r}
#| code-fold: true

library(dplyr)

schedule_path <- here::here("inst/extdata/epi-204-schedule-2026.xlsx")

# Combine the per-textbook reading-reference columns into a single labeled
# string, dropping any that are blank for a given session.
combine_readings <- function(...) {
  values <- c(...)
  labels <- c("Dunn & Smyth", "Dobson", "Vittinghoff", "Notes")
  parts <- mapply(
    function(label, value) {
      if (is.na(value) || trimws(value) == "") {
        return(NA_character_)
      }
      paste0(label, " ", trimws(value))
    },
    labels, values,
    USE.NAMES = FALSE
  )
  parts <- parts[!is.na(parts)]
  if (length(parts) == 0) {
    return(NA_character_)
  }
  paste(parts, collapse = "; ")
}

class_schedule <-
  readxl::read_excel(schedule_path, sheet = "Schedule") |>
  # Drop footnote rows and the tentative "Skipped?" row (no real session date).
  filter(!is.na(Date)) |>
  mutate(Date = format(as.Date(Date), "%b %d")) |>
  rowwise() |>
  mutate(
    Readings = combine_readings(
      `Dunn & Smyth sections`, `Dobson sections`,
      `Vittinghoff sections`, `Lecture notes **`
    )
  ) |>
  ungroup() |>
  transmute(
    Week, Date, Day,
    Topic = Topics,
    Subtopics,
    Deadlines = `Deadlines*`,
    `New work` = `New Assignments`,
    Lecturer,
    Readings
  )

class_schedule |>
  gt::gt() |>
  gt::sub_missing(missing_text = "") |>
  gt::opt_row_striping() |>
  gt::tab_options(table.font.size = gt::pct(80)) |>
  gt::tab_footnote(
    footnote = paste(
      "Homework assignments have an automatic 24-hour grace period",
      "unless otherwise announced."
    ),
    locations = gt::cells_column_labels(columns = Deadlines)
  )
```

Tentative schedule of class sessions (Spring 2026).
:::

## Assignments and exams

::: {#tbl-assignments}

```{r}
#| code-fold: true

readxl::read_excel(schedule_path, sheet = "Homework and Exams") |>
  filter(!is.na(Assignments)) |>
  mutate(
    Topics = gsub("[\r\n]+", " ", Topics),
    across(
      c(`Date Assigned`, `Date Due*`, `Date Returned (approx.)`),
      ~ format(as.Date(.x), "%b %d")
    )
  ) |>
  rename(
    `Date due` = `Date Due*`,
    `Date returned (approx.)` = `Date Returned (approx.)`
  ) |>
  gt::gt() |>
  gt::sub_missing(missing_text = "") |>
  gt::opt_row_striping() |>
  gt::tab_options(table.font.size = gt::pct(85))
```

Homework assignments and exams (Spring 2026).
:::
