:::{#def-cond-pdf}

#### Conditional probability density function

Let $X$ and $Y$ be jointly distributed continuous random variables
with joint density $\p(X = x,\, Y = y)$ and marginal density $\p(X = x)$.
The **conditional probability density function** of $Y$ given $X = x$
(for values of $x$ with $\p(X = x) > 0$) is:

$$\p(Y = y \mid X = x) \eqdef \frac{\p(X = x,\, Y = y)}{\p(X = x)}$$

:::

---

:::{#exm-cond-pdf}

#### Conditional PDF from a bivariate normal model of birthweight data

The `birthweight` dataset in the `dobson` package
(@dobson4e, Table 2.3)
records gestational age (weeks) and birthweight (grams)
for 12 boys and 12 girls.
Let $X$ be gestational age and $Y$ be birthweight,
pooling both sexes into $n = 24$ observations.

```{r}
#| code-fold: true
#| message: false
library(dobson)
data(birthweight)
ga <- c(birthweight[["boys gestational age"]],
        birthweight[["girls gestational age"]])
wt <- c(birthweight[["boys weight"]], birthweight[["girls weight"]])
mu_x <- mean(ga)
sigma_x <- sd(ga)
mu_y <- mean(wt)
sigma_y <- sd(wt)
rho <- cor(ga, wt)
beta <- cov(ga, wt) / var(ga)
intercept <- mu_y - beta * mu_x
x0 <- 40
marg_dens_x0 <- dnorm(x0, mean = mu_x, sd = sigma_x)
cond_mean <- intercept + beta * x0
cond_sd <- sigma_y * sqrt(1 - rho^2)
pander::pander(data.frame(
  parameter = c("mu_x", "sigma_x", "mu_y", "sigma_y", "rho"),
  estimate = round(c(mu_x, sigma_x, mu_y, sigma_y, rho), 4)
))
```

Modeling $(X, Y)$ as bivariate normal
with parameters set equal to these sample moments,
the joint density is
(a standard result; e.g. @CaseBerg01):

$$
\p(X=x,\,Y=y) = \frac{1}{2\pi\sigma_X\sigma_Y\sqrt{1-\rho^2}}
\e{-\frac{1}{2(1-\rho^2)}
\sb{\frac{(x-\mu_X)^2}{\sigma_X^2}
- \frac{2\rho(x-\mu_X)(y-\mu_Y)}{\sigma_X\sigma_Y}
+ \frac{(y-\mu_Y)^2}{\sigma_Y^2}}}
$$

A further standard fact about the bivariate normal (@CaseBerg01)
is that the marginal distribution of $X$ is
$X \sim \ndist{\mu_X, \sigma_X^2}$.
At $x = `r x0`$ weeks (a full-term pregnancy),
$\mu_X = `r round(mu_x, 4)`$ and $\sigma_X = `r round(sigma_x, 4)`$, so:

$$
\ba
\p(X=`r x0`) &= \frac{1}{\sigma_X\sqrt{2\pi}}
  \e{-\frac{(`r x0`-\mu_X)^2}{2\sigma_X^2}}
\\&= \frac{1}{`r round(sigma_x, 4)`\sqrt{2\pi}}
  \e{-\frac{(`r x0`-`r round(mu_x, 4)`)^2}{2(`r round(sigma_x, 4)`)^2}}
\\&\approx `r round(marg_dens_x0, 4)`
\ea
$$

By @def-cond-pdf, dividing the joint density by this marginal density
and simplifying the exponent
(completing the square in $y$; @CaseBerg01)
gives the conditional PDF of $Y$ given $X = `r x0`$,
which is itself normal
with mean shifted along the regression line
and variance reduced by a factor of $1-\rho^2$:

$$
\ba
\p(Y=y \mid X=`r x0`) &= \frac{\p(X=`r x0`,\,Y=y)}{\p(X=`r x0`)}
\\&= \frac{1}{\sigma_Y\sqrt{2\pi(1-\rho^2)}}
  \e{-\frac{\paren{y - \sb{\mu_Y + \rho\frac{\sigma_Y}{\sigma_X}(`r x0`-\mu_X)}}^2}{2\sigma_Y^2(1-\rho^2)}}
\ea
$$

So $Y \mid X = `r x0` \sim \ndist{`r round(cond_mean, 2)`,\, `r round(cond_sd, 2)`^2}$:
the conditional mean, `r round(cond_mean, 2)` g,
is exactly the fitted regression line's prediction at $x=`r x0`$
($`r round(intercept, 4)` + `r round(beta, 4)` \times `r x0` = `r round(cond_mean, 2)`$),
matching `R`'s `lm(wt ~ ga)` fit directly:

```{r}
#| code-fold: true
pander::pander(coef(lm(wt ~ ga)))
```

:::
