:::{#def-cond-mixed}

#### Conditional expectation: mixed case

Suppose exactly one of $X, Y$ is discrete and the other is continuous.
Write $\p(X = x,\, Y = y)$ for their **joint density-mass function**:
a probability density in the continuous variable
and a probability mass in the discrete variable.

**$X$ discrete, $Y$ continuous.**
Here $\p(X=x,\,Y=y)$ is, for each fixed $x$,
a probability density in $y$,
scaled so that $\int_{y} \p(X=x,\,Y=y)\,dy = \P(X=x)$.
The conditional PDF of $Y$ given $X = x$
(for values of $x$ with $\P(X = x) > 0$) is:

$$\p(Y = y \mid X = x) \eqdef \frac{\p(X = x,\, Y = y)}{\P(X = x)}$$

and the conditional expectation of $Y$ given $X = x$ is:

$$\E{Y \mid X = x} \eqdef \int_{y \in \rangef{Y}} y \cd \p(Y = y \mid X = x)\, dy$$

**$X$ continuous, $Y$ discrete.**
Here $\p(X=x,\,Y=y)$ is, for each fixed $y$,
a probability density in $x$,
scaled so that $\sum_{y} \p(X=x,\,Y=y) = \p(X=x)$.
The conditional PMF of $Y$ given $X = x$
(for values of $x$ with $\p(X = x) > 0$) is:

$$\P(Y = y \mid X = x) \eqdef \frac{\p(X = x,\, Y = y)}{\p(X = x)}$$

and the conditional expectation of $Y$ given $X = x$ is:

$$\E{Y \mid X = x} \eqdef \sum_{y \in \rangef{Y}} y \cd \P(Y = y \mid X = x)$$

:::

---

:::{#exm-cond-mixed}

#### Conditional expectation, one discrete variable and one continuous variable

**$X$ discrete, $Y$ continuous.**
The `plasma` dataset in the `dobson` package
(@dobson4e, Table 6.25)
records plasma inorganic phosphate levels (mg/dL)
one hour after a glucose tolerance test,
for hyperinsulinemic obese (`H-O`) and control (`C`) participants.
Let $X$ be group and $Y$ be phosphate level.

```{r}
#| code-fold: true
#| message: false
library(dobson)
library(dplyr)
data(plasma)
plasma_subset <- plasma |> filter(Group %in% c("H-O", "C"))
plasma_summary <- plasma_subset |>
  summarise(n = n(), mean = mean(phosphate), sd = sd(phosphate), .by = Group)
n_ho <- plasma_summary$n[plasma_summary$Group == "H-O"]
n_c <- plasma_summary$n[plasma_summary$Group == "C"]
mean_ho <- plasma_summary$mean[plasma_summary$Group == "H-O"]
sd_ho <- plasma_summary$sd[plasma_summary$Group == "H-O"]
mean_c <- plasma_summary$mean[plasma_summary$Group == "C"]
sd_c <- plasma_summary$sd[plasma_summary$Group == "C"]
pander::pander(plasma_summary)
```

Modeling phosphate as approximately normal within each group,
with parameters set equal to each group's sample mean and SD,
and $\P(X = \text{H-O}) = `r n_ho`/`r n_ho + n_c`$:

$$\p(Y = y \mid X = \text{H-O}) \approx \ndist{`r round(mean_ho, 2)`,\, `r round(sd_ho, 2)`^2}$$
$$\p(Y = y \mid X = \text{C}) \approx \ndist{`r round(mean_c, 2)`,\, `r round(sd_c, 2)`^2}$$

By @def-cond-mixed, since the mean of a normal random variable
is its own location parameter:

$$\E{Y \mid X = \text{H-O}} = `r round(mean_ho, 2)` \text{ mg/dL}, \qquad
  \E{Y \mid X = \text{C}} = `r round(mean_c, 2)` \text{ mg/dL}$$

**$X$ continuous, $Y$ discrete.**
The `senility` dataset in the `dobson` package
(@dobson4e, Table 7.8)
records, for 54 elderly people,
a WAIS (Wechsler Adult Intelligence Scale) score
and whether symptoms of senility were present.
Let $X$ be WAIS score and $Y$ indicate senility symptoms.

```{r}
#| code-fold: true
#| message: false
data(senility)
senility_fit <- glm(s ~ x, data = senility, family = binomial)
b0 <- coef(senility_fit)[["(Intercept)"]]
b1 <- coef(senility_fit)[["x"]]
pander::pander(coef(senility_fit))
```

Modeling $\P(Y = 1 \mid X = x)$ with the fitted logistic regression:

$$\logit \P(Y = 1 \mid X = x) = `r sprintf("%.4f", b0)` - `r sprintf("%.4f", abs(b1))`\, x$$

By @def-cond-mixed, since $Y \mid X=x$ is Bernoulli:

$$
\ba
\E{Y \mid X = x} &= 0 \cd \P(Y=0 \mid X=x) + 1 \cd \P(Y=1 \mid X=x)
\\&= \P(Y=1 \mid X=x)
\\&= \frac{1}{1 + \e{-(`r sprintf("%.4f", b0)` - `r sprintf("%.4f", abs(b1))`\, x)}}
\ea
$$

```{r}
#| code-fold: true
senility_p10 <- predict(senility_fit,
                        newdata = data.frame(x = 10), type = "response")
senility_p15 <- predict(senility_fit,
                        newdata = data.frame(x = 15), type = "response")
```

At $x = 10$: $\E{Y \mid X=10} = `r round(senility_p10, 2)`$.
At $x = 15$: $\E{Y \mid X=15} = `r round(senility_p15, 2)`$ —
a lower WAIS score is associated with higher predicted probability
of senility symptoms.

:::
