When the posterior distribution has a known closed form,
we can sample from it directly and compute summaries analytically.
In most real-world models, however,
the normalizing constant $p(\vy)$ is intractable,
making direct sampling and closed-form summaries impossible.
**Markov Chain Monte Carlo (MCMC)** methods provide a practical solution:
instead of sampling independent draws from the exact posterior,
they generate a *correlated* sequence of draws whose long-run distribution
converges to the posterior [@dobson4e, Chapter 13, p. 287].

## Why the normalizing constant is computationally challenging

The posterior is $p(\vth \mid \vy) \propto p(\vy \mid \vth)\, p(\vth)$.
The normalizing constant is

$$p(\vy) = \int p(\vy \mid \vth)\, p(\vth)\, d\vth,$$

a high-dimensional integral that typically has no closed form.
Computing it numerically becomes infeasible as the dimension of $\vth$ grows
[@dobson4e, Chapter 13, p. 287].

MCMC sidesteps this problem:
many algorithms require only the *ratio*
$p(\vth^* \mid \vy) / p(\vth \mid \vy)$,
in which the normalizing constant cancels.

## Monte Carlo integration

Before introducing Markov chains,
it helps to understand **Monte Carlo integration**
[@dobson4e, Chapter 13, p. 290].
If we could draw $M$ independent samples
$\vth^{(1)}, \ldots, \vth^{(M)} \siid p(\vth \mid \vy)$,
we could approximate any posterior expectation:

$$\E{g(\vth) \mid \vy} \approx \frac{1}{M} \sum_{m=1}^{M} g\!\left(\vth^{(m)}\right).$$

In particular, the posterior mean,
posterior variance,
and credible-interval endpoints
are all special cases of this formula.

:::{#exm-mc-bernoulli}
#### Monte Carlo posterior summaries for a Bernoulli model

Recall the Beta-Bernoulli model with
$n = 91$ observations and $r = 55$ successes.
The posterior is $\pi \mid \vy \sim \operatorname{Beta}(56, 37)$.
We draw $M = 5{,}000$ samples and compare Monte Carlo summaries with exact values.

```{r}
#| code-fold: show
#| label: exm-mc-bernoulli-code
set.seed(42)
n_obs <- 91L
r_obs <- 55L
a_post <- 1L + r_obs
b_post <- 1L + n_obs - r_obs
M <- 5000L
pi_mc <- rbeta(n = M, shape1 = a_post, shape2 = b_post)
round(c(
  exact_mean  = a_post / (a_post + b_post),
  mc_mean     = mean(pi_mc),
  exact_lower = qbeta(0.025, a_post, b_post),
  mc_lower    = unname(quantile(pi_mc, 0.025)),
  exact_upper = qbeta(0.975, a_post, b_post),
  mc_upper    = unname(quantile(pi_mc, 0.975))
), 4)
```
:::

## Markov chains

A **Markov chain** is a sequence of random variables
$\vth^{(1)}, \vth^{(2)}, \ldots$ such that
the distribution of $\vth^{(t+1)}$ depends only on $\vth^{(t)}$
(the *Markov property*).

:::{#def-markov-chain}
#### Markov chain

A sequence of random variables $\vth^{(1)}, \vth^{(2)}, \ldots$
is a **Markov chain** if, for all $t$,

$$p(\vth^{(t+1)} \mid \vth^{(t)}, \vth^{(t-1)}, \ldots, \vth^{(1)}) =
p(\vth^{(t+1)} \mid \vth^{(t)}).$$
:::

Under mild conditions,
a Markov chain converges to a unique **stationary distribution**.
MCMC algorithms are designed so that the stationary distribution
is the target posterior $p(\vth \mid \vy)$
[@dobson4e, Chapter 13, p. 291].

## The Metropolis--Hastings algorithm

The **Metropolis--Hastings (MH) algorithm** is the most general MCMC method
[@dobson4e, Chapter 13, p. 293].
At each iteration $t$:

1. Propose a new value $\vth^*$ from a **proposal distribution** $q(\vth^* \mid \vth^{(t)})$.
2. Compute the **acceptance ratio**

$$\alpha = \frac{p(\vy \mid \vth^*)\, p(\vth^*)}{p(\vy \mid \vth^{(t)})\, p(\vth^{(t)})}
\cdot \frac{q(\vth^{(t)} \mid \vth^*)}{q(\vth^* \mid \vth^{(t)})}.$$

3. Accept the proposal with probability $\min(1, \alpha)$:
   if accepted, set $\vth^{(t+1)} = \vth^*$.
   otherwise set $\vth^{(t+1)} = \vth^{(t)}$.

Because the normalizing constant $p(\vy)$ appears in both numerator and denominator,
it cancels in the ratio,
so only the unnormalized posterior is needed.

:::{#def-mh}
#### Metropolis--Hastings algorithm

Given a target posterior $p(\vth \mid \vy) \propto p(\vy \mid \vth)\, p(\vth)$
and a proposal distribution $q(\cdot \mid \vth)$,
one step of the **MH algorithm** proceeds as:

1. Draw $\vth^* \sim q(\cdot \mid \vth^{(t)})$.
2. Compute $\alpha = \min\!\left(1,\,
   \frac{p(\vy \mid \vth^*)\, p(\vth^*)}{p(\vy \mid \vth^{(t)})\, p(\vth^{(t)})}
   \cdot \frac{q(\vth^{(t)} \mid \vth^*)}{q(\vth^* \mid \vth^{(t)})}\right)$.
3. Set $\vth^{(t+1)} = \vth^*$ with probability $\alpha$,
   else $\vth^{(t+1)} = \vth^{(t)}$.
:::

:::{#exm-mh-bernoulli}
#### Metropolis--Hastings for a Bernoulli model

We implement the MH algorithm for the same Beta-Bernoulli model
as in @exm-mc-bernoulli.
We use a normal random-walk proposal $\pi^* = \pi^{(t)} + \varepsilon$,
$\varepsilon \sim \ndist{0, \sigma^2}$.

```{r}
#| code-fold: show
#| label: exm-mh-bernoulli-code
log_post_unnorm <- function(pi, r, n) {
  if (pi <= 0 || pi >= 1) return(-Inf)
  r * log(pi) + (n - r) * log(1 - pi)
}

mh_bernoulli <- function(n_iter, r, n, pi_init = 0.5, sigma = 0.05) {
  chain <- numeric(n_iter)
  chain[1] <- pi_init
  for (i in seq(2, n_iter)) {
    pi_curr <- chain[i - 1]
    pi_prop <- pi_curr + rnorm(1, sd = sigma)
    log_alpha <- log_post_unnorm(pi_prop, r, n) -
      log_post_unnorm(pi_curr, r, n)
    chain[i] <- if (log(runif(1)) < log_alpha) pi_prop else pi_curr
  }
  chain
}

set.seed(1)
mh_chain <- mh_bernoulli(n_iter = 5000L, r = 55L, n = 91L)
```
:::

## The Gibbs sampler

The **Gibbs sampler** is a special case of the MH algorithm
for multiparameter models
[@dobson4e, Chapter 13, p. 300].
When $\vth = (\theta_1, \ldots, \theta_K)$,
each component is updated in turn by sampling from its
**full conditional distribution**
$p(\theta_k \mid \theta_{-k}, \vy)$,
where $\theta_{-k}$ denotes all components except $\theta_k$.
The resulting proposal is always accepted (acceptance probability 1).
Gibbs sampling requires that the full conditionals are available in closed form.
JAGS (Just Another Gibbs Sampler) automates this approach for a wide class of models.

## Burn-in {#sec-burnin}

:::{#def-burnin}
#### burn-in

The **burn-in** period is the initial segment of an MCMC chain
that is discarded before posterior summaries are computed.
During burn-in, the chain is moving toward the stationary distribution
and does not yet represent draws from the posterior
[@dobson4e, Chapter 13, p. 301].
:::

The length of the burn-in period is chosen by inspection of trace plots.
A common default is to discard the first 10--20% of iterations.

## Convergence diagnostics {#sec-mcmc-diagnostics}

Because MCMC chains are correlated and may take time to reach stationarity,
we must assess whether the chain has converged before using it for inference
[@dobson4e, Chapter 13, p. 302].

### Trace plots

A **trace plot** graphs the sampled value $\vth^{(t)}$ against iteration $t$.
A well-mixed chain shows rapid oscillation around a stable mean,
with no obvious trends or stuck regions.

:::{#exm-trace-plot}
#### Trace plot for MH chain

:::{#fig-trace-mh}
```{r}
#| code-fold: show
burnin <- 500L
df_chain <- data.frame(iteration = seq_along(mh_chain), pi = mh_chain)
ggplot(df_chain, aes(x = iteration, y = pi)) +
  geom_line(linewidth = 0.2, alpha = 0.7) +
  geom_vline(xintercept = burnin, linetype = "dashed", colour = "red") +
  labs(x = "Iteration", y = expression(pi)) +
  theme_bw()
```
MH chain trace plot; dashed line marks end of burn-in.
:::
:::

### Autocorrelation function (ACF)

Because successive draws in an MCMC chain are correlated,
the **effective sample size** (ESS) is less than the number of iterations.
The autocorrelation function (ACF) quantifies this dependence.
A rapidly decaying ACF indicates good mixing.

### Multiple chains and the Gelman--Rubin statistic

Running multiple chains from different starting points
provides a stronger convergence check than a single chain.
The **Gelman--Rubin statistic** $\hat{R}$
compares within-chain and between-chain variance.
$\hat{R} \approx 1$ indicates convergence
[@dobson4e, Chapter 13, p. 304].
A common rule of thumb is $\hat{R} < 1.1$.

## Posterior inference from MCMC samples

After discarding burn-in,
the remaining chain $\vth^{(B+1)}, \ldots, \vth^{(T)}$
is treated as a (correlated) sample from the posterior.
Posterior summaries are computed as sample statistics:

- **Posterior mean**: $\hmu = \frac{1}{T - B} \sum_{t=B+1}^{T} \vth^{(t)}$
- **Credible interval**: sample quantiles of the chain after burn-in

[@dobson4e, Chapter 13, p. 305].

:::{#exm-post-summary-mh}
#### Posterior summaries from MH chain

Using the chain from @exm-mh-bernoulli,
we discard the first 500 iterations as burn-in
and compute posterior summaries.

```{r}
#| code-fold: show
#| label: exm-post-summary-mh-code
burnin <- 500L
post_chain <- mh_chain[seq(burnin + 1L, length(mh_chain))]
round(c(
  mean  = mean(post_chain),
  lower = unname(quantile(post_chain, 0.025)),
  upper = unname(quantile(post_chain, 0.975))
), 4)
```
:::