::: notes
Applied exercises adapted from @james2021islr2e [§7.9] and
@hastie2009esl2e [§9.1].
:::

{{< slidebreak >}}

### Polynomial regression and step functions (ISLRv2 Applied 7) {#sec-exr-wage-poly}

<!-- Reviewer note (intentional, do not re-flag): no slidebreak between this
heading and the div below. The preceding slidebreak already opens a fresh
slide; the heading and this div intentionally share that slide to avoid a
title-only slide. -->
::::{#exr-wage-polynomial-cv}

Using the `Wage` dataset from the `ISLR2` package [@james2021islr2e]:

**(a)** Use 10-fold cross-validation to choose the polynomial degree $d^*$ for
predicting `wage` from `age`. Plot the CV errors for degrees 1 through 10.

**(b)** Fit the degree-$d^*$ polynomial and plot it with a 95% confidence band.

**(c)** Separately, choose the number of cut-points $K^*$ for a step-function
model using cross-validation. Plot the chosen step-function fit.

::::


{{< slidebreak >}}

::::{#sol-wage-polynomial-cv}

::: {.panel-tabset}

#### R

```{r}
#| code-fold: true
#| eval: false
library(ISLR2)
library(boot)
library(ggplot2)
set.seed(1)
cv_poly <- vapply(1:10, function(d) {
  cv.glm(Wage, glm(wage ~ poly(age, d), data = Wage), K = 10)$delta[1]
}, numeric(1))
best_d <- which.min(cv_poly)
cat("CV-selected degree:", best_d, "\n")
plot(1:10, cv_poly, type = "b", xlab = "Degree", ylab = "CV MSE",
     main = "Polynomial degree selection")
points(best_d, cv_poly[best_d], col = "red", pch = 19)
```

#### Python

```{python}
#| eval: false
#| python.reticulate: false
import numpy as np
from sklearn.preprocessing import PolynomialFeatures
from sklearn.linear_model import LinearRegression
from sklearn.pipeline import Pipeline
from sklearn.model_selection import cross_val_score
from ISLP import load_data
Wage = load_data("Wage")
X = Wage["age"].values.reshape(-1, 1)
y = Wage["wage"].values
cv_errs = [-cross_val_score(
    Pipeline([("poly", PolynomialFeatures(d, include_bias=False)),
              ("reg", LinearRegression())]),
    X, y, cv=10, scoring="neg_mean_squared_error").mean() for d in range(1, 11)]
best_d = int(np.argmin(cv_errs)) + 1
print(f"CV-selected degree: {best_d}")
```

:::

::::


{{< slidebreak >}}

### Backfitting algorithm (ESL §9.1.1 / ISLRv2 Applied 11) {#sec-exr-backfitting}

<!-- Reviewer note (intentional, do not re-flag): no slidebreak between this
heading and the div below. The preceding slidebreak already opens a fresh
slide; the heading and this div intentionally share that slide to avoid a
title-only slide. -->
::::{#exr-backfitting}

The **backfitting algorithm** [@hastie2009esl2e, §9.1.1] fits an additive model
$Y = \a + \sum_j f_j(X_j) + \eps$ by iteratively smoothing partial residuals.

For $p=2$ predictors with linear component smoothers $\hat f_j(X_j) = \hat\b_j X_j$:

1. Initialise: $\hat\a = \bar{y}$, $\hat f_1 = \hat f_2 = 0$.
2. Repeat until convergence (Gauss–Seidel order — re-centre each
   $\hat f_j$ immediately, so the centred $\hat f_1$ is used when
   computing $r_2$):
   a. $r_1 \leftarrow y - \hat\a - \hat f_2(x_2)$; regress $r_1$ on
      $x_1$ to update $\hat f_1$; $\hat f_1 \leftarrow \hat f_1 -
      \overline{\hat f_1}$.
   b. $r_2 \leftarrow y - \hat\a - \hat f_1(x_1)$; regress $r_2$ on
      $x_2$ to update $\hat f_2$; $\hat f_2 \leftarrow \hat f_2 -
      \overline{\hat f_2}$.

**(a)** Implement this on simulated data with known true coefficients.

**(b)** Verify the coefficients match `lm(y ~ x1 + x2)`.

::::


{{< slidebreak >}}

::::{#sol-backfitting}

::: {.panel-tabset}

#### R

```{r}
#| code-fold: true
#| eval: false
set.seed(1)
n <- 100
x1 <- rnorm(n)
x2 <- rnorm(n)
y <- 2 + 3 * x1 - 1.5 * x2 + rnorm(n, sd = 0.5)
alpha_hat <- mean(y)
f1 <- f2 <- rep(0, n)
for (iter in seq_len(500)) {
  r1 <- y - alpha_hat - f2
  b1 <- coef(lm(r1 ~ x1))[2]
  f1 <- b1 * x1
  f1 <- f1 - mean(f1)
  r2 <- y - alpha_hat - f1
  b2 <- coef(lm(r2 ~ x2))[2]
  f2 <- b2 * x2
  f2 <- f2 - mean(f2)
}
lm_fit <- lm(y ~ x1 + x2)
cat(sprintf("Backfitting: b1=%.4f, b2=%.4f\n", b1, b2))
cat(sprintf("lm():        b1=%.4f, b2=%.4f\n",
            coef(lm_fit)[2], coef(lm_fit)[3]))
```

#### Python

```{python}
#| eval: false
#| python.reticulate: false
import numpy as np
rng = np.random.default_rng(1)
n = 100; x1 = rng.standard_normal(n); x2 = rng.standard_normal(n)
y = 2 + 3*x1 - 1.5*x2 + rng.normal(0, 0.5, n)
alpha_hat = y.mean(); f1 = np.zeros(n); f2 = np.zeros(n)
for _ in range(500):
    r1 = y - alpha_hat - f2
    b1 = float(np.cov(x1, r1)[0, 1] / np.var(x1, ddof=1))
    f1 = b1*x1 - (b1*x1).mean()
    r2 = y - alpha_hat - f1
    b2 = float(np.cov(x2, r2)[0, 1] / np.var(x2, ddof=1))
    f2 = b2*x2 - (b2*x2).mean()
X_ols = np.column_stack([np.ones(n), x1, x2])
b_ols = np.linalg.lstsq(X_ols, y, rcond=None)[0]
print(f"Backfitting: b1={b1:.4f}, b2={b2:.4f}")
print(f"OLS:         b1={b_ols[1]:.4f}, b2={b_ols[2]:.4f}")
```

:::

The coefficients agree, confirming backfitting converges to OLS when
all component smoothers are linear [@hastie2009esl2e, §9.1.1].

::::
