::: notes
Lab adapted from @james2021islr2e [§7.8] and @james2023islp [§7.8].
:::

:::{.panel-tabset}
#### R
```{r}
#| label: lab-setup-r
#| eval: false
#| message: false
library(ISLR2)
library(splines)
library(mgcv)
library(ggplot2)
library(boot)
data(Wage)
head(Wage[, c("age", "wage", "education", "year")])
```

#### Python
The smoothing-spline and GAM sections also require `pygam`, which is
not part of the standard ISLP/sklearn ecosystem — install with
`pip install pygam` before running those blocks.

```{python}
#| eval: false
#| python.reticulate: false
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from ISLP import load_data
from sklearn.preprocessing import PolynomialFeatures
from sklearn.linear_model import LinearRegression, LogisticRegression
import statsmodels.api as sm
from patsy import dmatrix
Wage = load_data("Wage")
print(Wage[["age","wage","education","year"]].head())
```
:::

## Polynomial regression {.unnumbered}

Fit degree-1 through degree-5 polynomials, then use ANOVA to select degree [@james2021islr2e, §7.1].

:::{.panel-tabset}
#### R
```{r}
#| label: lab-poly-r
#| eval: false
#| code-fold: true
fit1 <- lm(wage ~ poly(age, 1), data = Wage)
fit2 <- lm(wage ~ poly(age, 2), data = Wage)
fit3 <- lm(wage ~ poly(age, 3), data = Wage)
fit4 <- lm(wage ~ poly(age, 4), data = Wage)
fit5 <- lm(wage ~ poly(age, 5), data = Wage)
anova(fit1, fit2, fit3, fit4, fit5)
```

The ANOVA p-values suggest degree 3 or 4 is adequate.

```{r}
#| label: lab-poly-plot-r
#| eval: false
#| code-fold: true
age_grid <- seq(min(Wage$age), max(Wage$age), length.out = 200)
pred <- predict(fit4, newdata = data.frame(age = age_grid), se.fit = TRUE)
df_pred <- data.frame(
  age = age_grid,
  fit = pred$fit,
  lo  = pred$fit - 2 * pred$se.fit,
  hi  = pred$fit + 2 * pred$se.fit
)
ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_ribbon(data = df_pred, aes(y = fit, ymin = lo, ymax = hi),
              alpha = 0.3, fill = "steelblue") +
  geom_line(data = df_pred, aes(y = fit), colour = "steelblue", linewidth = 1) +
  labs(title = "Degree-4 polynomial fit", x = "Age", y = "Wage")
```
#### Python
```{python}
#| eval: false
#| python.reticulate: false
from scipy import stats
results = {}
for deg in range(1, 6):
    X = np.column_stack([Wage["age"] ** k for k in range(1, deg + 1)])
    X = sm.add_constant(X)
    results[deg] = sm.OLS(Wage["wage"], X).fit()

# F-tests between consecutive nested models. compare_f_test() is
# called on the *unrestricted* (more flexible) model and given the
# restricted model as the argument.
for d in range(1, 5):
    f, p, df_diff = results[d + 1].compare_f_test(results[d])
    print(f"Degree {d} vs {d+1}: F={f:.3f}, p={p:.4f}")
```

```{python}
#| eval: false
#| python.reticulate: false
age_grid = np.linspace(Wage["age"].min(), Wage["age"].max(), 200)
X_grid = sm.add_constant(np.column_stack([age_grid ** k for k in range(1, 5)]))
pred4 = results[4].get_prediction(X_grid)
pred_df = pred4.summary_frame(alpha=0.05)

plt.figure(figsize=(7, 4))
plt.scatter(Wage["age"], Wage["wage"], alpha=0.15, s=8)
plt.plot(age_grid, pred_df["mean"], color="steelblue", linewidth=2)
plt.fill_between(age_grid, pred_df["mean_ci_lower"], pred_df["mean_ci_upper"],
                 alpha=0.3, color="steelblue")
plt.title("Degree-4 polynomial fit"); plt.xlabel("Age"); plt.ylabel("Wage")
plt.tight_layout(); plt.show()
```
:::

## Step functions {.unnumbered}

Cut `age` into four intervals and fit a piecewise-constant model [@james2021islr2e, §7.2].

:::{.panel-tabset}
#### R
```{r}
#| label: lab-step-r
#| eval: false
#| code-fold: true
Wage$age_cut <- cut(Wage$age, 4)
fit_step <- lm(wage ~ age_cut, data = Wage)
summary(fit_step)$coef
```
#### Python
```{python}
#| eval: false
#| python.reticulate: false
Wage["age_cut"] = pd.cut(Wage["age"], bins=4)
step_dummies = pd.get_dummies(Wage["age_cut"], drop_first=True).astype(float)
X_step = sm.add_constant(step_dummies)
fit_step = sm.OLS(Wage["wage"], X_step).fit()
print(fit_step.summary().tables[1])
```
:::

## Regression splines {.unnumbered}

Fit a cubic B-spline with knots at ages 25, 40, and 60, and a natural cubic spline with the same knots [@james2021islr2e, §7.4].

:::{.panel-tabset}
#### R
```{r}
#| label: lab-splines-r
#| eval: false
#| code-fold: true
# Redefine age_grid so the splines section runs in isolation
# (it's also defined in the poly-plot chunk above).
age_grid <- seq(min(Wage$age), max(Wage$age), length.out = 200)
knots <- c(25, 40, 60)
fit_bs <- lm(wage ~ bs(age, knots = knots), data = Wage)
fit_ns <- lm(wage ~ ns(age, knots = knots), data = Wage)

pred_bs <- predict(fit_bs, newdata = data.frame(age = age_grid), se.fit = TRUE)
pred_ns <- predict(fit_ns, newdata = data.frame(age = age_grid), se.fit = TRUE)

df_sp <- data.frame(
  age = rep(age_grid, 2),
  fit = c(pred_bs$fit, pred_ns$fit),
  lo  = c(pred_bs$fit - 2*pred_bs$se.fit, pred_ns$fit - 2*pred_ns$se.fit),
  hi  = c(pred_bs$fit + 2*pred_bs$se.fit, pred_ns$fit + 2*pred_ns$se.fit),
  type = rep(c("B-spline", "Natural spline"), each = 200)
)
ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_ribbon(data = df_sp, aes(y = fit, ymin = lo, ymax = hi, fill = type), alpha = 0.25) +
  geom_line(data = df_sp, aes(y = fit, colour = type), linewidth = 1) +
  labs(x = "Age", y = "Wage", colour = NULL, fill = NULL)
```
#### Python
```{python}
#| eval: false
#| python.reticulate: false
bs_mat = dmatrix("bs(age, knots=(25,40,60), include_intercept=False)",
                 {"age": Wage["age"]}, return_type="dataframe")
ns_mat = dmatrix("cr(age, knots=(25,40,60))",
                 {"age": Wage["age"]}, return_type="dataframe")

fit_bs_py = sm.OLS(Wage["wage"], bs_mat).fit()
fit_ns_py = sm.OLS(Wage["wage"], ns_mat).fit()

bs_grid = dmatrix("bs(age, knots=(25,40,60), include_intercept=False)",
                  {"age": age_grid}, return_type="dataframe")
ns_grid = dmatrix("cr(age, knots=(25,40,60))",
                  {"age": age_grid}, return_type="dataframe")

plt.figure(figsize=(7, 4))
plt.scatter(Wage["age"], Wage["wage"], alpha=0.15, s=6)
plt.plot(age_grid, fit_bs_py.predict(bs_grid), label="B-spline", linewidth=2)
plt.plot(age_grid, fit_ns_py.predict(ns_grid), label="Natural spline", linewidth=2)
plt.legend(); plt.xlabel("Age"); plt.ylabel("Wage")
plt.tight_layout(); plt.show()
```
:::

## Smoothing splines and local regression {.unnumbered}

A smoothing spline selects the penalty parameter $\l$ by generalized cross-validation (GCV); LOESS uses a local polynomial fit [@james2021islr2e, §7.5–7.6].

:::{.panel-tabset}
#### R
```{r}
#| label: lab-smooth-r
#| eval: false
#| code-fold: true
fit_ss_16 <- smooth.spline(Wage$age, Wage$wage, df = 16)
fit_ss_gcv <- smooth.spline(Wage$age, Wage$wage, cv = FALSE)  # GCV
fit_lo    <- loess(wage ~ age, data = Wage, span = 0.2)

age_seq <- seq(min(Wage$age), max(Wage$age), length.out = 200)
df_sm <- data.frame(
  age  = rep(age_seq, 3),
  fit  = c(predict(fit_ss_16, age_seq)$y,
           predict(fit_ss_gcv, age_seq)$y,
           predict(fit_lo, data.frame(age = age_seq))),
  type = rep(c("Smooth spline df=16",
               paste0("Smooth spline GCV (df=",
                      round(fit_ss_gcv$df, 1), ")"),
               "LOESS span=0.2"),
             each = 200)
)
ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_line(data = df_sm, aes(y = fit, colour = type), linewidth = 1) +
  labs(x = "Age", y = "Wage", colour = NULL)
```
#### Python
```{python}
#| eval: false
#| python.reticulate: false
from pygam import LinearGAM, s as gam_s
from statsmodels.nonparametric.smoothers_lowess import lowess

# Smoothing spline via pyGAM
gam_smooth = LinearGAM(gam_s(0, n_splines=25)).gridsearch(
    Wage[["age"]].values, Wage["wage"].values)
XX = gam_smooth.generate_X_grid(term=0, n=200)
pdep, confi = gam_smooth.partial_dependence(term=0, X=XX, width=0.95)

# LOESS
lo_fit = lowess(Wage["wage"], Wage["age"], frac=0.2)

fig, axes = plt.subplots(1, 2, figsize=(12, 4))
axes[0].scatter(Wage["age"], Wage["wage"], alpha=0.15, s=6)
axes[0].plot(XX.ravel(), pdep, color="steelblue", linewidth=2, label="Smoothing spline (GCV)")
axes[0].fill_between(XX.ravel(), confi[:,0], confi[:,1], alpha=0.3)
axes[0].set_xlabel("Age"); axes[0].set_ylabel("Wage"); axes[0].legend()

axes[1].scatter(Wage["age"], Wage["wage"], alpha=0.15, s=6)
axes[1].plot(lo_fit[:,0], lo_fit[:,1], color="tomato", linewidth=2, label="LOESS span=0.2")
axes[1].set_xlabel("Age"); axes[1].set_ylabel("Wage"); axes[1].legend()
plt.tight_layout(); plt.show()
```
:::

## GAMs: continuous outcome {.unnumbered}

Fit three nested GAMs for `wage` on the `Wage` data and compare with AIC [@james2021islr2e, §7.7].
This lab follows the ISLRv2 code, which uses the `mgcv` default
`method = "GCV.Cp"` for smoothness selection; for new analyses prefer
`method = "REML"` (see @sec-smoothing-splines and @sec-gam-fitting).

:::{.panel-tabset}
#### R
```{r}
#| label: lab-gam-r
#| eval: false
#| code-fold: true
gam1 <- gam(wage ~ s(year, k = 4) + s(age, k = 5) + education,
            data = Wage)
gam2 <- gam(wage ~ year + s(age, k = 5) + education,
            data = Wage)
gam3 <- gam(wage ~ s(year, k = 4) + poly(age, 4) + education,
            data = Wage)

AIC(gam1, gam2, gam3)

par(mfrow = c(1, 3))
plot(gam1, shade = TRUE, seWithMean = TRUE,
     main = "GAM1: s(year) + s(age) + education")
```
#### Python
```{python}
#| eval: false
#| python.reticulate: false
from pygam import LinearGAM, s as gam_s, f as gam_f

edu_codes = Wage["education"].cat.codes.values
X_gam = np.column_stack([Wage["year"].values,
                         Wage["age"].values,
                         edu_codes])
y_gam = Wage["wage"].values

gam_py = LinearGAM(
    gam_s(0, n_splines=4) + gam_s(1, n_splines=5) + gam_f(2)
).gridsearch(X_gam, y_gam)

print("AIC:", gam_py.statistics_["AIC"])

fig, axes = plt.subplots(1, 3, figsize=(14, 4))
titles = ["year", "age", "education"]
for i, ax in enumerate(axes):
    XX = gam_py.generate_X_grid(term=i)
    pdep, confi = gam_py.partial_dependence(term=i, X=XX, width=0.95)
    ax.plot(XX[:, i], pdep, linewidth=2)
    ax.fill_between(XX[:, i], confi[:,0], confi[:,1], alpha=0.3)
    ax.set_xlabel(titles[i]); ax.set_ylabel("Partial effect")
plt.tight_layout(); plt.show()
```
:::

## GAMs: binary outcome {.unnumbered}

Predict whether a worker earns more than \$250k using a logistic GAM. We subset to workers
with at least some college education, following @james2021islr2e [§7.7].

:::{.panel-tabset}
#### R
```{r}
#| label: lab-gam-logistic-r
#| eval: false
#| code-fold: true
Wage$high <- factor(ifelse(Wage$wage > 250, "High", "Low"),
                    levels = c("Low", "High"))

fit_logit_gam <- gam(
  high ~ s(year, k = 4) + s(age, k = 5) + education,
  family = binomial,
  data = Wage,
  subset = (education != "1. < HS Grad")
)
summary(fit_logit_gam)

par(mfrow = c(1, 3))
plot(fit_logit_gam, shade = TRUE, seWithMean = TRUE,
     trans = plogis,
     main = "Logistic GAM (P(wage > 250k))")
```
#### Python
```{python}
#| eval: false
#| python.reticulate: false
from pygam import LogisticGAM, s as gam_s, f as gam_f

mask = Wage["education"] != "1. < HS Grad"
W2 = Wage[mask].copy()
W2["high"] = (W2["wage"] > 250).astype(int)
edu2 = W2["education"].cat.codes.values

X_log = np.column_stack([W2["year"].values,
                          W2["age"].values,
                          edu2])
y_log = W2["high"].values

gam_log = LogisticGAM(
    gam_s(0, n_splines=4) + gam_s(1, n_splines=5) + gam_f(2)
).gridsearch(X_log, y_log)

print("Training accuracy:", (gam_log.predict(X_log) == y_log).mean())

fig, axes = plt.subplots(1, 3, figsize=(14, 4))
titles = ["year", "age", "education"]
for i, ax in enumerate(axes):
    XX = gam_log.generate_X_grid(term=i)
    pdep, confi = gam_log.partial_dependence(term=i, X=XX, width=0.95)
    ax.plot(XX[:, i], pdep, linewidth=2)
    ax.fill_between(XX[:, i], confi[:,0], confi[:,1], alpha=0.3)
    ax.set_xlabel(titles[i]); ax.set_ylabel("Partial effect (log-odds)")
plt.tight_layout(); plt.show()
```
:::
