::: notes
Adapted from @wood2017generalized [§6].
:::

The standard tool for fitting GAMs in R is the `mgcv` package
[@wood2017generalized]. We'll use `mgcv::gam()`, which fits GAMs by
**penalized likelihood** with automatic smoothness selection via
REML or GCV.

## The `gam()` call

The basic syntax mirrors `lm()`/`glm()`, with smoothers introduced by
`s()`:

```{r}
#| eval: false
library(mgcv)

m <- gam(
  y ~ s(age) + s(bmi) + sex + s(year, k = 10),
  data = d,
  family = binomial(),
  method = "REML"
)
```

Reading the formula:

- `s(age)`, `s(bmi)`, `s(year, k = 10)` are **smooth terms**. Each
  introduces a basis expansion plus a smoothness penalty. The
  default basis (`bs = "tp"`, thin-plate regression spline) is
  almost always a good choice. `k = 10` requests a basis of
  dimension up to 10 — large enough that the smoothness penalty,
  not the basis, controls flexibility.
- `sex` is a **parametric term**, treated exactly as in a GLM.

You can mix smooth and parametric terms freely. Parametric terms
contribute conventional regression coefficients (with $t$/$z$
statistics in the model summary); smooth terms contribute fitted
curves with effective-degrees-of-freedom summaries.

## Useful `s()` arguments

| Argument | What it controls | When to change from default |
|:--|:--|:--|
| `bs` | basis type (`"tp"`, `"cr"`, `"ps"`, ...) | rarely; `"tp"` is the default and the safe choice |
| `k` | basis dimension (max EDF + 1) | when `gam.check()` flags `k-index < 1` |
| `by` | factor for per-level smooths | predictor-by-factor interactions |
| `m` | order of the penalty derivative | rarely; default penalizes 2nd derivative |
| `pc` | constraint point | when you need the smooth to pass through 0 at a chosen $x$ |

## Smoothness selection: REML vs GCV

`mgcv::gam()` picks the smoothness parameter(s) $\l$ automatically by
optimizing one of:

- **REML** (restricted maximum likelihood). Treats the penalty as
  arising from a normal prior on the spline coefficients and selects
  $\l$ by the marginal likelihood. **Not the default** — `gam()`
  defaults to `method = "GCV.Cp"`, so you have to pass
  `method = "REML"` explicitly (as in the examples in this
  section). Generally **more stable** than GCV, less prone to
  undersmoothing.
- **ML** (marginal likelihood). Similar to REML but with a slightly
  different effective sample size adjustment.
- **GCV** (generalized cross-validation). The default. Computes
  faster than REML but can occasionally select a much wigglier
  smooth than is justified by the data, especially when $n$ is
  small or signal is weak.

**Recommendation: use `method = "REML"`** for routine work. If
you're comparing GAMs to GLMs via AIC, also set `select = TRUE` to
turn on smoothness selection at the term level. `select = TRUE`
adds an extra null-space penalty to each smooth (separate from the
wiggliness penalty) that can shrink an *entire* smooth term to zero
— effectively functioning as variable selection
[@wood2017generalized, §5.4]. Without it, the ordinary wiggliness
penalty can only shrink a smooth down to a straight line, never to
identically zero; `select = TRUE` extends that floor all the way
down. This makes AIC comparisons more honest because the GAM is
then allowed to "discover" that a predictor's smooth shouldn't be
there at all.

## Checking the fit

After fitting, run `gam.check()` to verify:

```{r}
#| eval: false
gam.check(m)
```

Output includes:

- A **convergence diagnostic** (REML/GCV optimization should converge
  to a stationary point — usually it does without intervention).
- **Basis-dimension check** (`k'`, `edf`, `k-index`, `p-value`) for
  each smooth. If `edf` is close to `k'`, the smooth is hitting
  its maximum allowed flexibility — bump `k` up and refit. If
  `k-index < 1` with a small p-value, the basis may be too small.
- **Residual diagnostics**: QQ plot, residuals vs. linear predictor,
  histogram of residuals, observed vs. fitted. Same intuition as for
  a GLM.

## Plotting smooths

```{r}
#| eval: false
plot(m, pages = 1, shade = TRUE)
```

produces a panel of partial-effect plots — one per smooth term —
with a 2-SE pointwise confidence band. The y-axis is on the **linear
predictor scale** by default, centered to mean zero (the smooth's
contribution to $g(\E{Y})$ above its mean), not on the response
scale.

For a publication-quality plot on the response scale, use
`predict(m, type = "response", se.fit = TRUE)` on a grid of
$X$ values with the other predictors held at typical values.

