::: notes
Adapted from @james2021islr2e [§7.6].
:::

**Local regression**, also called **LOESS** (LOcally Estimated
Scatterplot Smoothing) or **LOWESS**, takes yet another approach to
nonlinear regression: instead of fitting a single global model with
flexible basis functions, it fits **many local models**, one for
each point on the $x$-axis where we want a prediction.

{{< slidebreak >}}

## The local fit

<!-- Reviewer note (intentional, do not re-flag): no slidebreak between this
heading and the div below. The preceding slidebreak already opens a fresh
slide; the heading and this div intentionally share that slide to avoid a
title-only slide. -->
::::{#def-local-regression}
#### Local regression at $x_0$

To estimate the regression function $f(x_0)$ at a target point $x_0$:

1. Choose a **bandwidth** $s \in (0, 1]$, also called the **span**.
   Take the fraction $s$ of training observations with $x_i$ closest
   to $x_0$ — this is the **local neighborhood**.
2. **Weight** the neighborhood observations using a kernel that
   gives higher weight to points closer to $x_0$ and zero weight to
   points outside the neighborhood. The tricube kernel
   $K(u) = (1 - |u|^3)^3 \indicp{|u| \le 1}$ is the standard choice.
3. **Fit a low-degree polynomial** (constant, linear, or quadratic)
   by weighted least squares to the neighborhood, using the kernel
   weights. The fitted value at $x_0$ is the prediction
   $\hat f(x_0)$.

To get the entire fitted curve, repeat for a dense grid of $x_0$
values and connect the resulting predictions.
::::

{{< slidebreak >}}

::::{#exm-local-regression}
#### LOESS fit on age--cholesterol

To smooth the relationship between age and total cholesterol in the
WCGS data using LOESS with a span of 0.5:

```{r}
#| eval: false
m_loess <- loess(chol ~ age, data = wcgs, span = 0.5)
```

At age 50, the local neighborhood is the 50% of observations with
ages closest to 50. Each observation in that window is weighted by
the tricube kernel --- down-weighted as its age moves away from 50,
zero weight outside the window. The predicted cholesterol at age 50
is the fitted value from the weighted least-squares linear regression
over that neighborhood.
::::

## Strengths

- **Conceptually simple.** No basis functions, no penalty matrices.
  Just "fit a line through the nearby points, and call that the
  curve at this $x$."
- **Adaptive to local structure.** Regions with rapid change get
  their own local polynomial, just like flat regions do.
- **Robust to outliers** in the standard `loess()` implementation,
  which iteratively down-weights observations with large residuals.

## Limitations

- **No global model.** Local regression doesn't produce a closed-form
  $\hat f(x)$ — predictions at any new $x$ are computed on demand by
  re-running the local fit at that point, but there is no parametric
  formula for the curve. Inference for the entire function (e.g. "is this exposure associated with
  outcome at all") needs special tooling.
- **Hard to extend to many predictors.** With $p$ predictors, the
  local neighborhood is in $p$-dimensional space, and the curse of
  dimensionality bites hard: even for moderate $p$, "the nearest
  fraction $s$ of the data" is geometrically nothing like local. LOESS
  is rarely used with more than 2–4 predictors.
- **Bandwidth selection.** Just like the smoothing parameter $\l$ in
  smoothing splines, $s$ has to be chosen. Cross-validation works
  but is slower.

## Local regression in R

```{r}
#| eval: false
m_loess <- loess(y ~ x, data = d, span = 0.5)
grid <- data.frame(x = seq(min(d$x), max(d$x), length.out = 200))
grid$fit <- predict(m_loess, newdata = grid)

# Inside ggplot2, geom_smooth() with method = "loess"
# is the standard scatterplot smoother:
ggplot(d, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "loess", span = 0.5)
```

For most modeling work in this chapter we'll prefer the
smoothing-spline approach via `mgcv::s()`, which produces a global
fit with full inferential machinery. LOESS is best thought of as a
**diagnostic smoother** — an unobtrusive way to overlay the local
mean of a scatterplot to check whether a parametric fit is missing
nonlinearity.
