::: notes
This section illustrates the methods of @sec-polynomial-regression
through @sec-local-regression on a shared synthetic dataset, so
the strengths and failure modes can be compared side-by-side.
:::

```{r}
#| label: method-comparison-setup
#| include: false
set.seed(204)
library(ggplot2)

n <- 200
x_lim <- c(0, 2 * pi)
truth <- function(x) sin(x) + 0.3 * sin(3 * x)
demo_data <- data.frame(
  x = runif(n, x_lim[1], x_lim[2])
)
demo_data$y <- truth(demo_data$x) + rnorm(n, sd = 0.4)

grid <- data.frame(x = seq(x_lim[1], x_lim[2], length.out = 400))
grid$truth <- truth(grid$x)

predict_with <- function(model) {
  predict(model, newdata = grid)
}

theme_set(theme_minimal())
```

We generate $n = `r n`$ noisy observations from a smooth nonlinear
function

$$f(x) = \sin(x) + 0.3 \sin(3 x), \quad x \in [0, 2\pi],$$

with additive Gaussian noise ($\s = 0.4$). The truth is bumpy
enough to expose differences between the methods but not so wild
that any one of them obviously breaks.

```{r}
#| label: method-comparison-data-fits
#| include: false
# Fit each method on the same demo_data; predictions on `grid` for
# overlay. Comments on each fit explain the design choice.

# (1) Linear baseline.
fit_lin <- lm(y ~ x, data = demo_data)
grid$linear <- predict_with(fit_lin)

# (2) Quadratic polynomial.
fit_poly2 <- lm(y ~ poly(x, 2), data = demo_data)
grid$poly2 <- predict_with(fit_poly2)

# (3) Degree-12 polynomial — deliberately high degree to surface
# Runge-style boundary swinging.
fit_poly12 <- lm(y ~ poly(x, 12), data = demo_data)
grid$poly12 <- predict_with(fit_poly12)

# (4) Step function with 5 bins on equal-quantile cuts. Pin the
# outer breakpoints to x_lim so the grid endpoints don't fall
# outside breaks[1] / breaks[length(breaks)] — otherwise cut()
# returns NA at the boundaries, leaving the step line truncated.
breaks <- quantile(demo_data$x, probs = seq(0, 1, length.out = 6))
breaks[c(1L, length(breaks))] <- x_lim
demo_data$bin <- cut(demo_data$x, breaks = breaks,
                     include.lowest = TRUE)
grid$bin <- cut(grid$x, breaks = breaks, include.lowest = TRUE)
fit_step <- lm(y ~ bin, data = demo_data)
grid$step <- predict_with(fit_step)

# (5) Natural cubic spline, 6 df (knots at internal quantiles).
fit_ns <- lm(y ~ splines::ns(x, df = 6), data = demo_data)
grid$ns <- predict_with(fit_ns)

# (6) Smoothing spline via mgcv::gam, REML smoothness selection.
fit_gam <- mgcv::gam(
  y ~ s(x, bs = "tp", k = 20),
  data = demo_data,
  method = "REML"
)
grid$gam <- predict_with(fit_gam)

# (7) LOESS, default span.
fit_loess <- loess(y ~ x, data = demo_data, span = 0.5)
grid$loess <- predict_with(fit_loess)
```

The fitted curves are shown in @fig-method-comparison-overlay.

::::{#fig-method-comparison-overlay layout-ncol="2"}

:::{#fig-method-comparison-poly}

```{r}
#| label: method-comparison-poly-plot
#| code-fold: true
ggplot(grid, aes(x = x)) +
  geom_point(
    data = demo_data, aes(y = y),
    alpha = 0.3, size = 0.8
  ) +
  geom_line(aes(y = truth, color = "Truth"), linewidth = 0.9) +
  geom_line(aes(y = linear, color = "Linear"),
            linewidth = 0.7, linetype = "longdash") +
  geom_line(aes(y = poly2, color = "Quadratic"), linewidth = 0.7) +
  geom_line(aes(y = poly12, color = "Degree 12"),
            linewidth = 0.7) +
  scale_color_manual(
    name = NULL,
    values = c("Truth" = "black", "Linear" = "grey50",
               "Quadratic" = "tomato", "Degree 12" = "steelblue")
  ) +
  labs(x = "x", y = "y") +
  theme(legend.position = "bottom")
```

Polynomial fits. Quadratic underfits the central inflection; the
degree-12 polynomial chases the noise and swings wildly near the
boundaries (Runge's phenomenon).
:::

:::{#fig-method-comparison-step}

```{r}
#| label: method-comparison-step-plot
#| code-fold: true
ggplot(grid, aes(x = x)) +
  geom_point(
    data = demo_data, aes(y = y),
    alpha = 0.3, size = 0.8
  ) +
  geom_line(aes(y = truth, color = "Truth"), linewidth = 0.9) +
  geom_step(aes(y = step, color = "Step (5 bins)"),
            linewidth = 0.7) +
  scale_color_manual(
    name = NULL,
    values = c("Truth" = "black", "Step (5 bins)" = "tomato")
  ) +
  labs(x = "x", y = "y") +
  theme(legend.position = "bottom")
```

Step function with 5 quantile-spaced bins. Captures the broad shape
but is discontinuous at each cutpoint — almost never the right
model for a continuous exposure.
:::

:::{#fig-method-comparison-splines}

```{r}
#| label: method-comparison-splines-plot
#| code-fold: true
ggplot(grid, aes(x = x)) +
  geom_point(
    data = demo_data, aes(y = y),
    alpha = 0.3, size = 0.8
  ) +
  geom_line(aes(y = truth, color = "Truth"), linewidth = 0.9) +
  geom_line(
    aes(y = ns, color = "Natural cubic spline (df = 6)"),
    linewidth = 0.7
  ) +
  geom_line(
    aes(y = gam, color = "Smoothing spline (REML, k = 20)"),
    linewidth = 0.7
  ) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Truth" = "black",
      "Natural cubic spline (df = 6)" = "tomato",
      "Smoothing spline (REML, k = 20)" = "steelblue"
    )
  ) +
  labs(x = "x", y = "y") +
  theme(legend.position = "bottom")
```

Splines. Both spline fits track the truth closely without the
boundary swings of the degree-12 polynomial. The smoothing spline
(REML-selected $\l$) has slightly less peak-to-peak amplitude
because the penalty shrinks toward a straight line.
:::

:::{#fig-method-comparison-loess}

```{r}
#| label: method-comparison-loess-plot
#| code-fold: true
ggplot(grid, aes(x = x)) +
  geom_point(
    data = demo_data, aes(y = y),
    alpha = 0.3, size = 0.8
  ) +
  geom_line(aes(y = truth, color = "Truth"), linewidth = 0.9) +
  geom_line(aes(y = loess, color = "LOESS (span = 0.5)"),
            linewidth = 0.7) +
  scale_color_manual(
    name = NULL,
    values = c("Truth" = "black", "LOESS (span = 0.5)" = "tomato")
  ) +
  labs(x = "x", y = "y") +
  theme(legend.position = "bottom")
```

LOESS with the default 50% span. Similar shape to the splines but
without a global closed-form parametric representation — the curve
is computed on demand at any evaluation point rather than stored as
a parametric formula.
:::

Comparison of nonlinear regression methods on synthetic data
$y = \sin(x) + 0.3\sin(3x) + \eps$, $\eps \sim \ndist{0, 0.4^2}$,
$n = `r n`$. Splines (both regression and smoothing) and LOESS
track the truth closely; quadratic polynomial underfits the multi-
modal structure; degree-12 polynomial overfits and swings at
boundaries.
::::

## Takeaways

- **Polynomials beyond degree 3–4 are usually a bad idea.** The
  degree-12 fit is a clean illustration: it's interpolating noise in
  the middle of the data and swinging beyond the data range. With 13
  parameters (intercept + 12 polynomial terms) it has more flexibility
  than the natural cubic spline (`df = 6`), yet uses that flexibility
  on noise rather than on the curve's shape — a reminder that global
  polynomials concentrate their degrees of freedom in the wrong
  places.
- **Splines and LOESS are essentially indistinguishable on this
  example.** Both produce smooth curves close to the truth. The
  practical differences are in how easily they extend to multi-
  predictor settings (splines do, LOESS doesn't beyond 2-D) and in
  how inference works (splines have full Wald/Bayesian machinery,
  LOESS doesn't).
- **REML-selected smoothing spline doesn't require a degrees-of-
  freedom choice.** The `df = 6` for the natural cubic spline was a
  modeler choice; the REML smoothing spline picked its own
  smoothness from the data.
- **Step functions are for categorization, not smoothing.** The 5-
  bin step function is the worst fit on a smooth-truth dataset, but
  on real epidemiologic data where bins have substantive meaning
  (clinical BMI categories, age groups), the interpretability
  benefit can outweigh the within-bin information loss.
