::: notes
Adapted from @james2021islr2e [§7.4] and @wood2017generalized [§5].
:::

A **regression spline** is a piecewise polynomial that's constrained
to be smooth at the points where the pieces meet ("knots"). The
result is much more flexible than a single global polynomial but
much smoother than a step function.

{{< slidebreak >}}

## Piecewise polynomials and knots

<!-- Reviewer note (intentional, do not re-flag): no slidebreak between this
heading and the div below. The preceding slidebreak already opens a fresh
slide; the heading and this div intentionally share that slide to avoid a
title-only slide. -->
::::{#def-piecewise-polynomial}
#### Piecewise polynomial

Choose **knots** $\xi_1 < \xi_2 < \cdots < \xi_K$ on the support of
$X$. A **piecewise polynomial of degree $d$** with knots
$\xi_1, \ldots, \xi_K$ is a function that is a polynomial of degree
$d$ on each of the intervals
$(-\infty, \xi_1], (\xi_1, \xi_2], \ldots, (\xi_K, \infty)$ — but
the polynomials in different intervals can have different
coefficients.
::::

{{< slidebreak >}}

::::{#exm-piecewise-polynomial}
#### Unconstrained piecewise polynomial parameter count

Without any continuity constraint, an unconstrained piecewise
polynomial has $(K+1)(d+1)$ free coefficients ($d+1$ per piece, $K+1$
pieces) and can have **jumps and corners** at every knot — basically
$K+1$ independent polynomial fits glued side by side. That's almost
always *too* flexible.
::::

## Continuity and smoothness constraints

Splines pay for their flexibility by adding **smoothness
constraints** at the knots.

{{< slidebreak >}}

::::{#def-spline}
#### Spline

A **spline of degree $d$** with knots $\xi_1, \ldots, \xi_K$ is a
piecewise polynomial of degree $d$ that is continuous, has
continuous first derivative, ..., has continuous $(d-1)$-th
derivative at every knot.
::::

{{< slidebreak >}}

::::{#exm-spline}
#### Effective parameter count for a cubic spline

Each continuity constraint removes one degree of freedom. A
piecewise-cubic ($d = 3$) with $K$ knots has $(K+1) \cdot 4 = 4K + 4$
unconstrained parameters. Requiring continuity of the function and
its first two derivatives at each of the $K$ knots imposes $3K$
constraints, leaving $4K + 4 - 3K = K + 4$ effective parameters.
::::

The most common choice in practice is $d = 3$ — the **cubic
spline** — because human eyes can't see discontinuities in third or
higher derivatives, so a cubic spline already looks "smooth" to the
viewer.

## Truncated power basis

One concrete way to construct a degree-$d$ spline with knots
$\xi_1, \ldots, \xi_K$ is the **truncated power basis**:

$$\b_0 + \b_1 X + \b_2 X^2 + \cdots + \b_d X^d
+ \sum_{k=1}^K \b_{d+k} (X - \xi_k)_+^d,$$

where $(u)_+ = \max(u, 0)$. The first $d+1$ terms are a global
polynomial; each truncated-power term $(X - \xi_k)_+^d$ adds a "kink"
of order $d$ that turns on at knot $\xi_k$. The basis has $K + d + 1$
columns — one for the intercept, $d$ for the global polynomial, and
$K$ for the knot-activated terms — matching the $K + 4$ effective
parameters of a cubic spline (when $d = 3$).

This basis is conceptually clean — every basis function has an
obvious geometric meaning — but **numerically poorly conditioned**
when knots are close together or far from zero. The design matrix
$\tp{\mat{B}} \mat{B}$ can be nearly singular, so OLS
estimates are unstable. For numerical work, the **B-spline basis**
(equivalent space of splines, locally supported basis functions) is
the standard choice. R's `splines::bs()` returns a B-spline basis.

## Natural cubic splines

Cubic splines often **misbehave at the boundary**: with no data
beyond the outermost knots, the cubic terms can swing wildly past the
last observed $x$. The **natural cubic spline** fixes this by adding
four more constraints — the function is required to be **linear**
beyond the outermost knots, which is equivalent to forcing the
second and third derivatives to vanish at each boundary knot:

$$f''(\xi_1) = f'''(\xi_1) = f''(\xi_K) = f'''(\xi_K) = 0.$$

These four boundary constraints (linearity past each end of the
data) reduce the effective parameter count from $K + 4$ to $K$. The
fit is identical to a cubic spline in the *interior* but is forced
to be linear in the tails — almost always a better-behaved
extrapolation than an unconstrained cubic.

R provides natural cubic splines via `splines::ns()`. For routine
work, **natural cubic splines are the default recommendation** over
unconstrained cubic splines.

## Knot placement

A spline fit depends on three choices:

1. **Number of knots** $K$. More knots = more flexibility = more risk
   of overfitting. Common practice: pick $K$ so that the resulting
   degrees of freedom is in the 4–8 range for most exposures, or
   tune $K$ by cross-validation.
2. **Location of knots** $\xi_1, \ldots, \xi_K$. The standard
   recommendation is to place knots at **quantiles of the $X$
   distribution** (e.g. for $K = 4$: at the 20th, 40th, 60th, 80th
   percentiles). This puts more knots where there's more data — the
   exact opposite of what equally-spaced knots would do.
3. **Boundary handling** (cubic vs natural cubic).

The `df` argument to `ns()` or `bs()` automates (1) and (2): R
chooses knots at the appropriate sample quantiles to produce a basis
of the requested dimension.

## Regression splines in R

```{r}
#| eval: false
library(splines)

# Natural cubic spline with 4 degrees of freedom
# (knots auto-placed at quantiles)
m_ns <- lm(y ~ ns(x, df = 4), data = d)

# Cubic B-spline basis with 6 degrees of freedom
m_bs <- lm(y ~ bs(x, df = 6), data = d)

# In a GLM, just substitute glm() and a family
m_glm <- glm(y ~ ns(x, df = 4),
             data = d, family = binomial())
```

Both `ns()` and `bs()` return numeric matrices that slot directly
into `lm()`/`glm()` formulas. The `df` argument is the **exact**
degrees of freedom (basis dimension), in contrast to the `k`
argument of `mgcv::s()` introduced later in this chapter, which is
an **upper bound** that REML penalization can reduce to the actual
effective degrees of freedom. With `splines::ns(df = 4)` you always
get 4 df; with `mgcv::s(x, k = 20)` you might end up with EDF 3 or
EDF 15 depending on the data.

The fitted curve and pointwise CIs come from `predict()` with
`se.fit = TRUE` in the usual way:

```{r}
#| eval: false
grid  <- data.frame(x = seq(min(d$x), max(d$x), length.out = 200))
pred  <- predict(m_ns, newdata = grid, se.fit = TRUE)
grid$fit   <- pred$fit
grid$lower <- pred$fit - 1.96 * pred$se.fit
grid$upper <- pred$fit + 1.96 * pred$se.fit
```
