We'll fit a logistic GAM to the Western Collaborative Group Study (WCGS)
data [@rosenman1975coronary],
the coronary-heart-disease cohort introduced in the
[logistic regression chapter](logistic-regression.qmd).
The outcome is incident coronary heart disease (`chd69`);
the continuous risk factors we smooth are
age (`age`),
body mass index (`bmi`),
serum cholesterol (`chol`),
and systolic blood pressure (`sbp`) —
predictors whose effect on CHD risk need not be linear on the log-odds scale.

## Setup {.unnumbered}

```{r}
#| eval: false
library(mgcv)        # gam(), s()
library(ggplot2)
library(gratia)      # tidy plots for gam fits (optional)

# WCGS ships with this package:
wcgs <- readRDS(system.file("extdata", "wcgs.rds", package = "rme"))

# Binary 0/1 outcome, and drop the one known cholesterol outlier (645):
wcgs <- subset(wcgs, !is.na(chol) & chol < 645)
wcgs$chd <- as.integer(wcgs$chd69 == "Yes")
```

## Linear baseline {.unnumbered}

Start with an ordinary logistic GLM as a baseline:

```{r}
#| eval: false
m_lin <- glm(
  chd ~ age + bmi + chol + sbp,
  family = binomial(), data = wcgs
)
summary(m_lin)
```

Each coefficient is a single log-odds increment per unit of the
predictor — one slope for the entire range of age, BMI, cholesterol,
and blood pressure. That's a strong assumption: it forces risk to
climb at a constant rate across the whole range of each exposure.

## GAM with smooth terms {.unnumbered}

Now refit with a smooth on each continuous predictor:

```{r}
#| eval: false
m_gam <- gam(
  chd ~ s(age) + s(bmi) + s(chol) + s(sbp),
  family = binomial(),
  data = wcgs,
  method = "REML"
)
summary(m_gam)
```

The summary reports an effective degrees of freedom (EDF) for each
smooth. A smooth whose EDF stays close to 1 has collapsed to a
straight line — the data give no evidence of curvature for that
predictor — while an EDF meaningfully above 1 signals a nonlinear
effect. In this dataset most of the smooths sit near 1 (the linear
baseline already captures them well), with cholesterol the predictor
most likely to show curvature; treat the actual EDFs from your run,
not these expectations, as the evidence.

## Reading the EDFs {.unnumbered}

For each smooth, ask:

- **EDF $\approx 1$**: the smooth has collapsed to a straight line.
  No evidence the relationship is nonlinear. You can drop the
  smoother and use a linear term without loss.
- **EDF in 2–5**: meaningfully nonlinear. The fitted curve is more
  than just a line but doesn't have lots of bumps.
- **EDF $> 5$**: substantially nonlinear, possibly with multiple
  inflections. Check whether the data actually support that wiggle
  (look at the plot; check `gam.check()`'s `k-index`).

If `gam.check(m_gam)` flags any smooth as having `edf` close to
`k'`, bump `k` up:

```{r}
#| eval: false
m_gam2 <- gam(
  chd ~ s(age) + s(bmi) + s(chol, k = 20) + s(sbp),
  family = binomial(),
  data = wcgs,
  method = "REML"
)
```

## Plotting partial-effect curves {.unnumbered}

```{r}
#| eval: false
plot(m_gam, pages = 1, shade = TRUE, seWithMean = TRUE)
```

produces one panel per smooth, each showing the partial effect on
the log-odds scale with a 2-SE band. `seWithMean = TRUE` adds the
variance of the smooth's overall level (its centering offset) to
the pointwise standard errors. Without it, the band is the variance
of the *deviation* of the smooth around its mean; with it, the band
is the variance of the smooth itself — the right interpretation when
comparing partial-effect curves across predictors on a publication
figure.

For a cleaner ggplot rendering, `gratia::draw()` is a one-line
alternative:

```{r}
#| eval: false
gratia::draw(m_gam)
```

## Translating to the response scale {.unnumbered}

Partial-effect plots are on the **link scale** (log-odds for
binary). To plot predicted probability against a single predictor —
here cholesterol, the predictor with the clearest curvature — hold
the others at typical values, build the Wald CI on the **link** scale,
then back-transform with `plogis()`:

```{r}
#| eval: false
grid <- data.frame(
  chol = seq(min(wcgs$chol), max(wcgs$chol), length.out = 200),
  age  = median(wcgs$age),
  bmi  = median(wcgs$bmi),
  sbp  = median(wcgs$sbp)
)
pred <- predict(m_gam, newdata = grid,
                type = "link", se.fit = TRUE)
grid$prob  <- plogis(pred$fit)
grid$lower <- plogis(pred$fit - 1.96 * pred$se.fit)
grid$upper <- plogis(pred$fit + 1.96 * pred$se.fit)

ggplot(grid, aes(x = chol, y = prob)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line() +
  labs(x = "Serum cholesterol (mg/dL)",
       y = "Predicted probability of CHD",
       caption = "Other predictors held at sample medians") +
  theme_minimal()
```

We compute the CI on the link scale and back-transform because the
delta-method SE on `type = "response"` doesn't respect the [0, 1]
boundary — Wald intervals built from it can extend outside (0, 1)
near the tails of the curve and have non-symmetric coverage. Working
on the link scale and applying the monotone `plogis()` transformation
keeps the interval inside (0, 1) by construction and preserves
asymmetry around the boundaries (wider near 0 and 1, narrower near
the middle), which is the right qualitative shape for a probability.

## Comparing GLM and GAM {.unnumbered}

```{r}
#| eval: false
# Refit the GAM with method = "ML" for AIC comparison.
# REML and ML likelihoods are on different scales; comparing AIC
# from glm() (always ML) with a REML-fitted gam() mixes criteria.
# Use select = TRUE to enable term-level smoothness selection (the
# null-space penalty), so a smooth that's really just a line — or
# zero — isn't penalized as more complex than the linear baseline.
m_gam_ml <- gam(
  chd ~ s(age) + s(bmi) + s(chol) + s(sbp),
  family = binomial(),
  data = wcgs,
  method = "ML",
  select = TRUE
)
AIC(m_lin, m_gam_ml)
```

Compare the two fits on AIC. When most smooths sit at EDF near 1 —
as several do here — the GLM is the parsimonious choice and the AIC
gap will be small or may even favor the GLM. The GAM earns its extra
complexity only when one or more smooths show clear curvature and
the curve plots reveal a substantively interesting shape; if they
do, the GAM is the better report.

The refit with `method = "ML"` is required only for the AIC
comparison; the REML fit `m_gam` above remains the preferred fit for
final inference and plotting [@wood2017generalized, §6.2].

## Reporting in a manuscript {.unnumbered}

A typical methods-section sentence is:

> Continuous predictors were modeled using thin-plate regression
> splines (the default basis for `mgcv::s()`) via penalized restricted
> maximum likelihood (`mgcv` *version*, R *version*;
> @wood2017generalized). The number of basis functions per predictor
> was set to 10 and reduced by penalization to the effective degrees
> of freedom shown in Table X.

Replace *version* with the actual versions you used — these can
drift between drafts.

Pair this with a table giving each smooth's EDF and approximate
p-value, plus a figure of the partial-effect curves on the
link scale (with the response-scale plot of the predictor of
primary interest as a supplementary figure).
