{{< include latex-macros/macros.qmd >}}

## Bias–variance trade-off and overfitting {#sec-pred-sel-bias-var}

When building a prediction model,
adding more predictors generally **reduces bias**:
the model better captures the true relationships in the population.
But it also **increases variance**:
more parameters to estimate means more sampling variability,
and the model can start to fit noise in the training data.

:::{#def-overfitting}
### Overfitting

**Overfitting** occurs when a model fits the training data well
but predicts poorly for new observations.
It results from including too many predictors
relative to the effective sample size.

:::

A good prediction model minimizes the sum of bias and variance —
the **bias–variance trade-off**.
Adding a predictor helps only if
the reduction in bias outweighs the increase in variance.

## Numerical example of overfitting {#sec-pred-sel-overfit-example}

The table below gives a simple numerical example.
A parsimonious model and an intentionally overfit model
are both trained on the same development set.
The overfit model adds many noise predictors
that have no true relationship with LDL.

::: {#tbl-overfit-example}

```{r}
#| label: overfit-example

set.seed(204)
n_noise_predictors <- 20

hers_base <-
  hers_ldl |>
  dplyr::select(LDL, HT, age, statins) |>
  tidyr::drop_na()

n_overfit <- nrow(hers_base)
idx_overfit <- sample(seq_len(n_overfit), size = floor(0.7 * n_overfit))

train_overfit <- hers_base[idx_overfit, ]
valid_overfit <- hers_base[-idx_overfit, ]

for (j in seq_len(n_noise_predictors)) {
  zname <- paste0("z", j)
  train_overfit[[zname]] <- rnorm(nrow(train_overfit))
  valid_overfit[[zname]] <- rnorm(nrow(valid_overfit))
}

model_parsimonious <- lm(LDL ~ HT + age + statins, data = train_overfit)
noise_terms <- paste0("z", seq_len(n_noise_predictors))
overfit_formula <- reformulate(c("HT", "age", "statins", noise_terms), "LDL")
model_overfit <- lm(
  overfit_formula,
  data = train_overfit
)

rmse <- function(y, yhat) {
  sqrt(mean((y - yhat)^2))
}

tibble::tibble(
  Model = c(
    "Parsimonious model",
    paste0("Overfit model (+", n_noise_predictors, " noise predictors)")
  ),
  Training_RMSE = c(
    rmse(
      train_overfit$LDL,
      predict(model_parsimonious, newdata = train_overfit)
    ),
    rmse(train_overfit$LDL, predict(model_overfit, newdata = train_overfit))
  ),
  Validation_RMSE = c(
    rmse(
      valid_overfit$LDL,
      predict(model_parsimonious, newdata = valid_overfit)
    ),
    rmse(valid_overfit$LDL, predict(model_overfit, newdata = valid_overfit))
  )
) |>
  dplyr::mutate(
    dplyr::across(c(Training_RMSE, Validation_RMSE), ~ round(.x, 2))
  )
```

Numerical example showing overfitting:
lower training error but worse validation error.

:::

In this example,
the overfit model usually has lower training RMSE
but higher validation RMSE,
which is the defining pattern of overfitting.

## Measures of prediction error {#sec-pred-sel-pe}

To select a model that minimizes prediction error,
we need a measure of PE that does not simply increase
with the number of predictors.

For **continuous outcomes**:

- **Adjusted $R^2$** penalizes the number of predictors:
$$\bar R^2 = 1 - \frac{n-1}{n-p-1}(1 - R^2)$$
Adding a variable increases $\bar R^2$ only if it reduces
the residual variance more than the penalty.

- **AIC** and **BIC** impose steeper penalties:
$$\text{AIC} = -2\hat\ell + 2p, \quad
  \text{BIC} = -2\hat\ell + p\log n$$
(see @sec-pred-sel-aic-bic).

For **binary outcomes**,
common measures include
**balanced accuracy** and the **C-statistic**.

**Balanced accuracy**:
$$\text{BA} = \frac{\text{Sensitivity} + \text{Specificity}}{2}$$
where
$$\text{Sensitivity} = \Pr(\hat Y = 1 \mid Y = 1), \quad
\text{Specificity} = \Pr(\hat Y = 0 \mid Y = 0).$$
If $J$ is **Youden's $J$ statistic**,
$$J = \text{Sensitivity} + \text{Specificity} - 1,$$
then
$$\text{BA} = \frac{J + 1}{2}.$$
So BA is a linear transformation of $J$,
and BA and $J$ are equivalent for comparing models.

**C-statistic** (area under the ROC curve, AUC):
$$
\ba
C \eqdef {} & \Pr(\hat{\pi}_i > \hat{\pi}_j \mid Y_i = 1, Y_j = 0) \\
& + \frac{1}{2}\Pr(\hat{\pi}_i = \hat{\pi}_j \mid Y_i = 1, Y_j = 0)
\ea
$$
where $\hat{\pi}$ is the predicted event probability.
An empirical estimator is
$$
\hat{C}
= \frac{1}{n_1 n_0}
\sum_{i:Y_i=1}
\sum_{j:Y_j=0}
\left[
I(\hat{\pi}_i > \hat{\pi}_j)
+ \frac{1}{2}I(\hat{\pi}_i = \hat{\pi}_j)
\right].
$$
It measures discrimination:
the model's ability to rank events above non-events.

For **survival outcomes**, the analogous measure is the **C-index**,
defined as
$$
\hat C_{\text{index}}
= \frac{N_{\text{concordant}} + 0.5\,N_{\text{tied}}}
{N_{\text{comparable}}},
$$
where pairs are comparable when one subject is observed to fail
before the other is known to fail or be censored,
and concordance means the subject with earlier failure
has the higher predicted risk score.

## Optimism and cross-validation {#sec-pred-sel-cv}

Naive estimates of PE evaluated on the same data used
to fit the model are **optimistic**:
they overstate the model's ability to predict new data.
For example, $R^2$ always increases when a predictor is added,
even if the predictor is pure noise.

**Cross-validation** provides a less optimistic estimate of PE
by evaluating predictions on observations
not used to estimate the model.

:::{#def-kfold}
### $k$-fold cross-validation

In **$k$-fold cross-validation**:

1. The data are randomly divided into $k$ equal-sized subsets ("folds").
2. For each fold $i = 1, \ldots, k$:
   - Fit the model using all data *except* fold $i$.
   - Compute predicted values for the observations in fold $i$.
3. Compute a summary PE measure across all folds.

Values of $k = 5$ or $k = 10$ are typical.

:::

Using cross-validation to select predictors avoids overfitting
because the predictive performance is evaluated on held-out data
that did not influence the model fit.

## Cross-validation for LASSO in the HERS data {#sec-pred-sel-cv-hers}

The LASSO (see @sec-pred-sel-lasso) uses cross-validation
to select the penalty parameter $\lambda$
that minimizes prediction error.
The `cv.glmnet` function implements 10-fold cross-validation,
fitting the LASSO for a grid of $\lambda$ values and evaluating
mean squared error on held-out folds.
The fitted object provides two key $\lambda$ values:
`lambda.min` (minimizes CV mean squared error) and
`lambda.1se` (largest $\lambda$ within one standard error of the minimum).
See @sec-pred-sel-lasso for the HERS example
with coefficient paths and CV results.

::: notes

The two recommended values of $\lambda$ represent different
points on the bias–variance trade-off:

- `lambda.min` minimizes cross-validated MSE,
  yielding a model with good predictive accuracy.
- `lambda.1se` uses a larger penalty
  (farther from zero, more shrinkage),
  resulting in a sparser, more conservative model
  that is often preferred for interpretability.

:::

## Development and validation sets {#sec-pred-sel-split}

The most transparent approach to measuring prediction error
is to split the data into a **development set**
(used to fit the model)
and a **validation set**
(used only to evaluate predictions).

::: notes

When a single data set must be split,
a common allocation is two-thirds to development
and one-third to validation.
External validation using a completely independent sample
(from a different time period or study site)
provides the most rigorous assessment of generalizability.

:::

```{r}
#| label: split-sample-ldl

set.seed(42)
n_hers <- nrow(hers_ldl)
train_idx <- sample(seq_len(n_hers), size = floor(0.67 * n_hers))

hers_train <- hers_ldl[train_idx, ]
hers_valid <- hers_ldl[-train_idx, ]

# Fit model on training data
model_train <- lm(
  LDL ~ HT + age + BMI + diabetes + statins,
  data = hers_train
)

# Evaluate on validation data
pred_valid <- predict(model_train, newdata = hers_valid)
rmse_valid <- sqrt(mean((hers_valid$LDL - pred_valid)^2, na.rm = TRUE))
rmse_train <- sqrt(mean(residuals(model_train)^2))

cat(
  "Training RMSE:", round(rmse_train, 2), "\n",
  "Validation RMSE:", round(rmse_valid, 2), "\n"
)
```

::: notes

If the validation RMSE is much larger than the training RMSE,
this is a sign of overfitting.
In well-specified models with adequate sample sizes,
the two values are typically close.

:::
